### 2018-07-18 notebook/helicoverpa_notes.Rmd
Initialise notebook

---
title: "The helicoverpa package"
author: "Brad Duthie and Rose McKeon"
output:
  html_document: default
  pdf_document:
    fig_caption: yes
  word_document:
    fig_caption: yes
linestretch: 1
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \linenumbers
- \bibliographystyle{amnatnat}
bibliography: references.bib
---

```{r, echo=FALSE}
library(shiny)
library(rhandsontable)
```

********************************************************************************

> **This notebook tracks progress on the development of [helicoverpa](https://bradduthie.github.io/helicoverpa/) R package, a BBSRC [FAPESP Newton Funded](http://www.fapesp.br/en/9552) Project to model how the spatial scale of heterogeneity in fungal isolate application and crop plant cultivation affects variation in the selection landscape for biopesticide resistance.**

*******************************************************************************

<a name="contents">Contents:</a>
================================================================================

[Project updates](#updates)

 - **2018**

     - **JUL** [18](#u-18-JUL-2018),

********************************************************************************

<a name="updates">Project updates:</a>
================================================================================

> <a name="u-18-JUL-2018">Update: 18 JUL 2018</a>

I have now initialised this lab notebook for the helicoverpa R package and model.

References

### 2018-07-23 notebook/helicoverpa_notes.Rmd
Attempt to add a comment section

@@
 
-     - **JUL** [18](#u-18-JUL-2018),
+     - **JUL** [18](#u-18-JUL-2018), **JUL** [23](#u-23-JUL-2018)
+
+[Project comments](#comments)
 
@@
 
+> <a name="u-23-JUL-2018">Update: 23 JUL 2018</a>
+
+I have attempted to set this notebook up with [Disqus](https://disqus.com/) commenting enabled with [the universal code optoins](https://helicoverpa.disqus.com/admin/install/platforms/universalcode/) (embedded HTML), which will ideally allow everyone on the project to make comments on the notes.
+
+Data structures
+--------------------------------------------------------------------------------
+
+There are two data structures that we need to think about in particular, and these include the **pests**, which will be represented individually
+
 > <a name="u-18-JUL-2018">Update: 18 JUL 2018</a>
@@
 
+Comments {.comments}
+--------------------------------------------------------------------------------
+
+<div id="disqus_thread"></div>
+<script>
+
+/**
+*  RECOMMENDED CONFIGURATION VARIABLES: EDIT AND UNCOMMENT THE SECTION BELOW TO INSERT DYNAMIC VALUES FROM YOUR PLATFORM OR CMS.
+*  LEARN WHY DEFINING THESE VARIABLES IS IMPORTANT: https://disqus.com/admin/universalcode/#configuration-variables*/
+/*
+var disqus_config = function () {
+this.page.url = PAGE_URL; https://bradduthie.github.io/helicoverpa/notebook/helicoverpa_notes.html
+this.page.identifier = PAGE_IDENTIFIER; // Replace PAGE_IDENTIFIER with your page's unique identifier variable
+};
+*/
+(function() { // DON'T EDIT BELOW THIS LINE
+var d = document, s = d.createElement('script');
+s.src = 'https://helicoverpa.disqus.com/embed.js';
+s.setAttribute('data-timestamp', +new Date());
+(d.head || d.body).appendChild(s);
+})();
+</script>
+<noscript>Please enable JavaScript to view the <a href="https://disqus.com/?ref_noscript">comments powered by Disqus.</a></noscript>
+
 References
+--------------------------------------------------------------------------------

### 2018-07-23 notebook/helicoverpa_notes.Rmd
Fix TOC
Add pest data structure notes

@@
 
-     - **JUL** [18](#u-18-JUL-2018), **JUL** [23](#u-23-JUL-2018)
+     - **JUL** [18](#u-18-JUL-2018), [23](#u-23-JUL-2018)
 
@@
 
-I have attempted to set this notebook up with [Disqus](https://disqus.com/) commenting enabled with [the universal code optoins](https://helicoverpa.disqus.com/admin/install/platforms/universalcode/) (embedded HTML), which will ideally allow everyone on the project to make comments on the notes.
+I have attempted to set this notebook up with [Disqus](https://disqus.com/) commenting enabled with [the universal code options](https://helicoverpa.disqus.com/admin/install/platforms/universalcode/) (embedded HTML), which will ideally allow everyone on the project to make comments on the notes.
 
@@
 
-There are two data structures that we need to think about in particular, and these include the **pests**, which will be represented individually
+There are two data structures that we need to think about in particular, and these include the **pests**, which will be represented individually, and the **landscape**, which will be a large 3D array.
+The biopesticide will not be modelled discretely, but through a change in landscape values.
+
+**Pests**
+
+Pests will need to be represented by some sort of data structure.
+We could create a custom structure in C, but this would get messy if and when we decide to convert the model into an R package so that users can simulate by calling an R function.
+I really don't see why pests cannot be represented by a large table, which is how I have always represented individuals in the past, and I already [have the code](https://github.com/bradduthie/gmse/blob/master/src/resource.c#L344) for reading R tables into C.
+The way that this will work then is to have each individual represented by a different row in the table, with columns corresponding to individual traits.
+Some traits that will be important include:
+
+- Individual ID
+- Sex (female or male)
+- Traits 1-T: This includes traits such as dispersal ability, location (x and y), fecundity, and resistance to different biopesticides
+- Loci 1-L: This will be a diploid system with allele values ($A^{v}_{i}$) and dominance coefficients $A^{h}_{v}$, the notation of which will likely change, but three `double` values will be used for each haploid loci: 1.
+allele number ($i$), value ($v$), and dominance ($h$).
+Since alleles are diploid, we will need $6L$ `double` columns in total, which will be the last columns of the table (this is just easier)
+
+The idea of each individual having its own ID and sex is fairly straightforward, but what I'm thinking we can do with the traits will be a bit different.
+We want the evolving genome to affect resistance through a evolving traits, such that the genome affects traits underlying resistance, which then affect the evolution of the genetic architecture itself.
+The idea then is that each trait will be modelled as a `double` real number that is 'visible' to selection -- in most cases we will want the value of each trait will be a function of the genome, but this wouldn't necessarily be a requirement (e.g., if we just wanted to fix dispersal ability to some value, we could set it to whatever we desire in the relevant row and column -- this will also be useful for location on the landscape, which will be represented by two values x and y).
+We could then write a specific modular function `build_traits`, which loops through each individual and derives a single number from the individual's genome, which is then placed in the relevant trait column (1-T).
+Trait columns can by dynamic -- we can have anywhere from 1 to T of them, and the value of T will need to be stored somewhere -- better, if in the code we define ID and Sex as traits, we can just have loci start at T (or T + 1 in R).
+
+By having a modular `build_traits` function, we can keep the code flexible enough to have different ways of deriving single trait values for any given genome.
+The genome itself is probably easiest to model by breaking it down into L blocks of six, taking up 6L total columns of the 2D **pest** array.
+The first three values of a block will be the number (1), value (2), and dominance (3) for an allele, while the next three values will be the same for a homologous allele at the locus.
+The dominance values could be compared to calculate which allele has a stronger effect, similar to @Duthie2016a [[code](https://github.com/bradduthie/Inbreeding)], and this could be pre-set or randomised in some way for different allele types.
+The values could mean whatever we want them to, but a good starting point would be to think of them as additively effecting one or more trait as in @Duthie2017 [[code](https://github.com/bradduthie/pre_post_cop)].
+We could of course just have the allele numbers in the genome, with a separate lookup table to match them with values and dominance coefficients, but I actually think this would be slower and more difficult to code.
+The `build_traits` function will loop through all of the loci, gathering in the relevant values to build each trait -- a sub-function (e.g., `calc_trait`) could do this for an individual trait, identifying where and what in the genome should be used to make a single trait such as resistance to a particular biopesticide.
+So, for example, the pest data structure for one (non ID or Sex) trait and one locus would look like this:
+
+```{r}
+simple_pests <- matrix(data = 0, nrow = 4, ncol = 9);
+colnames(simple_pests) <- c("ID", "Sex", "Trait_1", "L1A1_i", "L1A1_v", 
+                            "L1A1_h", "L1A2_i", "L1A2_v", "L1A2_h");
+simple_pests[,1] <- 1:4;
+simple_pests[,2] <- c(0, 1, 0, 1);
+simple_pests[,3] <- runif(n = 4);
+simple_pests[,4] <- sample(x = 1:2, size = 4, replace = TRUE);
+simple_pests[,7] <- sample(x = 1:2, size = 4, replace = TRUE);
+print(simple_pests);
+```
+
+I've not included the allele values or dominance coefficients for alleles 1 and 2.
+Note that in the columns 'L1' refers to locus 1, and 'A1' and 'A2' refer to alleles 1 and 2, respectively.
+As mentioned earlier, these will need to be represented by doubles in C, and the table could get quite big (well over 1000 columns, and potentially tens of thousands of rows).
+
+**Landscape**
+
+The landscape will be modelled using a three dimensional array.
 

### 2018-07-23 notebook/helicoverpa_notes.Rmd
Add landscape structure notes

@@
 As mentioned earlier, these will need to be represented by doubles in C, and the table could get quite big (well over 1000 columns, and potentially tens of thousands of rows).
+The benefit of doing all this is that we can potentially use multiple different types of infection models, as desired (e.g., gene for gene, quantative, threshold).
 
@@
 The landscape will be modelled using a three dimensional array.
+We want space to be explicit, and the easiest way to do this is to have the first two dimensions of the array correspond to the x and y locations on a region of landscape.
+Interactions can happen at the scale of landscape cells (array elements), and distance between cells in any location can allow for explicit spatial processes (typically, count the number of cells in any of 8 directions, including diagonally, but other neighbour distances could be used, or cells could be hexagonal if there is good reason to do it).
+Having different cells with different values allows for landscape level heterogeneity (could also make values autocorrelated as desired as in @Duthie2013).
+Perhaps each cell starts with having some ID for the farm, as in the case of the 8 by 8 landscape below (actually x and y dimensions will be a variable in the model -- I usually use 100 by 100, with a torus landscape that wraps around itself to avoid edge effects).
+
+```{r}
+land_IDs <- matrix(data = 0, nrow = 8, ncol = 8);
+land_IDs[1:2, 1:8] <- 1;
+land_IDs[3:6, 1:5] <- 2;
+land_IDs[3:8, 6:8] <- 3;
+land_IDs[7:8, 1:3] <- 4;
+land_IDs[7:8, 4:5] <- 5;
+print(land_IDs);
+```
+
+So each of five farms or types of farms (or something else) are represented in the landscape.
+But we need lots of pieces of information for each cell, including what is planted on the cell, local environmental variables, and biopesticide applied.
+To get this, we can then add multiple layers on the z axis of a 3D array.
+So the landscape layer above is just the 'top', with multiple layers beneath coding critical information in a `land` array.
+Assume, for now that we need three layers (we'll actually need a lot more): (1) `land_IDs`, (2) crop used, (3) biopesticide 1 applied.
+
+```{r}
+land      <- array(data = 0, dim = c(8, 8, 3));
+land[,,1] <- land_IDs;
+land[,,2] <- sample(x = 1:3, size = 64, replace = TRUE);
+land[,,3] <- round(x = runif(n = 64, min = 0, max = 1), digits = 3);
+print(land);
+```
+
+The details aren't so important for now; what matters is the idea that we can represent information about landscape cells dynamically using array layers, which can therefore correspond to any number of things that we want to affect pests.
+The above structures can be initialised in R and diverted to a toy model that could help accomplish the short-term goals of the project, but also diverted to C to do the more intense simulatoin involving complex genetic architecture (we'll want a standalone C version too for speed and high-performance cluster computing).
+The toy model is probably the fist step -- to have something loop over time steps showing that something (e.g., allelic diversity) is maintained under variable landscapes, but is not when landscapes are homogenous (or when dispersal is infinite).
+Probably best to simplify the genetics as much as possible (e.g., 1 locus, fewer than 5 alleles) for this, then return some numbers that can be extracted from these data structures to build a graphic.
+
+I see these two structures (**PEST** and **LAND**) as being the two most important ones to get right early on because these are the ones that will be used in almost every part of the model.
+The code will otherwise be written modularly, meaning that if we don't like the way that we're, e.g., deriving pest traits from their genetic architecture, or setting dispersal rules, or modelling sex, then we can simply rewrite the specific function that does these things and plug it back into the model.
+With this, some list or vector is also needed to hold parameter values (**PARA**), at lesat from the transition from R to C.
 

### 2018-07-30 notebook/helicoverpa_notes.Rmd
Some ideas about code structure

@@
 
-*******************************************************************************
+********************************************************************************
+
+<center>**[Go to comments](#comments)**</center>
+
+********************************************************************************
 
@@
 
-     - **JUL** [18](#u-18-JUL-2018), [23](#u-23-JUL-2018)
+     - **JUL** [18](#u-18-JUL-2018), [23](#u-23-JUL-2018), [30](#u-30-JUL-2018)
 
@@
 ================================================================================
+
+> <a name="u-30-JUL-2018">Update: 30 JUL 2018</a>
+
+**Some general thoughts about code structure**
+
+There are some high-level functions that will be useful to think about before coding; that is, functions that need to do something in sequence during simulation initialisation and then during each time step of the simulation.
+These functions will call subfunctions that run key routines -- I think that all of these function, and the outer function (e.g., `simulate_resistance`, which runs through multiple time steps to test a particular set of parameter combinations), should be written in C for efficiency.
+An R function can call `simulate_resistance` and return the results to R, or an outer C function can also call it and print results to a text file (faster and easier with the computing cluster).
+The point is to have a single funtion that everything feeds to that does the work by calling multiple subfunctions -- this workhorse function `simulate_resistance` will contain the main loop over $N$ `time_steps`.
+
+Prior to the loop, there will be some initialisation of the landscape `LAND` and the pests `PEST`.
+This will include all of the values in each layer `LAND` based on parameter initialisation.
+The functions `initialise_land` and `initialise_pest` are therefore called once at the start of the simulation; they will need to decide the distribution of landscape values and pests, and the sizes of each.
+It will not be necessary to initialise the `PEST` traits, only their genomes (based on allele and loci parameters, dominance, etc.) and non-genetic attributes such as location and ID (sex might or might not be based on genome).
+Before starting the big loop, the key structures of `PEST` and `LAND` need to otherwise be built.
+Any other key data structures will need to be initialised -- I am thinking that the `simulate_resistance` function, or maybe some small other function, should send a vector of information so that everything can be stored handily, then individual parameter values can be set in the small outer function then compiled into the vector for ease of use.
+
+After initialisation, one big loop will run the simulation over multiple time steps (some pseudocode below).
+
+```
+simulate_resistance(parameter_values){
+    error_checks(parameter_values);
+    initialise_land(LAND);
+    initialise_pest(PEST);
+    for(time_step = 0; time_step < time_max; time_step++){
+        build_pest_traits(PEST, build_type);
+        set_crops(LAND, crop_parameters);
+        set_biopesticide(LAND, biopesticide_parameters);
+        move_pest(PEST, move_parameters);
+        feed_pest(PEST, LAND, feeding_parameters);
+        kill_pest(PEST, LAND, death_parameters);
+        reproduce_pest(PEST, evolution_parameters);
+        collect_PEST_data(PEST);
+        collect_LAND_data(LAND);
+        summarise_time_step(PEST, LAND);
+        status_check(PEST, LAND);
+    }
+    summarise_simulation();
+}
+```
+
+Multiple events will occur in each time step, including (but not limited to) the following: The `build_pest_traits()` function will simply read in and out the `PEST` array, building the traits from the pest genome according to some `build_type` in each generation.
+The `set_crops` function will assign `LAND` cel layer values basedon on the crop parameter values underlying, e.g., spatial heterogeneity; similarly, `set_biopesticide` will affect `LAND` cell layer values.
+Then the `move_pest` function will cause the pests to move according to some sort of disperal rules set in `parameter_values`.
+Note that the landscape and the pests have not interacted in the time step yet (though we can imagine the landscape affecting pest movement rules), and will now do so in `feed_pest`, where they will affect crops, followed by `kill_pest` and `reproduce_pest`, where they will suffer mortality and reproduce as a consequence of their interactions.
+**We might want to think about the order of these operations.** After pest reproduction (which will include sex and genetic -- but not trait -- transmission), then summary data on the pests and landscape need to be collected (could even have an option to print everything out at this point, but it might be a lot of information) -- followed by a sort of grand summary of the time step.
+When the whole simulation has finished, then the results can be printed.
+
+Note also that some error checks and status checks will be useful, since we'll want to make this function able to run from R.
+The error checks will just make sure nothing will crash the simulation (some of this might be better programmed in R instead of C), while the `status_check` might look to see, e.g., if pests have gone extinct, or have some sort of rule for termination of the simulation if something happens.
+
+I suspect that the above structure will need rearranging and rethinking, but it's a starting point.
+Subfunctions will be the difficult part, coding efficient ways of building traits from the genome, movement rules, reproduction, etc.
 
@@
 
-Comments {.comments}
+Comments {#comments}
 --------------------------------------------------------------------------------

### 2018-08-13 R/toy_model.R
Draft toy

@@ -6,18 +6,19 @@
 # but also don't want to worry about not being able to name functions like
 # "initialise_land" the same here as elsewhere in the eventual bigger model.
 # ==============================================================================
-toy_simulate_resistance <- function(generations = 20,   
-                                    xdim = 2, 
-                                    ydim = 2, 
-                                    pathogens = 1, 
-                                    crops = 1, 
-                                    pest_init = 100, 
-                                    crop_rotate = "static", 
-                                    path_rotate = "static", 
-                                    pest_move_pr = 0.1,
-                                    pest_move_dist = 1, 
-                                    fecundity = 2, 
-                                    cell_K = 100){
+toy_simulate_resistance <- function(generations = 20,       # Generations to sim
+                                    xdim = 2,               # Land dimension 1
+                                    ydim = 2,               # Land dimension 2
+                                    pathogens = 1,          # Pathogen strains
+                                    crops = 1,              # Crop species
+                                    pest_init = 2000,       # Initial pests 
+                                    crop_rotate = "static", # Crops rotated
+                                    path_rotate = "static", # Pathogens rotated
+                                    pest_move_pr = 0.1,     # Pest movement
+                                    pest_move_dist = 1,     # Pest move distance
+                                    fecundity = 8,          # Offspring per fem
+                                    cell_K = 2000           # K per cell
+                                    ){
     
     if(pest_move_dist > xdim & pest_move_dist > ydim){
         pest_move_dist <- max(c(xdim, ydim)); # Avoids error
@@ -36,16 +37,16 @@ toy_simulate_resistance <- function(generations = 20,
         PEST <- toy_move_pest(PEST, LAND, pest_move_pr, pest_move_dist);        
         PEST <- toy_feed_pest(PEST, LAND);                                      
         if(toy_check_extinction(PEST, gen) == TRUE){ # Hate these if breaks here
-            print("feed"); break;
+            break;
         }
         PEST <- toy_kill_pest(PEST, LAND);
         if(toy_check_extinction(PEST, gen) == TRUE){
-            print("kill"); break;
+            break;
         }
         PEST <- toy_reproduce_pest(PEST, LAND, pathogens, crops, fecundity, 
                                    cell_K);
         if(toy_check_extinction(PEST, gen) == TRUE){
-            print("repr"); break;
[…]

### 2018-08-13 notebook/helicoverpa_notes.Rmd
Explanation of toy model

@@
 
+     - **AUG** [13](#u-13-AUG-2018)
      - **JUL** [18](#u-18-JUL-2018), [23](#u-23-JUL-2018), [30](#u-30-JUL-2018)
@@
 ================================================================================
+
+> <a name="u-18-AUG-2018">Update: 18 AUG 2018</a>
+
+**Visual stuff looks good!**
+
+Rose has created a very nice [visual display](http://r.rosemckeon.co.uk/helicoverpa/) of the landscape, with sliders allowing for multiple crop species and pathogen strains.
+
+**Toy model prototype**
+
+I have managed to complete a toy model today, which should be sufficient to produce the simulated data needed to get the key ideas across.
+I'm going to dump the code below, which is located in the [toy_model.R](https://github.com/bradduthie/helicoverpa/blob/dev/R/toy_model.R) data file, not yet pushed on to master.
+We'll need to merge the interface branch with the dev branch soon, which should be easy enough.
+The whole thing is in R, and includes all of the basic steps from the code structure mentioned on [30 JUL 2018](#u-30-JUL-2018), including mutation, recombination, dispersal, and pathogen and crop rotation (random or sequential).
+The output isn't in that helpful of a form yet, but it will get there soon.
+To skip the many lines of code from the [toy_model.R](https://github.com/bradduthie/helicoverpa/blob/dev/R/toy_model.R) file, [go here](#skip_AUG_code).
+
+To quickly summarise, there are 12 parameters we can play with in the toy simulation model:
+
+1. Number of generations to simulate (`generations`)
+2. X dimension (longitude), i.e., cell number, on the landscape (`xdim`)
+3. Y dimension (latitude), i.e., cell number, on the landscape (`ydim`)
+4. Total number of pathogen strains that can be applied (`pathogens`)
+5. Total number of crops that can be planted (`crops`)
+6. Initial number of pests (`pest_init`)
+7.
+How to rotate crops on the landscape (`crop_rotate`: where `crop_rotate = "static"` means don't rotate between generations, `crop_rotate = "rotate"` means sequence through each crop one at a time over generations, and `crop_rotate = "random"` means apply a random crop in each generation -- all done independently for each landscape cell)
+8.
+How to rotate pathogen application on the landscape (`path_rotate`: where `path_rotate = "static"` means don't rotate between generations, `path_rotate = "rotate"` means sequence through each pathogen one at a time over generations, and `path_rotate = "random"` means apply a random pathogen in each generation -- all done independently for each landscape cell)
+9. Probability that a pest will move to a new cell in the generation (`pest_move_pr`)
+10.
+Distance a pest will move up to in any direction from its natal cell (`pest_move_dist`; note, the landscape is a torus, and there is therefore no edge -- pests that move off of one side end up on the other)
+11. Number of offspring each *female* pest produces (`fecundity`)
+12. Carrying capacity of pests *per cell* (`cell_K`)
+
+Default parameter values are shown below.
+Note that the model is an individual-based simulation with an *extremely* simple diploid genetic architecture with two loci (one for pathogens, and one for crops).
+If there are $N$ pathogens, then there are $N$ alleles for pathogen resistance, and pests must have at least one allele to resist (else they die).
+Similarly, if there are $N$ crops, then there are $N$ alleles for being able to eat crops, and pests must have at least one to eat (else they die).
+
+```{r}
+toy_simulate_resistance <- function(generations = 20,       # Generations to sim
+                                    xdim = 2,               # Land dimension 1
+                                    ydim = 2,               # Land dimension 2
+                                    pathogens = 1,          # Pathogen strains
+                                    crops = 1,              # Crop species
+                                    pest_init = 2000,       # Initial pests 
+                                    crop_rotate = "static", # Crops rotated
+                                    path_rotate = "static", # Pathogens rotated
+                                    pest_move_pr = 0.1,     # Pest movement
+                                    pest_move_dist = 1,     # Pest move distance
+                                    fecundity = 8,          # Offspring per fem
+                                    cell_K = 2000           # K per cell
+                                    ){
+
+    if(pest_move_dist > xdim & pest_move_dist > ydim){
+        pest_move_dist <- max(c(xdim, ydim)); # Avoids error
+    }
+    # Start initialising the landscape and pests
+    LAND <- toy_initialise_land(xdim  = xdim, ydim = ydim, 
+                                pathogens = pathogens, crops = crops);
+    PEST <- toy_initialise_pest(LAND, N = pest_init, p_al = pathogens, 
+                                c_al = crops);
+    # Start the generations
+    PEST_DATA   <- NULL;
+    gen         <- 1;
+    while(gen < generations){
+        LAND <- toy_set_crops(LAND, crops, crop_rotate);                        
+        LAND <- toy_set_paths(LAND, pathogens, path_rotate);                    
+        PEST <- toy_move_pest(PEST, LAND, pest_move_pr, pest_move_dist);        
+        PEST <- toy_feed_pest(PEST, LAND);                                      
+        if(toy_check_extinction(PEST, gen) == TRUE){ # Hate these if breaks here
+            break;
+        }
+        PEST <- toy_kill_pest(PEST, LAND);
+        if(toy_check_extinction(PEST, gen) == TRUE){
+            break;
+        }
+        PEST <- toy_reproduce_pest(PEST, LAND, pathogens, crops, fecundity, 
+                                   cell_K);
+        if(toy_check_extinction(PEST, gen) == TRUE){
+            break;
+        }
+        PEST_DATA[[gen]] <- PEST;
+        gen <- gen + 1;
+    }
+    return(PEST_DATA);
+}
+
+summarise_pest_data <- function(PEST_DATA){
+    # Density estimates
+    den_list <- unlist(lapply(PEST_DATA, dim));
+    den_list <- den_list[c(TRUE, FALSE)];
+    return(den_list);
+}
+
+toy_check_extinction <- function(PEST, gen){
+    if(is.vector(PEST) == TRUE){
+        print(paste("Extinction occurred in generation", gen));
+        return(TRUE);
+    }
+    if(dim(PEST)[1] < 4){
+        print(paste("Extinction occurred in generation", gen));
+        return(TRUE);
+    }
+    if(sum(PEST[,2] == 0) < 1 | sum(PEST[,2] == 1) < 1){
+        print(paste("Extinction occurred in generation", gen));
+        return(TRUE);
+    }
+    return(FALSE);
+}
+
+# Initialise the landscape
+toy_initialise_land <- function(xdim = 2, ydim = 2, pathogens = 1, crops = 1){
+    LAND      <- array(data = 0, dim = c(xdim, ydim, 3));
+    p_values  <- sample(x = 1:pathogens, size = xdim * ydim, replace = TRUE);
+    p_layer   <- matrix(data = p_values, ncol = xdim, nrow = ydim);
+    c_values  <- sample(x = 1:crops, size = xdim * ydim, replace = TRUE);
+    c_layer   <- matrix(data = c_values, ncol = xdim, nrow = ydim);
+    LAND[,,2] <- p_layer; # Just occurred to me that we might want a layer for
+    LAND[,,3] <- c_layer; # carrying capacity of the cell in LAND later
+    return(LAND);
+}
+
+# Initialise some *very* simple pests. Each allele is just going to map one to
+# one for whether a crop can be attacked or a pathogen resisted. I am not even
+# going to add traits affecting dispersal in the toy version -- this can be
+# uniform for all individuals
+toy_initialise_pest <- function(LAND, N = 10, p_al = 1, c_al = 1){
+    xdim     <- dim(LAND)[1]; # This is needed to put the pests on some
+    ydim     <- dim(LAND)[2]; # landscape cell
+    PEST     <- matrix(data = 0, nrow = N, ncol = 8);
+    PEST[,1] <- 1:N;                                           # ID
+    PEST[,2] <- sample(x = c(0, 1), size = N, replace = TRUE); # Sex
+    PEST[,3] <- sample(x = 1:xdim,  size = N, replace = TRUE); # x-location
+    PEST[,4] <- sample(x = 1:ydim,  size = N, replace = TRUE); # y-location
+    PEST[,5] <- sample(x = 1:p_al,  size = N, replace = TRUE); # p allele 1
+    PEST[,6] <- sample(x = 1:p_al,  size = N, replace = TRUE); # p allele 2
+    PEST[,7] <- sample(x = 1:c_al,  size = N, replace = TRUE); # c allele 1
+    PEST[,8] <- sample(x = 1:c_al,  size = N, replace = TRUE); # c allele 2
+    return(PEST);
+}
+
+# Just going to make this a simple function at first, randomly changing crops
+toy_set_crops <- function(LAND, crops = 1, type = "static"){
+    if(crops == 1 | type == "static"){
+        return(LAND);
+    }
+    if(type == "rotate"){
+        old_crop                   <- LAND[,,3];
+        new_crop                   <- old_crop + 1;
+        new_crop[new_crop > crops] <- 1;
+        LAND[,,3]                  <- new_crop;
+    }else{ # Else it just randomises -- can think about fancier ways later
+        xdim      <- dim(LAND)[1];
+        ydim      <- dim(LAND)[2];
+        new_cval  <- sample(x = 1:crops, size = xdim * ydim, replace = TRUE);
+        new_crop  <- matrix(data = new_cval, nrow = xdim, ncol = ydim);
+        LAND[,,3] <- new_crop;
+    }
+    return(LAND);
+}
+
+# Ditto here -- just a simple function changing pathogens
+toy_set_paths <- function(LAND, paths = 1, type = "static"){
+    if(paths == 1 | type == "static"){
+        return(LAND);
+    }
+    if(type == "rotate"){
+        old_path                   <- LAND[,,3];
+        new_path                   <- old_path + 1;
+        new_path[new_path > paths] <- 1;
+        LAND[,,2]                  <- new_path;
+    }else{ # Else it just randomises -- can think about fancier ways later
+        xdim      <- dim(LAND)[1];
+        ydim      <- dim(LAND)[2];
+        new_pval  <- sample(x = 1:paths, size = xdim * ydim, replace = TRUE);
+        new_path  <- matrix(data = new_pval, nrow = xdim, ncol = ydim);
+        LAND[,,2] <- new_path;
+    }
+    return(LAND);
+}
+
+# Just a function to move pests at a given value, randomly on the landscape. The
+# `prob` is just probability of moving in a time step, while `dist` is just the
+# maximum distance moved in cells, in any direction.
+toy_move_pest <- function(PEST, LAND, prob = 0.1, dist = 1){
+    xdim                     <- dim(LAND)[1];
+    ydim                     <- dim(LAND)[2];
+    pests                    <- dim(PEST)[1];
+    to_move                  <- rbinom(n = pests, size = 1, pr = prob);
+    move_x                   <- sample(x = -dist:dist, size = pests, 
+                                       replace = TRUE);
+    move_y                   <- sample(x = -dist:dist, size = pests, 
+                                       replace = TRUE);
+    PEST[to_move == 1, 3]    <- PEST[to_move == 1, 3] + move_x[to_move == 1];
+    PEST[to_move == 1, 4]    <- PEST[to_move == 1, 4] + move_y[to_move == 1];
+    # The ifs below make a torus landscape so that pests don't move off of it
+    # In C, this will be done with a loop that looks cleaner; the below is just
+    # avoiding using a loop in R, though it would probably be fine.
+    if(length(PEST[PEST[,3] < 1, 3]) > 0){ 
+        PEST[PEST[,3] < 1, 3] <- PEST[PEST[,3] < 1, 3] + xdim;
+    }
+    if(length(PEST[PEST[,3] > xdim, 3]) > 0){
+        PEST[PEST[,3] > xdim, 3] <- PEST[PEST[,3] > xdim, 3] - xdim;
+    }
+    if(length(PEST[PEST[,4] < 1, 4]) > 0){
+        PEST[PEST[,4] < 1, 4] <- PEST[PEST[,4] < 1, 4] + ydim;
+    }
+    if(length(PEST[PEST[,4] > ydim, 4]) > 0){
+        PEST[PEST[,4] > ydim, 4] <- PEST[PEST[,4] > ydim, 4] - ydim;
+    }
+    return(PEST);
+}
+
+# Need to now feed the pests and remove those that don't feed (crop unavailable)
+toy_feed_pest <- function(PEST, LAND){
+    # I'm just going to use a loop here, else matching to LAND cells is rough
+    pests <- dim(PEST)[1];
+    eaten <- rep(x = 0, times = pests); 
+    for(i in 1:pests){
+        x_loc <- PEST[i, 3];
+        y_loc <- PEST[i, 4];
+        food  <- LAND[x_loc, y_loc, 3];
+        if(PEST[i, 7] == food | PEST[i, 8] == food){
+            eaten[i] <- 1;
+        }
+    }
+    PEST <- PEST[eaten == 1,]; # You don't eat, you don't live
+    return(PEST);
+}
+
+# Then need to have pests resist pathogens, kill those that cannot
+toy_kill_pest <- function(PEST, LAND){
+    # I'm just going to use a loop here, else matching to LAND cells is rough
+    pests    <- dim(PEST)[1];
+    survived <- rep(x = 0, times = pests);
+    for(i in 1:pests){
+        x_loc <- PEST[i, 3];
+        y_loc <- PEST[i, 4];
+        patho <- LAND[x_loc, y_loc, 2];
+        if(PEST[i, 5] == patho | PEST[i, 6] == patho){
+            survived[i] <- 1;
+        }
+    }
+    PEST <- PEST[survived == 1,]; # You don't eat, you don't live
+    return(PEST);
+}
+
+# Then, reproduce pests that are left
+toy_reproduce_pest <- function(PEST, LAND, pa, cr, births = 2, K = 100){
+    x_dim           <- dim(LAND)[1];
+    y_dim           <- dim(LAND)[2];
+    offspring       <- NULL;
+    total_offspring <- 0; # This and the below are just to avoid an rbind(),
+    cell            <- 1; # which is a massive memory sink
+    lst_ID          <- max(PEST[,1]);
+    for(xloc in 1:x_dim){ # Double loop is just much cleaner here
+        for(yloc in 1:y_dim){
+            locals     <- which(PEST[,3] == xloc & PEST[,4] == yloc);
+            local_offs <- NULL;
+            if( length(locals) > 1 ){
+                local_offs <- toy_breed_locals(PEST, locals, births, K, lst_ID);
+            }
+            if(length(local_offs) > 0){
+                lst_ID            <- lst_ID + dim(local_offs)[1];
+                total_offspring   <- total_offspring + dim(local_offs)[1];
+                offspring[[cell]] <- local_offs;
+            }else{
+                offspring[[cell]] <- NULL;
+            }
+            cell              <- cell + 1;
+        }
+    }
+    offspring <- build_new_pest(offspring, total_offspring, pa, cr);
+    return(offspring);
+}
+
+# Need to recombine the genomes correctly in the offspring
+toy_breed_locals <- function(PEST, locals, births, K, last_ID){
+    loc_PEST <- PEST[locals,]; # Need two of each sex (allee effect)
+    if(sum(loc_PEST[,2] == 0) < 2 | sum(loc_PEST[,2] == 1) < 2){
+        return(NULL);
+    }
+    females       <- loc_PEST[loc_PEST[,2] == 0,];
+    males         <- loc_PEST[loc_PEST[,2] == 1,];
+    new_offs      <- dim(females)[1] * floor(births);  
+    offspring     <- matrix(data = 0, nrow = new_offs, ncol = 8);     
+    offspring[,1] <- (last_ID + 1):(last_ID + new_offs);
+    offspring[,2] <- sample(x = c(0, 1), size = new_offs, replace = TRUE);
+    offspring[,3] <- loc_PEST[1, 3];
+    offspring[,4] <- loc_PEST[1, 4];
+    # Below grabs all of the alleles from females and males
+    p_fem_alleles <- c(females[,5], females[,6]);
+    p_mal_alleles <- c(males[,5], males[,6]);
+    c_fem_alleles <- c(females[,7], females[,8]);
+    c_mal_alleles <- c(males[,7], males[,8]);
+    # Now add them to the offspring randomly, one from female and one from male
+    for(i in 1:new_offs){
+        if(runif(n = 1) < 0.5){
+            offspring[i, 5] <- sample(x = p_fem_alleles, size = 1);
+            offspring[i, 6] <- sample(x = p_mal_alleles, size = 1);
+        }else{
+            offspring[i, 5] <- sample(x = p_mal_alleles, size = 1);
+            offspring[i, 6] <- sample(x = p_fem_alleles, size = 1);            
+        }
+        if(runif(n = 1) < 0.5){
+            offspring[i, 7] <- sample(x = c_fem_alleles, size = 1);
+            offspring[i, 8] <- sample(x = c_mal_alleles, size = 1);
+        }else{
+            offspring[i, 7] <- sample(x = c_mal_alleles, size = 1);
+            offspring[i, 8] <- sample(x = c_fem_alleles, size = 1);            
+        }
+    }
+    if(dim(offspring)[1] > K){
+        offspring <- offspring[1:K,];
+    }
+    return(offspring);
+}
+
+# Merge the different layers into a new pest array
+build_new_pest <- function(offspring, total_offspring, pa, cr, mutation = 0.01){
+    new_PEST   <- matrix(data = 0, nrow = total_offspring, ncol = 8);
+    start_row  <- 1; # Again, avoids an rbind memory issue
+    for(cell in 1:length(offspring)){
+        if(length(offspring[[cell]]) > 0){
+            cell_offs       <- dim(offspring[[cell]])[1];
+            rows            <- start_row:(start_row + cell_offs - 1);
+            new_PEST[rows,] <- offspring[[cell]];
+            start_row       <- start_row + cell_offs;
+        }
+    }
+    mu5 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    mu6 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    mu7 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    mu8 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    new_PEST[mu5, 5] <- sample(x = 1:cr, size = length(mu5), replace = TRUE);
+    new_PEST[mu6, 6] <- sample(x = 1:cr, size = length(mu6), replace = TRUE);
+    new_PEST[mu7, 7] <- sample(x = 1:pa, size = length(mu7), replace = TRUE);
+    new_PEST[mu8, 8] <- sample(x = 1:pa, size = length(mu8), replace = TRUE);
+    return(new_PEST);
+}
+```
+
+<a name="skip_AUG_code">Okay</a>, so what can the above code do?
+I'll get some fancy results later, but for now, let's just look at many simulations across different `pathogens` and `crops` to see how it affects pest global density.
+Below I run and plot the situation with just one pathogen and one crop.
+Note that when there is only 1 pathogen or crop, set rotation to `static`.
+
+```{r}
+sim1 <- toy_simulate_resistance(pathogens = 1, crops = 1, cell_K = 2000, 
+                                pest_init = 2000, fecundity = 8, 
+                                generations = 20, pest_move_pr = 0.1, 
+                                crop_rotate = "static", path_rotate = "static");
+dat1 <- summarise_pest_data(sim1);
+plot(x = 1:length(dat1), y = dat1, type = "b", pch = 20, lwd = 2, cex.lab = 1.5,
+     cex.axis = 1.5, ylim = c(0, 8000), xlim = c(1, 19), xlab = "Generation",
+     ylab = "Pest density");
+```
+
+Next, we can see what happens when we instead have 3 pathogens under otherwise identical conditions. In each generation, one of three pathogens is randomly applied to each landscape cell.
+
+```{r}
+sim2 <- toy_simulate_resistance(pathogens = 3, crops = 1, cell_K = 2000, 
+                                pest_init = 2000, fecundity = 8, 
+                                generations = 20, pest_move_pr = 0.1, 
+                                crop_rotate = "static", path_rotate = "random");
+dat2 <- summarise_pest_data(sim2);
+plot(x = 1:length(dat2), y = dat2, type = "b", pch = 20, lwd = 2, cex.lab = 1.5,
+     cex.axis = 1.5, ylim = c(0, 8000), xlim = c(1, 19), xlab = "Generation",
+     ylab = "Pest density");
+```
+
+Next, we can see what happens when we instead have 3 pathogens under otherwise identical conditions. In each generation, one of three pathogens is randomly applied to each landscape cell.
+
+```{r}
+sim3 <- toy_simulate_resistance(pathogens = 3, crops = 3, cell_K = 2000, 
+                                pest_init = 2000, fecundity = 8, 
+                                generations = 20, pest_move_pr = 0.1, 
+                                crop_rotate = "random", path_rotate = "random");
+dat3 <- summarise_pest_data(sim3);
+plot(x = 1:length(dat3), y = dat3, type = "b", pch = 20, lwd = 2, cex.lab = 1.5,
+     cex.axis = 1.5, ylim = c(0, 8000), xlim = c(1, 19), xlab = "Generation",
+     ylab = "Pest density");
+```
+
+As is evident, pathogen and crop rotation decreases pest density.
+More on this later, including diversity results, but this demonstrates a working toy model.
+As the heterogeneity of pathogens and crops increases, pest resistance seems to go down.
 

### 2018-08-14 R/toy_model.R
Summarise allele frequencies

@@ -29,7 +29,7 @@ toy_simulate_resistance <- function(generations = 20,       # Generations to sim
     LAND <- toy_initialise_land(xdim  = xdim, ydim = ydim, 
                                 pathogens = pathogens, crops = crops);
     PEST <- toy_initialise_pest(LAND, N = pest_init, p_al = path_alleles, 
-                                crop_alleles = crops);
+                                c_al = crop_alleles);
     # Start the generations
     PEST_DATA   <- NULL;
     gen         <- 1;
@@ -58,9 +58,26 @@ toy_simulate_resistance <- function(generations = 20,       # Generations to sim
 
 summarise_pest_data <- function(PEST_DATA){
     # Density estimates
-    den_list <- unlist(lapply(PEST_DATA, dim));
-    den_list <- den_list[c(TRUE, FALSE)];
-    return(den_list);
+    den_list  <- unlist(lapply(PEST_DATA, dim));
+    den_vect  <- den_list[c(TRUE, FALSE)];
+    gens      <- length(PEST_DATA);
+    p_alleles <- max(c(PEST_DATA[[1]][,5], PEST_DATA[[1]][,6]));
+    c_alleles <- max(c(PEST_DATA[[1]][,7], PEST_DATA[[1]][,8])); 
+    p_tabl    <- matrix(data = 0, nrow = gens, ncol = p_alleles);
+    c_tabl    <- matrix(data = 0, nrow = gens, ncol = c_alleles);
+    # Allele frequencies
+    for(gen in 1:length(PEST_DATA)){
+        for(allele in 1:p_alleles){
+            p_tabl[gen, allele]  <- sum(PEST_DATA[[gen]][,5:6] == allele);
+        }
+        for(allele in 1:c_alleles){
+            c_tabl[gen, allele]  <- sum(PEST_DATA[[gen]][,7:8] == allele);
+        }
+    }
+    p_sum <- apply(X = p_tabl, MARGIN = 1, FUN = sum);
+    c_sum <- apply(X = c_tabl, MARGIN = 1, FUN = sum);
+    
+    return(list(densities = den_vect, pathogen_fr = p_tabl, crop_fr = c_tabl));
 }
 
 toy_check_extinction <- function(PEST, gen){

### 2018-08-14 R/toy_model.R
Minor edits

@@ -67,16 +67,17 @@ summarise_pest_data <- function(PEST_DATA){
     c_tabl    <- matrix(data = 0, nrow = gens, ncol = c_alleles);
     # Allele frequencies
     for(gen in 1:length(PEST_DATA)){
+        total_p_alleles <- length(PEST_DATA[[gen]][,5:6]);
+        total_c_alleles <- length(PEST_DATA[[gen]][,7:8]);
         for(allele in 1:p_alleles){
-            p_tabl[gen, allele]  <- sum(PEST_DATA[[gen]][,5:6] == allele);
+            allele_count         <- sum(PEST_DATA[[gen]][,5:6] == allele);
+            p_tabl[gen, allele]  <- allele_count / total_p_alleles
         }
         for(allele in 1:c_alleles){
-            c_tabl[gen, allele]  <- sum(PEST_DATA[[gen]][,7:8] == allele);
+            allele_count         <- sum(PEST_DATA[[gen]][,7:8] == allele);
+            c_tabl[gen, allele]  <- allele_count / total_c_alleles
         }
     }
-    p_sum <- apply(X = p_tabl, MARGIN = 1, FUN = sum);
-    c_sum <- apply(X = c_tabl, MARGIN = 1, FUN = sum);
-    
     return(list(densities = den_vect, pathogen_fr = p_tabl, crop_fr = c_tabl));
 }
 

### 2018-08-14 notebook/helicoverpa_notes.Rmd
Fix link
Update notebook

@@
 
-     - **AUG** [13](#u-13-AUG-2018)
+     - **AUG** [13](#u-13-AUG-2018), [14](#u-14-AUG-2018)
      - **JUL** [18](#u-18-JUL-2018), [23](#u-23-JUL-2018), [30](#u-30-JUL-2018)
@@
 
-> <a name="u-18-AUG-2018">Update: 18 AUG 2018</a>
+> <a name="u-14-AUG-2018">Update: 14 AUG 2018</a>
+
+```{r, echo = FALSE}
+toy_simulate_resistance <- function(generations = 20,       # Generations to sim
+                                    xdim = 2,               # Land dimension 1
+                                    ydim = 2,               # Land dimension 2
+                                    pathogens = 1,          # Pathogen strains
+                                    crops = 1,              # Crop species
+                                    path_alleles = 3,       # Pathogen alleles
+                                    crop_alleles = 3,       # Crop alleles
+                                    pest_init = 2000,       # Initial pests 
+                                    crop_rotate = "static", # Crops rotated
+                                    path_rotate = "static", # Pathogens rotated
+                                    pest_move_pr = 0.1,     # Pest movement
+                                    pest_move_dist = 1,     # Pest move distance
+                                    fecundity = 8,          # Offspring per fem
+                                    cell_K = 2000           # K per cell
+                                    ){
+
+    if(pest_move_dist > xdim & pest_move_dist > ydim){
+        pest_move_dist <- max(c(xdim, ydim)); # Avoids error
+    }
+    # Start initialising the landscape and pests
+    LAND <- toy_initialise_land(xdim  = xdim, ydim = ydim, 
+                                pathogens = pathogens, crops = crops);
+    PEST <- toy_initialise_pest(LAND, N = pest_init, p_al = path_alleles, 
+                                c_al = crop_alleles);
+    # Start the generations
+    PEST_DATA   <- NULL;
+    gen         <- 1;
+    while(gen < generations){
+        LAND <- toy_set_crops(LAND, crops, crop_rotate);                        
+        LAND <- toy_set_paths(LAND, pathogens, path_rotate);                    
+        PEST <- toy_move_pest(PEST, LAND, pest_move_pr, pest_move_dist);        
+        PEST <- toy_feed_pest(PEST, LAND);                                      
+        if(toy_check_extinction(PEST, gen) == TRUE){ # Hate these if breaks here
+            break;
+        }
+        PEST <- toy_kill_pest(PEST, LAND);
+        if(toy_check_extinction(PEST, gen) == TRUE){
+            break;
+        }
+        PEST <- toy_reproduce_pest(PEST, LAND, path_alleles, crop_alleles, 
+                                   fecundity, cell_K);
+        if(toy_check_extinction(PEST, gen) == TRUE){
+            break;
+        }
+        PEST_DATA[[gen]] <- PEST;
+        gen <- gen + 1;
+    }
+    return(PEST_DATA);
+}
+
+summarise_pest_data <- function(PEST_DATA){
+    # Density estimates
+    den_list  <- unlist(lapply(PEST_DATA, dim));
+    den_vect  <- den_list[c(TRUE, FALSE)];
+    gens      <- length(PEST_DATA);
+    p_alleles <- max(c(PEST_DATA[[1]][,5], PEST_DATA[[1]][,6]));
+    c_alleles <- max(c(PEST_DATA[[1]][,7], PEST_DATA[[1]][,8])); 
+    p_tabl    <- matrix(data = 0, nrow = gens, ncol = p_alleles);
+    c_tabl    <- matrix(data = 0, nrow = gens, ncol = c_alleles);
+    # Allele frequencies
+    for(gen in 1:length(PEST_DATA)){
+        total_p_alleles <- length(PEST_DATA[[gen]][,5:6]);
+        total_c_alleles <- length(PEST_DATA[[gen]][,7:8]);
+        for(allele in 1:p_alleles){
+            allele_count         <- sum(PEST_DATA[[gen]][,5:6] == allele);
+            p_tabl[gen, allele]  <- allele_count / total_p_alleles
+        }
+        for(allele in 1:c_alleles){
+            allele_count         <- sum(PEST_DATA[[gen]][,7:8] == allele);
+            c_tabl[gen, allele]  <- allele_count / total_c_alleles
+        }
+    }
+    return(list(densities = den_vect, pathogen_fr = p_tabl, crop_fr = c_tabl));
+}
+
+toy_check_extinction <- function(PEST, gen){
+    if(is.vector(PEST) == TRUE){
+        print(paste("Extinction occurred in generation", gen));
+        return(TRUE);
+    }
+    if(dim(PEST)[1] < 4){
+        print(paste("Extinction occurred in generation", gen));
+        return(TRUE);
+    }
+    if(sum(PEST[,2] == 0) < 1 | sum(PEST[,2] == 1) < 1){
+        print(paste("Extinction occurred in generation", gen));
+        return(TRUE);
+    }
+    return(FALSE);
+}
+
+# Initialise the landscape
+toy_initialise_land <- function(xdim = 2, ydim = 2, pathogens = 1, crops = 1){
+    LAND      <- array(data = 0, dim = c(xdim, ydim, 3));
+    p_values  <- sample(x = 1:pathogens, size = xdim * ydim, replace = TRUE);
+    p_layer   <- matrix(data = p_values, ncol = xdim, nrow = ydim);
+    c_values  <- sample(x = 1:crops, size = xdim * ydim, replace = TRUE);
+    c_layer   <- matrix(data = c_values, ncol = xdim, nrow = ydim);
+    LAND[,,2] <- p_layer; # Just occurred to me that we might want a layer for
+    LAND[,,3] <- c_layer; # carrying capacity of the cell in LAND later
+    return(LAND);
+}
+
+# Initialise some *very* simple pests. Each allele is just going to map one to
+# one for whether a crop can be attacked or a pathogen resisted. I am not even
+# going to add traits affecting dispersal in the toy version -- this can be
+# uniform for all individuals
+toy_initialise_pest <- function(LAND, N = 10, p_al = 1, c_al = 1){
+    xdim     <- dim(LAND)[1]; # This is needed to put the pests on some
+    ydim     <- dim(LAND)[2]; # landscape cell
+    PEST     <- matrix(data = 0, nrow = N, ncol = 8);
+    PEST[,1] <- 1:N;                                           # ID
+    PEST[,2] <- sample(x = c(0, 1), size = N, replace = TRUE); # Sex
+    PEST[,3] <- sample(x = 1:xdim,  size = N, replace = TRUE); # x-location
+    PEST[,4] <- sample(x = 1:ydim,  size = N, replace = TRUE); # y-location
+    PEST[,5] <- sample(x = 1:p_al,  size = N, replace = TRUE); # p allele 1
+    PEST[,6] <- sample(x = 1:p_al,  size = N, replace = TRUE); # p allele 2
+    PEST[,7] <- sample(x = 1:c_al,  size = N, replace = TRUE); # c allele 1
+    PEST[,8] <- sample(x = 1:c_al,  size = N, replace = TRUE); # c allele 2
+    return(PEST);
+}
+
+# Just going to make this a simple function at first, randomly changing crops
+toy_set_crops <- function(LAND, crops = 1, type = "static"){
+    if(crops == 1 | type == "static"){
+        return(LAND);
+    }
+    if(type == "rotate"){
+        old_crop                   <- LAND[,,3];
+        new_crop                   <- old_crop + 1;
+        new_crop[new_crop > crops] <- 1;
+        LAND[,,3]                  <- new_crop;
+    }else{ # Else it just randomises -- can think about fancier ways later
+        xdim      <- dim(LAND)[1];
+        ydim      <- dim(LAND)[2];
+        new_cval  <- sample(x = 1:crops, size = xdim * ydim, replace = TRUE);
+        new_crop  <- matrix(data = new_cval, nrow = xdim, ncol = ydim);
+        LAND[,,3] <- new_crop;
+    }
+    return(LAND);
+}
+
+# Ditto here -- just a simple function changing pathogens
+toy_set_paths <- function(LAND, paths = 1, type = "static"){
+    if(paths == 1 | type == "static"){
+        return(LAND);
+    }
+    if(type == "rotate"){
+        old_path                   <- LAND[,,3];
+        new_path                   <- old_path + 1;
+        new_path[new_path > paths] <- 1;
+        LAND[,,2]                  <- new_path;
+    }else{ # Else it just randomises -- can think about fancier ways later
+        xdim      <- dim(LAND)[1];
+        ydim      <- dim(LAND)[2];
+        new_pval  <- sample(x = 1:paths, size = xdim * ydim, replace = TRUE);
+        new_path  <- matrix(data = new_pval, nrow = xdim, ncol = ydim);
+        LAND[,,2] <- new_path;
+    }
+    return(LAND);
+}
+
+# Just a function to move pests at a given value, randomly on the landscape. The
+# `prob` is just probability of moving in a time step, while `dist` is just the
+# maximum distance moved in cells, in any direction.
+toy_move_pest <- function(PEST, LAND, prob = 0.1, dist = 1){
+    xdim                     <- dim(LAND)[1];
+    ydim                     <- dim(LAND)[2];
+    pests                    <- dim(PEST)[1];
+    to_move                  <- rbinom(n = pests, size = 1, pr = prob);
+    move_x                   <- sample(x = -dist:dist, size = pests, 
+                                       replace = TRUE);
+    move_y                   <- sample(x = -dist:dist, size = pests, 
+                                       replace = TRUE);
+    PEST[to_move == 1, 3]    <- PEST[to_move == 1, 3] + move_x[to_move == 1];
+    PEST[to_move == 1, 4]    <- PEST[to_move == 1, 4] + move_y[to_move == 1];
+    # The ifs below make a torus landscape so that pests don't move off of it
+    # In C, this will be done with a loop that looks cleaner; the below is just
+    # avoiding using a loop in R, though it would probably be fine.
+    if(length(PEST[PEST[,3] < 1, 3]) > 0){ 
+        PEST[PEST[,3] < 1, 3] <- PEST[PEST[,3] < 1, 3] + xdim;
+    }
+    if(length(PEST[PEST[,3] > xdim, 3]) > 0){
+        PEST[PEST[,3] > xdim, 3] <- PEST[PEST[,3] > xdim, 3] - xdim;
+    }
+    if(length(PEST[PEST[,4] < 1, 4]) > 0){
+        PEST[PEST[,4] < 1, 4] <- PEST[PEST[,4] < 1, 4] + ydim;
+    }
+    if(length(PEST[PEST[,4] > ydim, 4]) > 0){
+        PEST[PEST[,4] > ydim, 4] <- PEST[PEST[,4] > ydim, 4] - ydim;
+    }
+    return(PEST);
+}
+
+# Need to now feed the pests and remove those that don't feed (crop unavailable)
+toy_feed_pest <- function(PEST, LAND){
+    # I'm just going to use a loop here, else matching to LAND cells is rough
+    pests <- dim(PEST)[1];
+    eaten <- rep(x = 0, times = pests); 
+    for(i in 1:pests){
+        x_loc <- PEST[i, 3];
+        y_loc <- PEST[i, 4];
+        food  <- LAND[x_loc, y_loc, 3];
+        if(PEST[i, 7] == food | PEST[i, 8] == food){
+            eaten[i] <- 1;
+        }
+    }
+    PEST <- PEST[eaten == 1,]; # You don't eat, you don't live
+    return(PEST);
+}
+
+# Then need to have pests resist pathogens, kill those that cannot
+toy_kill_pest <- function(PEST, LAND){
+    # I'm just going to use a loop here, else matching to LAND cells is rough
+    pests    <- dim(PEST)[1];
+    survived <- rep(x = 0, times = pests);
+    for(i in 1:pests){
+        x_loc <- PEST[i, 3];
+        y_loc <- PEST[i, 4];
+        patho <- LAND[x_loc, y_loc, 2];
+        if(PEST[i, 5] == patho | PEST[i, 6] == patho){
+            survived[i] <- 1;
+        }
+    }
+    PEST <- PEST[survived == 1,]; # You don't eat, you don't live
+    return(PEST);
+}
+
+# Then, reproduce pests that are left
+toy_reproduce_pest <- function(PEST, LAND, pa, cr, births = 2, K = 100){
+    x_dim           <- dim(LAND)[1];
+    y_dim           <- dim(LAND)[2];
+    offspring       <- NULL;
+    total_offspring <- 0; # This and the below are just to avoid an rbind(),
+    cell            <- 1; # which is a massive memory sink
+    lst_ID          <- max(PEST[,1]);
+    for(xloc in 1:x_dim){ # Double loop is just much cleaner here
+        for(yloc in 1:y_dim){
+            locals     <- which(PEST[,3] == xloc & PEST[,4] == yloc);
+            local_offs <- NULL;
+            if( length(locals) > 1 ){
+                local_offs <- toy_breed_locals(PEST, locals, births, K, lst_ID);
+            }
+            if(length(local_offs) > 0){
+                lst_ID            <- lst_ID + dim(local_offs)[1];
+                total_offspring   <- total_offspring + dim(local_offs)[1];
+                offspring[[cell]] <- local_offs;
+            }else{
+                offspring[[cell]] <- NULL;
+            }
+            cell              <- cell + 1;
+        }
+    }
+    offspring <- build_new_pest(offspring, total_offspring, pa, cr);
+    return(offspring);
+}
+
+# Need to recombine the genomes correctly in the offspring
+toy_breed_locals <- function(PEST, locals, births, K, last_ID){
+    loc_PEST <- PEST[locals,]; # Need two of each sex (allee effect)
+    if(sum(loc_PEST[,2] == 0) < 2 | sum(loc_PEST[,2] == 1) < 2){
+        return(NULL);
+    }
+    females       <- loc_PEST[loc_PEST[,2] == 0,];
+    males         <- loc_PEST[loc_PEST[,2] == 1,];
+    new_offs      <- dim(females)[1] * floor(births);  
+    offspring     <- matrix(data = 0, nrow = new_offs, ncol = 8);     
+    offspring[,1] <- (last_ID + 1):(last_ID + new_offs);
+    offspring[,2] <- sample(x = c(0, 1), size = new_offs, replace = TRUE);
+    offspring[,3] <- loc_PEST[1, 3];
+    offspring[,4] <- loc_PEST[1, 4];
+    # Below grabs all of the alleles from females and males
+    p_fem_alleles <- c(females[,5], females[,6]);
+    p_mal_alleles <- c(males[,5], males[,6]);
+    c_fem_alleles <- c(females[,7], females[,8]);
+    c_mal_alleles <- c(males[,7], males[,8]);
+    # Now add them to the offspring randomly, one from female and one from male
+    for(i in 1:new_offs){
+        if(runif(n = 1) < 0.5){
+            offspring[i, 5] <- sample(x = p_fem_alleles, size = 1);
+            offspring[i, 6] <- sample(x = p_mal_alleles, size = 1);
+        }else{
+            offspring[i, 5] <- sample(x = p_mal_alleles, size = 1);
+            offspring[i, 6] <- sample(x = p_fem_alleles, size = 1);            
+        }
+        if(runif(n = 1) < 0.5){
+            offspring[i, 7] <- sample(x = c_fem_alleles, size = 1);
+            offspring[i, 8] <- sample(x = c_mal_alleles, size = 1);
+        }else{
+            offspring[i, 7] <- sample(x = c_mal_alleles, size = 1);
+            offspring[i, 8] <- sample(x = c_fem_alleles, size = 1);            
+        }
+    }
+    if(dim(offspring)[1] > K){
+        offspring <- offspring[1:K,];
+    }
+    return(offspring);
+}
+
+# Merge the different layers into a new pest array
+build_new_pest <- function(offspring, total_offspring, pa, cr, mutation = 0.01){
+    new_PEST   <- matrix(data = 0, nrow = total_offspring, ncol = 8);
+    start_row  <- 1; # Again, avoids an rbind memory issue
+    for(cell in 1:length(offspring)){
+        if(length(offspring[[cell]]) > 0){
+            cell_offs       <- dim(offspring[[cell]])[1];
+            rows            <- start_row:(start_row + cell_offs - 1);
+            new_PEST[rows,] <- offspring[[cell]];
+            start_row       <- start_row + cell_offs;
+        }
+    }
+    mu5 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    mu6 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    mu7 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    mu8 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    new_PEST[mu5, 5] <- sample(x = 1:cr, size = length(mu5), replace = TRUE);
+    new_PEST[mu6, 6] <- sample(x = 1:cr, size = length(mu6), replace = TRUE);
+    new_PEST[mu7, 7] <- sample(x = 1:pa, size = length(mu7), replace = TRUE);
+    new_PEST[mu8, 8] <- sample(x = 1:pa, size = length(mu8), replace = TRUE);
+    return(new_PEST);
+}
+```
+
+**More playing around with the toy model**
+
+Following up on [yesterday](#skip_AUG_code), I have tweaked two functions to allow us to manipulate the number of alleles underlying pathogen resistance and crop feeding.
+In effect, this allows us to see how selection can sweep through to nearly fix an allele for resistance to a particular pathogen or for feeding on a particular crop (since only one allele is needed, other neutral alleles can hide in the heterozygotes).
+But, when crops are rotated, or pathogens are rotated, we have higher genetic diversity -- note that we are making an implicit assumption of a strong negative genetic correlation; pests can only have two alleles for resistance, e.g., and if there are three pathogens, then all pests must necessarily be susceptible to 1-2 pathogens.
+
+```{r}
+# ==============================================================================
+# This is just a toy version of the model, for the presentation
+# ==============================================================================
+# Note that I'm adding 'toy' at the start of each function; this is just in case
+# we actually want to keep some of these functions in the eventual R package, 
+# but also don't want to worry about not being able to name functions like
+# "initialise_land" the same here as elsewhere in the eventual bigger model.
+# ==============================================================================
+toy_simulate_resistance <- function(generations = 20,       # Generations to sim
+                                    xdim = 2,               # Land dimension 1
+                                    ydim = 2,               # Land dimension 2
+                                    pathogens = 1,          # Pathogen strains
+                                    crops = 1,              # Crop species
+                                    path_alleles = 3,       # Pathogen alleles
+                                    crop_alleles = 3,       # Crop alleles
+                                    pest_init = 2000,       # Initial pests 
+                                    crop_rotate = "static", # Crops rotated
+                                    path_rotate = "static", # Pathogens rotated
+                                    pest_move_pr = 0.1,     # Pest movement
+                                    pest_move_dist = 1,     # Pest move distance
+                                    fecundity = 8,          # Offspring per fem
+                                    cell_K = 2000           # K per cell
+                                    ){
+
+    if(pest_move_dist > xdim & pest_move_dist > ydim){
+        pest_move_dist <- max(c(xdim, ydim)); # Avoids error
+    }
+    # Start initialising the landscape and pests
+    LAND <- toy_initialise_land(xdim  = xdim, ydim = ydim, 
+                                pathogens = pathogens, crops = crops);
+    PEST <- toy_initialise_pest(LAND, N = pest_init, p_al = path_alleles, 
+                                c_al = crop_alleles);
+    # Start the generations
+    PEST_DATA   <- NULL;
+    gen         <- 1;
+    while(gen < generations){
+        LAND <- toy_set_crops(LAND, crops, crop_rotate);                        
+        LAND <- toy_set_paths(LAND, pathogens, path_rotate);                    
+        PEST <- toy_move_pest(PEST, LAND, pest_move_pr, pest_move_dist);        
+        PEST <- toy_feed_pest(PEST, LAND);                                      
+        if(toy_check_extinction(PEST, gen) == TRUE){ # Hate these if breaks here
+            break;
+        }
+        PEST <- toy_kill_pest(PEST, LAND);
+        if(toy_check_extinction(PEST, gen) == TRUE){
+            break;
+        }
+        PEST <- toy_reproduce_pest(PEST, LAND, path_alleles, crop_alleles, 
+                                   fecundity, cell_K);
+        if(toy_check_extinction(PEST, gen) == TRUE){
+            break;
+        }
+        PEST_DATA[[gen]] <- PEST;
+        gen <- gen + 1;
+    }
+    return(PEST_DATA);
+}
+
+summarise_pest_data <- function(PEST_DATA){
+    # Density estimates
+    den_list  <- unlist(lapply(PEST_DATA, dim));
+    den_vect  <- den_list[c(TRUE, FALSE)];
+    gens      <- length(PEST_DATA);
+    p_alleles <- max(c(PEST_DATA[[1]][,5], PEST_DATA[[1]][,6]));
+    c_alleles <- max(c(PEST_DATA[[1]][,7], PEST_DATA[[1]][,8])); 
+    p_tabl    <- matrix(data = 0, nrow = gens, ncol = p_alleles);
+    c_tabl    <- matrix(data = 0, nrow = gens, ncol = c_alleles);
+    # Allele frequencies
+    for(gen in 1:length(PEST_DATA)){
+        total_p_alleles <- length(PEST_DATA[[gen]][,5:6]);
+        total_c_alleles <- length(PEST_DATA[[gen]][,7:8]);
+        for(allele in 1:p_alleles){
+            allele_count         <- sum(PEST_DATA[[gen]][,5:6] == allele);
+            p_tabl[gen, allele]  <- allele_count / total_p_alleles
+        }
+        for(allele in 1:c_alleles){
+            allele_count         <- sum(PEST_DATA[[gen]][,7:8] == allele);
+            c_tabl[gen, allele]  <- allele_count / total_c_alleles
+        }
+    }
+    return(list(densities = den_vect, pathogen_fr = p_tabl, crop_fr = c_tabl));
+}
+```
+
+I'm going to run the same analysis as [yesterday](#skip_AUG_code), but now show how allele frequencies change over generations.
+I've not broken this down by landscape cell yet, but this will be easy to do (just let me know the format the data would be best presented in).
+In each case, now there will be 3 alleles underlying pathogen resistance, and 3 alleles underlying crop feeding.
+
+```{r}
+sim1 <- toy_simulate_resistance(pathogens = 1, crops = 1, cell_K = 2000, 
+                                pest_init = 2000, fecundity = 8, 
+                                generations = 20, pest_move_pr = 0.1, 
+                                crop_rotate = "static", path_rotate = "static");
+dat1 <- summarise_pest_data(sim1);
+plot(x = 1:length(dat1$densities), y = dat1$densities, type = "b", pch = 20, 
+     lwd = 2, cex.lab = 1.5, cex.axis = 1.5, ylim = c(0, 8000), xlim = c(1, 19),
+     xlab = "Generation", ylab = "Pest density");
+plot(x = 1:length(dat1$pathogen_fr[,1]), y = dat1$pathogen_fr[,1], type = "b", 
+     pch = 20, lwd = 2, cex.lab = 1.5, cex.axis = 1.5, ylim = c(0, 1), 
+     xlim = c(1, 19), xlab = "Generation", ylab = "Allele frequency");
+points(x = 1:length(dat1$pathogen_fr[,2]), y = dat1$pathogen_fr[,2], type = "b", 
+     pch = 20, lwd = 2, col = "red");
+points(x = 1:length(dat1$pathogen_fr[,3]), y = dat1$pathogen_fr[,3], type = "b", 
+     pch = 20, lwd = 2, col = "blue");
+```
+
+Next, we can see what happens when we instead have 3 pathogens under otherwise identical conditions. In each generation, one of three pathogens is randomly applied to each landscape cell.
+
+```{r}
+sim2 <- toy_simulate_resistance(pathogens = 3, crops = 1, cell_K = 2000, 
+                                pest_init = 2000, fecundity = 8, 
+                                generations = 20, pest_move_pr = 0.1, 
+                                crop_rotate = "static", path_rotate = "random");
+dat2 <- summarise_pest_data(sim2);
+plot(x = 1:length(dat2$densities), y = dat2$densities, type = "b", pch = 20, 
+     lwd = 2, cex.lab = 1.5, cex.axis = 1.5, ylim = c(0, 8000), xlim = c(1, 19),
+     xlab = "Generation", ylab = "Pest density");
+plot(x = 1:length(dat2$pathogen_fr[,1]), y = dat2$pathogen_fr[,1], type = "b", 
+     pch = 20, lwd = 2, cex.lab = 1.5, cex.axis = 1.5, ylim = c(0, 1), 
+     xlim = c(1, 19), xlab = "Generation", ylab = "Allele frequency");
+points(x = 1:length(dat2$pathogen_fr[,2]), y = dat2$pathogen_fr[,2], type = "b", 
+     pch = 20, lwd = 2, col = "red");
+points(x = 1:length(dat2$pathogen_fr[,3]), y = dat2$pathogen_fr[,3], type = "b", 
+     pch = 20, lwd = 2, col = "blue");
+```
+
+Next, we can see what happens when we instead have 3 pathogens *and* 3 crops under otherwise identical conditions.
+In each generation, one of three pathogens and one of three crops are randomly applied to each landscape cell.
+
+```{r}
+sim3 <- toy_simulate_resistance(pathogens = 3, crops = 3, cell_K = 2000, 
+                                pest_init = 2000, fecundity = 8, 
+                                generations = 20, pest_move_pr = 0.1, 
+                                crop_rotate = "random", path_rotate = "random");
+dat3 <- summarise_pest_data(sim3);
+plot(x = 1:length(dat3$densities), y = dat3$densities, type = "b", pch = 20, 
+     lwd = 2, cex.lab = 1.5, cex.axis = 1.5, ylim = c(0, 8000), xlim = c(1, 19),
+     xlab = "Generation", ylab = "Pest density");
+plot(x = 1:length(dat3$pathogen_fr[,1]), y = dat3$pathogen_fr[,1], type = "b", 
+     pch = 20, lwd = 2, cex.lab = 1.5, cex.axis = 1.5, ylim = c(0, 1), 
+     xlim = c(1, 19), xlab = "Generation", ylab = "Allele frequency");
+points(x = 1:length(dat3$pathogen_fr[,2]), y = dat3$pathogen_fr[,2], type = "b", 
+     pch = 20, lwd = 2, col = "red");
+points(x = 1:length(dat3$pathogen_fr[,3]), y = dat3$pathogen_fr[,3], type = "b", 
+     pch = 20, lwd = 2, col = "blue");
+```
+
+As is evident, pathogen and crop rotation decreases pest density and increases the diversity of alleles underlying pathogen resistance and crop feeding.
+
+> <a name="u-13-AUG-2018">Update: 13 AUG 2018</a>
 
@@
 
-Next, we can see what happens when we instead have 3 pathogens under otherwise identical conditions. In each generation, one of three pathogens is randomly applied to each landscape cell.
+Next, we can see what happens when we instead have 3 pathogens *and* 3 crops under otherwise identical conditions.
+In each generation, one of three pathogens and one of three crops are randomly applied to each landscape cell.
 

### 2018-08-27 assets/dist/scripts/main.js
improved sidebar transistions and device compatability

@@ -1,5 +1,5 @@
 !function(t,e){"function"==typeof define&&define.amd?define([],e):"object"==typeof exports?module.exports=e():t.Tether=e()}(this,function(){"use strict";function t(t,e){if(!(t instanceof e))throw new TypeError("Cannot call a class as a function")}function e(t){var i=t.getBoundingClientRect(),n={};for(var s in i)n[s]=i[s];if(t.ownerDocument!==document){var o=t.ownerDocument.defaultView.frameElement;if(o){var r=e(o);n.top+=r.top,n.bottom+=r.top,n.left+=r.left,n.right+=r.left}}return n}function i(t){var e=getComputedStyle(t)||{},i=e.position,n=[];if("fixed"===i)return[t];for(var s=t;(s=s.parentNode)&&s&&1===s.nodeType;){var o=void 0;try{o=getComputedStyle(s)}catch(t){}if(void 0===o||null===o)return n.push(s),n;var r=o,a=r.overflow,l=r.overflowX;/(auto|scroll|overlay)/.test(a+r.overflowY+l)&&("absolute"!==i||["relative","absolute","fixed"].indexOf(o.position)>=0)&&n.push(s)}return n.push(t.ownerDocument.body),t.ownerDocument!==document&&n.push(t.ownerDocument.defaultView),n}function n(){C&&document.body.removeChild(C),C=null}function s(t){var i=void 0;t===document?(i=document,t=document.documentElement):i=t.ownerDocument;var n=i.documentElement,s=e(t),o=A();return s.top-=o.top,s.left-=o.left,void 0===s.width&&(s.width=document.body.scrollWidth-s.left-s.right),void 0===s.height&&(s.height=document.body.scrollHeight-s.top-s.bottom),s.top=s.top-n.clientTop,s.left=s.left-n.clientLeft,s.right=i.body.clientWidth-s.width-s.left,s.bottom=i.body.clientHeight-s.height-s.top,s}function o(t){return t.offsetParent||document.documentElement}function r(){if(S)return S;var t=document.createElement("div");t.style.width="100%",t.style.height="200px";var e=document.createElement("div");a(e.style,{position:"absolute",top:0,left:0,pointerEvents:"none",visibility:"hidden",width:"200px",height:"150px",overflow:"hidden"}),e.appendChild(t),document.body.appendChild(e);var i=t.offsetWidth;e.style.overflo
[…]

### 2018-08-27 assets/scripts/main.js

@@ -82,7 +82,7 @@
 				$grid.masonry();
 				// the sidebar
 				$('#toggle-controls').click(function(){			
-		      $('.wrapper').toggleClass('toggle');
+		      $('.wrapper').toggleClass('open-controls');
 		      $('.wrapper #content').one('webkitTransitionEnd otransitionend oTransitionEnd msTransitionEnd transitionend', function(e) {
 				    // rearrarange helicoverpa when the transistion ends
 						$grid.masonry();

### 2018-08-27 R/toy_model.R
Update toy model with json function

@@ -56,6 +56,24 @@ toy_simulate_resistance <- function(generations = 20,       # Generations to sim
     return(PEST_DATA);
 }
 
+results_to_json <- function(sim, printit = TRUE, filename = "sim.json"){
+    if("package:jsonlite" %in% search() == FALSE){
+        stop("Error: Need to load the R package 'jsonlite'")
+    }
+    colnames(sim) <- c("ID", "sex", "xloc", "yloc", "p_al_1", "p_al_2", 
+                       "c_al_1", "c_al_2");
+    modsim <- list( traits = colnames(sim), 
+                    values = unname(apply(sim, 1, 
+                                          function(x) as.data.frame(t(x))))
+    );
+    sim_json <- toJSON(list(traits = names(modsim), values = modsim), 
+                       pretty = TRUE);
+    if(printit == TRUE){
+        write(sim_json, filename);
+    }
+    return(sim_json);
+}
+
 summarise_pest_data <- function(PEST_DATA){
     # Density estimates
     den_list  <- unlist(lapply(PEST_DATA, dim));

### 2018-08-27 notebook/helicoverpa_notes.Rmd
Update notes

@@
 
-     - **AUG** [13](#u-13-AUG-2018), [14](#u-14-AUG-2018)
+     - **AUG** [13](#u-13-AUG-2018), [14](#u-14-AUG-2018), [27](#u-27-AUG-2018)
      - **JUL** [18](#u-18-JUL-2018), [23](#u-23-JUL-2018), [30](#u-30-JUL-2018)
@@
 ================================================================================
+
+> <a name="u-27-AUG-2018">Update: 27 AUG 2018</a>
+
+```{r, echo = FALSE}
+# ==============================================================================
+# This is just a toy version of the model, for the presentation
+# ==============================================================================
+# Note that I'm adding 'toy' at the start of each function; this is just in case
+# we actually want to keep some of these functions in the eventual R package, 
+# but also don't want to worry about not being able to name functions like
+# "initialise_land" the same here as elsewhere in the eventual bigger model.
+# ==============================================================================
+toy_simulate_resistance <- function(generations = 20,       # Generations to sim
+                                    xdim = 2,               # Land dimension 1
+                                    ydim = 2,               # Land dimension 2
+                                    pathogens = 1,          # Pathogen strains
+                                    crops = 1,              # Crop species
+                                    path_alleles = 3,       # Pathogen alleles
+                                    crop_alleles = 3,       # Crop alleles
+                                    pest_init = 2000,       # Initial pests 
+                                    crop_rotate = "static", # Crops rotated
+                                    path_rotate = "static", # Pathogens rotated
+                                    pest_move_pr = 0.1,     # Pest movement
+                                    pest_move_dist = 1,     # Pest move distance
+                                    fecundity = 8,          # Offspring per fem
+                                    cell_K = 2000           # K per cell
+                                    ){
+
+    if(pest_move_dist > xdim & pest_move_dist > ydim){
+        pest_move_dist <- max(c(xdim, ydim)); # Avoids error
+    }
+    # Start initialising the landscape and pests
+    LAND <- toy_initialise_land(xdim  = xdim, ydim = ydim, 
+                                pathogens = pathogens, crops = crops);
+    PEST <- toy_initialise_pest(LAND, N = pest_init, p_al = path_alleles, 
+                                c_al = crop_alleles);
+    # Start the generations
+    PEST_DATA   <- NULL;
+    gen         <- 1;
+    while(gen < generations){
+        LAND <- toy_set_crops(LAND, crops, crop_rotate);                        
+        LAND <- toy_set_paths(LAND, pathogens, path_rotate);                    
+        PEST <- toy_move_pest(PEST, LAND, pest_move_pr, pest_move_dist);        
+        PEST <- toy_feed_pest(PEST, LAND);                                      
+        if(toy_check_extinction(PEST, gen) == TRUE){ # Hate these if breaks here
+            break;
+        }
+        PEST <- toy_kill_pest(PEST, LAND);
+        if(toy_check_extinction(PEST, gen) == TRUE){
+            break;
+        }
+        PEST <- toy_reproduce_pest(PEST, LAND, path_alleles, crop_alleles, 
+                                   fecundity, cell_K);
+        if(toy_check_extinction(PEST, gen) == TRUE){
+            break;
+        }
+        PEST_DATA[[gen]] <- PEST;
+        gen <- gen + 1;
+    }
+    return(PEST_DATA);
+}
+
+results_to_json <- function(sim, filename = "sim.json"){
+    if("package:jsonlite" %in% search() == FALSE){
+        stop("Error: Need to load the R package 'jsonlite'")
+    }
+    colnames(sim) <- c("ID", "sex", "xloc", "yloc", "p_al_1", "p_al_2", 
+                       "c_al_1", "c_al_2");
+    modsim <- list( traits = colnames(sim), 
+                    values = unname(apply(sim, 1, 
+                                          function(x) as.data.frame(t(x))))
+              );
+    sim_json <- toJSON(list(traits = names(modsim), values = modsim), 
+                   pretty = TRUE);                   
+    write(sim_json, filename);
+    return(sim_json);
+}
+
+summarise_pest_data <- function(PEST_DATA){
+    # Density estimates
+    den_list  <- unlist(lapply(PEST_DATA, dim));
+    den_vect  <- den_list[c(TRUE, FALSE)];
+    gens      <- length(PEST_DATA);
+    p_alleles <- max(c(PEST_DATA[[1]][,5], PEST_DATA[[1]][,6]));
+    c_alleles <- max(c(PEST_DATA[[1]][,7], PEST_DATA[[1]][,8])); 
+    p_tabl    <- matrix(data = 0, nrow = gens, ncol = p_alleles);
+    c_tabl    <- matrix(data = 0, nrow = gens, ncol = c_alleles);
+    # Allele frequencies
+    for(gen in 1:length(PEST_DATA)){
+        total_p_alleles <- length(PEST_DATA[[gen]][,5:6]);
+        total_c_alleles <- length(PEST_DATA[[gen]][,7:8]);
+        for(allele in 1:p_alleles){
+            allele_count         <- sum(PEST_DATA[[gen]][,5:6] == allele);
+            p_tabl[gen, allele]  <- allele_count / total_p_alleles
+        }
+        for(allele in 1:c_alleles){
+            allele_count         <- sum(PEST_DATA[[gen]][,7:8] == allele);
+            c_tabl[gen, allele]  <- allele_count / total_c_alleles
+        }
+    }
+    return(list(densities = den_vect, pathogen_fr = p_tabl, crop_fr = c_tabl));
+}
+
+toy_check_extinction <- function(PEST, gen){
+    if(is.vector(PEST) == TRUE){
+        print(paste("Extinction occurred in generation", gen));
+        return(TRUE);
+    }
+    if(dim(PEST)[1] < 4){
+        print(paste("Extinction occurred in generation", gen));
+        return(TRUE);
+    }
+    if(sum(PEST[,2] == 0) < 1 | sum(PEST[,2] == 1) < 1){
+        print(paste("Extinction occurred in generation", gen));
+        return(TRUE);
+    }
+    return(FALSE);
+}
+
+# Initialise the landscape
+toy_initialise_land <- function(xdim = 2, ydim = 2, pathogens = 1, crops = 1){
+    LAND      <- array(data = 0, dim = c(xdim, ydim, 3));
+    p_values  <- sample(x = 1:pathogens, size = xdim * ydim, replace = TRUE);
+    p_layer   <- matrix(data = p_values, ncol = xdim, nrow = ydim);
+    c_values  <- sample(x = 1:crops, size = xdim * ydim, replace = TRUE);
+    c_layer   <- matrix(data = c_values, ncol = xdim, nrow = ydim);
+    LAND[,,2] <- p_layer; # Just occurred to me that we might want a layer for
+    LAND[,,3] <- c_layer; # carrying capacity of the cell in LAND later
+    return(LAND);
+}
+
+# Initialise some *very* simple pests. Each allele is just going to map one to
+# one for whether a crop can be attacked or a pathogen resisted. I am not even
+# going to add traits affecting dispersal in the toy version -- this can be
+# uniform for all individuals
+toy_initialise_pest <- function(LAND, N = 10, p_al = 1, c_al = 1){
+    xdim     <- dim(LAND)[1]; # This is needed to put the pests on some
+    ydim     <- dim(LAND)[2]; # landscape cell
+    PEST     <- matrix(data = 0, nrow = N, ncol = 8);
+    PEST[,1] <- 1:N;                                           # ID
+    PEST[,2] <- sample(x = c(0, 1), size = N, replace = TRUE); # Sex
+    PEST[,3] <- sample(x = 1:xdim,  size = N, replace = TRUE); # x-location
+    PEST[,4] <- sample(x = 1:ydim,  size = N, replace = TRUE); # y-location
+    PEST[,5] <- sample(x = 1:p_al,  size = N, replace = TRUE); # p allele 1
+    PEST[,6] <- sample(x = 1:p_al,  size = N, replace = TRUE); # p allele 2
+    PEST[,7] <- sample(x = 1:c_al,  size = N, replace = TRUE); # c allele 1
+    PEST[,8] <- sample(x = 1:c_al,  size = N, replace = TRUE); # c allele 2
+    return(PEST);
+}
+
+# Just going to make this a simple function at first, randomly changing crops
+toy_set_crops <- function(LAND, crops = 1, type = "static"){
+    if(crops == 1 | type == "static"){
+        return(LAND);
+    }
+    if(type == "rotate"){
+        old_crop                   <- LAND[,,3];
+        new_crop                   <- old_crop + 1;
+        new_crop[new_crop > crops] <- 1;
+        LAND[,,3]                  <- new_crop;
+    }else{ # Else it just randomises -- can think about fancier ways later
+        xdim      <- dim(LAND)[1];
+        ydim      <- dim(LAND)[2];
+        new_cval  <- sample(x = 1:crops, size = xdim * ydim, replace = TRUE);
+        new_crop  <- matrix(data = new_cval, nrow = xdim, ncol = ydim);
+        LAND[,,3] <- new_crop;
+    }
+    return(LAND);
+}
+
+# Ditto here -- just a simple function changing pathogens
+toy_set_paths <- function(LAND, paths = 1, type = "static"){
+    if(paths == 1 | type == "static"){
+        return(LAND);
+    }
+    if(type == "rotate"){
+        old_path                   <- LAND[,,3];
+        new_path                   <- old_path + 1;
+        new_path[new_path > paths] <- 1;
+        LAND[,,2]                  <- new_path;
+    }else{ # Else it just randomises -- can think about fancier ways later
+        xdim      <- dim(LAND)[1];
+        ydim      <- dim(LAND)[2];
+        new_pval  <- sample(x = 1:paths, size = xdim * ydim, replace = TRUE);
+        new_path  <- matrix(data = new_pval, nrow = xdim, ncol = ydim);
+        LAND[,,2] <- new_path;
+    }
+    return(LAND);
+}
+
+# Just a function to move pests at a given value, randomly on the landscape. The
+# `prob` is just probability of moving in a time step, while `dist` is just the
+# maximum distance moved in cells, in any direction.
+toy_move_pest <- function(PEST, LAND, prob = 0.1, dist = 1){
+    xdim                     <- dim(LAND)[1];
+    ydim                     <- dim(LAND)[2];
+    pests                    <- dim(PEST)[1];
+    to_move                  <- rbinom(n = pests, size = 1, pr = prob);
+    move_x                   <- sample(x = -dist:dist, size = pests, 
+                                       replace = TRUE);
+    move_y                   <- sample(x = -dist:dist, size = pests, 
+                                       replace = TRUE);
+    PEST[to_move == 1, 3]    <- PEST[to_move == 1, 3] + move_x[to_move == 1];
+    PEST[to_move == 1, 4]    <- PEST[to_move == 1, 4] + move_y[to_move == 1];
+    # The ifs below make a torus landscape so that pests don't move off of it
+    # In C, this will be done with a loop that looks cleaner; the below is just
+    # avoiding using a loop in R, though it would probably be fine.
+    if(length(PEST[PEST[,3] < 1, 3]) > 0){ 
+        PEST[PEST[,3] < 1, 3] <- PEST[PEST[,3] < 1, 3] + xdim;
+    }
+    if(length(PEST[PEST[,3] > xdim, 3]) > 0){
+        PEST[PEST[,3] > xdim, 3] <- PEST[PEST[,3] > xdim, 3] - xdim;
+    }
+    if(length(PEST[PEST[,4] < 1, 4]) > 0){
+        PEST[PEST[,4] < 1, 4] <- PEST[PEST[,4] < 1, 4] + ydim;
+    }
+    if(length(PEST[PEST[,4] > ydim, 4]) > 0){
+        PEST[PEST[,4] > ydim, 4] <- PEST[PEST[,4] > ydim, 4] - ydim;
+    }
+    return(PEST);
+}
+
+# Need to now feed the pests and remove those that don't feed (crop unavailable)
+toy_feed_pest <- function(PEST, LAND){
+    # I'm just going to use a loop here, else matching to LAND cells is rough
+    pests <- dim(PEST)[1];
+    eaten <- rep(x = 0, times = pests); 
+    for(i in 1:pests){
+        x_loc <- PEST[i, 3];
+        y_loc <- PEST[i, 4];
+        food  <- LAND[x_loc, y_loc, 3];
+        if(PEST[i, 7] == food | PEST[i, 8] == food){
+            eaten[i] <- 1;
+        }
+    }
+    PEST <- PEST[eaten == 1,]; # You don't eat, you don't live
+    return(PEST);
+}
+
+# Then need to have pests resist pathogens, kill those that cannot
+toy_kill_pest <- function(PEST, LAND){
+    # I'm just going to use a loop here, else matching to LAND cells is rough
+    pests    <- dim(PEST)[1];
+    survived <- rep(x = 0, times = pests);
+    for(i in 1:pests){
+        x_loc <- PEST[i, 3];
+        y_loc <- PEST[i, 4];
+        patho <- LAND[x_loc, y_loc, 2];
+        if(PEST[i, 5] == patho | PEST[i, 6] == patho){
+            survived[i] <- 1;
+        }
+    }
+    PEST <- PEST[survived == 1,]; # You don't eat, you don't live
+    return(PEST);
+}
+
+# Then, reproduce pests that are left
+toy_reproduce_pest <- function(PEST, LAND, pa, cr, births = 2, K = 100){
+    x_dim           <- dim(LAND)[1];
+    y_dim           <- dim(LAND)[2];
+    offspring       <- NULL;
+    total_offspring <- 0; # This and the below are just to avoid an rbind(),
+    cell            <- 1; # which is a massive memory sink
+    lst_ID          <- max(PEST[,1]);
+    for(xloc in 1:x_dim){ # Double loop is just much cleaner here
+        for(yloc in 1:y_dim){
+            locals     <- which(PEST[,3] == xloc & PEST[,4] == yloc);
+            local_offs <- NULL;
+            if( length(locals) > 1 ){
+                local_offs <- toy_breed_locals(PEST, locals, births, K, lst_ID);
+            }
+            if(length(local_offs) > 0){
+                lst_ID            <- lst_ID + dim(local_offs)[1];
+                total_offspring   <- total_offspring + dim(local_offs)[1];
+                offspring[[cell]] <- local_offs;
+            }else{
+                offspring[[cell]] <- NULL;
+            }
+            cell              <- cell + 1;
+        }
+    }
+    offspring <- build_new_pest(offspring, total_offspring, pa, cr);
+    return(offspring);
+}
+
+# Need to recombine the genomes correctly in the offspring
+toy_breed_locals <- function(PEST, locals, births, K, last_ID){
+    loc_PEST <- PEST[locals,]; # Need two of each sex (allee effect)
+    if(sum(loc_PEST[,2] == 0) < 2 | sum(loc_PEST[,2] == 1) < 2){
+        return(NULL);
+    }
+    females       <- loc_PEST[loc_PEST[,2] == 0,];
+    males         <- loc_PEST[loc_PEST[,2] == 1,];
+    new_offs      <- dim(females)[1] * floor(births);  
+    offspring     <- matrix(data = 0, nrow = new_offs, ncol = 8);     
+    offspring[,1] <- (last_ID + 1):(last_ID + new_offs);
+    offspring[,2] <- sample(x = c(0, 1), size = new_offs, replace = TRUE);
+    offspring[,3] <- loc_PEST[1, 3];
+    offspring[,4] <- loc_PEST[1, 4];
+    # Below grabs all of the alleles from females and males
+    p_fem_alleles <- c(females[,5], females[,6]);
+    p_mal_alleles <- c(males[,5], males[,6]);
+    c_fem_alleles <- c(females[,7], females[,8]);
+    c_mal_alleles <- c(males[,7], males[,8]);
+    # Now add them to the offspring randomly, one from female and one from male
+    for(i in 1:new_offs){
+        if(runif(n = 1) < 0.5){
+            offspring[i, 5] <- sample(x = p_fem_alleles, size = 1);
+            offspring[i, 6] <- sample(x = p_mal_alleles, size = 1);
+        }else{
+            offspring[i, 5] <- sample(x = p_mal_alleles, size = 1);
+            offspring[i, 6] <- sample(x = p_fem_alleles, size = 1);            
+        }
+        if(runif(n = 1) < 0.5){
+            offspring[i, 7] <- sample(x = c_fem_alleles, size = 1);
+            offspring[i, 8] <- sample(x = c_mal_alleles, size = 1);
+        }else{
+            offspring[i, 7] <- sample(x = c_mal_alleles, size = 1);
+            offspring[i, 8] <- sample(x = c_fem_alleles, size = 1);            
+        }
+    }
+    if(dim(offspring)[1] > K){
+        offspring <- offspring[1:K,];
+    }
+    return(offspring);
+}
+
+# Merge the different layers into a new pest array
+build_new_pest <- function(offspring, total_offspring, pa, cr, mutation = 0.01){
+    new_PEST   <- matrix(data = 0, nrow = total_offspring, ncol = 8);
+    start_row  <- 1; # Again, avoids an rbind memory issue
+    for(cell in 1:length(offspring)){
+        if(length(offspring[[cell]]) > 0){
+            cell_offs       <- dim(offspring[[cell]])[1];
+            rows            <- start_row:(start_row + cell_offs - 1);
+            new_PEST[rows,] <- offspring[[cell]];
+            start_row       <- start_row + cell_offs;
+        }
+    }
+    mu5 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    mu6 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    mu7 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    mu8 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    new_PEST[mu5, 5] <- sample(x = 1:cr, size = length(mu5), replace = TRUE);
+    new_PEST[mu6, 6] <- sample(x = 1:cr, size = length(mu6), replace = TRUE);
+    new_PEST[mu7, 7] <- sample(x = 1:pa, size = length(mu7), replace = TRUE);
+    new_PEST[mu8, 8] <- sample(x = 1:pa, size = length(mu8), replace = TRUE);
+    return(new_PEST);
+}
+```
+
+**Transferring simulation results to json**
+
+I've written a short bit of code to link the simulation results to the code.
+The input for this is a single *generation* of the model (this could be easily changed).
+So if we run the `toy_simulate_resistance` function [as before](#u-14-AUG-2018), we get a list with elements being each generation.
+
+```{r}
+sim <- toy_simulate_resistance(crops = 3, pathogens = 3);
+```
+
+We can look at just the last generation below.
+
+```{r}
+head(sim[[19]]);
+```
+
+The code below sticks all of this into the json format (for the one generation), prints the json output to a file if `printit == TRUE`, and returns the json output.
+
+```{r}
+results_to_json <- function(sim, printit = TRUE, filename = "sim.json"){
+    if("package:jsonlite" %in% search() == FALSE){
+        stop("Error: Need to load the R package 'jsonlite'")
+    }
+    colnames(sim) <- c("ID", "sex", "xloc", "yloc", "p_al_1", "p_al_2", 
+                       "c_al_1", "c_al_2");
+    modsim <- list( traits = colnames(sim), 
+                    values = unname(apply(sim, 1, 
+                                          function(x) as.data.frame(t(x))))
+              );
+    sim_json <- toJSON(list(traits = names(modsim), values = modsim), 
+                   pretty = TRUE);
+    if(printit == TRUE){
+        write(sim_json, filename);
+    }
+    return(sim_json);
+}
+```
+
+Using the function below to print and return the json. This relies on the `jsonlite` package, which must be installed and loaded.
+
+```{r}
+library(jsonlite);
+sim_json <- results_to_json(sim = sim[[19]], 
+                            filename = "../data/sample_sim.json");
+```
+
+The `sim_json` output should be easy to modify as necessary.
+I've not added it here because it's much too large for the notebook, but it can be accessed on the dev branch in the file [sample_sim.json](https://github.com/bradduthie/helicoverpa/blob/dev/data/sample_sim.json).
 

### 2018-08-27 R/toy_model.R
Modify toy to get the right json output

@@ -32,6 +32,7 @@ toy_simulate_resistance <- function(generations = 20,       # Generations to sim
                                 c_al = crop_alleles);
     # Start the generations
     PEST_DATA   <- NULL;
+    LAND_DATA   <- NULL;
     gen         <- 1;
     while(gen < generations){
         LAND <- toy_set_crops(LAND, crops, crop_rotate);                        
@@ -51,19 +52,52 @@ toy_simulate_resistance <- function(generations = 20,       # Generations to sim
             break;
         }
         PEST_DATA[[gen]] <- PEST;
+        LAND_DATA[[gen]] <- LAND;
         gen <- gen + 1;
     }
-    return(PEST_DATA);
+    return(list(PEST_DATA = PEST_DATA, LAND_DATA = LAND_DATA));
 }
 
-results_to_json <- function(sim, printit = TRUE, filename = "sim.json"){
+results_to_json <- function(pest, land, printit = TRUE, filename = "sim.json"){
     if("package:jsonlite" %in% search() == FALSE){
         stop("Error: Need to load the R package 'jsonlite'")
     }
-    colnames(sim) <- c("ID", "sex", "xloc", "yloc", "p_al_1", "p_al_2", 
-                       "c_al_1", "c_al_2");
-    modsim <- list( traits = colnames(sim), 
-                    values = unname(apply(sim, 1, 
+    inds   <- dim(pest)[1];
+    p_geno <- rep(x = 0, times = inds);
+    c_geno <- rep(x = 0, times = inds);
+    path   <- rep(x = 0, times = inds);
+    crop   <- rep(x = 0, times = inds);
+    r_path <- rep(x = 0, times = inds);
+    r_crop <- rep(x = 0, times = inds);
+    for(i in 1:inds){ # Doing this a bit lazily, without refactoring the rest
+        p_geno[i] <- as.numeric( paste(pest[i,5], pest[i,6], sep = ""));
+        c_geno[i] <- as.numeric( paste(pest[i,7], pest[i,8], sep = ""));
+        xloc      <- pest[i, 3];
+        yloc      <- pest[i, 4];
+        path[i]   <- land[xloc, yloc, 2];
+        crop[i]   <- land[xloc, yloc, 3];
+        if(pest[i,5] == path[i] | pest[i,5] == path[i]){
+            r_path <- 1;
+        }
+        if(pest[i,7] == crop[i] | pest[i,8] == crop[i]){
+            r_crop <- 1;
[…]

### 2018-08-27 R/toy_model.R
Update toy model for improved json output

@@ -381,6 +381,3 @@ build_new_pest <- function(offspring, total_offspring, pa, cr, mutation = 0.01){
 }
 
 
-
-
-

### 2018-08-27 notebook/helicoverpa_notes.Rmd
Notebook update

@@
 > <a name="u-27-AUG-2018">Update: 27 AUG 2018</a>
+
+**Update to transferring to JSON**
 
@@
     PEST_DATA   <- NULL;
+    LAND_DATA   <- NULL;
+    gen         <- 1;
+    while(gen < generations){
+        LAND <- toy_set_crops(LAND, crops, crop_rotate);                        
+        LAND <- toy_set_paths(LAND, pathogens, path_rotate);                    
+        PEST <- toy_move_pest(PEST, LAND, pest_move_pr, pest_move_dist);        
+        PEST <- toy_feed_pest(PEST, LAND);                                      
+        if(toy_check_extinction(PEST, gen) == TRUE){ # Hate these if breaks here
+            break;
+        }
+        PEST <- toy_kill_pest(PEST, LAND);
+        if(toy_check_extinction(PEST, gen) == TRUE){
+            break;
+        }
+        PEST <- toy_reproduce_pest(PEST, LAND, path_alleles, crop_alleles, 
+                                   fecundity, cell_K);
+        if(toy_check_extinction(PEST, gen) == TRUE){
+            break;
+        }
+        PEST_DATA[[gen]] <- PEST;
+        LAND_DATA[[gen]] <- LAND;
+        gen <- gen + 1;
+    }
+    return(list(PEST_DATA = PEST_DATA, LAND_DATA = LAND_DATA));
+}
+
+results_to_json <- function(pest, land, printit = TRUE, filename = "sim.json"){
+    if("package:jsonlite" %in% search() == FALSE){
+        stop("Error: Need to load the R package 'jsonlite'")
+    }
+    inds   <- dim(pest)[1];
+    p_geno <- rep(x = 0, times = inds);
+    c_geno <- rep(x = 0, times = inds);
+    path   <- rep(x = 0, times = inds);
+    crop   <- rep(x = 0, times = inds);
+    r_path <- rep(x = 0, times = inds);
+    r_crop <- rep(x = 0, times = inds);
+    for(i in 1:inds){ # Doing this a bit lazily, without refactoring the rest
+        p_geno[i] <- as.numeric( paste(pest[i,5], pest[i,6], sep = ""));
+        c_geno[i] <- as.numeric( paste(pest[i,7], pest[i,8], sep = ""));
+        xloc      <- pest[i, 3];
+        yloc      <- pest[i, 4];
+        path[i]   <- land[xloc, yloc, 2];
+        crop[i]   <- land[xloc, yloc, 3];
+        if(pest[i,5] == path[i] | pest[i,5] == path[i]){
+            r_path <- 1;
+        }
+        if(pest[i,7] == crop[i] | pest[i,8] == crop[i]){
+            r_crop <- 1;
+        }
+    }
+    data      <- matrix(data = 0, nrow = inds, ncol = 10);
+    data[,1]  <- pest[,1];
+    data[,2]  <- pest[,2];
+    data[,3]  <- pest[,3];
+    data[,4]  <- pest[,4];
+    data[,5]  <- path;
+    data[,6]  <- crop;
+    data[,7]  <- p_geno;
+    data[,8]  <- c_geno;
+    data[,9]  <- r_path;
+    data[,10] <- r_crop;
+    colnames(data) <- c("ID", "sex", "xloc", "yloc", "path", "crop", 
+                        "p_geno", "c_geno", "resist_path", "eat_crop");
+    modsim <- list( traits = colnames(data), 
+                    values = unname(apply(data, 1, 
+                                          function(x) as.data.frame(t(x))))
+    );
+    sim_json <- toJSON(list(traits = names(modsim), values = modsim), 
+                       pretty = TRUE);
+    if(printit == TRUE){
+        write(sim_json, filename);
+    }
+    return(sim_json);
+}
+
+summarise_pest_data <- function(PEST_DATA){
+    # Density estimates
+    den_list  <- unlist(lapply(PEST_DATA, dim));
+    den_vect  <- den_list[c(TRUE, FALSE)];
+    gens      <- length(PEST_DATA);
+    p_alleles <- max(c(PEST_DATA[[1]][,5], PEST_DATA[[1]][,6]));
+    c_alleles <- max(c(PEST_DATA[[1]][,7], PEST_DATA[[1]][,8])); 
+    p_tabl    <- matrix(data = 0, nrow = gens, ncol = p_alleles);
+    c_tabl    <- matrix(data = 0, nrow = gens, ncol = c_alleles);
+    # Allele frequencies
+    for(gen in 1:length(PEST_DATA)){
+        total_p_alleles <- length(PEST_DATA[[gen]][,5:6]);
+        total_c_alleles <- length(PEST_DATA[[gen]][,7:8]);
+        for(allele in 1:p_alleles){
+            allele_count         <- sum(PEST_DATA[[gen]][,5:6] == allele);
+            p_tabl[gen, allele]  <- allele_count / total_p_alleles
+        }
+        for(allele in 1:c_alleles){
+            allele_count         <- sum(PEST_DATA[[gen]][,7:8] == allele);
+            c_tabl[gen, allele]  <- allele_count / total_c_alleles
+        }
+    }
+    return(list(densities = den_vect, pathogen_fr = p_tabl, crop_fr = c_tabl));
+}
+
+toy_check_extinction <- function(PEST, gen){
+    if(is.vector(PEST) == TRUE){
+        print(paste("Extinction occurred in generation", gen));
+        return(TRUE);
+    }
+    if(dim(PEST)[1] < 4){
+        print(paste("Extinction occurred in generation", gen));
+        return(TRUE);
+    }
+    if(sum(PEST[,2] == 0) < 1 | sum(PEST[,2] == 1) < 1){
+        print(paste("Extinction occurred in generation", gen));
+        return(TRUE);
+    }
+    return(FALSE);
+}
+
+# Initialise the landscape
+toy_initialise_land <- function(xdim = 2, ydim = 2, pathogens = 1, crops = 1){
+    LAND      <- array(data = 0, dim = c(xdim, ydim, 3));
+    p_values  <- sample(x = 1:pathogens, size = xdim * ydim, replace = TRUE);
+    p_layer   <- matrix(data = p_values, ncol = xdim, nrow = ydim);
+    c_values  <- sample(x = 1:crops, size = xdim * ydim, replace = TRUE);
+    c_layer   <- matrix(data = c_values, ncol = xdim, nrow = ydim);
+    LAND[,,2] <- p_layer; # Just occurred to me that we might want a layer for
+    LAND[,,3] <- c_layer; # carrying capacity of the cell in LAND later
+    return(LAND);
+}
+
+# Initialise some *very* simple pests. Each allele is just going to map one to
+# one for whether a crop can be attacked or a pathogen resisted. I am not even
+# going to add traits affecting dispersal in the toy version -- this can be
+# uniform for all individuals
+toy_initialise_pest <- function(LAND, N = 10, p_al = 1, c_al = 1){
+    xdim     <- dim(LAND)[1]; # This is needed to put the pests on some
+    ydim     <- dim(LAND)[2]; # landscape cell
+    PEST     <- matrix(data = 0, nrow = N, ncol = 8);
+    PEST[,1] <- 1:N;                                           # ID
+    PEST[,2] <- sample(x = c(0, 1), size = N, replace = TRUE); # Sex
+    PEST[,3] <- sample(x = 1:xdim,  size = N, replace = TRUE); # x-location
+    PEST[,4] <- sample(x = 1:ydim,  size = N, replace = TRUE); # y-location
+    PEST[,5] <- sample(x = 1:p_al,  size = N, replace = TRUE); # p allele 1
+    PEST[,6] <- sample(x = 1:p_al,  size = N, replace = TRUE); # p allele 2
+    PEST[,7] <- sample(x = 1:c_al,  size = N, replace = TRUE); # c allele 1
+    PEST[,8] <- sample(x = 1:c_al,  size = N, replace = TRUE); # c allele 2
+    return(PEST);
+}
+
+# Just going to make this a simple function at first, randomly changing crops
+toy_set_crops <- function(LAND, crops = 1, type = "static"){
+    if(crops == 1 | type == "static"){
+        return(LAND);
+    }
+    if(type == "rotate"){
+        old_crop                   <- LAND[,,3];
+        new_crop                   <- old_crop + 1;
+        new_crop[new_crop > crops] <- 1;
+        LAND[,,3]                  <- new_crop;
+    }else{ # Else it just randomises -- can think about fancier ways later
+        xdim      <- dim(LAND)[1];
+        ydim      <- dim(LAND)[2];
+        new_cval  <- sample(x = 1:crops, size = xdim * ydim, replace = TRUE);
+        new_crop  <- matrix(data = new_cval, nrow = xdim, ncol = ydim);
+        LAND[,,3] <- new_crop;
+    }
+    return(LAND);
+}
+
+# Ditto here -- just a simple function changing pathogens
+toy_set_paths <- function(LAND, paths = 1, type = "static"){
+    if(paths == 1 | type == "static"){
+        return(LAND);
+    }
+    if(type == "rotate"){
+        old_path                   <- LAND[,,3];
+        new_path                   <- old_path + 1;
+        new_path[new_path > paths] <- 1;
+        LAND[,,2]                  <- new_path;
+    }else{ # Else it just randomises -- can think about fancier ways later
+        xdim      <- dim(LAND)[1];
+        ydim      <- dim(LAND)[2];
+        new_pval  <- sample(x = 1:paths, size = xdim * ydim, replace = TRUE);
+        new_path  <- matrix(data = new_pval, nrow = xdim, ncol = ydim);
+        LAND[,,2] <- new_path;
+    }
+    return(LAND);
+}
+
+# Just a function to move pests at a given value, randomly on the landscape. The
+# `prob` is just probability of moving in a time step, while `dist` is just the
+# maximum distance moved in cells, in any direction.
+toy_move_pest <- function(PEST, LAND, prob = 0.1, dist = 1){
+    xdim                     <- dim(LAND)[1];
+    ydim                     <- dim(LAND)[2];
+    pests                    <- dim(PEST)[1];
+    to_move                  <- rbinom(n = pests, size = 1, pr = prob);
+    move_x                   <- sample(x = -dist:dist, size = pests, 
+                                       replace = TRUE);
+    move_y                   <- sample(x = -dist:dist, size = pests, 
+                                       replace = TRUE);
+    PEST[to_move == 1, 3]    <- PEST[to_move == 1, 3] + move_x[to_move == 1];
+    PEST[to_move == 1, 4]    <- PEST[to_move == 1, 4] + move_y[to_move == 1];
+    # The ifs below make a torus landscape so that pests don't move off of it
+    # In C, this will be done with a loop that looks cleaner; the below is just
+    # avoiding using a loop in R, though it would probably be fine.
+    if(length(PEST[PEST[,3] < 1, 3]) > 0){ 
+        PEST[PEST[,3] < 1, 3] <- PEST[PEST[,3] < 1, 3] + xdim;
+    }
+    if(length(PEST[PEST[,3] > xdim, 3]) > 0){
+        PEST[PEST[,3] > xdim, 3] <- PEST[PEST[,3] > xdim, 3] - xdim;
+    }
+    if(length(PEST[PEST[,4] < 1, 4]) > 0){
+        PEST[PEST[,4] < 1, 4] <- PEST[PEST[,4] < 1, 4] + ydim;
+    }
+    if(length(PEST[PEST[,4] > ydim, 4]) > 0){
+        PEST[PEST[,4] > ydim, 4] <- PEST[PEST[,4] > ydim, 4] - ydim;
+    }
+    return(PEST);
+}
+
+# Need to now feed the pests and remove those that don't feed (crop unavailable)
+toy_feed_pest <- function(PEST, LAND){
+    # I'm just going to use a loop here, else matching to LAND cells is rough
+    pests <- dim(PEST)[1];
+    eaten <- rep(x = 0, times = pests); 
+    for(i in 1:pests){
+        x_loc <- PEST[i, 3];
+        y_loc <- PEST[i, 4];
+        food  <- LAND[x_loc, y_loc, 3];
+        if(PEST[i, 7] == food | PEST[i, 8] == food){
+            eaten[i] <- 1;
+        }
+    }
+    PEST <- PEST[eaten == 1,]; # You don't eat, you don't live
+    return(PEST);
+}
+
+# Then need to have pests resist pathogens, kill those that cannot
+toy_kill_pest <- function(PEST, LAND){
+    # I'm just going to use a loop here, else matching to LAND cells is rough
+    pests    <- dim(PEST)[1];
+    survived <- rep(x = 0, times = pests);
+    for(i in 1:pests){
+        x_loc <- PEST[i, 3];
+        y_loc <- PEST[i, 4];
+        patho <- LAND[x_loc, y_loc, 2];
+        if(PEST[i, 5] == patho | PEST[i, 6] == patho){
+            survived[i] <- 1;
+        }
+    }
+    PEST <- PEST[survived == 1,]; # You don't eat, you don't live
+    return(PEST);
+}
+
+# Then, reproduce pests that are left
+toy_reproduce_pest <- function(PEST, LAND, pa, cr, births = 2, K = 100){
+    x_dim           <- dim(LAND)[1];
+    y_dim           <- dim(LAND)[2];
+    offspring       <- NULL;
+    total_offspring <- 0; # This and the below are just to avoid an rbind(),
+    cell            <- 1; # which is a massive memory sink
+    lst_ID          <- max(PEST[,1]);
+    for(xloc in 1:x_dim){ # Double loop is just much cleaner here
+        for(yloc in 1:y_dim){
+            locals     <- which(PEST[,3] == xloc & PEST[,4] == yloc);
+            local_offs <- NULL;
+            if( length(locals) > 1 ){
+                local_offs <- toy_breed_locals(PEST, locals, births, K, lst_ID);
+            }
+            if(length(local_offs) > 0){
+                lst_ID            <- lst_ID + dim(local_offs)[1];
+                total_offspring   <- total_offspring + dim(local_offs)[1];
+                offspring[[cell]] <- local_offs;
+            }else{
+                offspring[[cell]] <- NULL;
+            }
+            cell              <- cell + 1;
+        }
+    }
+    offspring <- build_new_pest(offspring, total_offspring, pa, cr);
+    return(offspring);
+}
+
+# Need to recombine the genomes correctly in the offspring
+toy_breed_locals <- function(PEST, locals, births, K, last_ID){
+    loc_PEST <- PEST[locals,]; # Need two of each sex (allee effect)
+    if(sum(loc_PEST[,2] == 0) < 2 | sum(loc_PEST[,2] == 1) < 2){
+        return(NULL);
+    }
+    females       <- loc_PEST[loc_PEST[,2] == 0,];
+    males         <- loc_PEST[loc_PEST[,2] == 1,];
+    new_offs      <- dim(females)[1] * floor(births);  
+    offspring     <- matrix(data = 0, nrow = new_offs, ncol = 8);     
+    offspring[,1] <- (last_ID + 1):(last_ID + new_offs);
+    offspring[,2] <- sample(x = c(0, 1), size = new_offs, replace = TRUE);
+    offspring[,3] <- loc_PEST[1, 3];
+    offspring[,4] <- loc_PEST[1, 4];
+    # Below grabs all of the alleles from females and males
+    p_fem_alleles <- c(females[,5], females[,6]);
+    p_mal_alleles <- c(males[,5], males[,6]);
+    c_fem_alleles <- c(females[,7], females[,8]);
+    c_mal_alleles <- c(males[,7], males[,8]);
+    # Now add them to the offspring randomly, one from female and one from male
+    for(i in 1:new_offs){
+        if(runif(n = 1) < 0.5){
+            offspring[i, 5] <- sample(x = p_fem_alleles, size = 1);
+            offspring[i, 6] <- sample(x = p_mal_alleles, size = 1);
+        }else{
+            offspring[i, 5] <- sample(x = p_mal_alleles, size = 1);
+            offspring[i, 6] <- sample(x = p_fem_alleles, size = 1);            
+        }
+        if(runif(n = 1) < 0.5){
+            offspring[i, 7] <- sample(x = c_fem_alleles, size = 1);
+            offspring[i, 8] <- sample(x = c_mal_alleles, size = 1);
+        }else{
+            offspring[i, 7] <- sample(x = c_mal_alleles, size = 1);
+            offspring[i, 8] <- sample(x = c_fem_alleles, size = 1);            
+        }
+    }
+    if(dim(offspring)[1] > K){
+        offspring <- offspring[1:K,];
+    }
+    return(offspring);
+}
+
+# Merge the different layers into a new pest array
+build_new_pest <- function(offspring, total_offspring, pa, cr, mutation = 0.01){
+    new_PEST   <- matrix(data = 0, nrow = total_offspring, ncol = 8);
+    start_row  <- 1; # Again, avoids an rbind memory issue
+    for(cell in 1:length(offspring)){
+        if(length(offspring[[cell]]) > 0){
+            cell_offs       <- dim(offspring[[cell]])[1];
+            rows            <- start_row:(start_row + cell_offs - 1);
+            new_PEST[rows,] <- offspring[[cell]];
+            start_row       <- start_row + cell_offs;
+        }
+    }
+    mu5 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    mu6 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    mu7 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    mu8 <- which(rbinom(n = total_offspring, size = 1, pr = mutation) == 1);
+    new_PEST[mu5, 5] <- sample(x = 1:cr, size = length(mu5), replace = TRUE);
+    new_PEST[mu6, 6] <- sample(x = 1:cr, size = length(mu6), replace = TRUE);
+    new_PEST[mu7, 7] <- sample(x = 1:pa, size = length(mu7), replace = TRUE);
+    new_PEST[mu8, 8] <- sample(x = 1:pa, size = length(mu8), replace = TRUE);
+    return(new_PEST);
+}
+```
+
+The function below has now been modified, along with the outer function for simulating resistance.
+
+```{r}
+toy_simulate_resistance <- function(generations = 20,       # Generations to sim
+                                    xdim = 2,               # Land dimension 1
+                                    ydim = 2,               # Land dimension 2
+                                    pathogens = 1,          # Pathogen strains
+                                    crops = 1,              # Crop species
+                                    path_alleles = 3,       # Pathogen alleles
+                                    crop_alleles = 3,       # Crop alleles
+                                    pest_init = 2000,       # Initial pests 
+                                    crop_rotate = "static", # Crops rotated
+                                    path_rotate = "static", # Pathogens rotated
+                                    pest_move_pr = 0.1,     # Pest movement
+                                    pest_move_dist = 1,     # Pest move distance
+                                    fecundity = 8,          # Offspring per fem
+                                    cell_K = 2000           # K per cell
+                                    ){
+
+    if(pest_move_dist > xdim & pest_move_dist > ydim){
+        pest_move_dist <- max(c(xdim, ydim)); # Avoids error
+    }
+    # Start initialising the landscape and pests
+    LAND <- toy_initialise_land(xdim  = xdim, ydim = ydim, 
+                                pathogens = pathogens, crops = crops);
+    PEST <- toy_initialise_pest(LAND, N = pest_init, p_al = path_alleles, 
+                                c_al = crop_alleles);
+    # Start the generations
+    PEST_DATA   <- NULL;
+    LAND_DATA   <- NULL;
+    gen         <- 1;
+    while(gen < generations){
+        LAND <- toy_set_crops(LAND, crops, crop_rotate);                        
+        LAND <- toy_set_paths(LAND, pathogens, path_rotate);                    
+        PEST <- toy_move_pest(PEST, LAND, pest_move_pr, pest_move_dist);        
+        PEST <- toy_feed_pest(PEST, LAND);                                      
+        if(toy_check_extinction(PEST, gen) == TRUE){ # Hate these if breaks here
+            break;
+        }
+        PEST <- toy_kill_pest(PEST, LAND);
+        if(toy_check_extinction(PEST, gen) == TRUE){
+            break;
+        }
+        PEST <- toy_reproduce_pest(PEST, LAND, path_alleles, crop_alleles, 
+                                   fecundity, cell_K);
+        if(toy_check_extinction(PEST, gen) == TRUE){
+            break;
+        }
+        PEST_DATA[[gen]] <- PEST;
+        LAND_DATA[[gen]] <- LAND;
+        gen <- gen + 1;
+    }
+    return(list(PEST_DATA = PEST_DATA, LAND_DATA = LAND_DATA));
+}
+
+results_to_json <- function(pest, land, printit = TRUE, filename = "sim.json"){
+    if("package:jsonlite" %in% search() == FALSE){
+        stop("Error: Need to load the R package 'jsonlite'")
+    }
+    inds   <- dim(pest)[1];
+    p_geno <- rep(x = 0, times = inds);
+    c_geno <- rep(x = 0, times = inds);
+    path   <- rep(x = 0, times = inds);
+    crop   <- rep(x = 0, times = inds);
+    r_path <- rep(x = 0, times = inds);
+    r_crop <- rep(x = 0, times = inds);
+    for(i in 1:inds){ # Doing this a bit lazily, without refactoring the rest
+        p_geno[i] <- as.numeric( paste(pest[i,5], pest[i,6], sep = ""));
+        c_geno[i] <- as.numeric( paste(pest[i,7], pest[i,8], sep = ""));
+        xloc      <- pest[i, 3];
+        yloc      <- pest[i, 4];
+        path[i]   <- land[xloc, yloc, 2];
+        crop[i]   <- land[xloc, yloc, 3];
+        if(pest[i,5] == path[i] | pest[i,5] == path[i]){
+            r_path <- 1;
+        }
+        if(pest[i,7] == crop[i] | pest[i,8] == crop[i]){
+            r_crop <- 1;
+        }
+    }
+    data      <- matrix(data = 0, nrow = inds, ncol = 10);
+    data[,1]  <- pest[,1];
+    data[,2]  <- pest[,2];
+    data[,3]  <- pest[,3];
+    data[,4]  <- pest[,4];
+    data[,5]  <- path;
+    data[,6]  <- crop;
+    data[,7]  <- p_geno;
+    data[,8]  <- c_geno;
+    data[,9]  <- r_path;
+    data[,10] <- r_crop;
+    colnames(data) <- c("ID", "sex", "xloc", "yloc", "path", "crop", 
+                        "p_geno", "c_geno", "resist_path", "eat_crop");
+    modsim <- list( traits = colnames(data), 
+                    values = unname(apply(data, 1, 
+                                          function(x) as.data.frame(t(x))))
+    );
+    sim_json <- toJSON(list(traits = names(modsim), values = modsim), 
+                       pretty = TRUE);
+    if(printit == TRUE){
+        write(sim_json, filename);
+    }
+    return(sim_json);
+}
+```
+
+We can run things the same way to generate the output.
+Note that the output from `toy_simulate_resistance` has changed, so now we need to read the list into `results_to_json`.
+The list contains the elements `PEST_DATA` and `LAND_DATA`.
+
+```{r}
+library(jsonlite);
+sim  <- toy_simulate_resistance(pathogens = 3, crops = 3);
+json <- results_to_json(pest = sim$PEST_DATA[[19]], land = sim$LAND_DATA[[19]],
+                        filename = "../data/sample_sim.json");
+```
+
+ The output is [here]([sample_sim.json](https://github.com/bradduthie/helicoverpa/blob/dev/data/sample_sim.json)).
+
+```{r, echo = FALSE}
+# ==============================================================================
+# This is just a toy version of the model, for the presentation
+# ==============================================================================
+# Note that I'm adding 'toy' at the start of each function; this is just in case
+# we actually want to keep some of these functions in the eventual R package, 
+# but also don't want to worry about not being able to name functions like
+# "initialise_land" the same here as elsewhere in the eventual bigger model.
+# ==============================================================================
+toy_simulate_resistance <- function(generations = 20,       # Generations to sim
+                                    xdim = 2,               # Land dimension 1
+                                    ydim = 2,               # Land dimension 2
+                                    pathogens = 1,          # Pathogen strains
+                                    crops = 1,              # Crop species
+                                    path_alleles = 3,       # Pathogen alleles
+                                    crop_alleles = 3,       # Crop alleles
+                                    pest_init = 2000,       # Initial pests 
+                                    crop_rotate = "static", # Crops rotated
+                                    path_rotate = "static", # Pathogens rotated
+                                    pest_move_pr = 0.1,     # Pest movement
+                                    pest_move_dist = 1,     # Pest move distance
+                                    fecundity = 8,          # Offspring per fem
+                                    cell_K = 2000           # K per cell
+                                    ){
+
+    if(pest_move_dist > xdim & pest_move_dist > ydim){
+        pest_move_dist <- max(c(xdim, ydim)); # Avoids error
+    }
+    # Start initialising the landscape and pests
+    LAND <- toy_initialise_land(xdim  = xdim, ydim = ydim, 
+                                pathogens = pathogens, crops = crops);
+    PEST <- toy_initialise_pest(LAND, N = pest_init, p_al = path_alleles, 
+                                c_al = crop_alleles);
+    # Start the generations
+    PEST_DATA   <- NULL;
     gen         <- 1;

### 2018-08-27 assets/dist/scripts/main.js
handled missing output and made sidebar toggling operational without patch data

@@ -1,5 +1,5 @@
 !function(t,e){"function"==typeof define&&define.amd?define([],e):"object"==typeof exports?module.exports=e():t.Tether=e()}(this,function(){"use strict";function t(t,e){if(!(t instanceof e))throw new TypeError("Cannot call a class as a function")}function e(t){var i=t.getBoundingClientRect(),n={};for(var s in i)n[s]=i[s];if(t.ownerDocument!==document){var o=t.ownerDocument.defaultView.frameElement;if(o){var r=e(o);n.top+=r.top,n.bottom+=r.top,n.left+=r.left,n.right+=r.left}}return n}function i(t){var e=getComputedStyle(t)||{},i=e.position,n=[];if("fixed"===i)return[t];for(var s=t;(s=s.parentNode)&&s&&1===s.nodeType;){var o=void 0;try{o=getComputedStyle(s)}catch(t){}if(void 0===o||null===o)return n.push(s),n;var r=o,a=r.overflow,l=r.overflowX;/(auto|scroll|overlay)/.test(a+r.overflowY+l)&&("absolute"!==i||["relative","absolute","fixed"].indexOf(o.position)>=0)&&n.push(s)}return n.push(t.ownerDocument.body),t.ownerDocument!==document&&n.push(t.ownerDocument.defaultView),n}function n(){C&&document.body.removeChild(C),C=null}function s(t){var i=void 0;t===document?(i=document,t=document.documentElement):i=t.ownerDocument;var n=i.documentElement,s=e(t),o=A();return s.top-=o.top,s.left-=o.left,void 0===s.width&&(s.width=document.body.scrollWidth-s.left-s.right),void 0===s.height&&(s.height=document.body.scrollHeight-s.top-s.bottom),s.top=s.top-n.clientTop,s.left=s.left-n.clientLeft,s.right=i.body.clientWidth-s.width-s.left,s.bottom=i.body.clientHeight-s.height-s.top,s}function o(t){return t.offsetParent||document.documentElement}function r(){if(S)return S;var t=document.createElement("div");t.style.width="100%",t.style.height="200px";var e=document.createElement("div");a(e.style,{position:"absolute",top:0,left:0,pointerEvents:"none",visibility:"hidden",width:"200px",height:"150px",overflow:"hidden"}),e.appendChild(t),document.body.appendChild(e);var i=t.offsetWidth;e.style.overflo
[…]

### 2018-08-27 assets/scripts/main.js

@@ -29,10 +29,13 @@
 			"helicoverpa": [{crop, pathogen, genotype, resistance},{crop, pathogen, genotype, resistance}]
 		---
 		*/
-		$.getJSON("./data/output.json", function(patches){
-			if($.isEmptyObject(patches)) {
+		$.getJSON('./data/output.json').done(function(patches) {
+	    if($.isEmptyObject(patches)) {
 				// Handle missing data...
-				console.log("No patch data");	
+				console.log("No patch data");	// the sidebar
+				$('#toggle-controls').click(function(){			
+		      $('.wrapper').toggleClass('open-controls');
+		    });
 			} else {
 				// If we have data Make the patches...
 				//console.log(patches);
@@ -88,7 +91,14 @@
 						$grid.masonry();
 				  });
 		    });
-			}
-		});
+			}			
+	  })
+	  .fail(function() {
+	    // Handle missing data...
+			console.log("No output file");	// the sidebar
+			$('#toggle-controls').click(function(){			
+	      $('.wrapper').toggleClass('open-controls');
+	    });
+	  });	
 	}); // end document ready
 })(jQuery); // Fully reference jQuery after this point.

### 2018-09-17 notebook/grant_part_2/duthie_paragraph.Rmd
Add phase 2 grant contributions

---
title: "Duthie Work Package"
author: "Brad Duthie"
date: "17 September 2018"
output:
  word_document:
    reference_docx: docx_template.docx
  pdf_document: default
  html_document: default
bibliography: references.bib
link-citations: yes
linkcolor: blue
csl: science.csl
biblio-style: apalike
---

Duthie Suitability
================================================================================

Brad Duthie is a theoretical biologist with a proven track record in modelling complex ecological and evolutionary systems (1-2). His previous research has included modelling evolutionary dynamics of inbreeding behaviour given complex genetic architectures and population pedigrees (3-5), ecological interactions among species on spatially explicit landscapes (6), and software development for modelling population management (7).

(1) Duthie, A. B., Abbott, K. C., & Nason, J. D. (2014). Trade-offs and coexistence: A lottery model applied to fig wasp communities. American Naturalist, 183(6), 826–841. https://doi.org/10.1086/675897

(2) Duthie, A. B., Lee, A. M., & Reid, J. M. (2016). Inbreeding parents should invest more resources in fewer offspring. Proceedings of The Royal Society B, 20161845. https://doi.org/10.1098/rspb.2016.184

(3) Duthie, A. B., Bocedi, G., & Reid, J. M. (2016). When does female multiple mating evolve to adjust inbreeding? Effects of inbreeding depression, direct costs, mating constraints, and polyandry as a threshold trait. Evolution, 70(9), 1927–1943. https://doi.org/10.1111/evo.13005

(4) Duthie, A. B., & Reid, J. M. (2016). Evolution of inbreeding avoidance and inbreeding preference through mate choice among interacting relatives. American Naturalist, 188(6), 000–000. https://doi.org/10.1086/688919

(5) Duthie, A. B., Bocedi, G., Germain, R. R., & Reid, J. M. (2018). Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance and associated polyandry. Journal of Evolutionary Biology. https://doi.org/10.1111/jeb.13189

(6) Duthie, A. B., & Falcy, M. R. (2013). The influence of habitat autocorrelation on plants and their seed-eating pollinators. Ecological Modelling, 251, 260–270. https://doi.org/10.1016/j.ecolmodel.2012.12.019

(7) Duthie, A. B., Cusack, J. J., Jones, I. L., Nilsen, E. B., Pozo, R. A., Rakotonarivo, O. S., … Bunnefeld, N. (2018). GMSE: an R package for generalised management strategy evaluation. Methods in Ecology and Evolution. https://doi.org/10.1101/221432

Duthie Work Package
================================================================================

To demonstrate how rotating biopesticide strains and crop species at the local scale can ultimately overcome pest resistance in the long-term at the landscape scale, we will simulate the full social-ecological system using an individual-based model (IBM). IBMs allow for interactions among discrete individuals (*H. armigera* pests) on a spatially-explicit landscape where key landscape properties (biopesticide application, crop species) vary locally [@Bocedi2014a; @Duthie2018]. Individuals can have complex genetic architectures that underly relevant quantitative traits such as bipesticide resistance, feeding efficiency, and dispersal behaviour [@Grimm2005; @DeAngelis2005; @Duthie2015]. Landscapes can be equally complex, allowing individual traits to interact with the unique set of landscape properties that occur on an landscape cell [@Duthie2013; @Peer2013; @Aben2014]. By parameterising pest genetic architecture with real estimates of dispersal and measures of genetic associations from Phase I/1a, then simulating different biopesticide application and crop choice regimes on landscape cells, we can use IBMs to model the evolution of bipesticide resistance *in silico* [@Stratonovitch2014].

A prototype model of the evolution of *H. armigera* biopesticide resistance has already been developed using a simplified genetic achitecture and landscape. Simulations from this prototype demonstrate that biopesticide and crop rotatation can overcome biopesticide resistance in principle (see http://r.rosemckeon.co.uk/helicoverpa/), but these results have yet to be demonstrated with a suitably realistic genetic architecture and landscape properties. This work package will include the required realism in an expanded model to simulate the long-term evolution of *H. armigera* resistance. The model will be built as an open-source R package, written in R and C, and made accessible to all on GitHub. The R package will be externally evaluated by the rOpenSci community, which will include review by two independent experts who will comment on code usability and quality.

<div id="refs"></div>

Duthie Costings
================================================================================

In theory, I believe that I have all the resources I need to do this -- basically a good computer and time to code. If there is a way to buy out my time for programming after 31 OCT 2020, it would be helpful. If there is room for it in the budget (though it's not really critical), it might be useful to include some money (ca 500 GBP) for cluster time and (maybe ca 250 GBP) for server time, if we're going to be releasing a shiny output online.

As mentioned in a previous meeting, I don't think I need postdoc time -- it would be great to collaboratively code with someone on this, but for it to be a net benefit, I think, in terms of efficiency in getting things done, we'd want to hire someone with biological modelling experience and programming experience in C or C++ (I can ask around -- know at least one good coder that might be looking for a job). I'm totally on board if it can be done, but we'd probably want the person to be on salary for at least a year? (or, if less, perhaps a remote worker could do -- collaboration via GitHub means that a potential postdoc wouldn't need to be physically in Stirling or Brazil).

### 2018-09-21 notebook/helicoverpa_notes.Rmd
Add new notebook entry showing some progress on Issue #19 getting dispersal and blocks to work

@@
 
+     - **SEP** [21](#u-21-SEP-2018)
      - **AUG** [13](#u-13-AUG-2018), [14](#u-14-AUG-2018), [27](#u-27-AUG-2018)
@@
 ================================================================================
+
+> <a name="u-21-SEP-2018">Update: 21 SEP 2018</a>
+
+**Success with the graphical display**
+
+While I did not update this journal during its development very much, the [graphical display](http://r.rosemckeon.co.uk/helicoverpa/) looks really good, and received positive feedback from colleagues in Brazil.
+
+**Simulations for Phase II**
+
+```{r, echo = FALSE}
+toy_initialise_land <- function(xdim = 2, ydim = 2, pathogens = 1, crops = 1){
+    LAND      <- array(data = 0, dim = c(xdim, ydim, 3));
+    p_values  <- sample(x = 1:pathogens, size = xdim * ydim, replace = TRUE);
+    p_layer   <- matrix(data = p_values, ncol = xdim, nrow = ydim);
+    c_values  <- sample(x = 1:crops, size = xdim * ydim, replace = TRUE);
+    c_layer   <- matrix(data = c_values, ncol = xdim, nrow = ydim);
+    LAND[,,2] <- p_layer; # Just occurred to me that we might want a layer for
+    LAND[,,3] <- c_layer; # carrying capacity of the cell in LAND later
+    return(LAND);
+}
+
+toy_initialise_pest <- function(LAND, N = 10, p_al = 1, c_al = 1){
+    xdim     <- dim(LAND)[1]; # This is needed to put the pests on some
+    ydim     <- dim(LAND)[2]; # landscape cell
+    PEST     <- matrix(data = 0, nrow = N, ncol = 8);
+    PEST[,1] <- 1:N;                                           # ID
+    PEST[,2] <- sample(x = c(0, 1), size = N, replace = TRUE); # Sex
+    PEST[,3] <- sample(x = 1:xdim,  size = N, replace = TRUE); # x-location
+    PEST[,4] <- sample(x = 1:ydim,  size = N, replace = TRUE); # y-location
+    PEST[,5] <- sample(x = 1:p_al,  size = N, replace = TRUE); # p allele 1
+    PEST[,6] <- sample(x = 1:p_al,  size = N, replace = TRUE); # p allele 2
+    PEST[,7] <- sample(x = 1:c_al,  size = N, replace = TRUE); # c allele 1
+    PEST[,8] <- sample(x = 1:c_al,  size = N, replace = TRUE); # c allele 2
+    return(PEST);
+}
+```
+
+While the toy model of the [graphical display](http://r.rosemckeon.co.uk/helicoverpa/) demonstrated the key points, a couple more functions are needed to do some initial simulations to get results to put in Phase II.
+First, new movement rules are needed for more realistic dispersal of *Helicoverpa* among cells.
+This is done now with the `toy_pois_move_pest` function.
+In the function below, for each pest, movement in a time step proceeds by (1) selecting how many jumps that the pest gets by sampling from a poisson distribution with mean `dist`, then moving up to `dist` cells away in any direction during each jump (uniform probability of all cells within `dist` of the original cell).
+
+```{r}
+# Simulate Poisson movement of toys -- a nasty loop here -- faster in C
+toy_pois_move_pest <- function(PEST, LAND, dist = 1){
+    pest_n   <- dim(PEST)[1];
+    xlim     <- dim(LAND)[2];
+    ylim     <- dim(LAND)[1];
+    distance <- rpois(n = pest_n, lambda = dist);
+    for(pest in 1:pest_n){
+        move <- distance[pest];
+        while(move > 0){
+            move_x  <- sample(x = -move:move, size = 1); 
+            move_y  <- sample(x = -move:move, size = 1);
+            PEST[pest, 3] <- PEST[pest, 3] + move_x;
+            PEST[pest, 4] <- PEST[pest, 4] + move_y;
+            move    <- move - 1;
+        }
+        while(PEST[pest, 3] < 1){
+            PEST[pest, 3] <- PEST[pest, 3] + xlim;
+        }
+        while(PEST[pest, 3] > xlim){
+            PEST[pest, 3] <- PEST[pest, 3] - xlim;
+        }
+        while(PEST[pest, 4] < 1){
+            PEST[pest, 4] <- PEST[pest, 4] + ylim;
+        }
+        while(PEST[pest, 4] > ylim){
+            PEST[pest, 4] <- PEST[pest, 4] - ylim;
+        }
+    }
+    return(PEST);
+}
+```
+
+We can check out the distribution of distance away from the original cell when the above function is implemented.
+
+```{r}
+LAND     <- toy_initialise_land(xdim = 50, ydim = 50, pathogens = 1, crops = 1);
+PEST_1   <- toy_initialise_pest(LAND, N = 10000);
+PEST_2   <- toy_pois_move_pest(PEST_1, LAND, dist = 5);
+diffx_1  <- (PEST_1[,3] - PEST_2[,3])^2;
+diffx_2  <- (PEST_1[,3] - PEST_2[,3] + 50)^2; # Account for the torus
+diffx_3  <- (PEST_1[,3] - PEST_2[,3] - 50)^2;
+diffy_1  <- (PEST_1[,4] - PEST_2[,4])^2;
+diffy_2  <- (PEST_1[,4] - PEST_2[,4] + 50)^2; # Account for the torus
+diffy_3  <- (PEST_1[,4] - PEST_2[,4] - 50)^2;
+diffx    <- rep(x = 0, times = 10000);
+diffy    <- rep(x = 0, times = 10000); 
+for(i in 1:10000){ # Assuming no one goes too far
+    diffx[i] <- min(c(diffx_1[i], diffx_2[i], diffx_3[i]));
+    diffy[i] <- min(c(diffy_1[i], diffy_2[i], diffy_3[i]));
+}
+dist_mv1 <- sqrt(diffx + diffy);
+par(mar = c(5, 5, 2, 1));
+hist(dist_mv1, main = "", col = "grey", cex.lab = 1.5, cex.axis = 1.5, 
+     xlab = "Total distance travelled (km)", ylab = "Count", 
+     ylim = c(0, 2500), xlim = c(0, 50));
+```
+
+The next part is to initialise the land in blocks so that we can control how big of an area is deadicated toward a paricular fungal biopesticide and crop area.
+We want to be able to control how many blocks are put down over the landscape.
+The code below does this, albeit in not the most efficient way.
+
+```{r}
+toy_block_land <- function(xdim, ydim, pathogens, crops, block_len){
+    cells     <- xdim * ydim;
+    cell_left <- cells %% (block_len * block_len);
+    if(cell_left > 0){
+        stop("Landscape cannot be divided into squares of size 'block_len'");
+    }
+    LAND      <- array(data = 0, dim = c(xdim, ydim, 3));
+    tx0       <- 1;
+    tx1       <- block_len;
+    ty0       <- 1;
+    ty1       <- block_len;
+    block_num <- 1;
+    while(tx0 <= xdim & ty0 <= ydim){
+        LAND[tx0:tx1, ty0:ty1, 1] <- block_num;
+        tx0                       <- tx1 + 1;
+        tx1                       <- tx1 + block_len;
+        if(tx0 > xdim){
+            tx0 <- 0;
+            tx1 <- block_len;
+            ty0 <- ty1 + 1;
+            ty1 <- ty1 + block_len;
+        }
+        block_num <- block_num + 1;
+    }
+    block_num  <- block_num - 1;
+    for(block in 1:block_num){
+        rand_path <- sample(x = 1:pathogens, size = 1);
+        rand_crop <- sample(x = 1:crops, size = 1);
+        for(xx in 1:xdim){
+            for(yy in 1:ydim){
+                if(LAND[xx, yy, 1] == block){
+                    LAND[xx, yy, 2] <- rand_path;
+                    LAND[xx, yy, 3] <- rand_crop;
+                }
+            }
+        }
+    }
+    return(LAND);
+}
+```
+
+One reason I don't mind all of the loops in the toy model is that they will need to be used when this is translated to C -- there's not really a way around it (I don't think), so there's no reason to worry too much about making everything elegant and avoiding R loops.
+It does take a while for landscapes to be initialised in the toy model, but not nearly as long as it takes most generations to run (at least, not ones with lots of pests).
+Here are some ways of using `toy_block_land` to make landscapes.
+I'll show fungal biopesticide blocks being put down randomly below
+
+**Block length 1**: Each cell is its own block, corresponding to applying fungal biopesticides in each 1 km by 1 km square on the landscape.
+
+```{r}
+LAND1 <- toy_block_land(xdim = 100, ydim = 100, pathogens = 3, crops = 3,
+                        block_len = 1);
+par(mar = c(1, 1, 1, 1));
+image(LAND1[,,2], xaxt = "n", yaxt = "n");
+```
+
+**Block length 5**: Each cell is its own block, corresponding to applying fungal biopesticides in each 2 km by 2 km square on the landscape.
+
+```{r}
+LAND5 <- toy_block_land(xdim = 100, ydim = 100, pathogens = 3, crops = 3,
+                        block_len = 5);
+par(mar = c(1, 1, 1, 1));
+image(LAND5[,,2], xaxt = "n", yaxt = "n");
+```
+
+**Block length 10**: Each cell is its own block, corresponding to applying fungal biopesticides in each 10 km by 10 km square on the landscape.
+
+```{r}
+LAND10 <- toy_block_land(xdim = 100, ydim = 100, pathogens = 3, crops = 3,
+                        block_len = 10);
+par(mar = c(1, 1, 1, 1));
+image(LAND10[,,2], xaxt = "n", yaxt = "n");
+```
+
+**Block length 25**: Each cell is its own block, corresponding to applying fungal biopesticides in each 2 km by 2 km square on the landscape.
+
+```{r}
+LAND25 <- toy_block_land(xdim = 100, ydim = 100, pathogens = 3, crops = 3,
+                        block_len = 25);
+par(mar = c(1, 1, 1, 1));
+image(LAND25[,,2], xaxt = "n", yaxt = "n");
+```
+
+Of course, for even the Phase II toy simulations, we'll want a bigger landscape. I'll try to do this for 1000 km by 1000 km and see how it goes.
 

### 2018-09-21 R/toy_model.R
Working incorporation of dispersal and block landscape design into the outer most function for Issue #19 -- need to make a more outer function

@@ -17,19 +17,27 @@ toy_simulate_resistance <- function(generations = 20,       # Generations to sim
                                     crop_rotate = "static", # Crops rotated
                                     path_rotate = "static", # Pathogens rotated
                                     pest_move_pr = 0.1,     # Pest movement
-                                    pest_move_dist = 1,     # Pest move distance
+                                    pest_move_dist = 5,     # Pest move distance
                                     fecundity = 8,          # Offspring per fem
                                     cell_K = 2000,          # K per cell
                                     print_gen = FALSE,      # Option print gen
-                                    pois_move = TRUE        # Kind of movement
+                                    pois_move = TRUE,       # Kind of movement
+                                    land_bloc = TRUE,       # Blocked land  
+                                    block_len = 1           # Length block
                                     ){
     
     if(pest_move_dist > xdim & pest_move_dist > ydim){
         pest_move_dist <- max(c(xdim, ydim)); # Avoids error
     }
     # Start initialising the landscape and pests
-    LAND <- toy_initialise_land(xdim  = xdim, ydim = ydim, 
+    if(land_bloc == TRUE){
+        LAND <- toy_block_land(xdim = xdim, ydim = ydim, 
+                               pathogens = pathogens, crops = crops, 
+                               block_len = block_len);
+    }else{
+        LAND <- toy_initialise_land(xdim  = xdim, ydim = ydim, 
                                 pathogens = pathogens, crops = crops);
+    }
     PEST <- toy_initialise_pest(LAND, N = pest_init, p_al = path_alleles, 
                                 c_al = crop_alleles);
     # Start the generations

### 2018-09-24 R/toy_model.R
Add complete code draft, add sims soon

@@ -6,15 +6,6 @@
 # but also don't want to worry about not being able to name functions like
 # "initialise_land" the same here as elsewhere in the eventual bigger model.
 # ==============================================================================
-
-###
-
-# TODO: Create a 'crop_freq' and 'path_freq'  to be used in the workhorse
-##      toy_simulate_resistance function. Use the trick of if gen %% crop_freq
-##      equals zero, then call toy_set_crops and toy_set_paths, changing stuff
-
-###
-
 replicate_toy_sims <- function(generations = 20,       # Generations to sim
                                xdim = 100,             # Land dimension 1
                                ydim = 100,             # Land dimension 2
@@ -33,10 +24,14 @@ replicate_toy_sims <- function(generations = 20,       # Generations to sim
                                pois_move = TRUE,       # Kind of movement
                                land_bloc = TRUE,       # Blocked land  
                                block_len = 1,          # Length block
-                               replicates = 1          # How many times
+                               rep_num   = 1,          # How many times
+                               crop_freq = 10,         # Rel freq crop change
+                               path_freq = 10,         # Rel freq path change
+                               print_it  = TRUE,       # Print to file
+                               print_file = "toy.csv"  # Print to file
                               ){
-  results <- NULL;
-  while(replicates > 0){
+  results    <- NULL;
+  while(rep_num > 0){
       sim <- toy_simulate_resistance(generations = generations,
                                      xdim = xdim,            
                                      ydim = ydim,             
@@ -54,16 +49,23 @@ replicate_toy_sims <- function(generations = 20,       # Generations to sim
                                      print_gen = print_gen,     
                                      pois_move = pois_move,    
                                      land_bloc = land_bloc,  
-                                     block_len = block_len);
+                                     block_len = block_len,
+                                     crop_freq = crop_freq,
+                                     path_freq = path_freq);
       sim_summ <- summarise_gens(sim);
       rows_tot <- 1;
-      if(is.matrix(sim_sum) == TRUE){
-          rows_tot <- dim(sim_summ)[1];
+      if(is.matrix(sim_summ) == TRUE){
+          rows_tot   <- dim(sim_summ)[1];
+          rep_col    <- rep(x = rep_num, times = rows_tot);
+          sim_summ   <- cbind(rep_col, sim_summ);
[…]

### 2018-10-15 notebook/DuthieCV_helicoverpa.Rmd
Ignore data in git

---
linestretch: 1
link-citations: yes
linkcolor: blue
output:
  word_document:
    reference_docx: docx_template.docx
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
---

**Name:** Alexander Bradley Duthie, PhD, FHEA

**1) Education/Training**

Year of Completion,  Title or activity, Institution
2007,               B.S. in Biology,                  Southern Illinois University Edwardsville
2007,               B.S. in Philosophy,               Southern Illinois University Edwardsville
2013,               PhD in Ecology & Evolution,       Iowa State University                    
2016,               Post-doctoral Research Fellow,    University of Aberdeen                
2017,               Post-doctoral Research Assistant, University of Stirling

**2) Professional History.**

Leverhulme Early Career Research Fellow (Nov. 2017 - Oct. 2020), University of Stirling

Lecturer in Environmental Modelling (Nov. 2020, proleptic), University of Stirling

**3) List of up to 10 of the most relevant scientific results, such as scientific articles, book chapters, patents (either filed or granted), registered software, or any other type of publication considered to be among the 10 most relevant to the researcher´s career.**

**Selected Scientific Articles:**

**Duthie, A. B.**, J. J. Cusack, J. Minderman, I. L. Jones, E. B. Nilsen, R. A. Pozo, O. S. Rakotonarivo, B. Van Moorter, and N. Bunnefeld. GMSE: an R package for generalised management strategy evaluation. *Methods in Ecology and Evolution*. *In press*. DOI: 10.1111/2041-210X.13091

Cusack, J. J., **A. B. Duthie**, O. S. Rakotonarivo, R. A. Pozo, T. H. E. Mason, J. M\r{a}nsson, L. Nilsson, R. McKenzie, I. M. Tombre, E. Eyth\'{o}rsson, J. Madsen, A. Tulloch, G. Churchill, J. Shaw, R. D. Hearn, S. Redpath, and N. Bunnefeld. 2018. Time series analysis reveals synchrony and asynchrony between conflict management effort and increasing large grazing bird populations in northern Europe. *Conservation Letters*. *In press*. DOI: 10.1111/conl.12450 **(1 citation)**

Paine, C. E. T., A. Deasey, and **A. B. Duthie**. 2018. Towards general mechanistic predictions of community dynamics. *Functional Ecology*. 32 (7):1681-1692. DOI: 10.1111/1365-2435.13096 **(2 citations)**

Redpath, S. M., Andren, Z. Baynham-Herd, N. Bunnefeld, **A. B. Duthie**, Frank, C. A. Garcia, A. Keane, J. M\r{a}nsson, L. Nilsson, C. R. J. Pollard, O. S. Rakotonarivo, C. F. Salk, and H. Travers. 2018. Games as tools to address conservation conflicts. *Trends in Ecology and Evolution*. 33 (6):415-426. DOI: 10.1016/j.tree.2018.03.005

**Duthie, A. B.**, G. Bocedi, R. R. Germain, and J. M. Reid. 2018. Evolution of pre-copulatory and post-copulatory strategies of inbreeding avoidance and associated polyandry. *Journal of Evolutionary Biology*. 31:31-45. DOI: 10.1111/jeb.13189 **(1 citation)**

**Duthie, A. B.**, A. M. Lee, and J. M. Reid. 2016. Inbreeding parents should invest more resources in fewer offspring. *Proceedings of the Royal Society B*. 282:20161845. DOI: 10.1098/rspb.2016.1845 **(7 citations)**

**Duthie, A. B.**, and J. M. Reid. 2016. Evolution of inbreeding avoidance and inbreeding preference through mate choice among interacting relatives. *American Naturalist*. 188 (6):651-667. DOI: 10.1086/688919 **(6 citations)**

**Duthie, A. B.**, G. Bocedi, and J. M. Reid. 2016. When does female multiple mating evolve to adjust inbreeding? Effects of inbreeding depression, direct costs, mating constraints, and polyandry as a threshold trait. *Evolution*. 70 (9):1927-1943. DOI: 10.1111/evo.13005 **(10 citations)**

**Duthie, A. B.**, K. C. Abbott, and J. D. Nason. 2014. Trade-offs and coexistence: A lottery model applied to fig wasp communities. *American Naturalist*. 183 (6):826-841. DOI: 10.1086/675897 **(7 citations)**

**Duthie, A. B.**, and M. R. Falcy. 2013. The influence of habitat autocorrelation on plants and their seed-eating pollinators. *Ecological Modelling*. 251:260-270. DOI: 10.1016/j.ecolmodel.2012.12.019 **(3 citations)**

**4) Current research grants awarded by any agency or company to the researcher.**

Leverhulme Early Career Fellowship (ECF-2016-376)
"Biodiversity and ecosystem function in ephemeral patches: a trait-based approach"
88,650 GBP
1 Nov. 2017 - 31 Oct. 2020
The Leverhulme Trust

**5) List of ongoing supervisions, with students holding fellowships.**

James Bellis (Masters, 2018-2019)
*Effects of inbreeding depression on ecosystem function*

Adrian Bach (PhD, 2018-2022)
*Using artificial intelligence to improve decision-making in conservation conflicts*
Funded by NERC IAPETUS Doctoral Training Partnership

**6) Academic quantitative indicators.**

1) Books: **0**
2) Publications in journals with selective editorial policy: **20**
3) Book chapters: **0**
4) Supervised and concluded Master's dissertations: **0**
5) Supervised and concluded Doctoral theses: **0**
6) Number of citations received in the international scientific literature according to ISI, Scopus or Google Scholar: **119** (Google Scholar)

**7) Links to the web pages MyResearcherID (ISI) or MyCitations (Google Scholar):**

https://scholar.google.co.uk/citations?user=l_pNM8QAAAAJ&hl=en

**8) Other information.** Other relevant biographical information for assessing your recent academic work (past five years), and documenting your experience and expertise in the field of knowledge of the proposed project.

**Relevant computer skills**

- Experience in statistical analysis, mathematical & individual-based modelling, genetic algorithms. 
- Programming ability in C, R, and MATLAB
- Knowledge of Linux, version control (git), and high performance computing clusters
- GitHub profile: https://github.com/bradduthie

### 2018-10-15 R/toy_analysis.R

@@ -38,7 +38,7 @@ mtext("Percentage of pests surviving on landscape cell", outer = TRUE, side = 2,
 
 
 
-x <- c(1, 5, 10, 25, 50, 100);
+x <- c(100, 2500, 10000, 62500, 250000, 1000000);
 y <- c(32.604, 43.496, 54.745, 74.626, 77.373, 88.182);
 
 par(mar = c(5, 5, 1, 1));
@@ -90,14 +90,16 @@ ymin <- c(min(b1), min(b5), min(b10), min(b25), min(b50), min(b100));
 ymax <- c(max(b1), max(b5), max(b10), max(b25), max(b50), max(b100));
 yl95 <- ymin; # (2.576 * ysd) / sqrt(yln);
 yu95 <- ymax; # (2.576 * ysd) / sqrt(yln);
-x    <- c(1, 5, 10, 25, 50, 100);
+options(scipen = 5);
+x    <- c(100, 2500, 10000, 62500, 250000, 1000000);
 
 
+# 100 hectares per km^2
 
-
-par(mar = c(5, 5, 1, 1));
-plot(x = x, y = yy, ylim = c(0, 100), xlab = "Monoculture block size (sq. km)", 
-     ylab = "Percent of pests killed per block", type = "b", cex.lab = 1.5, 
+par(mar = c(5, 5, 1.25, 1.25));
+plot(x = x, y = yy, ylim = c(0, 100), 
+     xlab = "Monoculture field size (hectares)", 
+     ylab = "Percent of pests killed per field", type = "b", cex.lab = 1.5, 
      cex = 1.0, pch = 20, lwd = 2, cex.axis = 1.5, log = "x");
 arrows(x0 = x, x1 = x, y0 = yu95, y1 = yl95, length = 0.1, code = 3, angle = 90,
        lwd = 2);

### 2020-01-28 notebook/helicoverpa_notes.Rmd
Add complete code draft, add sims soon
New entry explaining some ideas behind the genetic architecture

@@
 
+ - **2020**
+
+    - **JAN** [27](#u-27-JAN-2020)
+
+ - **2019**
+
+    - **MAY** [14](#u-14-MAY-2019)
+
  - **2018**
@@
 ================================================================================
+
+> <a name="u-27-JAN-2020">Update: 27 JAN 2020</a>
+
+Work Package 1C develops an individual-based model to investigate how landscape diversity affects the maintenance of susceptibility alleles, and will focus especially on the scale at which diversifying biopesticide strain use and crop species production can overcome pest resistance in the long term.
+In our [prototype model](https://bradduthie.github.io/helicoverpa/notebook/helicoverpa_notes.html#u-21-SEP-2018) written in R, we did this with a very simple genetic architecture that used matching alleles at a single locus to determine pesticide resistance and feeding ability across crops.
+While this was useful as an initial model to get a general idea of the scale at which different biopesticide application and crop production will need to be applied, the matching alleles model forces a strong negative genetic correlation between biopesticide resistance and feeding ability across crops.
+This is unrealistic, and a large part of this work package will be to try to understand the scale at which diversity is needed when the genetic architecture of pests is more realistic.
+A major modelling goal is to develop a method for constructing a more realistic genetic architecture for simulating pest management *in silico*, such that arbitrary and potentially evolving genetic correlations between pest traits (resistance, feeding efficiency) can be constructed.
+This entry is an initial plan for this genetic architecture; what follows is just the starting point, all subject to further revision.
+
+When modelling the evolution of pest resistance and crop feeding ability, we need to get from pest genome to pest traits.
+Everything in between is the challenging bit, but these two end points are actually fairly straightfoward.
+
+- **Genome** can be modelled as a vector of real numbers.
+Each individual pest gets its own genome, which is essentially a very long vector in which each element is a single locus.
+There are relevant details to fill in later.
+For example, there are several ways to model how alleles work.
+One of the simplest ways is an infinite alleles model, in which each locus holds a real number of some arbitrary value.
+New mutations are then introduced by randomly selecting a new real number from some distribution.
+A slightly more challenging way (computationally) is to have some fixed number of discrete alleles that are possible for an individual to have (e.g., 100 alleles at each locus), but I do not think that this is the best way to go unless there is some compelling reason for it.
+Ploidy level can also be determined in advance; the simplest is haploid, but diploid organisms can be created simply by pairing vector elements (loci) and having rules for homozygosity and heterozygosity.
+In the end, however, we will have a long vector of numbers (1000s to 10000s of elements) with values that affect pest traits in some way, and are expected to change over the course of the simulation (evolution).
+
+- **Traits** can be modelled, again, as some vector of real numbers.
+Each individual gets their own values for a fixed number of 'traits' (e.g., sex, location, dispersal ability, resistance to a particular biopesticide, ability to feed on a particular crop).
+Any number of traits are possible, making the model as complex as it needs to be, but the idea is that each pest has its own set of values that make it unique.
+Some traits will not be directly related to the genome (e.g., location on the landscape).
+But most trait values, and especially bipesticide resistance and feeding ability, will be some function of the underlying genome values.
+
+**The challenge is to get from genome to traits in a way that allows traits to be quantitative and correlated in some realistic and potentially evolving way.
+Below is a starting point, though the details are likely to change.
+Feel free to leave [comments here](#comments)**
+
+A diploid genome will be constructed with $N_{L}$ total loci, so that each pest genome will represented by a vector of length $2L$.
+Vector elements will be able to take any real number, thereby modelling potentially infinite alleles.
+Allele values (i.e., the numbers at the loci) will be used in a function that ultimately determines the values of up to $N_{T}$ pest traits.
+The simplest way to do this is to just have a single locus for a single trait with additive effects of alleles.
+For example, for a locus $L_{i}$ with alleles $A_{i1}$ and
+$A_{i2}$, the value of $T_{i} = 1/2(A_{i1} + A_{i2})$.
+More generally, any different rules for making a trait (e.g., complete or partial dominance) could also be modelled by setting a function $T_{i} = f(A_{i1}, A_{i2})$.
+Similarly, we can also make traits polygenic by having them determined by multiple loci.
+For example, we might want five traits, each of which take values that are determined by 10 total loci in the genome (20 alleles).
+So, e.g., $T_{1} = \sum_{i=1}^{10}(A_{i1} + A_{i2})$, $T_{2} = \sum_{i=11}^{20}(A_{i1} + A_{i2})$, etc.
+This would give us five traits, each of which takes a value that is the sum of 20 individual real numbers within the genome.
+
+One way to allow for correlated traits would be to have two separate trait values (e.g., $T_{1}$ and $T_{2}$) be determined by some overlapping set of loci.
+Correlation between traits could be adjusted by adjusting the size of this overlap.
+For complete correlation, e.g., the two traits would be determined by the exact same loci.
+Negative correlations could be constructed by inverting the effects of loci such that $L_{i} \to T_{1}$ but $-L_{i} \to T_{2}$.
+Still, this means that the correlations must be set *a priori*, and sets some restrictions on the kinds of correlations that are possible (e.g., precision as a consequence of the total number of alleles that can be overlapping versus non-overlapping).
+Hence, I propose a slightly different construction, which will allow us to go from genome to traits in a way that allows us to **set** genetic correlations without needing to **fix** them *a priori* through some pre-specified mechanism.
+First, here is a simplified version of what I want to try.
+
+**The basic idea**
+
+Assume a large population of individuals with only two haploid loci $L0$ and $L1$, and two traits $T0$ and $T1$.
+There are four possible links between $L_{i} \to T_{j}$, which we can call $X_{ij}$ describing the link between $L_{i}$ and $T_{j}$: $X_{00}$, $X_{01}$, $X_{10}$, and $X_{11}$.
+Assume that $X_{00} = 1$, $X_{01} = 0.3$, $X_{10} = 0$, and $X_{11} = \sqrt{1 - 0.3^{2}}$.
+That is,
+
+$$T_{0} = X_{00},$$
+
+while,
+
+$$T_{1} = \sqrt{1 - 0.3^{2}} X_{11} + 0.3 X_{01}.$$
+
+If we randomly sample $L0 \sim \mathcal{N}(0, 1)$ and $L1 \sim \mathcal{N}(0, 1)$, then trait variances equal $1$, while $cov(T_{0}, T_{1}) = 0.3$.
+We can confirm this by simulating a population of $N = 100000$ individuals.
+
+```{r}
+N    <- 100000;
+inds <- matrix(data = 0, nrow = N, ncol = 9);
+colnames(inds) <- c("ID", "L0", "L1", "X00", "X01", "X10", "X11", "T0", "T1");
+inds[,1]  <- 1:N;
+inds[,2]  <- rnorm(n = N, mean = 0, sd = 1);
+inds[,3]  <- rnorm(n = N, mean = 0, sd = 1);
+inds[,4]  <- 1;
+inds[,5]  <- 0.3;
+inds[,6]  <- 0;
+inds[,7]  <- sqrt(1 - 0.3^2);
+inds[,8]  <- (inds[,2] * inds[,4]) + (inds[,3] * inds[,6]);
+inds[,9]  <- (inds[,2] * inds[,5]) + (inds[,3] * inds[,7]);
+```
+
+So the individuals are in rows, while their attributes (IDs, genomes, interactions, traits) are in columns.
+
+```{r, echo = FALSE}
+head(inds);
+```
+
+We can look at the covariance between traits $T0$ and $T1$ in columns 8 and 9 (which might then code for 'resistance ability' or 'feeding efficiency', or something like it).
+
+```{r}
+cov(inds[,8:9]);
+```
+
+This is roughly what we want as our variance covariance matrix.
+One really simple way to acheive this for five traits, for example, could be just to consider five loci (or blocks of some number of loci) $L_{i}$ and all of the interactions between locus $i$ and $j$, $X_{ij}$ to get to each trait.
+
+![](images/loci_to_traits_simple.png)
+
+In the figure above, five loci $L_{i}$ are connected to each trait $T_{j}$ by arrows $X_{ij}$.
+Arrow values could be represented *within the genome*, as with the individual attributes `"X00"`, `"X01"`, `"X10"`, and `"X11"` in the toy example above.
+For a given desired covariance matrix of trait values, calculations of $X_{ij}$ become longer as trait number increases, but I should be able to write the code to get this for an arbitrary number of traits.
+
+This is a simple way of doing things, but I think that it still leaves something to be desired.
+If we want the genetic covariances themselves to evolve (rather than just the traits via changing allele frequences in $L_{i}$), then we are forced to do it in a simplified way in which genetic covariances will almost inevitably evolve to become positive so that a 'good' value of $T0$ and $T1$ can be assured by a joint change with a change in $L0$ by modifying $X_{01}$ accordingly.
+In other words, any variation stuck into the genome that modifies $X_{ij}$ values can be done so that there is directional selection increasing the fitness of $T_{j}$ via change in $L_{i}$, I think.
+What we really want is to tangle things so that changes along the path from alleles to traits is more complex.
+Below is my initial plan for increasing the complexity in such a way as to make the genetic architecture itself evolve a bit more realistically.
+
+**Adding some complexity**
+
+I propose the addition of multiple hidden layers beween loci and traits, effectively building something resembling a [neural network](https://en.wikipedia.org/wiki/Artificial_neural_network) ([see here for video]([https://www.youtube.com/watch?v=aircAruvnKk)), but perhaps loosely interpreting these layers as steps along biochemical or developmental pathways from alleles to traits affecting fitness.
+
+![](images/loci_to_traits_2.png)
+
+These hidden layers (blue) can now make each trait value a linear combination of all of the paths from each locus to each trait.
+For example, the value of $L0$ can now get to $T0$ in up to 100 different ways.
+Each link between nodes has its own value to modify the effect of a locus' value on the end value of each trait.
+For example if the link from $L0 \to H0 = 0.5$, $H0 \to H0 = 0.5$, and $H0 \to T0 = -1$, then this one complete path causes $L0$ to decrease $T0$ by $L0(0.5 \times 0.5 \times -1) = -0.25L0$.
+This effect would then be **added to the effects** of all 99 other possible paths from $L0$ to $T0$ (I think it would probably make sense in the end to restrict values of edges to $0 \leq e_{i} \leq 1$; we could also fix some edge values to be zero, perhaps giving a more realistic network structure).
+
+This looks like a mess, but that's kind of the point.
+Having many paths from loci to traits should give us the freedom to allow genetic covariance to evolve *with some restrictions*.
+As with the example above, we can have values for the 250 edges that form the paths from loci to traits, but the hidden layers give us more flexibility in setting evolutionary constraints.
+
+![](images/loci_to_traits_3.png)
+
+We could, for example, allow e1 to e250 above to be represented by potentially changing values in an individual's genome (through mutation, selection, and drift).
+Given this, selection across all L0-L9 and e1-e250 would almost certainly rapidly increase fitness of all traits.
+But we could also, e.g., block mutation from happening at any of the hidden layers after initialising, e.g., e1-e200 to random values (i.e., alleles would have polygenic effects that potentially lead to fitness tradeoffs, but only parts or proportions of the genetic architecture that we want to evolve actually could).
+Or we could change the initial genetic variation or mutation rates of edges relative to loci (e.g., make the genetic architecture itself evolve more or less rapidly).
+We could then ask several potentially interesting questions by controlling the extent to which genetic covariances can evolve and how processes such as drift might affect loci that underly traits versus loci that underly covariance among traits.
+
+To initialise a desired genetic covariance matrix, an evolutionary algorithm could be used.
+Basically, this would iteratively tweak the hidden layer edge values and check for a better or worse fit to the desired trait covariance matrix.
+The advantage of this would be that we could tell a function to build a genetic architecture that confroms to a covariance matrix without forcing it to specify *how*  (i.e., the genetic architecture would be 'found' rather than imposed *a priori*, reducing the number of assumptions that we need to make).
+
+Alternatively, I think it might be reasonable to introduce selection on traits by modelling a process of sequential invasion.
+That is, initialise all traits at some random value, but don't immediately impose selection.
+Hence, the traits and the genetic architecture underlying them could be random and selectively neutral for a burn-in period before introducing some sort of selective pressure (e.g., after 100 generations of burn-in, feeding efficiency gets turned on such that there is selection on feeding efficiency for one crop, $T0$).
+One by one, new selection pressures could arise.
+This could avoid having to make the assumption that the genetic architecture evolves as a consequence of early simultaneous selection for multiple traits -- it's more realistic to assume that some traits arise and evolve first (e.g., the ability to feed on one crop), then a new selective pressure arise after an evolved genome and genetic architecture already exists.
+
+<!---
+
+2A) First, we will use simulations (parameterised with real measures of genetic associations from Phase I and estimates of H.
+armigera dispersal) to model how the spatial scale of heterogeneity in fungal isolate application and crop plant cultivation affects variation in the selection landscape for biopesticide resistance; this will determine the correct deployment scales to achieve long-term, sustainable, evolution-proof pest control.
+2B) The modelling in 2A will inform experiments addressing the second objective in Phase II: to obtain on-farm estimates of how selection for biopesticide resistance is influenced by crop species and fungal isolate heterogeneity in the field; this will quantify the benefits of our system in a Brazilian context, relative to standard monocultures with homogenous biopesticide application.
+
+1A) To quantify how combining multiple fungal biopesticide isolates with a landscape of different crop species can make resistance-selection inconsistent, and thereby block resistance evolution.
+This experiment will demonstrate that genetic associations across habitats degrade with environmental distance, and quantify the magnitude of this decline for the range of habitats in our experiment.
+
+To demonstrate how rotating biopesticide strains and crop species at the local scale can ultimately overcome pest resistance in the long-term at the landscape scale, we will simulate the full social-ecological system using an individual-based model (IBM).
+IBMs allow for interactions among discrete individuals (*H.
+armigera* pests) on a spatially-explicit landscape where key landscape properties (biopesticide application, crop species) vary locally [@Bocedi2014a; @Duthie2018].
+Individuals can have complex genetic architectures that underly relevant quantitative traits such as bipesticide resistance, feeding efficiency, and dispersal behaviour [@Grimm2005; @DeAngelis2005; @Duthie2015].
+Landscapes can be equally complex, allowing individual traits to interact with the unique set of landscape properties that occur on an landscape cell [@Duthie2013; @Peer2013; @Aben2014].
+By parameterising pest genetic architecture with real estimates of dispersal and measures of genetic associations from Phase I/1a, then simulating different biopesticide application and crop choice regimes on landscape cells, we can use IBMs to model the evolution of bipesticide resistance *in silico* [@Stratonovitch2014].
+
+A prototype model of the evolution of *H.
+armigera* biopesticide resistance has already been developed using a simplified genetic achitecture and landscape.
+Simulations from this prototype demonstrate that biopesticide and crop rotatation can overcome biopesticide resistance in principle (see http://r.rosemckeon.co.uk/helicoverpa/), but these results have yet to be demonstrated with a suitably realistic genetic architecture and landscape properties.
+This work package will include the required realism in an expanded model to simulate the long-term evolution of *H.
+armigera* resistance.
+The model will be built as an open-source R package, written in R and C, and made accessible to all on GitHub.
+The R package will be externally evaluated by the rOpenSci community, which will include review by two independent experts who will comment on code usability and quality.
+
+A key question concerns the scale at which diversified biopesticide strain and crop species use
+can overcome pest resistance in the long-term. We have already developed a prototype model
+using a simplified genetic architecture and landscape (see progress report), and verified using
+preliminary simulations that biopesticide and crop heterogeneity can in principle overcome
+biopesticide resistance 65 , but these results have yet to be demonstrated with a suitably realistic
+genetic architecture (i.e., involving many genes) and landscape properties. We will implement this
+critical realism in an expanded model to simulate the long-term evolution of H. armigera resistance.
+We will simulate the genetics of real pests on a spatially explicit landscape system using an
+individual-based model (IBM). This IBM will allow for interactions among discrete individuals where key landscape properties (biopesticide application, crop species), and therefore the fitness of
+pesticide and crop specific alleles, vary locally 66 . The in-silico individuals have complex genetic
+architectures that underlie relevant quantitative traits such as biopesticide resistance, feeding
+efficiency, and dispersal behaviour 67–69 . Landscapes are equally complex, allowing individual traits
+to interact with the unique set of landscape properties on each landscape cell 25,70,71 (cells are the
+minimum scale at which land-use is homogenous). We will parameterise this general model with
+real estimates of H. armigera dispersal 4 and measures of genetic associations from Challenge 1a,
+then simulate different biopesticide application and crop choice regimes on landscape cells. We
+will model the evolution of biopesticide resistance as a function of the scale at which heterogeneity
+is applied (i.e., the number of adjacent cells that share identical crop and biopesticide properties) 7
+
+--->
+
+> <a name="u-14-MAY-2019">Update: 14 MAY 2019</a>
+
+**Plans for the full helicoverpa R package**
+
+Given successful funding of the ENDORSE project, these notes will now continue in development of work package C1c of the project. The following very rough timeline is suggested.
+
+- [ ] General software planning (language use, file organisation): SEP 2019 
+- [ ] Model planning (data structures, flow chart): AUG 2019 
+- [ ] Discussions with colleagues, revision, final planning: OCT 2019 
+- [ ] Ethical approval for software development: NOV 2019 
+- [ ] Coding and R package development: JAN-JUN 2020 
+- [ ] Code testing and simulations: JUL-SEP 2020 
+- [ ] Submission to rOpenSci: OCT 2020 
+- [ ] Publish R package on CRAN and GitHub: NOV 2020 
+- [ ] Final simulations and write-up: JAN-MAR 2021
+
+This timeline is subject to revision, but the initial steps are more detailed planning for the software and model, with input from colleagues across all work packages.
 

### 2021-05-14 notebook/presentation/12MAY2021_meeting.Rmd
Presentation updates

---
title: "Progress on Work Package 1C"
date: "12 May 2021"
output:
  beamer_presentation:
    theme: "default"
    colortheme: "default"
    fonttheme: "default"
  ioslides_presentation: default
  slidy_presentation: default
header-includes:
- \usepackage{hyperref}
colorlinks: true
linkcolor: blue
urlcolor: blue
---

## Objectives of Work Package 1C

```{r, echo = FALSE}
load("initial_output.Rdata");
GEI <- read.csv("GEI.csv");
vcv <- mg2[[2]];
rownames(vcv) <- colnames(vcv);
```

- Determine how the scale of landscape diversity affects the maintenance of susceptibility alleles \pause

- Use individual-based modelling (i.e., agent-based modelling) to simulate genetic architecture and landscapes \pause

- Simulate dynamic application of biopesticide application and crop rotation *in silico* \pause

- Parameterise with data from *Helicoverpa armigera* and data from Work Package 1A

## Simulate genetic architecture

```{r, echo = FALSE}
print(GEI[1:16,]);
```

## Simulate genetic architecture

Correlation matrix derived from Work Package 1A

\vspace{3mm}

\footnotesize

```{r, echo = FALSE}
print(round(vcv,3));
```

\vspace{3mm}

\normalsize

Need to create a quantitative genetic architecture that produces these trait correlations from allele values

##

![](traits.png)

##

![](loci.png)

##

![](hidden.png)

##

![](network.png)

##

![](arrows.png)

##

![](network.png)

## Getting from loci to traits

Use an evolutionary algorithm to find arrow values

1. Initialise a population of N random networks \pause
2. Crossover between networks (swap values) \pause
3. Mutation of network values \pause
4. Get network fitnesses (how close to empirical matrix?) \pause
5. Tournament to select highest fitness networks \pause
6. High fitness networks reproduce \pause
7. N offspring networks replace parents \pause
8. Repeat 2-7 until termination criteria satisfied

End result is a set of network values that gets close the empirical matrix when loci have standard random normal values

## The helicoverpa package

An R function `mine_gmatrix` can find network values given:

- A square empirical matrix
- Arbitrary loci number
- Arbitrary number of hidden layers
- Various algorithm input options

\vspace{3mm}

```{r, eval = FALSE}
eg_net <- mine_gmatrix(gmatrix = gmt, loci = 18, 
                       layers = 6, mu_rate = 0.1, 
                       mu_sd = 0.04, net_N = 8000, 
                       test_N = 1000);
```

## The helicoverpa package

```
===============================================
Initialising gmatrix mining...                 
===============================================
Gen: 0	 Stress: -1.300040	 Min: -1.300051
Gen: 1	 Stress: -1.300052	 Min: -1.300150
Gen: 2	 Stress: -1.300119	 Min: -1.300985
Gen: 3	 Stress: -1.300434	 Min: -1.303620
Gen: 4	 Stress: -1.301580	 Min: -1.311596
Gen: 5	 Stress: -1.305397	 Min: -1.332303
Gen: 6	 Stress: -1.316018	 Min: -1.388752
Gen: 7	 Stress: -1.342745	 Min: -1.537480
Gen: 8	 Stress: -1.400662	 Min: -1.643793
Gen: 9	 Stress: -1.483798	 Min: -1.720355
Gen: 10	 Stress: -1.552275	 Min: -1.832071
Gen: 11	 Stress: -1.580273	 Min: -1.863808
```

## Building the empirical matrix from random normal loci

\scriptsize

```{r, echo = FALSE}
round(cor(mg2[[6]]), 3);
```

\vspace{2mm} \pause

\hrule

```{r, echo = FALSE}
round(mg2[[2]], 3);
```

## Next steps

- Further improve the evolutionary algorithm
- Model *Helicoverpa* with random initial alleles
- Simulate *Helicoverpa* on a landscape
- Observe evolution given changing pesticide and crop scenarios
- Complete R package and publications

### 2021-05-14 notebook/scratch.R

@@ -6,15 +6,26 @@
 
 mat <- read.csv("notebook/gmat.csv");
 gmt <- as.matrix(mat);
-mg  <- mine_gmatrix(gmatrix = gmt, paras = c(18, 3, 1000, 4000, 0.4, 0.02, 40, 0.2, 800, 20, -5.3));
+mg  <- mine_gmatrix(gmatrix = gmt, paras = c(18, 6, 1000, 10000, 0.05, 0.01, 460, 0.05, 2500, 50, -5.3));
 
 
+
+#==============================================================================#
+#   IMPORTANT THINGS TO DO WITH THE mine_gmatrix FUNCTION
+#==============================================================================#
 # Scale the variation of the initialised network to the gmatrix input.
 # Either that, or allow the variation to be set in the R function
 # Maybe set it to the digit just below the highest (e.g., 1 then 0.1, 10 then 1)
+#
+# Have an option for finding the correlation matrix rather that VCV
+# Figure out how to scale the trait values appropriately
+#
+# Have an option for diploid genetic architecture
+#==============================================================================#
 
 
-N      <- 10000;
+
+N      <- 100000;
 loci   <- 18;
 inds   <- matrix(data = rnorm(n = (N*loci), sd = 1), nrow = N, ncol = loci);
 trts   <- inds %*% mg[[5]];

### 2021-05-14 src/mine_gmatrix.c
Make some minor changes to the code, now working to retain best strategy

@@ -1249,6 +1249,7 @@ SEXP mine_gmatrix(SEXP PARAS, SEXP GMATRIX){
       for(i = 0; i < traits; i++){
         for(j = 0; j < traits; j++){
           genome_ptr[vec_pos] = win_net[k][i][j];
+          vec_pos++;
         }
       }
     }

### 2021-05-28 R/mine_gmatrix.R
Refactor to make it easier to put R options in, and remove the hard code for the 0.1 standard devation of network initialisation

@@ -2,13 +2,71 @@
 #'
 #' Mine networks for establishing the link between genome and g-matrix.
 #'
-#'@param paras Doesn't matter what this is yet, it will run.
+#'@param loci The number of loci that individuals in the model will have
+#'@param layers The number of layers in the network from loci to traits
+#'@param indivs The number of individuals to test the covariance matrix
+#'@param npsize The size of the network population in the evolutionary algorithm
+#'@param mu_pr The probability of a network value to mutate
+#'@param mu_sd The standard deviation of mutation effect size
+#'@param max_gen The maximum number of generations of the evolutionary algorithm
+#'@param pr_cross The probability of a crossover occurring for a network
+#'@sampleK The number of networks sampled to take part in tournament selection
+#'@chooseK The number of winners in tournament selection
+#'@term_cri Stress criteria (logged) for the evolutionary algorithm terminating
+#'@sd_ini StDev of initialised networked values (best if < 0.1*max gmat value)
 #'@param gmatrix G-matrix that the evolutionary algorithm will match
 #'@return A set of values that will produce a desired G-matrix
 #'@export
-mine_gmatrix <- function(paras = c(40, 10, 1000, 1000, 0.2, 0.01, 100, 0.2, 40, 
-                                   2, -5.3), 
+mine_gmatrix <- function(loci     = 18, 
+                         layers   = 6, 
+                         indivs   = 1000,
+                         npsize   = 2000, 
+                         mu_pr    = 0.05, 
+                         mu_sd    = 0.01,
+                         max_gen  = 1000,
+                         pr_cross = 0.05,
+                         sampleK  = 40,
+                         chooseK  = 4,
+                         term_cri = -5.3,
+                         sd_ini   = 0.1,
                          gmatrix){
+    
+    paras <- c(loci, layers, indivs, npsize, mu_pr, mu_sd, max_gen, pr_cross,
+               sampleK, chooseK, term_cri, sd_ini);
+    
+    if(loci < 2 | loci %% 1 > 0){
+        stop("ERROR: 'loci' needs to be an integer value above 1.")
+    }
+    if(layers < 1 | layers %% 1 > 0){
[…]

### 2021-05-28 src/mine_gmatrix.c

@@ -754,7 +754,6 @@ void mutation_net(double ****netpop, int npsize, int layers, int traits,
   }
 }
 
-
 /* =============================================================================
  * This seeds a 4D array with standard random normal values; The array is used
  * to model an evolving population of 3D networks that move from loci values to
@@ -765,8 +764,10 @@ void mutation_net(double ****netpop, int npsize, int layers, int traits,
  *     npsize: The size of the population evolving in the evolutionary algorithm
  *     layers: The number of layers in an individual loci to trait network
  *     traits: The number of traits that an individual has
+ *     sd_ini: StDev of initialised network values
  * ========================================================================== */
-void ea_net_ini(double ****netpop, int npsize, int layers, int traits){
+void ea_net_ini(double ****netpop, int npsize, int layers, int traits, 
+                double sd_ini){
   
   int k, l, i, j;
   
@@ -774,13 +775,12 @@ void ea_net_ini(double ****netpop, int npsize, int layers, int traits){
     for(l = 0; l < layers; l++){
       for(i = 0; i < traits; i++){
         for(j = 0; j < traits; j++){
-          netpop[k][l][i][j] = rnorm(0, 0.1);
+          netpop[k][l][i][j] = rnorm(0, sd_ini);
         }
       }
     }
   }
 }
-
 
 /* =============================================================================
  * This seeds a 3D array with standard random normal values; The array is used
@@ -791,15 +791,17 @@ void ea_net_ini(double ****netpop, int npsize, int layers, int traits){
  *     npsize: The size of the population evolving in the evolutionary algorithm
  *     loci:   The number of loci in the model (rows in individual matrices)
  *     traits: The number of traits in the model (cols in individual matrices)
+ *     sd_ini: StDev of initialised network values
  * ========================================================================== */
-void ea_ltn_ini(double ***ltnpop, int npsize, int loci, int traits){
+void ea_ltn_ini(double ***ltnpop, int npsize, int loci, int traits, 
+                double sd_ini){
   
   int k, i, j;
   
   for(k = 0; k < npsize; k++){
     for(i = 0; i < loci; i++){
       for(j = 0; j < traits; j++){
-        ltnpop[k][i][j] = rnorm(0, 0.1);
+        ltnpop[k][i][j] = rnorm(0, sd_ini);
[…]

### 2021-05-28 notebook/scratch.R
Leftover additions to the notebook explaining the mine gmatrix

@@ -6,7 +6,7 @@
 
 mat <- read.csv("notebook/gmat.csv");
 gmt <- as.matrix(mat);
-mg  <- mine_gmatrix(gmatrix = gmt, paras = c(18, 6, 1000, 10000, 0.05, 0.01, 460, 0.05, 2500, 50, -5.3));
+mg  <- mine_gmatrix(gmatrix = gmt, paras = c(18, 6, 1000, 20000, 0.05, 0.01, 3200, 0.05, 4000, 50, -8));
 
 
 

### 2021-05-28 R/mine_gmatrix.R
Add an option to allow the stress check to use a correlation between traits instead of the covariance

@@ -14,6 +14,7 @@
 #'@chooseK The number of winners in tournament selection
 #'@term_cri Stress criteria (logged) for the evolutionary algorithm terminating
 #'@sd_ini StDev of initialised networked values (best if < 0.1*max gmat value)
+#'@use_cor Compare the correlation matrix rather than the covariance matrix
 #'@param gmatrix G-matrix that the evolutionary algorithm will match
 #'@return A set of values that will produce a desired G-matrix
 #'@export
@@ -29,10 +30,11 @@ mine_gmatrix <- function(loci     = 18,
                          chooseK  = 4,
                          term_cri = -5.3,
                          sd_ini   = 0.1,
+                         use_cor  = FALSE,
                          gmatrix){
     
     paras <- c(loci, layers, indivs, npsize, mu_pr, mu_sd, max_gen, pr_cross,
-               sampleK, chooseK, term_cri, sd_ini);
+               sampleK, chooseK, term_cri, sd_ini, use_cor);
     
     if(loci < 2 | loci %% 1 > 0){
         stop("ERROR: 'loci' needs to be an integer value above 1.")

### 2021-05-28 src/mine_gmatrix.c

@@ -244,8 +244,9 @@ double stress_VCV(double **gmatrix, int traits, double **VCV){
  *     rows:    Rows in the data frame
  *     cols:    Columns in the data frame (and VCV dimensions)
  *     VCV:     The variance covariance matrix
+ *     use_cor: Whether (1) or not (0) the correlation matrix should be found
  * ========================================================================== */
-void calc_VCV(double **mat, int rows, int cols, double **VCV){
+void calc_VCV(double **mat, int rows, int cols, double **VCV, int use_cor){
   
   int i, j, k;
   double *means, N;
@@ -270,6 +271,18 @@ void calc_VCV(double **mat, int rows, int cols, double **VCV){
       }
       VCV[i][j] = VCV[i][j] / (N - 1);
       VCV[j][i] = VCV[i][j];
+    }
+  }
+  
+  if(use_cor > 0){
+    for(i = 0; i < cols; i++){
+      for(j = 0; j < i; j++){
+        VCV[i][j] = VCV[i][j] / (sqrt(VCV[i][i]) * sqrt(VCV[j][j]));
+        VCV[j][i] = VCV[i][j];
+      }
+    }
+    for(i = 0; i < cols; i++){
+      VCV[i][i] = 1;
     }
   }
 
@@ -400,12 +413,13 @@ int get_rand_int(int from, int to){
 void get_vcv(double **loc2net, double ***net, double **gmatrix, double **VCV, 
              int traits, double *paras){
   
-  int indivs, loci, layers, row, col;
+  int indivs, loci, layers, use_cor, row, col;
   double stress, **T, **L, **net_sum, **loci_to_traits;
   
   loci     = (int) paras[0]; /* Number of loci for an individual */
   layers   = (int) paras[1]; /* Layers in the network from loci to trait */
   indivs   = (int) paras[2]; /* Individuals in the population */
+  use_cor  = (int) paras[12]; /* Whether the correlation matrix is used */
 
   T  = malloc(indivs * sizeof(double *));
   for(row = 0; row < indivs; row++){
@@ -440,7 +454,7 @@ void get_vcv(double **loc2net, double ***net, double **gmatrix, double **VCV,
   matrix_multiply(L, loci_to_traits, indivs, loci, loci, traits, T);
 
   /* Calculate the variance covariance of traits */
-  calc_VCV(T, indivs, traits, VCV);
+  calc_VCV(T, indivs, traits, VCV, use_cor);
 
[…]

### 2021-08-10 notebook/helicoverpa_notes.Rmd
Add notebook entry

@@
 
+ - **2021**
+
+    - **JUL** [23](#u-23-JUL-2021)
+
  - **2020**
@@
 ================================================================================
+
+> <a name="u-23-JUL-2021">Update: 23 JUL 2021</a>
+
+I just want to walk through some of the work and simulations that I have been running in the past two weeks.
+I am not going to get into the details, but I just want to show what the model can do and roughly what kind of output we might expect for a larger set of simulations.
+Everything here is reproducible using the [helicoverpa](https://github.com/bradduthie/helicoverpa) package, which can be downloaded from the R console using the code below.
+
+```{r, eval = FALSE}
+install.packages("devtools"); # If you've not got devtools already installed
+library(devtools);
+install_github("bradduthie/helicoverpa");
+```
+
+There are no real safeguards or error messagese built into the package yet, so it's probably best to keep to a small simulations when experimenting.
+There are a few key points I want to make before presenting the model.
+
+1.
+The code is written in a very modular way.
+We can take bits out and rearrange *most* things without too much effort, so please do make suggestions if you think something is important to change.
+Most changes will take very little time for me to implement now that the broad structure of the code is written.
+
+2.
+The results that I present below are really just about getting a feel for what the model can do, and for developing an intuition about what kind of results we might expect when we set things up more formally.
+It's also another layer of checking to make sure that there are no major bugs in the code.
+Hence, the simulations that I have run reflect my own testing of what the model can do rather than the best test of any specific question.
+
+3.
+I'm not going to explain all of the things that the software can do since I don't think this would be terribly productive.
+I just want to give a sense of what we're working with and what the possibilities are in relation to ENDORSE.
+There are some hidden hooks in here that were just too tempting not to code, which can be used for future papers.
+If I've done a good job, then it should be possible to use this one R package to address a range of potentially interesting questions.
+I don't need to take the lead in everything, and I would be delighted if folks were to use the software for individual-based modelling.
+
+Here are the basics of how to run simulations.
+You really only need to know two R functions, each of which can take a long list of arguments.
+The heavy lifting of both functions is done in C rather than R.
+The first function is the [mine_gmatrix](https://github.com/bradduthie/helicoverpa/blob/master/R/mine_gmatrix.R#L21) function.
+The second function is the [run_farm_sim](https://github.com/bradduthie/helicoverpa/blob/master/R/sim_farming.R#L60) function.
+The role of [mine_gmatrix](https://github.com/bradduthie/helicoverpa/blob/master/R/mine_gmatrix.R#L21) is to generate a genome that produces an arbitrary number of phenotypic traits with an arbitrary covariance structure given a set of allele values with a standard normal distribution.
+The role of [run_farm_sim](https://github.com/bradduthie/helicoverpa/blob/master/R/sim_farming.R#L60) is to do everything else -- that is, to generate individuals with the mined genomes simulate the ecology and evolution of the system over time.
+The output of the simulation is the creation of one or more CSV files in the directory where the simulation is run (more on this later).
+
+The `mine_gmatrix` function
+-----------------------------------------
+
+The goal of the `mine_gmatrix` function is to get from standard normal (mean = 0, stdev = 1) allele values at some number of loci (`loci`) through a network of some number of hidden layers (`layers`) to produce a set of traits that have some variance covariance structure (`gmatrix`, which is a square covariance matrix).
+
+![](images/network.png)
+
+The figure above shows an example with 18 loci and 9 traits, separated by 6 hidden layers.
+Each of the black arrows can be represented by a number indicating how much one locus (green) or node (blue) should contribute to the next (mathematically, arrows between two sets of nodes are represented by matrices that get [multiplied together](https://github.com/bradduthie/helicoverpa/blob/master/src/mine_gmatrix.c#L413)).
+These numbers become part of the genome for individuals, determining what a newly born individual's traits will be given their allele values at each locus.
+For now, an evolutionary algorithm is used to find a set of numbers that will produce traits with a desired covariance structure given random standard normal allele values.
+How long it takes to find these numbers depends on the number of loci, layers, and traits, and how much error we are willing to accept.
+Here's a very simple example of running the function with four loci, two layers, and two traits that have a variance of 1 and a covariance of -0.5.
+
+```{r, eval = FALSE}
+library(helicoverpa);
+gmt       <- matrix(data = 0, nrow = 2, ncol = 2);
+gmt[1, 1] <- 1;
+gmt[2, 2] <- 1;
+gmt[1, 2] <- -0.5;
+gmt[2, 1] <- -0.5;
+```
+
+```{r, echo = FALSE}
+library(helicoverpa);
+gmt       <- matrix(data = 0, nrow = 2, ncol = 2);
+gmt[1, 1] <- 1;
+gmt[2, 2] <- 1;
+gmt[1, 2] <- -0.5;
+gmt[2, 1] <- -0.5;
+```
+
+The covariance matrix is shown below.
+
+```{r, echo = FALSE}
+print(gmt);
+```
+
+This matrix is set as the argument `gmatrix = gmt`.
+The number of loci is set with `loci = 4`, and the number of layers is set with `layers = 2` (there is no need to set trait number because the function figures this out from the `gmatrix`).
+The rest of the arguments specify how the evolutionary algorithm is to run:
+
+| Argument     | What it does                                                 |
+|--------------|--------------------------------------------------------------|
+| npsize       | How big a population of networks evolve in the algorithm     |
+| indivs       | How many random normals to test the fitness of each network  |
+| max_gen      | The maximum number of generations allowed to evolve          |
+| sampleK      | Number of randomly selected networks to compete              |
+| chooseK      | Number of tournament winners (highest fitness)               |
+| mu_pr        | Probability of a value in a network mutating                 |
+| pr_cross     | Probability of crossing over between networks                |
+| mu_sd        | Standard deviation of mutation effect size (mean is 0)       |
+| term_cri     | Logged stress criteria for termination                       |
+
+The details here are not terribly important, but the key concepts are explained in more detail in @Hamblin2013 and @Duthie2018 in one of the [GMSE R package vignettes](https://confoobio.github.io/gmse/articles/SI1.html).
+Basically, what we're doing is creating a random population of `npsize` networks like the one in the figure above with random arrow values.
+We're then letting this network evolve over time with selection for networks that do a better job of producing the `gmatrix` from a standard normal loci values.
+Evolution stops after `max_gen` generations have passed or the mean squared deviation of elements from the evolved variance covariance matrix and `gmatrix` is less than `exp(term_cri)`.
+Finding a genome is a bit of an art relying on trying different parameter combinations and seeing what works.
+Big `npsize` (higher than 10000) with low mutation rates and standard deviations (both below 0.01) are slow and steady, but usually necessary for a large `gmatrix`.
+For smaller `gmatrix`, smaller populations and higher mutation and crossover rates are fine and quite fast.
+For the matrix `gmt` above, we can probably get away with the following:
+
+```{r, eval = FALSE}
+mg  <- mine_gmatrix(gmatrix = gmt, loci = 4, layers = 2, npsize = 2000, 
+                    indivs = 2000, max_gen = 800, sampleK = 1200, chooseK = 6,
+                    mu_pr = 0.3, pr_cross = 0.3, mu_sd = 0.05, term_cri = -4);
+```
+
+I'm being extremely relaxed here with the `term_cri` and `max_gen`, but I just want to show a quick example.
+The function prints off each generation, the mean stress, and the minimum stress (best) genome found.
+This is mostly just to reassure you that it's doing the job and has not crashed.
+The output in `mg` is a list with seven elements:
+
+1. The parameters input into `mine_gmatrix`.
+2. The specified `gmatrix`.
+3. The first set of arrows from loci to network (green to blue above) in matrix form.
+4. An array made up of `layers` square matrices, representing all of the rest of the arrows in the figure above.
+5. The summed effects that each loci has on each trait.
+6. The end variance covariance matrix.
+7. The whole network (figure above) as a genome to be inserted in individuals.
+8. The stress of the final network (mean squared deviation from `gmatrix`).
+
+Here is the output for `mg` above.
+
+```{r, echo = FALSE}
+mg <- readRDS(file = "data/2021_JUL_23_data/2021_JUL_23_mg.rds");
+```
+
+```{r}
+print(mg);
+```
+
+Note that the variance covariance matrix in `[[6]]` is reasonable, but probably not good enough for real simulations.
+It took less than 3 minutes to get, but the time taken appears to scale exponentially with `gmatrix` size and `term_cri`.
+But if we accept this result, we have all that we need to move ahead with the next function for actual simulation.
+
+The `run_farm_sim` function
+------------------------------------------------
+
+This function runs a single replicate simulation from start to finish.
+It generates a landscape divided into some arbitrary number of `farms` and places new individuals on them.
+A series of events then happen in pretty much any order that we want them to happen over the course of `time_steps` discrete time steps.
+
+First, individuals are initialised with their genomes as found in `mine_gmatrix`.
+They are given random allele values for their loci.
+Other options for individuals include the following (this is not exhaustive):
+
+- The reproduction type `repro` can be set as `asexual`, `sexual` (all individuals female and male with optional `selfing`), or `biparental` (females and males separate).
+- A `max_age` that an individual can be before death.
+- A minimum and maximum age at which the individual can move (`min_age_move` and `max_age_move`).
+- A minimum and maximum age at which the individual can reproduce (`min_age_reproduce` and `max_age_reproduce`).
+- A minimum and maximum age at which the individual can feed and imbibe pesticide (`min_age_feed`).
+- The amount of food consumed during feeding on a cell (`food_consume`, which can be a vector of up to length 10 -- an element for each possible food type).
+- The amount of pesticide consumed during feeding on a cell (`pesticide_consume`, which can be a vector of up to length 10 -- an element for each possible pesticide type).
+- The distance in cells an individual can travel when moving (`move_distance`).
+- The amount food that an individual needs to consume to survive (`food_need_surve`) and the age at which this threshold is assessed (`age_food_threshold`).
+- The amount of pesticide imbibed that an individual can tolerate before death (`pesticide_tolerated_repr`) and the age at which this threshold is assessed (`age_pesticide_threshold`).
+- The probability of mutation (`mutation_pr`) and crossover (`crossover_pr`). 
+- How much of the genome can mutate (`net_mu_layers`).
+- How many neutral loci individuals have (`neutral_loci`).
+
+When the simulation starts, the following happens:
+
+- The `crop_number` crops (up to 10 possible) grown on `farms` unique farms are changed at `crop_rotation_time` intervals (rules for changing can be adjusted).
+- The `pesticide_number` pesticides (up to 10 possible) applied on each farm is changed at `pesticide_rotation_time` intervals (here too, rules for changing can be adjusted).
+- Individuals age by a time step (this is fixed; all individuals do it at the start of a time step, but individuals can be as old as we like)
+- If individuals are at least `min_age_feed` and not over `max_age_feed`, then they eat on the landscape cell in which they are located.
+The amount of food that they consume on the cell can be set as `food_consume`, which can be a vector of up to 10 elements (one for each food type).
+Food consumption rates can also be specified as a trait (more on that later).
+Food consumption occurs in a random order of individuals.
+- Similarly, if individuals at least`min_age_feed` and not over `max_age_feed` (same as feeding) imbibe biopesticde on the landscape cell in which they are located at a rate set by `pesticide_consume`, which can also be a vector of up to 10 elements (one for each pesticide type).
+Biopesticide consumption rates can also be specified as a trait.
+Biopesticide consumption occurs in a random order of individuals.
+- Individuals move on the landscape if they are at least `min_age_move` (there are more complex rules available for feeding and moving; the simplest is just moving once in any direction).
+- The number of offspring that an individual produces is calculated, potentially based on their food consumption, mate access, parameter values, or traits.
+- New offspring are made, potentially with recombination in the case of sexual reproduction.
+- Individuals die according to some criteria (food and biopesticide consumption, age).
+
+The simulation ends after a fixed number of time steps (`time_steps`) have passed, or if the population goes extinct.
+I'll now show a very simplified example of a simulation being run, then a more complicated example.
+
+A simple example of the simulation
+=====================================
+
+In this simulation, I will just consider a very simple scenario of two traits, which will model the rate at which food is consumed (Trait 1) and the rate at which pesticide is imbibed (Trait 2).
+Just to get started, let each of these traits have a variance of 1 and a covariance of about 0.5 to model the idea that the rate at which you consume pesticide increases with the rate at which you consume food.
+
+```{r, echo = FALSE}
+mg_simple <- readRDS(file = "data/2021_JUL_23_data/2021_JUL_23_mg_simple.rds");
+print(mg_simple[[2]]);
+```
+
+Using the `mine_gmatrix` function with similar parameter values tried above, I got a genome that produces the following matrix given random standard normal allele values.
+
+```{r, echo = FALSE}
+mg_simple[[6]];
+```
+
+A convenient part of the code that I've not yet mentioned is that we can replace several numerical parameter arguments with evolving traits.
+This is possible for `move_distance`, `food_needed_surv`, `pesticide_tolerated_surv`, `food_needed_repr`, `pesticide_tolerated_repr`, `mating_distance`, `lambda_value` (raw reproduction value), `movement_bouts`, `food_consume`, and `pesticide_consume`.
+If, for example, we want movement distance in cells to be defined by trait one, we just set `move_distance = "T1"` (in this case, the value will be floored since cells are discrete).
+With that in mind, we can run the following simulation.
+
+```{r, eval = FALSE}
+sim <- run_farm_sim(mine_output              = mg,
+                    N                        = 1000,
+                    neutral_loci             = 1000,
+                    xdim                     = 40,
+                    ydim                     = 40,
+                    repro                    = "sexual",
+                    max_age                  = 5,
+                    selfing                  = FALSE,
+                    food_consume             = "T1",
+                    pesticide_consume        = c("T2", 0),
+                    food_needed_surv         = 1,
+                    food_needed_repr         = 1,
+                    reproduction_type        = "food_based",
+                    pesticide_tolerated_surv = 0,
+                    pesticide_rotation_type  = 2,
+                    min_age_reproduce        = 3,
+                    max_age_feed             = 2,
+                    farms                    = 4,
+                    time_steps               = 120,
+                    mutation_pr              = 0.01,
+                    crossover_pr             = 0.01,
+                    net_mu_layers            = 0,
+                    crop_rotation_time       = 4,
+                    pesticide_rotation_time  = 4,
+                    crop_per_cell            = 2,
+                    pesticide_per_cell       = 0,
+                    crop_number              = 1,
+                    pesticide_number         = 2,
+                    K_on_birth               = 1000000,
+                    print_last               = TRUE
+);
+```
+
+Note that the `mine_output` is just the whole vector output we got from the `mine_gmatrix` function.
+`N` is the initialsed number of individuals, `xdim` and `ydim` set the landscape dimensions (so we have a 40 by 40 square landscape).
+Individuals are sexual but cannot self.
+Note how `food_consume = "T1"`, so the amount of food consumed by each individual will be determined by its value in trait 1 (had we set `food_consume = 1`, then all individuals would consume 1 unit of food; note that individuals cannot consume negative food).
+The amount of pesticide 1 consumed by an individual will be determined by `"T2"`, while no individuals consume pesticide 2 (which can just be a trick for making some of the 4 `farms` not apply any pesticide -- since anything applied has no effect).
+Note that we indicate one crop by `crop_number = 1` and two pesticides by `pesticide_number = 2` (the second one being irrelevant), and each cell applies 0 pesticide (`pesticide_per_cell = 0`), so this is actually a null model in which no pesticide is applied at all.
+
+The argument `K_on_birth` limits the number of eggs that can be made in the population; if the population goes over, then eggs are randomly removed (this prevents the time consuming process of some individuals being initialised unnecessarily -- if they were just going to die immediately anyway).
+The argument `print_last` indicates that we want to print all of the information (everything, genome and all) for each individual in the last generation.
+The resulting file is often in the 100s of MBs.
+
+After the simulation runs, we get a file `population_data.csv` that gives information for each generation on the time step, population size, mean age, mean food consumed, mean pesticide consumed, mortality rate, mean food consumed of each type, mean pesticide consumed of each type, mean trait values, and mean inbreeding coefficient.
+Here's a plot of population size, trait 1 value, trait 2 value, and mortality rate.
+
+```{r, echo = FALSE}
+dat <- read.csv("data/2021_JUL_23_data/2021_JUL_23_population_data_no_pesticide.csv");
+par(mfrow = c(2, 2));
+plot(x = dat$time_step, y = dat$population_size, 
+     xlab = "Time", ylab = "Population Size", type = "l",
+     lwd = 2);
+plot(x = dat$time_step, y = dat$trait1_mean_value, 
+     xlab = "Time", ylab = "Trait 1 mean", type = "l",
+     lwd = 2);
+plot(x = dat$time_step, y = dat$trait2_mean_value, 
+     xlab = "Time", ylab = "Trait 2 mean", type = "l",
+     lwd = 2);
+plot(x = dat$time_step, y = dat$mortality_rate, 
+     xlab = "Time", ylab = "Mortality rate", type = "l",
+     lwd = 2);
+```
+
+This makes sense (ignore the small spikes caused by the age structure).
+The population size increases, the trait 1 mean increases as individuals evolve to consume more food (which gives higher reproductive output), and trait 2 gets dragged along due to the correlation with trait 1 (despite not having any effect).
+Now we can see what happens when some of the farms start to apply pesticide by running the same model with `pesticide_per_cell = 0`.
+
+```{r, echo = FALSE}
+dat <- read.csv("data/2021_JUL_23_data/2021_JUL_23_population_data_pesticide.csv");
+par(mfrow = c(2, 2));
+plot(x = dat$time_step, y = dat$population_size, 
+     xlab = "Time", ylab = "Population Size", type = "l",
+     lwd = 2);
+plot(x = dat$time_step, y = dat$trait1_mean_value, 
+     xlab = "Time", ylab = "Trait 1 mean", type = "l",
+     lwd = 2);
+plot(x = dat$time_step, y = dat$trait2_mean_value, 
+     xlab = "Time", ylab = "Trait 2 mean", type = "l",
+     lwd = 2);
+plot(x = dat$time_step, y = dat$mortality_rate, 
+     xlab = "Time", ylab = "Mortality rate", type = "l",
+     lwd = 2);
+```
+
+Note that the population size grows more slowly now, and that while trait 1 underlying food consumption evolves to be positive, the value is not as high as it was without the pesticide.
+This is because trait 2 now evolves to be a negative value, and traits 1 and 2 are positively correlated.
+
+![](images/2021_JUL_23/2021_JUL_23_img0.png)
+
+The plot above shows traits 1 and 2 for all individuals in the population where no pesticide is applied.
+The correlation here is 0.3219144.
+We can compare this to a plot of the same traits for the simulation in which pesticide was applied.
+
+![](images/2021_JUL_23/2021_JUL_23_img1.png)
+
+Note the shifted position of the y-axis, in particular, as selection pushes pesticide consumption trait values down while pushing food consumption values up.
+Even though the network itself has not changed (i.e., standard random normals will still give individuals with traits of the original correlation structure), the correlation between trait 1 and 2 is now 0.0577854.
+This is interesting, to me at least.
+I'm also not entirely sure if I understand what is maintaining the variance in traits, nor did I expect the realised correlation to change so much just as a consequence of alleles being selected.
+Note that if you set `net_mu_layers = 4` so that the entire genome can evolve, you get a strong negative covariance in 400 generations.
+
+![](images/2021_JUL_23/2021_JUL_23_img2.png)
+
+We can have a look how each trait is correlated with fitness, just measured by offspring production. Here is offspring production against the food consumption trait.
+
+![](images/2021_JUL_23/2021_JUL_23_img3.png)
+
+And here is offspring production against pesticide consumption trait.
+
+![](images/2021_JUL_23/2021_JUL_23_img4.png)
+
+The correlations are 0.0636287 and 0.0300173 for food consumption (trait 1) and pesticide consumption (trait 2) with offspring produced, respectively.
+This would seem to suggest that selection is quite weak after 120 generations?
+We could dive deeper into the mechanics of this if we wanted to, looking at how much food each individual consumed, exactly, and how much pesticide each individual consumed (and when, even, if needed).
+I'm tempted to let these run for a while longer to see if the correlation between traits drops to negative, and wondering if there is some reason for us to see a pattern of a 'trade-on' despite a hard coded loci to trait genome that causes a trade-off.
+Maybe there's a reason to expect mutation of allele values alone to easily find a way to create this pattern, meaning that constraints are not as important as would be intuitive?
+I definitely need to think more about it.
+
+A more complex simulation
+=====================================
+
+I have now also looked through the following simulation scenarios in which there are four crops and two traits.
+
+```{r, eval = FALSE}
+library(helicoverpa);
+set.seed(Sys.time());
+mg <- readRDS(file = "mg.rds");
+
+sim <- run_farm_sim(mine_output              = mg,
+                    N                        = 10000,
+                    neutral_loci             = 1000,
+                    xdim                     = 100,
+                    ydim                     = 100,
+                    repro                    = "biparental",
+                    max_age                  = 10,
+                    selfing                  = FALSE,
+                    food_consume             = c("T1", "T2"),
+                    pesticide_consume        = c("T3", "T4"),
+                    food_needed_surv         = 1,
+                    food_needed_repr         = 1,
+                    reproduction_type        = "food_based",
+                    pesticide_tolerated_surv = 0,
+                    pesticide_rotation_type  = 2,
+                    crop_rotation_type       = 2,
+                    min_age_reproduce        = 8,
+                    max_age_feed             = 6,
+                    lambda_value             = 1.5,
+                    farms                    = 16,
+                    time_steps               = 600,
+                    mutation_pr              = 0.01,
+                    crossover_pr             = 0.01,
+                    net_mu_layers            = 0,
+                    crop_rotation_time       = 24,
+                    pesticide_rotation_time  = 24,
+                    crop_per_cell            = 2,
+                    pesticide_per_cell       = 2,
+                    crop_number              = 2,
+                    pesticide_number         = 2,
+                    K_on_birth               = 1000000,
+                    min_age_move             = 8,
+                    age_food_threshold       = 7,
+                    min_age_feed             = 1,
+                    pesticide_start          = 200,
+                    );
+```
+
+These are a bit more high powered, but the general idea is that we now have four traits, two underlying crop consumption (one for each crop) and two underlying pesticide consumption (one for each pesticide).
+Here is what the covariance matrix was set to be.
+
+```{r, echo = FALSE}
+mg2 <- readRDS(file = "data/2021_JUL_23_data/2021_JUL_23_mg2.rds");
+print(mg2[[2]]);
+```
+
+Here is what the mining process came up with at some reasonable constraints.
+
+```{r, echo = FALSE}
+print(mg2[[6]]);
+```
+
+The general idea is that if you are good at consuming one crop, then you are probably not so good at consuming another crop.
+If you are good at consuming either crop, then you probably also imbibe a lot of pesticide.
+And if you are good at avoiding pesticide consumption in one trait, then you probably are in the other.
+I've run simulations above for all combinations of the following:
+
+- Reproductive type (`asexual`, `sexual`, and `biparental`)
+- No rotation of crops or pesticide
+- Only rotation of pesticide
+- Rotation of both crops and pesticide
+
+When rotated, each of 16 `farms` independently selects a crop and a pesticide at random.
+Below, I'm just going to show plots of population size, mortality rate, and each of the four trait values over time.
+I've run all of these simulations for 300 generations, and I'm considering the first 100 as a burn in before the pesticide is added.
+The pesticide is then added for the first time on time step 210.
+I've indicated these time steps with red vertical lines on the plots below.
+Note that these are just single runs -- we would need to run many replicates to make robust inferences.
+
+**No rotation: Asexual population**
+
+```{r, echo = FALSE}
+sim1 <- read.csv("data/2021_JUL_23_data/A_no_rotation/population_data_asexual.csv");
+par(mfrow = c(3, 2), mar = c(4, 4, 0.5, 1));
+plot(x = sim1$time_step, y = sim1$population_size, 
+     xlab = "Time", ylab = "Population Size", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim1$time_step, y = sim1$mortality_rate, 
+     xlab = "Time", ylab = "Mortality rate", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim1$time_step, y = sim1$trait1_mean_value, 
+     xlab = "Time", ylab = "Trait 1 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim1$time_step, y = sim1$trait2_mean_value, 
+     xlab = "Time", ylab = "Trait 2 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim1$time_step, y = sim1$trait3_mean_value, 
+     xlab = "Time", ylab = "Trait 3 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim1$time_step, y = sim1$trait4_mean_value, 
+     xlab = "Time", ylab = "Trait 4 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+```
+
+**No rotation: Sexual population**
+
+```{r, echo = FALSE}
+sim2 <- read.csv("data/2021_JUL_23_data/A_no_rotation/population_data_sexual.csv");
+par(mfrow = c(3, 2), mar = c(4, 4, 0.5, 1));
+plot(x = sim2$time_step, y = sim2$population_size, 
+     xlab = "Time", ylab = "Population Size", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim2$time_step, y = sim2$mortality_rate, 
+     xlab = "Time", ylab = "Mortality rate", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim2$time_step, y = sim2$trait1_mean_value, 
+     xlab = "Time", ylab = "Trait 1 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim2$time_step, y = sim2$trait2_mean_value, 
+     xlab = "Time", ylab = "Trait 2 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim2$time_step, y = sim2$trait3_mean_value, 
+     xlab = "Time", ylab = "Trait 3 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim2$time_step, y = sim2$trait4_mean_value, 
+     xlab = "Time", ylab = "Trait 4 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+```
+
+**No rotation: Biparental population**
+
+```{r, echo = FALSE}
+sim3 <- read.csv("data/2021_JUL_23_data/A_no_rotation/population_data_sexual.csv");
+par(mfrow = c(3, 2), mar = c(4, 4, 0.5, 1));
+plot(x = sim3$time_step, y = sim3$population_size, 
+     xlab = "Time", ylab = "Population Size", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim3$time_step, y = sim3$mortality_rate, 
+     xlab = "Time", ylab = "Mortality rate", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim3$time_step, y = sim3$trait1_mean_value, 
+     xlab = "Time", ylab = "Trait 1 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim3$time_step, y = sim3$trait2_mean_value, 
+     xlab = "Time", ylab = "Trait 2 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim3$time_step, y = sim3$trait3_mean_value, 
+     xlab = "Time", ylab = "Trait 3 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim3$time_step, y = sim3$trait4_mean_value, 
+     xlab = "Time", ylab = "Trait 4 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+```
+
+**Pesticide rotation only: Asexual population**
+
+```{r, echo = FALSE}
+sim4 <- read.csv("data/2021_JUL_23_data/B_rot_pesticide/population_data_asexual.csv");
+par(mfrow = c(3, 2), mar = c(4, 4, 0.5, 1));
+plot(x = sim4$time_step, y = sim4$population_size, 
+     xlab = "Time", ylab = "Population Size", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim4$time_step, y = sim4$mortality_rate, 
+     xlab = "Time", ylab = "Mortality rate", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim4$time_step, y = sim4$trait1_mean_value, 
+     xlab = "Time", ylab = "Trait 1 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim4$time_step, y = sim4$trait2_mean_value, 
+     xlab = "Time", ylab = "Trait 2 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim4$time_step, y = sim4$trait3_mean_value, 
+     xlab = "Time", ylab = "Trait 3 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim4$time_step, y = sim4$trait4_mean_value, 
+     xlab = "Time", ylab = "Trait 4 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+```
+
+**Pesticide rotation only: Sexual population**
+
+```{r, echo = FALSE}
+sim5 <- read.csv("data/2021_JUL_23_data/B_rot_pesticide/population_data_sexual.csv");
+par(mfrow = c(3, 2), mar = c(4, 4, 0.5, 1));
+plot(x = sim5$time_step, y = sim5$population_size, 
+     xlab = "Time", ylab = "Population Size", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim5$time_step, y = sim5$mortality_rate, 
+     xlab = "Time", ylab = "Mortality rate", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim5$time_step, y = sim5$trait1_mean_value, 
+     xlab = "Time", ylab = "Trait 1 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim5$time_step, y = sim5$trait2_mean_value, 
+     xlab = "Time", ylab = "Trait 2 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim5$time_step, y = sim5$trait3_mean_value, 
+     xlab = "Time", ylab = "Trait 3 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim5$time_step, y = sim5$trait4_mean_value, 
+     xlab = "Time", ylab = "Trait 4 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+```
+
+**Pesticide rotation only: Biparental population**
+
+```{r, echo = FALSE}
+sim6 <- read.csv("data/2021_JUL_23_data/B_rot_pesticide/population_data_biparental.csv");
+par(mfrow = c(3, 2), mar = c(4, 4, 0.5, 1));
+plot(x = sim6$time_step, y = sim6$population_size, 
+     xlab = "Time", ylab = "Population Size", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim6$time_step, y = sim6$mortality_rate, 
+     xlab = "Time", ylab = "Mortality rate", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim6$time_step, y = sim6$trait1_mean_value, 
+     xlab = "Time", ylab = "Trait 1 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim6$time_step, y = sim6$trait2_mean_value, 
+     xlab = "Time", ylab = "Trait 2 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim6$time_step, y = sim6$trait3_mean_value, 
+     xlab = "Time", ylab = "Trait 3 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim6$time_step, y = sim6$trait4_mean_value, 
+     xlab = "Time", ylab = "Trait 4 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+```
+
+**Crop and pesticide rotation: Asexual population**
+
+```{r, echo = FALSE}
+sim7 <- read.csv("data/2021_JUL_23_data/C_rotate_all/population_data_asexual.csv");
+par(mfrow = c(3, 2), mar = c(4, 4, 0.5, 1));
+plot(x = sim7$time_step, y = sim7$population_size, 
+     xlab = "Time", ylab = "Population Size", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim7$time_step, y = sim7$mortality_rate, 
+     xlab = "Time", ylab = "Mortality rate", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim7$time_step, y = sim7$trait1_mean_value, 
+     xlab = "Time", ylab = "Trait 1 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim7$time_step, y = sim7$trait2_mean_value, 
+     xlab = "Time", ylab = "Trait 2 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim7$time_step, y = sim7$trait3_mean_value, 
+     xlab = "Time", ylab = "Trait 3 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim7$time_step, y = sim7$trait4_mean_value, 
+     xlab = "Time", ylab = "Trait 4 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+```
+
+**Crop and pesticide rotation: Sexual population**
+
+```{r, echo = FALSE}
+sim8 <- read.csv("data/2021_JUL_23_data/C_rotate_all/population_data_sexual.csv");
+par(mfrow = c(3, 2), mar = c(4, 4, 0.5, 1));
+plot(x = sim8$time_step, y = sim8$population_size, 
+     xlab = "Time", ylab = "Population Size", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim8$time_step, y = sim8$mortality_rate, 
+     xlab = "Time", ylab = "Mortality rate", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim8$time_step, y = sim8$trait1_mean_value, 
+     xlab = "Time", ylab = "Trait 1 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim8$time_step, y = sim8$trait2_mean_value, 
+     xlab = "Time", ylab = "Trait 2 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim8$time_step, y = sim8$trait3_mean_value, 
+     xlab = "Time", ylab = "Trait 3 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim8$time_step, y = sim8$trait4_mean_value, 
+     xlab = "Time", ylab = "Trait 4 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+```
+
+**Crop and pesticide rotation: Biparental population**
+
+```{r, echo = FALSE}
+sim9 <- read.csv("data/2021_JUL_23_data/C_rotate_all/population_data_biparental.csv");
+par(mfrow = c(3, 2), mar = c(4, 4, 0.5, 1));
+plot(x = sim9$time_step, y = sim9$population_size, 
+     xlab = "Time", ylab = "Population Size", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim9$time_step, y = sim9$mortality_rate, 
+     xlab = "Time", ylab = "Mortality rate", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim9$time_step, y = sim9$trait1_mean_value, 
+     xlab = "Time", ylab = "Trait 1 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim9$time_step, y = sim9$trait2_mean_value, 
+     xlab = "Time", ylab = "Trait 2 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim9$time_step, y = sim9$trait3_mean_value, 
+     xlab = "Time", ylab = "Trait 3 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+plot(x = sim9$time_step, y = sim9$trait4_mean_value, 
+     xlab = "Time", ylab = "Trait 4 mean", type = "l",
+     lwd = 2);
+abline(v = 100, col = "red", lty = "dotted", lwd = 0.8);
+abline(v = 210, col = "red", lty = "dotted", lwd = 0.8);
+```
+
+Initial conclusions
+=====================================
+
+These simulations show some interesting dynamics, though we will need to think more carefully before running them at a larger scale.
+Despite having population sizes on the order of tens of thousands, these ran quite quickly, with most simulations finishing up within 10-20 minutes.
+It would not be difficult to run a huge number in 1-2 months if necessary.
+Annoyingly, when we increase by another order of magnitude, the pace is glacial, with me stopping after trying it after 72 hours and no more than 70 generations.
+We might be able to get a bit more power through some creative refactoring of the code, and a bit if I made a version entirely in C.
+The longest step by far is sexual reproduction and allocating new offspring with diploid genomes.
+Haploid genomes and asexual populations are much, much faster in comparison.
+
+I am thinking that it might be worthwhile to monitor the allelic diversity in the population in some way.
+Individuals are initialised with unique alleles, which at first I feared my be a bit unrealistic, but the population size starts out small, so perhaps not so much.
+We can always allow more burn-in to be safe.
+
+It is nice to recover the sudden drop given the application of pesticide followed by the evolution of resistance and increase in population soon after.
+The strong selection is also reflected in the evolution of traits such that both pesticide traits are negative (indicating resistance).
+
+One thing that concerned me at first was that only one of the crop feeding traits is positive in each simulation.
+This means that the population is quickly specialising on one type of food resource.
+But this is, I think, exactly what we would predict from theory in an evolving population.
+Evolutionary branching should only happen under some specific circumstances, which we would need to parameterise explicitly if we wanted to recreate.
+Perhaps we need to have a bit weaker of a negative correlation for feeding, or model multiple specialist populations?
 

### 2021-08-10 notebook/scratch.R

@@ -18,18 +18,20 @@ gmt[3, 4] <- -0.5;
 gmt[4, 3] <- -0.5;
 gmt[3, 3] <-  1.0;
 gmt[4, 4] <-  1.0;
-mg  <- mine_gmatrix(gmatrix = gmt, loci = 6, indivs = 2000, npsize = 12000, 
-                    max_gen = 400, sampleK = 1200, chooseK = 6, layers = 4,
+mg  <- mine_gmatrix(gmatrix = gmt, loci = 12, indivs = 2000, npsize = 12000, 
+                    max_gen = 2400, sampleK = 1200, chooseK = 6, layers = 6,
                     mu_pr = 0.05, pr_cross = 0.05, mu_sd = 0.01, 
                     term_cri = -12);
 
+
+mg  <- readRDS("mg.rds");
 sim <- run_farm_sim(mine_output              = mg,
-                    N                        = 10000,
+                    N                        = 1000,
                     neutral_loci             = 1000,
-                    xdim                     = 40,
-                    ydim                     = 40,
+                    xdim                     = 144,
+                    ydim                     = 144,
                     repro                    = "sexual",
-                    max_age                  = 10,
+                    max_age                  = 4,
                     selfing                  = FALSE,
                     food_consume             = c("T1", "T2"),
                     pesticide_consume        = c("T3", "T4"),
@@ -39,29 +41,32 @@ sim <- run_farm_sim(mine_output              = mg,
                     pesticide_tolerated_surv = 0,
                     pesticide_rotation_type  = 2,
                     crop_rotation_type       = 2,
-                    min_age_reproduce        = 8,
-                    max_age_feed             = 6,
+                    min_age_reproduce        = 4,
+                    max_age_feed             = 2,
                     lambda_value             = 1.5,
-                    farms                    = 16,
-                    time_steps               = 120,
-                    mutation_pr              = 0.01,
-                    crossover_pr             = 0.01,
+                    farms                    = 24,
+                    time_steps               = 400,
+                    mutation_pr              = 0.001,
+                    crossover_pr             = 0.1,
                     net_mu_layers            = 0,
-                    crop_rotation_time       = 24,
-                    pesticide_rotation_time  = 24,
+                    crop_rotation_time       = 12,
+                    pesticide_rotation_time  = 12,
                     crop_per_cell            = 2,
                     pesticide_per_cell       = 2,
                     crop_number              = 2,
                     pesticide_number         = 2,
[…]

### 2021-08-19 notebook/helicoverpa_notes.Rmd
add a new notebook entry with some initial results

@@
 
+    - **AUG** [19](#u-19-AUG-2021) 
     - **JUL** [23](#u-23-JUL-2021)
@@
 ================================================================================
+
+> <a name="u-19-AUG-2021">Update: 19 AUG 2021</a>
+
+The helicoverpa package will be submitted to [CRAN](https://cran.r-project.org/) sometime before early August.
+What this will require is some appropriate testing functions and documentation, and maybe one or two vignettes that show how the package works.
+I'll also make a website similar to the [GMSE website](https://confoobio.github.io/gmse/).
+We will want to decide if the package name should actually be 'helicoverpa', or if we should switch to something else ([endorse](https://cran.r-project.org/package=endorse) is unfortunately taken; I'm not sure if 'ENDORSE' in all caps would work, but probably best to not try this and avoid confusion).
+I will update with more details about this later.
+
+For now, I want to focus on some simulations that I have recently run with the idea of laying the foundations for a general paper on landscape heterogeneity and pesticide resistance (not so much focusing on the details of the *Helicoverpa* system).
+I've not yet checked to see how novel the general ideas are because this started more as a verification of the model (the code is working as intended, as far as I can tell).
+
+The model focuses on the way that negative genetic correlations affect niche differentiation across different ploidy and mating types.
+The idea is that sex can act as a constraint to adaptation due to outbreeding depression in a way that cannot happen given asexual reproduction.
+Given a heterogeneous landscape, specifically in the context of the evolution of pesticide resistance and crop specialisation, asexually reproducing pests can get the upper hand by specialising on specific combinations of crop and pesticide and ensuring that their descendants will keep these specialised genotypes intact.
+In contrast, sexually reproducing organisms have a risk of outbreeding depression, breeding with a conspecific that is specialised to different conditions and having offspring that is not good at dealing with any environment.
+I have not seen this point in the literature, and I need to figure out the extent to which it has or has not been made and modelled to determine whether or not it is actually novel.
+
+What follows is a lot of simulation results from four different types of species set by the `repro` argument in the R package:
+
+- **asexual**: A haploid organism that reproduces asexually.
+- **selfing**: A monoecious diploid organism that reproduces sexually and can self-fertilise.
+- **non-selfing**: A monoecious diploid organism that reproduces sexually and cannot self-fertilise.
+- **biparental**: A dioecious diploid organism that reproduces sexually.
+
+For each of these different organism types, I included two different crop types and two pesticide types, and I ran three different pesticide and crop rotation procedures on a landscape with 24 farms for 480 time steps (maybe very loosely thinking of 12 time steps as a year):
+
+- **no_rotation**: No rotation of either crop or pesticide. Farms randomly select a crop type and a pesticide to use at the start of the simulation and use it for the whole simulation.
+- **pesticide_rotation**: Crop is not rotated, but pesticide is randomly selected every 12 time steps.
+Farms randomly select a crop type and pesticide to use at the start of the simulation, and farmers use the same crop each time step.
+- **both**: Rotation of both crop and pesticide.
+Farms randomly select a crop and pesticide to use at the start of the simulation, and then randomly select a new crop and pesticide to use every 12 time steps.
+
+```{r, echo = FALSE}
+mg_2021_AUG_19 <- readRDS("data/2021_AUG_19_data/mg.rds");
+```
+
+I'm deliberately not having a crop rotation only simulation as it seems unrealistic and potentially a bit redundant, at least for now.
+The details of the simulation are as follows.
+The genome was created with the `mine_gmatrix` settings below.
+
+```{r, eval = FALSE}
+mg  <- mine_gmatrix(gmatrix = gmt, loci = 12, indivs = 2000, npsize = 12000, 
+                    max_gen = 2400, sampleK = 1200, chooseK = 6, layers = 6,
+                    mu_pr = 0.05, pr_cross = 0.05, mu_sd = 0.01, 
+                    term_cri = -12);
+```
+
+The output of `mg` is below. Note that list element 2 shows the target variance covariance matrix of four traits. Element 6 below gives the mined matrix.
+
+```{r, echo = FALSE}
+print(mg_2021_AUG_19);
+```
+
+This output was used to run simulation with the following parameter values (asexual without rotation is shown below).
+
+```{r, eval = FALSE}
+sim <- run_farm_sim(mine_output              = mg,
+                    N                        = 1000,
+                    neutral_loci             = 1000,
+                    xdim                     = 100,
+                    ydim                     = 100,
+                    repro                    = "asexual",
+                    max_age                  = 6,
+                    selfing                  = TRUE,
+                    food_consume             = c("T1", "T2"),
+                    pesticide_consume        = c("T3", "T4"),
+                    food_needed_surv         = 1,
+                    food_needed_repr         = 1,
+                    reproduction_type        = "food_based",
+                    pesticide_tolerated_surv = 0,
+                    pesticide_rotation_type  = 2,
+                    crop_rotation_type       = 2,
+                    min_age_reproduce        = 4,
+                    max_age_feed             = 3,
+                    lambda_value             = 1.5,
+                    farms                    = 24,
+                    time_steps               = 480,
+                    mutation_pr              = 0.001,
+                    crossover_pr             = 0.1,
+                    net_mu_layers            = 0,
+                    crop_rotation_time       = 12,
+                    pesticide_rotation_time  = 12,
+                    crop_per_cell            = 8,
+                    pesticide_per_cell       = 0.4,
+                    crop_number              = 2,
+                    pesticide_number         = 2,
+                    K_on_birth               = 1000000,
+                    min_age_move             = 4,
+                    age_food_threshold       = 2,
+                    min_age_feed             = 1,
+                    pesticide_start          = 100,
+                    print_last               = TRUE,
+                    mating_distance          = 2,
+);
+```
+
+I have left out some arguments that were set to their default options.
+Note that we have a 100 by 100 landscape.
+Two traits determine how much of crop an individual consumes at a time ("T1" and "T2"), and two separate traits determine how much of two types of pesticide an individual imbibes ("T3", "T4").
+Individuals start feeding in their first time step and stop feeding after their third time step.
+On their four time step, individuals can move and reproduce, and they have a maximum age of 6 (though mortality happens at the end of a time step).
+Each cell has 8 units of crop that an individual can eat, and the crop is replaced each time the land is rotated (i.e., individuals can keep eating and feeding on it until there is either nothing left on a cell or the farmer plants a new crop after 12 time steps).
+Individuals must eat at least 1 unit to survive and reproduce, and can starve on or after their second time step.
+Individuals also cannot tolerate any pesticide (note, 'zero' is kind of arbitrary, but it's an easy number to use).
+
+The following shows images of all of the different scenarios.
+For each scenario, I will show an image of the landscape on the last time step.
+Different coloured squares show different farms, and the points on top of them are pests.
+I will also show the dynamics of pest abundance and traits over time, and I will show scatterplots of the correlations between traits.
+
+Note that the pesticide application *starts* on time step 100 to allow a period of burn-in.
+
+**Asexual, no rotation**
+
+![](images/2021_AUG_19/asexual_no_rotation_space.png)
+
+![](images/2021_AUG_19/asexual_no_rotation_time.png)
+
+![](images/2021_AUG_19/asexual_no_rotation_traits.png)
+
+**Asexual, pesticide rotation**
+
+![](images/2021_AUG_19/asexual_pesticide_rotation_space.png)
+
+![](images/2021_AUG_19/asexual_pesticide_rotation_time.png)
+
+![](images/2021_AUG_19/asexual_pesticide_rotation_traits.png)
+
+**Asexual, both rotation**
+
+![](images/2021_AUG_19/asexual_both_rotation_space.png)
+
+![](images/2021_AUG_19/asexual_both_rotation_time.png)
+
+![](images/2021_AUG_19/asexual_both_rotation_traits.png)
+
+**Selfing, no rotation**
+
+![](images/2021_AUG_19/sexual_selfing_no_rotation_space.png)
+
+![](images/2021_AUG_19/sexual_selfing_no_rotation_time.png)
+
+![](images/2021_AUG_19/sexual_selfing_no_rotation_traits.png)
+
+**Selfing, pesticide rotation**
+
+![](images/2021_AUG_19/sexual_selfing_pesticide_rotation_space.png)
+
+![](images/2021_AUG_19/sexual_selfing_pesticide_rotation_time.png)
+
+![](images/2021_AUG_19/sexual_selfing_pesticide_rotation_traits.png)
+
+**Selfing, both rotation**
+
+![](images/2021_AUG_19/sexual_selfing_both_rotation_space.png)
+
+![](images/2021_AUG_19/sexual_selfing_both_rotation_time.png)
+
+![](images/2021_AUG_19/sexual_selfing_both_rotation_traits.png)
+
+**Non-selfing, no rotation**
+
+![](images/2021_AUG_19/sexual_noselfing_no_rotation_space.png)
+
+![](images/2021_AUG_19/sexual_noselfing_no_rotation_time.png)
+
+![](images/2021_AUG_19/sexual_noselfing_no_rotation_traits.png)
+
+**Non-selfing, pesticide rotation**
+
+![](images/2021_AUG_19/sexual_noselfing_pesticide_rotation_space.png)
+
+![](images/2021_AUG_19/sexual_noselfing_pesticide_rotation_time.png)
+
+![](images/2021_AUG_19/sexual_noselfing_pesticide_rotation_traits.png)
+
+**Non-selfing, both rotation**
+
+![](images/2021_AUG_19/sexual_noselfing_both_rotation_space.png)
+
+![](images/2021_AUG_19/sexual_noselfing_both_rotation_time.png)
+
+![](images/2021_AUG_19/sexual_noselfing_both_rotation_traits.png)
+
+**Biparental, no rotation**
+
+![](images/2021_AUG_19/biparental_no_rotation_space.png)
+
+![](images/2021_AUG_19/biparental_no_rotation_time.png)
+
+![](images/2021_AUG_19/biparental_no_rotation_traits.png)
+
+**Biparental, pesticide rotation**
+
+![](images/2021_AUG_19/biparental_pesticide_rotation_space.png)
+
+![](images/2021_AUG_19/biparental_pesticide_rotation_time.png)
+
+![](images/2021_AUG_19/biparental_pesticide_rotation_traits.png)
+
+**Biparental, both rotation**
+
+![](images/2021_AUG_19/biparental_both_rotation_space.png)
+
+![](images/2021_AUG_19/biparental_both_rotation_time.png)
+
+![](images/2021_AUG_19/biparental_both_rotation_traits.png)
+
+It's worth mentioning that the reason for the oscillating population size is due to individuals sometimes consuming all of the food and starting to decrease in population density before the next crop is put down.
+To get a full picture on what is going on, I will need to run multiple replicate simulations with different mined genomes (i.e., same genetic correlations, but potentially achieved in a different way).
+I will also want to run the same simulations with an identity matrix (i.e., no genetic correlations) to act as a control so that we can understand the role of the negative genetic correlations in all of these results.
+
+I have not really given all of these figures a hard look to try to think more clearly about what's going on yet -- at least not apart from the outbreeding depression idea.
 

### 2021-08-19 notebook/scratch.R

@@ -2,6 +2,7 @@
 # Example with two traits
 ################################################################################
 library(helicoverpa);
+set.seed(Sys.time());
 gmt <- matrix(data = 0, nrow = 4, ncol = 4);
 diag(gmt) <- 1;
 gmt[1, 2] <-  -0.5;
@@ -18,10 +19,10 @@ gmt[3, 4] <- -0.5;
 gmt[4, 3] <- -0.5;
 gmt[3, 3] <-  1.0;
 gmt[4, 4] <-  1.0;
-mg  <- mine_gmatrix(gmatrix = gmt, loci = 12, indivs = 2000, npsize = 12000, 
+mg  <- mine_gmatrix(gmatrix = gmt, loci = 12, indivs = 2000, npsize = 18000, 
                     max_gen = 2400, sampleK = 1200, chooseK = 6, layers = 6,
-                    mu_pr = 0.05, pr_cross = 0.05, mu_sd = 0.01, 
-                    term_cri = -12);
+                    mu_pr = 0.01, pr_cross = 0.01, mu_sd = 0.01, 
+                    term_cri = -9);
 
 
 mg  <- readRDS("mg.rds");
@@ -74,6 +75,14 @@ sim <- run_farm_sim(mine_output              = mg,
 ################################################################################
 ################################################################################
 ################################################################################
+
+## OLD CALL:
+
+mg  <- mine_gmatrix(gmatrix = gmt, loci = 12, indivs = 2000, npsize = 12000, 
+                    max_gen = 2400, sampleK = 1200, chooseK = 6, layers = 6,
+                    mu_pr = 0.05, pr_cross = 0.05, mu_sd = 0.01, 
+                    term_cri = -12);
+
 
 # Initialising a file
 #library(helicoverpa)

### 2021-08-19 notebook/testing/sim.R

@@ -1,0 +1,124 @@
+library("helicoverpa");
+set.seed(Sys.time());
+mg  <- readRDS("mg.rds");
+sim <- run_farm_sim(mine_output              = mg,
+                    N                        = 1000,
+                    neutral_loci             = 1000,
+                    xdim                     = 144,
+                    ydim                     = 144,
+                    repro                    = "biparental",
+                    max_age                  = 6,
+                    selfing                  = FALSE,
+                    food_consume             = c("T1", "T2"),
+                    pesticide_consume        = c("T3", "T4"),
+                    food_needed_surv         = 1,
+                    food_needed_repr         = 1,
+                    reproduction_type        = "food_based",
+                    pesticide_tolerated_surv = 0,
+                    pesticide_rotation_type  = 1,
+                    crop_rotation_type       = 1,
+                    min_age_reproduce        = 4,
+                    max_age_feed             = 3,
+                    lambda_value             = 1.5,
+                    farms                    = 24,
+                    time_steps               = 480,
+                    mutation_pr              = 0.001,
+                    crossover_pr             = 0.1,
+                    net_mu_layers            = 0,
+                    crop_rotation_time       = 12,
+                    pesticide_rotation_time  = 12,
+                    crop_per_cell            = 8,
+                    pesticide_per_cell       = 0.4,
+                    crop_number              = 2,
+                    pesticide_number         = 2,
+                    print_inds               = FALSE, # Careful with this one
+                    K_on_birth               = 1000000,
+                    min_age_move             = 4,
+                    age_food_threshold       = 2,
+                    min_age_feed             = 1,
+                    pesticide_start          = 100,
+                    print_last               = TRUE,
[…]

### 2021-08-19 notebook/helicoverpa_notes.Rmd
Fix typo in the notes

@@
 
-The helicoverpa package will be submitted to [CRAN](https://cran.r-project.org/) sometime before early August.
-What this will require is some appropriate testing functions and documentation, and maybe one or two vignettes that show how the package works.
+The helicoverpa package will be submitted to [CRAN](https://cran.r-project.org/) sometime before early October What this will require is some appropriate testing functions and documentation, and maybe one or two vignettes that show how the package works.
 I'll also make a website similar to the [GMSE website](https://confoobio.github.io/gmse/).

### 2021-08-26 notebook/helicoverpa_notes.Rmd
Add a notebook entry with output for the null simulations

@@
 
-    - **AUG** [19](#u-19-AUG-2021) 
+    - **AUG** [19](#u-19-AUG-2021), [26](#u-26-AUG-2021)
     - **JUL** [23](#u-23-JUL-2021)
@@
 ================================================================================
+
+> <a name="u-26-AUG-2021">Update: 26 AUG 2021</a>
+
+I have now run the same analysis that I did on [19 AUG 2021](#u-19-AUG-2021), but with traits that evolve independently.
+This was done by setting a covariance matrix to be the identity matrix for four traits.
+Here is the `mine_gmatrix` output that was used for the simulations.
+
+```{r, echo = FALSE}
+mg_2021_AUG_26 <- readRDS("data/2021_AUG_26_data/mg_n1.rds");
+print(mg_2021_AUG_26);
+```
+
+Note that the parameter values for mining the matrix were identical to those chosen in the simulations from [19 AUG 2021](#u-19-AUG-2021).
+The results are as expected; all traits evolve in response to selection to feed on both crop types and resist both pesticide types.
+
+**Asexual, no rotation**
+
+![](images/2021_AUG_26/asexual_no_rotation_space_NULL.png)
+
+![](images/2021_AUG_26/asexual_no_rotation_time_NULL.png)
+
+![](images/2021_AUG_26/asexual_no_rotation_traits_NULL.png)
+
+**Asexual, pesticide rotation**
+
+![](images/2021_AUG_26/asexual_pesticide_rotation_space_NULL.png)
+
+![](images/2021_AUG_26/asexual_pesticide_rotation_time_NULL.png)
+
+![](images/2021_AUG_26/asexual_pesticide_rotation_traits_NULL.png)
+
+**Asexual, both rotation**
+
+![](images/2021_AUG_26/asexual_both_rotation_space_NULL.png)
+
+![](images/2021_AUG_26/asexual_both_rotation_time_NULL.png)
+
+![](images/2021_AUG_26/asexual_both_rotation_traits_NULL.png)
+
+**Selfing, no rotation**
+
+![](images/2021_AUG_26/sexual_selfing_no_rotation_space_NULL.png)
+
+![](images/2021_AUG_26/sexual_selfing_no_rotation_time_NULL.png)
+
+![](images/2021_AUG_26/sexual_selfing_no_rotation_traits_NULL.png)
+
+**Selfing, pesticide rotation**
+
+![](images/2021_AUG_26/sexual_selfing_pesticide_rotation_space_NULL.png)
+
+![](images/2021_AUG_26/sexual_selfing_pesticide_rotation_time_NULL.png)
+
+![](images/2021_AUG_26/sexual_selfing_pesticide_rotation_traits_NULL.png)
+
+**Selfing, both rotation**
+
+![](images/2021_AUG_26/sexual_selfing_both_rotation_space_NULL.png)
+
+![](images/2021_AUG_26/sexual_selfing_both_rotation_time_NULL.png)
+
+![](images/2021_AUG_26/sexual_selfing_both_rotation_traits_NULL.png)
+
+**Non-selfing, no rotation**
+
+![](images/2021_AUG_26/sexual_noselfing_no_rotation_space_NULL.png)
+
+![](images/2021_AUG_26/sexual_noselfing_no_rotation_time_NULL.png)
+
+![](images/2021_AUG_26/sexual_noselfing_no_rotation_traits_NULL.png)
+
+**Non-selfing, pesticide rotation**
+
+![](images/2021_AUG_26/sexual_noselfing_pesticide_rotation_space_NULL.png)
+
+![](images/2021_AUG_26/sexual_noselfing_pesticide_rotation_time_NULL.png)
+
+![](images/2021_AUG_26/sexual_noselfing_pesticide_rotation_traits_NULL.png)
+
+**Non-selfing, both rotation**
+
+![](images/2021_AUG_26/sexual_noselfing_both_rotation_space_NULL.png)
+
+![](images/2021_AUG_26/sexual_noselfing_both_rotation_time_NULL.png)
+
+![](images/2021_AUG_26/sexual_noselfing_both_rotation_traits_NULL.png)
+
+**Biparental, no rotation**
+
+![](images/2021_AUG_26/biparental_no_rotation_space_NULL.png)
+
+![](images/2021_AUG_26/biparental_no_rotation_time_NULL.png)
+
+![](images/2021_AUG_26/biparental_no_rotation_traits_NULL.png)
+
+**Biparental, pesticide rotation**
+
+![](images/2021_AUG_26/biparental_pesticide_rotation_space_NULL.png)
+
+![](images/2021_AUG_26/biparental_pesticide_rotation_time_NULL.png)
+
+![](images/2021_AUG_26/biparental_pesticide_rotation_traits_NULL.png)
+
+**Biparental, both rotation**
+
+![](images/2021_AUG_26/biparental_both_rotation_space_NULL.png)
+
+![](images/2021_AUG_26/biparental_both_rotation_time_NULL.png)
+
+![](images/2021_AUG_26/biparental_both_rotation_traits_NULL.png)
+
+For full simulations, I think it will be important to have each replicate (and I'm including all of the above as one replicate) with its own mined gmatrix.
 
@@
 json <- results_to_json(pest = sim$PEST_DATA[[19]], land = sim$LAND_DATA[[19]],
-                        filename = "../data/sample_sim.json");
+                        filename = "toy_model/data/sample_sim.json");
 ```
@@
 sim_json <- results_to_json(sim = sim[[19]], 
-                            filename = "../data/sample_sim.json");
+                            filename = "toy_model/data/sample_sim.json");
 ```

### 2021-09-04 R/gmatrices.R
Add some mine_gmatrix outputs to the R package to be used as examples in the vignette and by users

@@ -1,0 +1,96 @@
+#' Sample mine_gmatrix outputs
+#'
+#' Output from mine_gmatrix that maps random normals from 12 loci to four traits
+#' with a covariance structure of the identity matrix (replicate 1)
+#'
+#' @format A list of 8 elements used in individual-based model initialisation
+"mg_n1"
+
+#' Sample mine_gmatrix outputs
+#'
+#' Output from mine_gmatrix that maps random normals from 12 loci to four traits
+#' with a covariance structure of the identity matrix (replicate 2)
+#'
+#' @format A list of 8 elements used in individual-based model initialisation
+"mg_n2"
+
+#' Sample mine_gmatrix outputs
+#'
+#' Output from mine_gmatrix that maps random normals from 12 loci to four traits
+#' with a covariance structure of the identity matrix (replicate 3)
+#'
+#' @format A list of 8 elements used in individual-based model initialisation
+"mg_n3"
+
+#' Sample mine_gmatrix outputs
+#'
+#' Output from mine_gmatrix that maps random normals from 12 loci to four traits
+#' with a covariance structure  of the identity matrix (replicate 4)
+#'
+#' @format A list of 8 elements used in individual-based model initialisation
+"mg_n4"
+
+#' Sample mine_gmatrix outputs
+#'
+#' Output from mine_gmatrix that maps random normals from 12 loci to four traits
+#' with a covariance structure of the identity matrix (replicate 5)
+#'
+#' @format A list of 8 elements used in individual-based model initialisation
+"mg_n5"
+
+
+#' Sample mine_gmatrix outputs
+#'
+#' Output from mine_gmatrix that maps random normals from 12 loci to four traits
+#' with a covariance structure including negative and positive values 
+#' (replicate 1)
+#'
+#' @format A list of 8 elements used in individual-based model initialisation
+"mg_v1"
+
+#' Sample mine_gmatrix outputs
+#'
+#' Output from mine_gmatrix that maps random normals from 12 loci to four traits
+#' with a covariance structure including negative and positive values 
+#' (replicate 2)
+#'
+#' @format A list of 8 elements used in individual-based model initialisation
+"mg_v2"
+
+
+#' Sample mine_gmatrix outputs
+#'
+#' Output from mine_gmatrix that maps random normals from 12 loci to four traits
+#' with a covariance structure including negative and positive values 
+#' (replicate 3)
+#'
[…]

### 2021-09-04 notebook/scratch.R
Finish the section on mine_gmatrix function in the vignette

@@ -319,4 +319,15 @@ sim <- run_farm_sim(mine_output              = mg,
                     pesticide_start          = 200,
 );
 
+##########################
+##########################
+##########################
+
+
+
+
+
+
+
+
 

### 2021-09-04 src/feeding.c
Introduce a function to refresh consumption in each time step and thereby resolve issue #38 -- where individuals were getting credit for food that they had already consumed and therefore should not be double counting

@@ -202,6 +202,43 @@ void feed(double **pests, double *paras, double ***land, int ind){
 }
 
 /* =============================================================================
+ * Processes all of the food consumed by individuals
+ *     pests: The array holding the parent's information
+ *     paras: The paras vector that holds global information
+ * ========================================================================== */
+void refresh_consumed(double **pests, double *paras){
+
+  int ind, N, consumed_col10;
+  int consumed_col1, consumed_col2, consumed_col3, consumed_col4, consumed_col5;
+  int consumed_col6, consumed_col7, consumed_col8, consumed_col9;
+
+  N              = (int) paras[101];
+  consumed_col1  = (int) paras[58];
+  consumed_col2  = (int) paras[59];
+  consumed_col3  = (int) paras[60];
+  consumed_col4  = (int) paras[61];
+  consumed_col5  = (int) paras[62];
+  consumed_col6  = (int) paras[63];
+  consumed_col7  = (int) paras[64];
+  consumed_col8  = (int) paras[65];
+  consumed_col9  = (int) paras[66];
+  consumed_col10 = (int) paras[67];
+  
+  for(ind = 0; ind < N; ind++){
+      pests[ind][consumed_col1]  = 0.0;
+      pests[ind][consumed_col2]  = 0.0;
+      pests[ind][consumed_col3]  = 0.0;
+      pests[ind][consumed_col4]  = 0.0;
+      pests[ind][consumed_col5]  = 0.0;
+      pests[ind][consumed_col6]  = 0.0;
+      pests[ind][consumed_col7]  = 0.0;
+      pests[ind][consumed_col8]  = 0.0;
+      pests[ind][consumed_col9]  = 0.0;
+      pests[ind][consumed_col10] = 0.0;
+  }
+}
+
+/* =============================================================================
  * Feeds all individuals on the x-y location where it is located
  *     pests: The array holding the parent's information
  *     paras: The paras vector that holds global information
@@ -210,6 +247,8 @@ void feed(double **pests, double *paras, double ***land, int ind){
 void feeding(double **pests, double *paras, double ***land){
   
   int ind, N, *not_fed, N_count;
+  
+  refresh_consumed(pests, paras);
   
   N       = (int) paras[101];
   not_fed = malloc(N * sizeof(int));

### 2021-09-04 src/feeding.h

@@ -6,4 +6,6 @@
 
 void feed(double **pests, double *paras, double ***land, int ind);
 
+void refresh_consumed(double **pests, double *paras);
+
 void feeding(double **pests, double *paras, double ***land);

### 2021-09-04 src/age.c
Introduce metabolism into the aging function, which burns some of the consumed food

@@ -7,12 +7,20 @@
  * ========================================================================== */
 void age_pests(double **pests, double *paras){
   
-  int ind, N, age_col;
+  int ind, N, age_col, meta_col, cons_col;
   
-  age_col = (int) paras[3];
-  N       = (int) paras[101];
+  age_col  = (int) paras[3];
+  N        = (int) paras[101];
+  meta_col = (int) paras[86];
+  cons_col = (int) paras[14];
 
   for(ind = 0; ind < N; ind++){
+    if(pests[ind][age_col] > 0){
+        pests[ind][cons_col] -= pests[ind][meta_col];
+        if(pests[ind][cons_col] < 0){
+            pests[ind][cons_col] = 0;
+        }
+    }
     pests[ind][age_col]++;
   }
   

### 2021-09-04 R/initialise.R
Introduce the metabolism function in R so that it can also be modelled as a trait of an individual

@@ -42,6 +42,7 @@
 #'@param mortality_type Type of mortality (currently only one option)
 #'@param age_food_threshold Age at which food threshold is enacted
 #'@param age_pesticide_threshold Age at which pesticide threshold is enacted
+#'@param metabolism The amount of consumed food lost each time step
 #'@return A two-dimensional array of individuals for simulation
 #'@examples
 #'gmt       <- matrix(data = 0, nrow = 2, ncol = 2);
@@ -80,7 +81,8 @@ initialise_inds <- function(mine_output,
                             feed_while_moving = FALSE,
                             mortality_type = 0,
                             age_food_threshold = NA,
-                            age_pesticide_threshold = NA){
+                            age_pesticide_threshold = NA,
+                            metabolism = 0){
   
   food      <- rep(x = 0, times = 10);
   pesticide <- rep(x = 0, times = 10);
@@ -99,8 +101,16 @@ initialise_inds <- function(mine_output,
     age_pesticide_threshold <- 0;
   }
   
+  if(N < 2){
+    stop("ERROR: Must initialise with at least two individuals.");
+  }
+  
+  if(neutral_loci < 10){
+    stop("ERROR: Must initialise with at least 10 neutral loci.");
+  }
+  
   if(repro != "asexual" & repro != "sexual" & repro != "biparental"){
-    stop("ERROR: Must specify 'repro' as asexual, sexual, or biparental.")
+    stop("ERROR: Must specify 'repro' as asexual, sexual, or biparental.");
   }
   if(repro == "sexual" | repro == "biparental"){
     inds      <- build_sexual(mine_output, N, neutral_loci);
@@ -185,6 +195,7 @@ initialise_inds <- function(mine_output,
   inds[, 81] <-  max_age;
   inds[, 83] <-  age_food_threshold;
   inds[, 84] <-  age_pesticide_threshold;
+  inds[, 87] <-  metabolism;
   
   return(inds);
 }

### 2021-09-04 R/sim_farming.R

@@ -81,6 +81,7 @@
 #'@param immigration_rate Mean number of immigrants per time step
 #'@param get_f_coef Get the inbreeding coefficient (not for asexual)
 #'@param get_stats Get population level statistics in a CSV printout
+#'@param metabolism The rate at which food consumed is burned in a time step
 #'@return The output in the R console is a list with two elements; the first 
 #'element is a vector of parameter values used by the model, and the second 
 #'element is the landscape in the simulation. The most relevant output will be
@@ -164,7 +165,8 @@ run_farm_sim <- function(mine_output,
                          pesticide_start = 0,
                          immigration_rate = 0,
                          get_f_coef = FALSE,
-                         get_stats = TRUE){
+                         get_stats = TRUE,
+                         metabolism = 0){
   
     land <- make_landscape(rows = ydim, cols = xdim, depth = 21, farms = farms);
   
@@ -176,6 +178,7 @@ run_farm_sim <- function(mine_output,
     mating_distance_T           <- check_is_trait(mating_distance);
     lambda_value_T              <- check_is_trait(lambda_value);
     movement_bouts_T            <- check_is_trait(movement_bouts);
+    metabolism_T                <- check_is_trait(metabolism);
     
     # Check to see if these are traits
     food_consume_T <- rep(x = 0, times = length(food_consume));
@@ -226,6 +229,11 @@ run_farm_sim <- function(mine_output,
     move_bout <- movement_bouts;
     if(movement_bouts_T == TRUE){
         move_bout <- 0;
+    }
+    
+    metab_rate <- metabolism;
+    if(metabolism_T == TRUE){
+        metab_rate <- 0;
     }
     
     if(N < 5){
@@ -283,7 +291,8 @@ run_farm_sim <- function(mine_output,
                             feed_while_moving        = feed_while_moving,
                             mortality_type           = mortality_type,
                             age_food_threshold       = age_food_threshold,
-                            age_pesticide_threshold  = age_pesticide_threshold);
+                            age_pesticide_threshold  = age_pesticide_threshold,
+                            metabolism               = metab_rate);
     
     sim_results <- sim_crops(pests                    = pest, 
                              land                     = land,
@@ -325,7 +334,8 @@ run_farm_sim <- function(mine_output,
                              pesticide_start          = pesticide_start,
                              immigration_rate         = immigration_rate,
                              get_f_coef               = get_f_coef,
[…]

### 2021-09-04 tests/testthat/test-sim_farming.R

@@ -39,7 +39,8 @@ test_that("Simulations are run and output is produced", {
                               print_inds = FALSE, print_gens = FALSE,
                               print_last = FALSE, K_on_birth = 1000,
                               pesticide_start = 0, immigration_rate = 0,
-                              get_f_coef = FALSE, get_stats = FALSE);
+                              get_f_coef = FALSE, get_stats = FALSE,
+                              metabolism = 0);
     expect_equal(length(sim[[1]]), 173);
     expect_equal(dim(sim[[2]]), c(40, 40, 21));
 })

### 2021-09-04 src/age.c
Just get the floored metabolic rate done so that this does not come back to be a problem for Issue #38 -- I will still want to do the baseline rate tomorrow

@@ -8,6 +8,7 @@
 void age_pests(double **pests, double *paras){
   
   int ind, N, age_col, meta_col, cons_col;
+  double meta_rate;
   
   age_col  = (int) paras[3];
   N        = (int) paras[101];
@@ -15,8 +16,12 @@ void age_pests(double **pests, double *paras){
   cons_col = (int) paras[14];
 
   for(ind = 0; ind < N; ind++){
+    meta_rate = pests[ind][meta_col];
+    if(meta_rate < 0){
+      meta_rate = 0.0;
+    }
     if(pests[ind][age_col] > 0){
-        pests[ind][cons_col] -= pests[ind][meta_col];
+        pests[ind][cons_col] -= meta_rate;
         if(pests[ind][cons_col] < 0){
             pests[ind][cons_col] = 0;
         }

### 2021-09-04 src/age.c
This is not an important commit -- I'm just cleaning some files up

@@ -28,5 +28,4 @@ void age_pests(double **pests, double *paras){
     }
     pests[ind][age_col]++;
   }
-  
 }

### 2021-09-05 src/reproduction.c
Add the threshold of pesticide consumed for reproduction

@@ -153,25 +153,31 @@ void count_offspring(double **pests, double *paras, int row){
  * ========================================================================== */
 void calculate_offspring(double **pests, double *paras){
   
-  int ind, N, tot_offspring, birth_K, off_col, sex_col, sex;
-  int age_col, age, min_age_col, max_age_col, min_age, max_age;
+  int ind, N, tot_offspring, birth_K, off_col, sex_col, sex, pest_thresh_col;
+  int age_col, age, min_age_col, max_age_col, min_age, max_age, pest_consum_col;
+  double pesticide_thresh, pesticide_consumed;
 
-  age_col     = (int) paras[3];
-  sex_col     = (int) paras[4];
-  off_col     = (int) paras[10];
-  min_age_col = (int) paras[35];
-  max_age_col = (int) paras[36];
-  N           = (int) paras[101];
-  birth_K     = (int) paras[167];
+  age_col         = (int) paras[3];
+  sex_col         = (int) paras[4];
+  off_col         = (int) paras[10];
+  pest_consum_col = (int) paras[15];
+  pest_thresh_col = (int) paras[19];
+  min_age_col     = (int) paras[35];
+  max_age_col     = (int) paras[36];
+  N               = (int) paras[101];
+  birth_K         = (int) paras[167];
   
-  paras[106] = 0.0; /* Start with no offspring */
+  paras[106]         = 0.0; /* Start with no offspring */
   
   for(ind = 0; ind < N; ind++){
-      sex     = (int) pests[ind][sex_col];
-      age     = (int) pests[ind][age_col];
-      min_age = (int) pests[ind][min_age_col];
-      max_age = (int) pests[ind][max_age_col];
-      if(age >= min_age && age <= max_age && sex < 3){
+      sex                = (int) pests[ind][sex_col];
+      age                = (int) pests[ind][age_col];
+      min_age            = (int) pests[ind][min_age_col];
+      max_age            = (int) pests[ind][max_age_col];
+      pesticide_thresh   = pests[ind][pest_thresh_col];
+      pesticide_consumed = pests[ind][pest_consum_col];
+      if(age >= min_age && age <= max_age && sex < 3 && 
+         pesticide_consumed < pesticide_thresh){
           count_offspring(pests, paras, ind);
       }
   }

### 2021-09-05 R/initialise.R
Set up a baseline metabolic rate so that variation around the rate can be specified more easily. Now, e.g., a baseline metabolic rate of 1 could be specified with individuals allowing to vary with a trait plus or minus 0.2

@@ -43,6 +43,7 @@
 #'@param age_food_threshold Age at which food threshold is enacted
 #'@param age_pesticide_threshold Age at which pesticide threshold is enacted
 #'@param metabolism The amount of consumed food lost each time step
+#'@param baseline_metabolism A fixed baseline rate added to 'metabolism'
 #'@return A two-dimensional array of individuals for simulation
 #'@examples
 #'gmt       <- matrix(data = 0, nrow = 2, ncol = 2);
@@ -82,7 +83,8 @@ initialise_inds <- function(mine_output,
                             mortality_type = 0,
                             age_food_threshold = NA,
                             age_pesticide_threshold = NA,
-                            metabolism = 0){
+                            metabolism = 0,
+                            baseline_metabolism = 0){
   
   food      <- rep(x = 0, times = 10);
   pesticide <- rep(x = 0, times = 10);
@@ -196,6 +198,7 @@ initialise_inds <- function(mine_output,
   inds[, 83] <-  age_food_threshold;
   inds[, 84] <-  age_pesticide_threshold;
   inds[, 87] <-  metabolism;
+  inds[, 88] <-  baseline_metabolism;
   
   return(inds);
 }

### 2021-09-05 R/sim_farming.R

@@ -82,6 +82,7 @@
 #'@param get_f_coef Get the inbreeding coefficient (not for asexual)
 #'@param get_stats Get population level statistics in a CSV printout
 #'@param metabolism The rate at which food consumed is burned in a time step
+#'@param baseline_metabolism A fixed baseline rate added to 'metabolism'
 #'@return The output in the R console is a list with two elements; the first 
 #'element is a vector of parameter values used by the model, and the second 
 #'element is the landscape in the simulation. The most relevant output will be
@@ -166,7 +167,8 @@ run_farm_sim <- function(mine_output,
                          immigration_rate = 0,
                          get_f_coef = FALSE,
                          get_stats = TRUE,
-                         metabolism = 0){
+                         metabolism = 0,
+                         baseline_metabolism = 0){
   
     land <- make_landscape(rows = ydim, cols = xdim, depth = 21, farms = farms);
   
@@ -292,7 +294,8 @@ run_farm_sim <- function(mine_output,
                             mortality_type           = mortality_type,
                             age_food_threshold       = age_food_threshold,
                             age_pesticide_threshold  = age_pesticide_threshold,
-                            metabolism               = metab_rate);
+                            metabolism               = metab_rate,
+                            baseline_metabolism      = baseline_metabolism);
     
     sim_results <- sim_crops(pests                    = pest, 
                              land                     = land,
@@ -508,7 +511,7 @@ sim_crops <- function(pests,
               84.0,   # 84) pests column for inbreeding coefficient
               85.0,   # 85) pests column for lamba adjustment
               86.0,   # 86) pests column for metabolic rate (food lost)
-              87.0,   # 87)
+              87.0,   # 87) pests column for baseline metabolic rate (fixed)
               88.0,   # 88)
               89.0,   # 89)
               90.0,   # 90)

### 2021-09-05 src/age.c

@@ -7,16 +7,17 @@
  * ========================================================================== */
 void age_pests(double **pests, double *paras){
   
-  int ind, N, age_col, meta_col, cons_col;
+  int ind, N, age_col, meta_col, cons_col, base_m_col;
   double meta_rate;
   
-  age_col  = (int) paras[3];
-  N        = (int) paras[101];
-  meta_col = (int) paras[86];
-  cons_col = (int) paras[14];
+  age_col    = (int) paras[3];
+  N          = (int) paras[101];
+  meta_col   = (int) paras[86];
+  base_m_col = (int) paras[87];
+  cons_col   = (int) paras[14];
 
   for(ind = 0; ind < N; ind++){
-    meta_rate = pests[ind][meta_col];
+    meta_rate = pests[ind][base_m_col] + pests[ind][meta_col];
     if(meta_rate < 0){
       meta_rate = 0.0;
     }

### 2021-09-05 src/mortality.c
Clean up some of the reproduction functions a bit and assign a threshold pesticide tolerance for reproduction, which had not yet been done

@@ -38,7 +38,7 @@ void mortality(double **pests, double *paras, int ind){
   
   switch(mortality_type){
       case 0: /* Mortality if too old, not enough food, too much pesticide */
-          if(age > max_age){
+          if(age >= max_age){
               pests[ind][mortality_col] = 1.0;
           }
           if(food_consumed < food_threshold && age >= food_age){
@@ -49,7 +49,7 @@ void mortality(double **pests, double *paras, int ind){
           }
           break;
       default: /* Mortality if too old, not enough food, too much pesticide */
-          if(age > max_age){
+          if(age >= max_age){
               pests[ind][mortality_col] = 1.0;
           }
           if(food_consumed < food_threshold && age >= food_age){

### 2021-09-05 src/reproduction.c

@@ -10,7 +10,7 @@
 int mate_in_range(double **pests, double *paras, int row, int mate_sex){
   
   int ind, N, range_col, range, sex_col;
-  int range_count, opp_sex, in_range, selfing, focal_sex;
+  int range_count, opp_sex, in_range, selfing, selfing_col;
   int age_col, min_age_col, max_age_col, age, min_age, max_age;
   
   sex_col   = (int) paras[4];
@@ -19,16 +19,16 @@ int mate_in_range(double **pests, double *paras, int row, int mate_sex){
   range     = pests[row][range_col];
   
   age_col         = (int) paras[3];
+  selfing_col     = (int) paras[26];
   min_age_col     = (int) paras[35];
   max_age_col     = (int) paras[36];
-  
-  focal_sex = (int) pests[ind][sex_col];
-  
+ 
   range_count = 0;
   for(ind = 0; ind < N; ind++){
     in_range = is_in_range(pests, row, ind, paras, range);
     opp_sex  = (int) pests[ind][sex_col];
     age      = (int) pests[ind][age_col];
+    selfing  = (int) pests[ind][selfing_col];
     min_age  = (int) pests[ind][min_age_col];
     max_age  = (int) pests[ind][max_age_col];
     if(in_range > 0 && opp_sex == mate_sex && age >= min_age && age <= max_age){
@@ -49,14 +49,11 @@ int mate_in_range(double **pests, double *paras, int row, int mate_sex){
  * ========================================================================== */
 int mate_available(double **pests, double *paras, int row){
   
-  int N, ind, sex, sex_col, mate_found,selfing_col, selfing;
+  int sex, sex_col, mate_found;
   
-  N            = (int) paras[101];
   sex_col      = (int) paras[4];
   sex          = (int) pests[row][sex_col];
-  selfing_col  = (int) paras[26];
-  selfing      = pests[row][selfing_col];
-  
+
   mate_found = 0;
   switch(sex){
       case 0:
@@ -92,12 +89,11 @@ int mate_available(double **pests, double *paras, int row){
  * ========================================================================== */
 void count_offspring(double **pests, double *paras, int row){
   
-  int N, repr_param_col, offspring, repr_type_col, repr_type;
+  int repr_param_col, offspring, repr_type_col, repr_type;
   int mate_access, mate_access_col, off_col;
   int food_consumed_col, food_needed_col, repr_incr_col;
   double food_consumed, food_needed, repr_incr, repr_param;
   
[…]

### 2021-09-05 R/initialise.R
Introduce a minimum and a maximum age for metabolism to take effect.

@@ -43,7 +43,9 @@
 #'@param age_food_threshold Age at which food threshold is enacted
 #'@param age_pesticide_threshold Age at which pesticide threshold is enacted
 #'@param metabolism The amount of consumed food lost each time step
-#'@param baseline_metabolism A fixed baseline rate added to 'metabolism'
+#'@param baseline_metabolism A fixed baseline rate added to 'metabolism'+
+#'@param min_age_metabolism The minimum age affected by metabolism
+#'@param max_age_metabolism The maximum age affected by metabolism
 #'@return A two-dimensional array of individuals for simulation
 #'@examples
 #'gmt       <- matrix(data = 0, nrow = 2, ncol = 2);
@@ -84,7 +86,9 @@ initialise_inds <- function(mine_output,
                             age_food_threshold = NA,
                             age_pesticide_threshold = NA,
                             metabolism = 0,
-                            baseline_metabolism = 0){
+                            baseline_metabolism = 0,
+                            min_age_metabolism = 1,
+                            max_age_metabolism = 9){
   
   food      <- rep(x = 0, times = 10);
   pesticide <- rep(x = 0, times = 10);
@@ -199,6 +203,8 @@ initialise_inds <- function(mine_output,
   inds[, 84] <-  age_pesticide_threshold;
   inds[, 87] <-  metabolism;
   inds[, 88] <-  baseline_metabolism;
+  inds[, 89] <-  min_age_metabolism;
+  inds[, 90] <-  max_age_metabolism;
   
   return(inds);
 }

### 2021-09-05 R/sim_farming.R

@@ -83,6 +83,8 @@
 #'@param get_stats Get population level statistics in a CSV printout
 #'@param metabolism The rate at which food consumed is burned in a time step
 #'@param baseline_metabolism A fixed baseline rate added to 'metabolism'
+#'@param min_age_metabolism The minimum age affected by metabolism
+#'@param max_age_metabolism The maximum age affected by metabolism
 #'@return The output in the R console is a list with two elements; the first 
 #'element is a vector of parameter values used by the model, and the second 
 #'element is the landscape in the simulation. The most relevant output will be
@@ -168,7 +170,9 @@ run_farm_sim <- function(mine_output,
                          get_f_coef = FALSE,
                          get_stats = TRUE,
                          metabolism = 0,
-                         baseline_metabolism = 0){
+                         baseline_metabolism = 0,
+                         min_age_metabolism  = 1,
+                         max_age_metabolism  = 9){
   
     land <- make_landscape(rows = ydim, cols = xdim, depth = 21, farms = farms);
   
@@ -295,7 +299,9 @@ run_farm_sim <- function(mine_output,
                             age_food_threshold       = age_food_threshold,
                             age_pesticide_threshold  = age_pesticide_threshold,
                             metabolism               = metab_rate,
-                            baseline_metabolism      = baseline_metabolism);
+                            baseline_metabolism      = baseline_metabolism,
+                            min_age_metabolism       = min_age_metabolism,
+                            max_age_metabolism       = max_age_metabolism);
     
     sim_results <- sim_crops(pests                    = pest, 
                              land                     = land,
@@ -512,8 +518,8 @@ sim_crops <- function(pests,
               85.0,   # 85) pests column for lamba adjustment
               86.0,   # 86) pests column for metabolic rate (food lost)
               87.0,   # 87) pests column for baseline metabolic rate (fixed)
-              88.0,   # 88)
-              89.0,   # 89)
+              88.0,   # 88) pests column for minimum age of metabolism effects
+              89.0,   # 89) pests column for maximum age of metabolism effects
               90.0,   # 90)
               91.0,   # 91)
               92.0,   # 92)

### 2021-09-05 src/age.c

@@ -7,26 +7,35 @@
  * ========================================================================== */
 void age_pests(double **pests, double *paras){
   
-  int ind, N, age_col, meta_col, cons_col, base_m_col;
-  double meta_rate;
+  int ind, N, age_col, meta_col, cons_col, base_m_col, min_age_col, max_age_col;
+  int age, min_age, max_age;
+  double meta_rate, base_metabolism, metabolism;
   
-  age_col    = (int) paras[3];
-  N          = (int) paras[101];
-  meta_col   = (int) paras[86];
-  base_m_col = (int) paras[87];
-  cons_col   = (int) paras[14];
+  age_col     = (int) paras[3];
+  N           = (int) paras[101];
+  meta_col    = (int) paras[86];
+  base_m_col  = (int) paras[87];
+  cons_col    = (int) paras[14];
+  min_age_col = (int) paras[88];
+  max_age_col = (int) paras[89];
 
   for(ind = 0; ind < N; ind++){
-    meta_rate = pests[ind][base_m_col] + pests[ind][meta_col];
-    if(meta_rate < 0){
-      meta_rate = 0.0;
-    }
-    if(pests[ind][age_col] > 0){
-        pests[ind][cons_col] -= meta_rate;
-        if(pests[ind][cons_col] < 0){
-            pests[ind][cons_col] = 0;
-        }
-    }
-    pests[ind][age_col]++;
+      age     = pests[ind][age_col];
+      min_age = pests[ind][min_age_col];
+      max_age = pests[ind][max_age_col];
+    
+      if(age >= min_age && age <= max_age){
+          base_metabolism = pests[ind][base_m_col];
+          metabolism      = pests[ind][meta_col];
+          meta_rate       = base_metabolism + metabolism;
+          if(meta_rate < 0){
+              meta_rate = 0.0;
+          }
+          pests[ind][cons_col] -= meta_rate;
+          if(pests[ind][cons_col] < 0){
+              pests[ind][cons_col] = 0;
+          }
+      }
+      pests[ind][age_col]++;
   }
 }

### 2021-09-05 R/initialise.R
Finish the first vignette, and boost the version number

@@ -39,6 +39,7 @@
 #'@param movement_bouts Number of bouts of movement per time step
 #'@param selfing If sexual reproduction, is selfing allowed? (TRUE/FALSE)
 #'@param feed_while_moving Do individuals feed after each movement bout?
+#'@param pesticide_while_moving Individuals consume pesticide after move bout?
 #'@param mortality_type Type of mortality (currently only one option)
 #'@param age_food_threshold Age at which food threshold is enacted
 #'@param age_pesticide_threshold Age at which pesticide threshold is enacted
@@ -82,6 +83,7 @@ initialise_inds <- function(mine_output,
                             movement_bouts = 1,
                             selfing = TRUE,
                             feed_while_moving = FALSE,
+                            pesticide_while_moving = FALSE,
                             mortality_type = 0,
                             age_food_threshold = NA,
                             age_pesticide_threshold = NA,
@@ -197,6 +199,7 @@ initialise_inds <- function(mine_output,
   inds[, 56] <-  pesticide[[9]];
   inds[, 57] <-  pesticide[[10]];
   inds[, 58] <-  feed_while_moving; # Do not eat on a bout
+  inds[, 79] <-  pesticide_while_moving;
   inds[, 80] <-  mortality_type;
   inds[, 81] <-  max_age;
   inds[, 83] <-  age_food_threshold;

### 2021-09-05 R/sim_farming.R

@@ -48,6 +48,7 @@
 #'@param movement_bouts Number of bouts of movement per time step
 #'@param selfing If sexual reproduction, is selfing allowed?
 #'@param feed_while_moving Do individuals feed after each movement bout?
+#'@param pesticide_while_moving Individuals consume pesticide after move bout?
 #'@param mortality_type Type of mortality (currently only one option)
 #'@param age_food_threshold Age at which food threshold is enacted
 #'@param age_pesticide_threshold Age at which pesticide threshold is enacted
@@ -136,6 +137,7 @@ run_farm_sim <- function(mine_output,
                          movement_bouts = 1,
                          selfing = TRUE,
                          feed_while_moving = FALSE,
+                         pesticide_while_moving = FALSE,
                          mortality_type = 0,
                          age_food_threshold = 0,
                          age_pesticide_threshold = 0,
@@ -295,6 +297,7 @@ run_farm_sim <- function(mine_output,
                             movement_bouts           = move_bout,
                             selfing                  = selfing,
                             feed_while_moving        = feed_while_moving,
+                            pesticide_while_moving   = pesticide_while_moving,
                             mortality_type           = mortality_type,
                             age_food_threshold       = age_food_threshold,
                             age_pesticide_threshold  = age_pesticide_threshold,

### 2021-09-06 notebook/scratch.R
Some material for the website and documentation

@@ -1,3 +1,4 @@
+# Switch to 'resevol'.
 ################################################################################
 # Example with two traits
 ################################################################################

### 2021-09-09 R/landscape.R
Change the name from 'helicoverpa' to 'resevol' wherever it is relevant

@@ -22,7 +22,7 @@
 #'@return the_land A cols by rows landscape with randomly distributed cell types
 #'@examples
 #'land <- make_landscape(rows = 10, cols = 10, depth = 2, farms = 4)
-#'@useDynLib helicoverpa
+#'@useDynLib resevol
 #'@importFrom stats rnorm rpois runif
 #'@export
 make_landscape <- function(rows, cols, depth = 21, farms = 4,  public_land = 0, 

### 2021-09-09 R/mine_gmatrix.R

@@ -41,7 +41,7 @@
 #'diag(gmt) <- 1;
 #'mg        <- mine_gmatrix(gmatrix = gmt, loci = 6, layers = 3, indivs = 100, 
 #'                          npsize = 100, max_gen = 8, prnt_out = FALSE);
-#'@useDynLib helicoverpa
+#'@useDynLib resevol
 #'@importFrom stats rnorm rpois runif rbinom
 #'@importFrom utils read.csv write.csv
 #'@export

### 2021-09-09 R/sim_farming.R

@@ -107,7 +107,7 @@
 #'                          repro = "asexual", time_steps = 10, 
 #'                          print_inds = FALSE, print_gens = FALSE,
 #'                          print_last = FALSE, get_stats = FALSE);
-#'@useDynLib helicoverpa
+#'@useDynLib resevol
 #'@importFrom stats rnorm rpois runif
 #'@export
 run_farm_sim <- function(mine_output,

### 2021-09-09 src/resevol_init.c

@@ -19,7 +19,7 @@ static const R_CallMethodDef CallEntries[] = {
     {NULL, NULL, 0}
 };
 
-void R_init_helicoverpa(DllInfo *dll)
+void R_init_resevol(DllInfo *dll)
 {
     R_registerRoutines(dll, NULL, CallEntries, NULL, NULL);
     R_useDynamicSymbols(dll, FALSE);

### 2021-09-09 tests/testthat.R

@@ -1,4 +1,4 @@
 library("testthat");
-library("helicoverpa");
+library("resevol");
 
-test_check("helicoverpa");
+test_check("resevol");

### 2021-09-09 tests/testthat/test-initialise.R

@@ -1,5 +1,5 @@
 library("testthat");
-library("helicoverpa");
+library("resevol");
 context("initialise tests");
 
 test_that("Correctly initialises different individual types", {

### 2021-09-09 tests/testthat/test-landscape.R

@@ -1,5 +1,5 @@
 library("testthat");
-library("helicoverpa");
+library("resevol");
 context("landscape tests");
 
 test_that("Landscape dimensions are correct", {

### 2021-09-09 tests/testthat/test-mine_gmatrix.R

@@ -1,5 +1,5 @@
 library("testthat");
-library("helicoverpa");
+library("resevol");
 context("mine_gmatrix tests");
 
 test_that("Correct number of elements in the mine_gmatrix output", {

### 2021-09-09 tests/testthat/test-sim_farming.R

@@ -1,5 +1,5 @@
 library("testthat");
-library("helicoverpa");
+library("resevol");
 context("Simulation run tests");
 
 test_that("Simulations are run and output is produced", {

### 2021-09-09 notebook/helicoverpa_notes.Rmd
More naming is switched over from helicoverpa to resevosevol

@@
 ---
-title: "The helicoverpa package"
+title: "The resevol package"
 author: "Brad Duthie and Rose McKeon"

### 2021-09-09 notebook/index.Rmd
More naming is switched over from helicoverpa to resevosevol

--- 2021-09-09 notebook/helicoverpa_notes.Rmd
@@
 
-> **This notebook tracks progress on the development of [helicoverpa](https://bradduthie.github.io/helicoverpa/) R package, a BBSRC [FAPESP Newton Funded](http://www.fapesp.br/en/9552) Project to model how the spatial scale of heterogeneity in fungal isolate application and crop plant cultivation affects variation in the selection landscape for biopesticide resistance.**
+> **This notebook tracks progress on the development of [resevol](https://bradduthie.github.io/resevol/) R package, a BBSRC [FAPESP Newton Funded](http://www.fapesp.br/en/9552) Project to model how the spatial scale of heterogeneity in fungal isolate application and crop plant cultivation affects variation in the selection landscape for biopesticide resistance.**
 

### 2021-09-11 docs/notebook/images/2021_AUG_19/sim.R
Add the notebook the the docs folder

@@ -1,0 +1,124 @@
+library("helicoverpa");
+set.seed(Sys.time());
+mg  <- readRDS("mg.rds");
+sim <- run_farm_sim(mine_output              = mg,
+                    N                        = 1000,
+                    neutral_loci             = 1000,
+                    xdim                     = 100,
+                    ydim                     = 100,
+                    repro                    = "sexual",
+                    max_age                  = 6,
+                    selfing                  = TRUE,
+                    food_consume             = c("T1", "T2"),
+                    pesticide_consume        = c("T3", "T4"),
+                    food_needed_surv         = 1,
+                    food_needed_repr         = 1,
+                    reproduction_type        = "food_based",
+                    pesticide_tolerated_surv = 0,
+                    pesticide_rotation_type  = 1,
+                    crop_rotation_type       = 1,
+                    min_age_reproduce        = 4,
+                    max_age_feed             = 3,
+                    lambda_value             = 1.5,
+                    farms                    = 24,
+                    time_steps               = 480,
+                    mutation_pr              = 0.001,
+                    crossover_pr             = 0.1,
+                    net_mu_layers            = 0,
+                    crop_rotation_time       = 12,
+                    pesticide_rotation_time  = 12,
+                    crop_per_cell            = 8,
+                    pesticide_per_cell       = 0.4,
+                    crop_number              = 2,
+                    pesticide_number         = 2,
+                    print_inds               = FALSE, # Careful with this one
+                    K_on_birth               = 1000000,
+                    min_age_move             = 4,
+                    age_food_threshold       = 2,
+                    min_age_feed             = 1,
+                    pesticide_start          = 100,
+                    print_last               = TRUE,
[…]

### 2021-09-18 R/initialise.R
Remove the dontrun wrap at the suggestion of CRAN

@@ -49,13 +49,11 @@
 #'@param max_age_metabolism The maximum age affected by metabolism
 #'@return A two-dimensional array of individuals for simulation
 #'@examples
-#'\dontrun{
 #'gmt       <- matrix(data = 0, nrow = 2, ncol = 2);
 #'diag(gmt) <- 1;
 #'mg        <- mine_gmatrix(gmatrix = gmt, loci = 4, layers = 2, indivs = 100, 
 #'                          npsize = 100, max_gen = 4, prnt_out = FALSE);
 #'inds      <- initialise_inds(mine_output = mg, N = 40, repro = "asexual");
-#'}
 #'@importFrom stats rnorm
 #'@export
 initialise_inds <- function(mine_output, 

### 2021-09-18 R/landscape.R

@@ -21,9 +21,7 @@
 #'@param farm_var Does the land distribution vary among farms (>=0, <1)
 #'@return the_land A cols by rows landscape with randomly distributed cell types
 #'@examples
-#'\dontrun{
 #'land <- make_landscape(rows = 10, cols = 10, depth = 2, farms = 4)
-#'}
 #'@useDynLib resevol
 #'@importFrom stats rnorm rpois runif
 #'@export

### 2021-09-18 R/mine_gmatrix.R

@@ -40,7 +40,7 @@
 #'gmt       <- matrix(data = 0, nrow = 3, ncol = 3);
 #'diag(gmt) <- 1;
 #'mg        <- mine_gmatrix(gmatrix = gmt, loci = 6, layers = 3, indivs = 100, 
-#'                          npsize = 100, max_gen = 8, prnt_out = FALSE);
+#'                          npsize = 100, max_gen = 2, prnt_out = FALSE);
 #'@useDynLib resevol
 #'@importFrom stats rnorm rpois runif rbinom
 #'@importFrom utils read.csv write.csv

### 2021-09-18 R/sim_farming.R

@@ -102,9 +102,9 @@
 #'gmt       <- matrix(data = 0, nrow = 2, ncol = 2);
 #'diag(gmt) <- 1;
 #'mg        <- mine_gmatrix(gmatrix = gmt, loci = 4, layers = 2, indivs = 100, 
-#'                          npsize = 100, max_gen = 4, prnt_out = FALSE);
+#'                          npsize = 100, max_gen = 2, prnt_out = FALSE);
 #'sim       <- run_farm_sim(mine_output = mg, N = 100, xdim = 40, ydim = 40, 
-#'                          repro = "asexual", time_steps = 10, 
+#'                          repro = "asexual", time_steps = 3, 
 #'                          print_inds = FALSE, print_gens = FALSE,
 #'                          print_last = FALSE, get_stats = FALSE);
 #'@useDynLib resevol

### 2021-09-18 src/parents.c
Fix a minor error in the code that was found by WinBuilder

@@ -363,6 +363,7 @@ void mutation_haploid(double **offspring, double *paras, int offspr){
             }
         }
     }
+    network_len++;
     
 }
 

### 2021-09-18 tests/testthat/test-initialise.R
Skip the tests on CRAN, as they're not really necessary

@@ -3,6 +3,7 @@ library("resevol");
 context("initialise tests");
 
 test_that("Correctly initialises different individual types", {
+    skip_on_cran();
     set.seed(Sys.time());
     gmt       <- matrix(data = 0, nrow = 2, ncol = 2);
     diag(gmt) <- 1;

### 2021-09-18 tests/testthat/test-landscape.R

@@ -3,17 +3,20 @@ library("resevol");
 context("landscape tests");
 
 test_that("Landscape dimensions are correct", {
+    skip_on_cran();
     land <- make_landscape(rows = 12, cols = 12, farms = 12);
     expect_equal(dim(land), c(12, 12, 21));
 })
 
 test_that("Number of farms is correct", {
+    skip_on_cran();
     land     <- make_landscape(rows = 12, cols = 12, farms = 12);
     farms    <- length(unique(as.vector(land[,,1])));
     expect_equal(farms, 12);
 })
 
 test_that("Landscape farm allocation is equal", {
+    skip_on_cran();
     land     <- make_landscape(rows = 12, cols = 12, farms = 12);
     farms    <- length(unique(as.vector(land[,,1])));
     farmland <- table(as.vector(land[,,1]));

### 2021-09-18 tests/testthat/test-mine_gmatrix.R

@@ -3,6 +3,7 @@ library("resevol");
 context("mine_gmatrix tests");
 
 test_that("Correct number of elements in the mine_gmatrix output", {
+    skip_on_cran();
     set.seed(Sys.time());
     gmt       <- matrix(data = 0, nrow = 2, ncol = 2);
     diag(gmt) <- 1;
@@ -12,6 +13,7 @@ test_that("Correct number of elements in the mine_gmatrix output", {
 })
 
 test_that("Correct element sizes or dimensions in the mine_gmatrix output", {
+    skip_on_cran();
     set.seed(Sys.time());
     gmt       <- matrix(data = 0, nrow = 2, ncol = 2);
     diag(gmt) <- 1;

### 2021-09-18 tests/testthat/test-sim_farming.R

@@ -3,6 +3,7 @@ library("resevol");
 context("Simulation run tests");
 
 test_that("Simulations are run and output is produced", {
+    skip_on_cran();
     set.seed(Sys.time());
     gmt       <- matrix(data = 0, nrow = 2, ncol = 2);
     diag(gmt) <- 1;

### 2022-01-06 notebook/helicoverpa_notes.Rmd
Add the updates for v0.2.0.9 to the NEWS.md and notebook

@@
 
+ - **2022**
+
+   - **JAN** [06](#u-06-JAN-2022)
+
  - **2021**
@@
 ================================================================================
+
+> <a name="u-06-JAN-2022">Update: 06 JAN 2022</a>
+
+There had for some time been an annoying issue of [resevol CRAN checks](https://cran.r-project.org/web/checks/check_results_resevol.html) having ERRORs on Windows systems with ix86 architecture.
+As it turns out, this was caused by some errors in specifying the memory for arrays, which was fixed in [this commit](https://github.com/bradduthie/resevol/commit/b12918045402fa0a4443e4f7620d7d9d481cf3aa).
+Just to make sure that everything is clear and correct, I have made the cast much clearer in [a new commit](https://github.com/bradduthie/resevol/commit/223fda3cbadb79cd712b587e7153a5a7d27b71c0).
+Tests using the [rhub package](https://r-hub.github.io/rhub/) and on [Winbuilder](https://win-builder.r-project.org/) confirm that everything works as intended.
+Valgrind found no memory leaks; everything appears fine, so I have updated the version to v0.2.0.9 and submitted the fix to CRAN.
 

### 2022-06-13 notebook/ms/ms.Rmd
Initialise the manuscript file

---
title: 'resevol: an R package for spatially explicit models of pesticide resistance given evolving pest genomes'
author: |
        | A. Bradley Duthie&#xb9;&#xb2;, et al.
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2]  alexander.duthie@stir.ac.uk,
  Biological and Environmental Sciences 3A149 University of Stirling Stirling, FK9
  4LA, UK'
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
  html_document: default
  word_document: default
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[R]{RESEVOL, AN R PACKAGE}
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: resevol.bib
biblio-style: apalike
---

******************************

<!---

1. Software: new packages or significant updates in functionality - significant is defined as “novel capabilities enabling new applications”, as opposed to refinements on existing functionalities.

2. At a minimum, everything the reader will need to install and run the software; this includes a URL to see the source in an anonymous way (no login required), a specification of the language(s) used and their version(s), and dependencies on binaries that are not bundled with the software itself nor handled by the language package manager.

3.  Although we do not wish to be prescriptive about the structure, we recommend that the paper includes the application description along with one or several worked examples. Please note that the typical introduction, methods, results, discussion structure isn’t required, and often is not appropriate, for Applications papers.

4. Style guide for R: http://adv-r.had.co.nz/Style.html

--->

Abstract
===================================================

1. Pesticide resistance is a serious problem.
2. We introduce the resevol R package, which runs individual-based models of pests with quantitative and covarying traits and three mating systems.
3. Simulations are on a landscape where crops and pesticides are changed.
4. We give an example of simulating pests with covarying traits
5. The resevol R package is open source under GNU Public License; source code and documents are freely available on GitHub.

**Key words:** *pest management*, *food security*, *ecological modelling*

Introduction
===================================================

- Pest resistance to pesticide is a serious wicked problem that affects food security.
- A quantitative genetic approach to pesticide resistance to maintaining resistance.
- The resevol R package addresses two problems: genetic architecture and landscape selection
- The R package can facilitate general questions and targetted ones

Covarying pest quantitative traits
===================================================

- The issue is that we want mechanistic, not phenomenological, trait covariances
- Solution is to use a genetic algorithm to find genome structures [@Hamblin2013]
- Genomes include values that then define a network from alleles to traits
- Example using the `mine_gmatrix` function, and its various arguments

Simulating landscape-level pesticide resistance
===================================================

- We want to model evolution of pesticide resistance across heterogenous landscape
- Track entire genomes, traits, and yield over time and across space
- Use covarying traits from `mine_gmatrix` to make evolving pests
- Options for 10 crops and 10 pesticides simultaneously, and rotation
- Function `run_farm_sim` runs simulations

Example of individual-based simulations
===================================================

- Example of evolving species with different crop and pesticide appications

Conclusions
===================================================

Availability
===================================================

The resevol R package can be downloaded from CRAN (https://cran.r-project.org/package=resevol) or GitHub (https://bradduthie.github.io/resevol/). The package is open source under GNU Public License.

Conclusions
===================================================

Acknowledgements
===================================================

This software was developed as part of the project for Enhancing Diversity to Overcome Resistance Evolution (ENDORSE). The ENDORSE project is a joint Newton funded international partnership between the Biotechnology and Biological Sciences Research Council (BBSRC) in the UK and the S&#xE3;o Paulo Research Foundation (FAPESP) in Brazil under BBSRC award reference BB/S018956/1 and FAPESP award reference 2018/21089-3. ENDORSE is a partnership among Universidade Estadual Paulista (UNESP), the University of Stirling (UoS), and the Centre for Agricultural and Biosciences International (CABI).

References
===================================================

### 2022-06-15 notebook/index.Rmd
I believe that I have now implemented the crop change using matrices, as introduced in Issue #43 -- the pesticide needs to be changed now too, and it absolutely must be tested @bradduthie -- make sure that the landscape values are accurate during the rotation

@@
 
+   - **JUN** [13](#u-13-JUN-2022)
    - **JAN** [06](#u-06-JAN-2022)
@@
 ================================================================================
+
+> <a name="u-13-JUN-2022">Update: 13 JUN 2022</a>
+
+As I prepare the pre-print and software note for the resevol package, I am thinking that might be worth doing just a bit more coding so that the publication can include four more features:
+
+1.
+**More explicit crop and pesticide rotation**, as mentioned in [Issue #43](https://github.com/bradduthie/resevol/issues/43).
+I think that this would be fairly easy to implement and worth the time.
+Currently, the `run_farm_sim` argument `crop_rotation_type` accepts either 1 or a 2.
+A 1 results in no rotation of crops on the landscape, and a 2 results in a random transition between types.
+I think that a custom transition matrix should be possible to insert, whereby each element should specify the probability of transitioning from crop *i* (row) to crop *j* (column).
+In the simplest case (e.g., could be `crop_rotation_type = 3`), this could just be a binary matrix that forms a cyclical graph.
+The current options 1 and 2 could even be subsumed in the transition matrix.
+2.
+**Allow custom landscapes.** I think that this should be possible by allowing users to put their own matrix in a function, with numbers on the matrix representing different farms (integers greater than zeros) or public land (zeros).
+A new `custom_landscape` function could check this matrix, then use it as the top layer of an array.
+3.
+**Custom plotting functions.** This one is tricky given the flexibility of simulations and therefore the complexity of the output, but it might be worth including at least some basic plotting functions, or summary functions.
+4.
+**Simulated annealing algorithm** for the `mine_gmatrix` option.
+This could be fairly easy to implement, and for some cases might actually find genome solutions faster than the evolutionary algorithm.
+
+My hope is that these different tasks could be done fairly quickly, and that they will make writing up the paper go more smoothly.
 
@@
 ```{r, echo = FALSE}
-library(helicoverpa);
+library(resevol);
 gmt       <- matrix(data = 0, nrow = 2, ncol = 2);

### 2022-06-15 notebook/ms/ms.Rmd
I believe that I have now implemented the crop change using matrices, as introduced in Issue #43 -- the pesticide needs to be changed now too, and it absolutely must be tested @bradduthie -- make sure that the landscape values are accurate during the rotation

@@
 - Options for 10 crops and 10 pesticides simultaneously, and rotation
-- Function `run_farm_sim` runs simulations
+- Function `run_farm_sim` runs simulations. Note unusual timing structure
 
@@
 
-- Example of evolving species with different crop and pesticide appications
+- Example of evolving species with different crop and pesticide applications
 

### 2022-06-15 R/landscape.R

@@ -37,7 +37,7 @@ make_landscape <- function(rows, cols, depth = 21, farms = 4,  public_land = 0,
     }
     
     who_owns       <- land_ssa(rows, cols, farms, public_land, farm_var);
-    farmland       <- who_owns;
+    farmland       <- who_owns - 1;
     
     extr_layers    <- rep(x = 0, times = (depth-1) * rows * cols);
     alldata        <- c(farmland, extr_layers);
@@ -62,3 +62,137 @@ land_ssa <- function(dim_x, dim_y, farms, public_land, land_var){
 run_landscape_a <- function(LANDSCAPE_PARAMETERS){
     .Call("build_ownership", LANDSCAPE_PARAMETERS);
 }
+
+crop_transitions <- function(rotation_type = 1, crop_number){
+    crop_N      <- crop_number;
+    custom_land <- is.matrix(rotation_type);
+    if(custom_land == TRUE){
+        check_dims  <- dim(rotation_type);
+        if(check_dims[1] != check_dims[2]){
+            stop("ERROR: Transition matrices must be square.")
+        }
+        if(check_dims[1] != crop_N){
+            stop("ERROR: Transition matrix dimensions must equal crop numbers");
+        }
+        check_probs <- apply(X = rotation_type, MARGIN = 1, FUN = sum);
+        bad_probs   <- sum(check_probs != 1);
+        if(bad_probs > 0){
+            stop("ERROR: Row values of transition matrices must sum to 1.")
+        }
+        tmat <- rotation_type;
+    }else{
+        tmat <- NA;
+        if(rotation_type == 1){
+            tmat       <- matrix(data = 0, nrow = crop_N, ncol = crop_N);
+            diag(tmat) <- 1;
+        }
+        if(rotation_type == 2){
+            tmat <- matrix(data = 1/crop_N, nrow = crop_N, ncol = crop_N);
+        }
+        if(rotation_type == 3){
+            tmat <- matrix(data = 0, nrow = crop_N, ncol = crop_N);
+            for(i in 1:crop_N){
+                if(i < crop_N){
+                    tmat[i, i + 1] <- 1;
+                }else{
+                    tmat[i, 1]     <- 1;
+                }
+            }
+        }
+        if(is.na(tmat[1]) == TRUE){
+            stop("ERROR: Could not build the transition matrix.")
+        }
+    }
+    return(tmat);
+}
+
+pesticide_transitions <- function(rotation_type = 1, pesticide_number){
[…]

### 2022-06-15 R/sim_farming.R

@@ -60,8 +60,12 @@
 #'@param net_mu_layers Layers of the network allowed to mutate
 #'@param net_mu_dir  Layers mutate from loci to (1) or traits back (0)
 #'@param mutation_direction Is mutation directional (unlikely to need)
+#'@param crop_init Initial crop type for each farm (either "random" or a 
+#'   vector of the same length as farm number with integer crop types)
 #'@param crop_rotation_type None (1) or random (2) rotation of crop type
 #'@param crop_rotation_time How frequently are the crops rotated?
+#'@param pesticide_init Initial crop type for each farm (either "random" or a 
+#'   vector of the same length as farm number with integer pesticide types)
 #'@param pesticide_rotation_type None (1) or random (2) rotation of pesticide
 #'@param pesticide_rotation_time How frequently are the pesticides rotated?
 #'@param crop_per_cell How much crop is put on a single cell?
@@ -149,8 +153,10 @@ run_farm_sim <- function(mine_output,
                          net_mu_layers = 0,
                          net_mu_dir = 0, 
                          mutation_direction = 0,
+                         crop_init = "random",
                          crop_rotation_type = 2,
                          crop_rotation_time = 1,
+                         pesticide_init = "random",
                          pesticide_rotation_type = 2,
                          pesticide_rotation_time = 1,
                          crop_per_cell = 1,
@@ -315,8 +321,10 @@ run_farm_sim <- function(mine_output,
                              net_mu_layers            = net_mu_layers, 
                              net_mu_dir               = net_mu_dir,
                              mutation_direction       = mutation_direction,
+                             crop_init                = crop_init,
                              crop_rotation_type       = crop_rotation_type,
                              crop_rotation_time       = crop_rotation_time,
+                             pesticide_init           = pesticide_init,
                              pesticide_rotation_type  = pesticide_rotation_type,
                              pesticide_rotation_time  = pesticide_rotation_time,
                              crop_per_cell            = crop_per_cell,
@@ -347,7 +355,8 @@ run_farm_sim <- function(mine_output,
                              immigration_rate         = immigration_rate,
                              get_f_coef               = get_f_coef,
                              get_stats                = get_stats,
-                             metabolism               = metabolism);
[…]

### 2022-06-15 src/land_change.c

@@ -171,6 +171,7 @@ double get_crop_val(double *paras){
  *     owner_count: Vector of the number
  *     max_own:     The highest ID of a land owner
  * ========================================================================== */
+/* SLATED FOR DELETION
 void no_crop_rot(double ***land, double *paras, int *owner_choice, int max_own){
     
     int i, j, xdim, ydim, crop, owner, own_layer, choice, layer;
@@ -200,6 +201,7 @@ void no_crop_rot(double ***land, double *paras, int *owner_choice, int max_own){
         }
     }
 }
+ */
 
 /* =============================================================================
  * Each owner randomly selects a crop to put down
@@ -208,6 +210,7 @@ void no_crop_rot(double ***land, double *paras, int *owner_choice, int max_own){
  *     owner_count: Vector of the number
  *     max_own:     The highest ID of a land owner
  * ========================================================================== */
+/* SLATED FOR DELETION
 void rand_crop(double ***land, double *paras, int *owner_choice, int max_own){
   
     int i, j, xdim, ydim, owner, own_layer, choice, layer;
@@ -232,7 +235,7 @@ void rand_crop(double ***land, double *paras, int *owner_choice, int max_own){
         }
     }
 }
-
+*/
 
 /* =============================================================================
  * Changes the crop in a specific way
@@ -240,6 +243,7 @@ void rand_crop(double ***land, double *paras, int *owner_choice, int max_own){
  *     paras:   The paras vector that holds global information
  *     max_own: The maximum ID of an owner
  * ========================================================================== */
+/* SLATED FOR DELETION
 void change_crop(double ***land, double *paras, int max_own){
   
   int i;
@@ -253,12 +257,12 @@ void change_crop(double ***land, double *paras, int max_own){
   }
   
   switch(crop_rotate_type){
-      case 0: /* Nothing happens */
+      case 0:
           break;
-      case 1: /* Heterogeneous pattern but no rotation */
+      case 1:
           no_crop_rot(land, paras, owner_choice, max_own);
           break;
-      case 2: /* Random selection independent for each owner */
+      case 2:
           rand_crop(land, paras, owner_choice, max_own);
           break;
       default:
@@ -267,6 +271,7 @@ void change_crop(double ***land, double *paras, int max_own){
   
   free(owner_choice);
 }
+*/
 
 /* =============================================================================
  * Cleans the landscape of all crops
[…]

### 2022-06-15 src/land_change.h

@@ -23,4 +23,5 @@ void clean_crops(double ***land, double *paras);
 
 void clean_landscape(double ***land, double *paras);
 
-void land_change(double ***land, double *paras, int ts);
+void land_change(double ***land, double *paras, int ts, double **C_init,
+                 double **C_change);

### 2022-06-15 src/resevol_init.c

@@ -10,12 +10,12 @@
 /* .Call calls */
 extern SEXP build_ownership(SEXP);
 extern SEXP mine_gmatrix(SEXP, SEXP);
-extern SEXP sim_farming(SEXP, SEXP, SEXP);
+extern SEXP sim_farming(SEXP, SEXP, SEXP, SEXP, SEXP, SEXP, SEXP);
 
 static const R_CallMethodDef CallEntries[] = {
     {"build_ownership",         (DL_FUNC) &build_ownership,     1},
     {"mine_gmatrix",            (DL_FUNC) &mine_gmatrix,        2},
-    {"sim_farming",             (DL_FUNC) &sim_farming,         3},
+    {"sim_farming",             (DL_FUNC) &sim_farming,         7},
     {NULL, NULL, 0}
 };
 

### 2022-06-15 src/sim_farming.c

@@ -17,9 +17,14 @@
  *  Inputs include:
  *      IND:     Array with individual pest data
  *      LAND:    Three dimensional array of landscape
- *      PARAS:   Nothing yet, but will hold the paramters of interest
+ *      PARAS:   Nothing yet, but will hold the parameters of interest
+ *      CROT:    Crop rotation matrix
+ *      PROT:    Pesticide rotation matrix
+ *      CINIT:   Initial crop choice for each farmer
+ *      PINIT:   Initial pesticide choice for each farmer
  * ===========================================================================*/
-SEXP sim_farming(SEXP IND, SEXP LAND, SEXP PARAS){
+SEXP sim_farming(SEXP IND, SEXP LAND, SEXP PARAS, SEXP CROT, SEXP PROT,
+                 SEXP CINIT, SEXP PINIT){
  
     /* SOME STANDARD DECLARATIONS OF KEY VARIABLES AND POINTERS               */
     /* ====================================================================== */
@@ -27,6 +32,8 @@ SEXP sim_farming(SEXP IND, SEXP LAND, SEXP PARAS){
     int    row, col;
     int    xloc, yloc, zloc;
     int    land_z, land_y, land_x;
+    int    CROT_d1, CROT_d2, PROT_d1, PROT_d2;
+    int    CINIT_d1, CINIT_d2, PINIT_d1, PINIT_d2;
     int    vec_pos;
     int    time_steps;
     int    ind_number;
@@ -41,16 +48,28 @@ SEXP sim_farming(SEXP IND, SEXP LAND, SEXP PARAS){
     int    get_stats;        /* Should print a CSV with statistics */
     int    *dim_IND;         /* Dimensions of the individual array */
     int    *dim_LAND;        /* Dimensions of the landscape */
+    int    *dim_CROT;        /* Dimensions of the crop transition matrix */
+    int    *dim_PROT;        /* Dimensions of the pesticide transition matrix */
+    int    *dim_CINIT;       /* Dimensions of the crop initialisation matrix */
+    int    *dim_PINIT;       /* Dimensions of the pesticide init matrix */
   
     double *imm_sample;
     double *paras_ptr;
     double *IND_ptr;
     double *LAND_ptr;
+    double *CROT_ptr;
+    double *PROT_ptr;
+    double *CINIT_ptr;
+    double *PINIT_ptr;
[…]

### 2022-06-15 src/utilities.c

@@ -204,3 +204,26 @@ int max_in_col(double **array, int rows, int col){
     return max_val;
 }
 
+/* =============================================================================
+ * Sample an element from a probability vector
+ *     vec: The vector
+ *     len: The length of the vector
+ * ========================================================================== */
+int sample_pr_vector(double *vec, int len){
+    
+    int i;
+    double u, S;
+    
+    u = runif(0, 1);
+    S = 0.0;
+
+    for(i = 0; i < len; i++){
+        S += vec[i];
+        if(S >= u){
+            break;
+        }
+    }
+    
+    
+    return 0;
+}

### 2022-06-15 src/utilities.h

@@ -19,3 +19,5 @@ int get_rand_int(int from, int to);
 void swap_arrays(void **ARRAY_A, void **ARRAY_B);
 
 int max_in_col(double **array, int rows, int col);
+
+int sample_pr_vector(double *vec, int len);

### 2022-06-28 notebook/index.Rmd
Some comments on how the new crop and pesticide rotation options are implemented

@@
 ```{r, echo=FALSE}
-library(shiny)
-library(rhandsontable)
+library("shiny");
+library("rhandsontable");
+library("resevol");
 ```
@@
 
-   - **JUN** [13](#u-13-JUN-2022)
+   - **JUN** [13](#u-13-JUN-2022), [28](#u-28-JUN-2022)
    - **JAN** [06](#u-06-JAN-2022)
@@
 ================================================================================
+
+> <a name="u-28-JUN-2022">Update: 28 JUN 2022</a>
+
+**Custom crop and pesticide initialisation and rotation**
+
+I have now introduced custom crop and pesticide initialisation and rotation, as I set out to do in the [13 JUN 2022](#u-13-JUN-2022) notes below.
+The code appears to be stable, and I have introduced two new automated tests in the package to make sure that the feature remains stable.
+Below is an example of how to have a custom rotation regime for crops and pesticides; I will start with crops.
+First, I will use the `mg_n1` gmatrix that comes with the `resevol` package (I could have also made a dummy one; the details are not important here).
+
+```{r}
+data("gmatrices");
+```
+
+Now I will run a simulation.
+The `run_farm_sim` takes some simple options for `crop_init` and `crop_rotation_type`.
+This includes `crop_init = "random"`, which randomly assigns each farm to one of the potential `crops`.
+There are also three `crop_rotation_type` arguments, including `1` (no crop rotation), `2` (random crop choice whenever it is time to change), and `3` (cycling through crops by number; i.e., 1, 2, 3, 1, 2, 3).
+I might add more preset options like this later, but here is how to run a simulation with two crops, random crop initialisation on farms, and random rotation between crops.
+
+```{r}
+sim       <- run_farm_sim(mine_output = mg_n1, N = 100, xdim = 4, ydim = 4, 
+                          repro = "asexual", time_steps = 10, farms = 4,
+                          crop_number        = 2,
+                          crop_init          = "random",
+                          crop_rotation_type = 2,
+                          print_inds = FALSE, print_gens = FALSE,
+                          print_last = FALSE, get_stats = FALSE);
+```
+
+The new feature is that `crop_init` and `crop_rotation_type` can take custom arguments, so potentially any initialisation and cycling of crop choices is possible.
+With `crop_init`, we can specify which farm gets what crop at the start of the simulation using a fector of length `farms`.
+If, for example, we want farms 1 and 4 to have crop 1, and farms 2 and 3 to have crop 2, we could run the below.
+
+```{r}
+sim       <- run_farm_sim(mine_output = mg_n1, N = 100, xdim = 4, ydim = 4, 
+                          repro = "asexual", time_steps = 10, farms = 4,
+                          crop_number        = 2,
+                          crop_init          = c(1, 2, 2, 1),
+                          crop_rotation_type = 2,
+                          print_inds = FALSE, print_gens = FALSE,
+                          print_last = FALSE, get_stats = FALSE);
+```
+
+Note that the position of farms on the landscape is shown below.
+
+```{r, echo = FALSE}
+print(sim[[2]][,,1]);
+```
+
+Hence, this landscape would start out with farms numbered 1 and 4, diagonal from one another, applying crop 1, and farms numbered 2 and 3, also diagonal, applying crop 2.
+Now, suppose that we want users of crop 1 to switch to crop 2 with a probability of 0.75, and retain the use of crop 1 with a probability of 0.25.
+Likewise, suppose that we want users of crop 2 to switch to crop 1 with a probability of 0.75, and retain the use of crop 2 with a probability of 0.25.
+We can now specify this with a square matrix to input in place of the number for the argument `crop_rotation_type`.
+
+```{r}
+change_mat <- matrix(data = c(0.25, 0.75, 0.75, 0.25), nrow = 2, ncol = 2);
+```
+
+```{r, echo = FALSE}
+print(change_mat);
+```
+
+Note that matrix elements $M_{i,j}$ specify the probability of changing from crop $i$ to crop $j$, meaning that row elements must sum to 1.
+This can be binary too, as in the default `crop_rotation_type = 3`.
+
+```{r, echo = FALSE}
+change_mat <- matrix(data = c(0, 1, 1, 0), nrow = 2, ncol = 2);
+print(change_mat);
+```
+
+The matrix above cycles back and forth between crops 1 and 2.
+Overall, this should give a lot more flexibility in terms of specifying how crops cycle over time.
+We can input the `change_mat` directly as an argument in `run_farm_sim`.
+
+```{r}
+sim       <- run_farm_sim(mine_output = mg_n1, N = 100, xdim = 4, ydim = 4, 
+                          repro = "asexual", time_steps = 10, farms = 4,
+                          crop_number        = 2,
+                          crop_init          = c(1, 2, 2, 1),
+                          crop_rotation_type = change_mat,
+                          print_inds = FALSE, print_gens = FALSE,
+                          print_last = FALSE, get_stats = FALSE);
+```
+
+Nevertheless, the cycling is restricted to Markov chains, meaning that it cannot include cycles of crops such as 1, 1, 2, 1, 1, 2, 1, 1, etc., where the cycling repeats complex patterns.
+This is mainly due to the awkwardness of trying to allow for such cycles given simulations that can have high numbers of time steps (i.e., I do not think it is worth figuring out a system that allows for these kinds of cycles as an option in the code).
+
+Pesticide rotation options work exactly the same way as crop rotation options, so I will not spend to much time on them. Here is an example of running one.
+
+```{r}
+sim       <- run_farm_sim(mine_output = mg_n1, N = 100, xdim = 4, ydim = 4, 
+                          repro = "asexual", time_steps = 10, farms = 4,
+                          pesticide_number        = 2,
+                          pesticide_init          = c(1, 2, 2, 1),
+                          pesticide_rotation_type = change_mat,
+                          print_inds = FALSE, print_gens = FALSE,
+                          print_last = FALSE, get_stats = FALSE);
+```
+
+In conclusion, this new feature will allow for a very flexible range of different crop and pesticide cycling options.
 

### 2022-06-28 R/landscape.R
Add code to allow custom landscapes specified by the user

@@ -13,6 +13,7 @@
 #' make farm size as even as possible. Specifying public land is possible, and
 #' adds sections of land that are not farms, but this is not recommended.
 #'
+#'@param terrain a layer of terrain that can be added to specify farm position
 #'@param rows The dimension of the other side of the landscape (e.g., Longitude)
 #'@param cols The dimension of one side of the landscape (e.g., Latitude)
 #'@param depth The number of layers in the 3D landscape
@@ -25,9 +26,35 @@
 #'@useDynLib resevol
 #'@importFrom stats rnorm rpois runif
 #'@export
-make_landscape <- function(rows, cols, depth = 21, farms = 4,  public_land = 0, 
-                           farm_var = 0){
+make_landscape <- function(terrain = NA, rows = NA, cols = NA, depth = 21, 
+                           farms = 4, public_land = 0, farm_var = 0){
     
+    if(is.na(terrain)[1] == FALSE){
+        if(is.matrix(terrain) == FALSE){
+            stop("Custom terrain needs to be a matrix of integers");
+        }
+        min_terrain <- min(terrain);
+        max_terrain <- max(terrain);
+        if(min_terrain < 1){
+            stop("Terrain values cannot be less than 1")
+        }
+        min_to_max  <- min_terrain:max_terrain;
+        terrain_in  <- sum(terrain %in% min_to_max == FALSE);
+        if(terrain_in > 0){
+            stop("Values in the custom terrain need to be natural numbers");
+        }
+        terrain_out <- sum(min_to_max %in% terrain == FALSE);
+        if(terrain_out > 0){
+            stop("At least one farm number is left out of the terrain");
+        }
+        rows <- dim(terrain)[1];
+        cols <- dim(terrain)[2];
+    }else{
+        if(is.na(rows) == TRUE | is.na(cols) == TRUE){
+            stop("Must specify landscape rows and cols if no custom terrain");
+        }
+    }
+
     the_land  <- NULL;
     if(rows < 2){
         stop("Landscape dimensions must be 2 by 2 or greater");   
@@ -36,8 +63,12 @@ make_landscape <- function(rows, cols, depth = 21, farms = 4,  public_land = 0,
         stop("Landscape dimensions must be 2 by 2 or greater");   
     }
     
-    who_owns       <- land_ssa(rows, cols, farms, public_land, farm_var);
[…]

### 2022-06-28 tests/testthat/test-landscape.R

@@ -23,6 +23,16 @@ test_that("Landscape farm allocation is equal", {
     expect_equal(as.vector(farmland), rep(12, farms));
 })
 
+test_that("Custom landscape can be added", {
+    skip_on_cran();
+    custom_terrain       <- matrix(data = 0, nrow = 8, ncol = 8);
+    custom_terrain[,1:2] <- 1;
+    custom_terrain[,2:6] <- 2;
+    custom_terrain[,7:8] <- 3;
+    land <- make_landscape(terrain = custom_terrain);
+    expect_equal(dim(land), c(8, 8, 21));
+})
+
 test_that("Crops rotate correctly over time", {
     skip_on_cran();
     gmt       <- matrix(data = 0, nrow = 2, ncol = 2);

### 2022-06-28 notebook/index.Rmd
A bit more to fix things up with an example landscape, and avoid a crash caused by arguments that are not compatible

@@
 In conclusion, this new feature will allow for a very flexible range of different crop and pesticide cycling options.
+
+**Custom landscape option**
+
+I have also added an option to customise the farming landscape, also as I set out to do in the [13 JUN 2022](#u-13-JUN-2022) notes below.
+The customisation is on the top layer of the landscape that differentiates independent farms.
+What is required is a matrix that includes a sequence of natural numbers in all elements, so if there are 4 farms, then all matrix elements must be either 1, 2, 3, or 4.
+Beyond this requirement, there is no restriction on where different farms are placed; the do not even need to be contiguous on the landscape.
+Custom terrains can be inserted using the `terrain` argument in `run_farm_sim`.
+Below makes a custom terrain with three farms.
+
+```{r}
+custom_terrain       <- matrix(data = 0, nrow = 8, ncol = 8);
+custom_terrain[,1:2] <- 1;
+custom_terrain[,2:6] <- 2;
+custom_terrain[,7:8] <- 3;
+```
+
+```{r, echo = FALSE}
+printf(custom_terrain);
+```
+
+We can insert the terrain into the landscape with the `terrain` argument.
+
+```{r}
+sim       <- run_farm_sim(mine_output = mg_n1, N = 100, xdim = 4, ydim = 4, 
+                          repro = "asexual", time_steps = 10, farms = 4,
+                          terrain = custom_terrain,
+                          print_inds = FALSE, print_gens = FALSE,
+                          print_last = FALSE, get_stats = FALSE);
+```
+
+Note that a custom terrain will override the arguments `farms`, `xdim`, and `ydim`, so even though the above sets `farms = 4`, there are actually only three farms on the landscape.
+I am tempted to put up a warning for this, but I am actually not sure if it is needed; my hope is that anyone customising the landscape would recognise this from the documentation.
+
+Also note that these terrain values do not necessarily need to be farms.
+Through the use of a custom landscape and pesticide rotation option, these cells could represent something like diversionary feeding sites or even buildings or rivers.
+Consider, for example, the following matrix.
+
+```{r}
+complex_terrain              <- matrix(data = 0, nrow = 12, ncol = 12)
+complex_terrain[1:2, 1:3]    <- 1;
+complex_terrain[3:7, 1:2]    <- 1;
+complex_terrain[8:12, 1:2]   <- 2;
+complex_terrain[1:2, 6:9]    <- 3;
+complex_terrain[3:6, 7:9]    <- 3;
+complex_terrain[1:6, 10:12]  <- 4;
+complex_terrain[10:12, 7]    <- 5;
+complex_terrain[7:12, 8:12]  <- 5;
+complex_terrain[3:12, 3:4]   <- 6;
+complex_terrain[7:9, 5]      <- 6;
+complex_terrain[1:2, 4:5]    <- 7;
+complex_terrain[3:6, 5:6]    <- 7;
+complex_terrain[7:9, 6:7]    <- 7;
+complex_terrain[10:12, 5:6]  <- 7;
+```
+
+```{r, echo = FALSE}
+print(complex_terrain);
+```
+
+We can visualise the terrain a bit more clearly with an image below.
+
+```{r, echo = FALSE}
+image(t(complex_terrain), xaxt = "n", yaxt = "n")
+```
+
+We can imagine the dark maroon colour as being a river (7), and the red region on the immediate left of the river as being a nature reserve (6).
+To get the simulations to run treating it this way, we would only need to spray a crop and pesticide that has no effect on the pests on the river (effectively making it an empty area of landscape in terms of feeding or resistance), and a pesticide but not crop in the nature reserve.
+First we can initialise the crop and pesticide vectors.
+Assume that there are 2 legitimate crops and pesticides that the actual farms (1-5) will use.
+
+```{r}
+initial_crops      <- c(1, 2, 1, 2, 1, 3, 4);
+initial_pesticides <- c(1, 1, 2, 2, 1, 3, 3);
+```
+
+We can then clarify the matrix so that there is only a cycle between 1 and 2, but never 3 and 4 for either the pesticide or the crops. First I will do the crops.
+
+```{r}
+crop_cycle <- matrix(data = 0, nrow = 4, ncol = 4);
+crop_cycle[1, 2] <- 1;
+crop_cycle[2, 1] <- 1;
+crop_cycle[3, 3] <- 1;
+crop_cycle[4, 4] <- 1;
+```
+
+```{r, echo = FALSE}
+print(crop_cycle);
+```
+
+Now I will do the pesticide.
+
+```{r}
+pesticide_cycle <- matrix(data = 0, nrow = 3, ncol = 3);
+pesticide_cycle[1, 2] <- 1;
+pesticide_cycle[2, 1] <- 1;
+pesticide_cycle[3, 3] <- 1;
+```
+
+```{r, echo = FALSE}
+print(pesticide_cycle);
+```
+
+So farms (1-5) will swap between crops 1 and 2, while the reserve (6) will maintain crop 3 and the river (7) will be 'crop' 4.
+And farms will swap between pesticides 1 and 2, while the reserve and river will hold pesticide 3.
+Then, to make crop 4 and pesticide 3 not really exist, we just make it impossible for the pests to consume these.
+
+```{r}
+sim       <- run_farm_sim(mine_output = mg_n1, N = 100, 
+                          repro = "asexual", time_steps = 10,
+                          terrain = complex_terrain,
+                          print_inds = FALSE, print_gens = FALSE,
+                          print_last = FALSE, get_stats = FALSE);
+```
 

### 2022-06-28 R/sim_farming.R

@@ -90,6 +90,7 @@
 #'@param baseline_metabolism A fixed baseline rate added to 'metabolism'
 #'@param min_age_metabolism The minimum age affected by metabolism
 #'@param max_age_metabolism The maximum age affected by metabolism
+#'@param terrain Insert a custom terrain of different farms (matrix)
 #'@return The output in the R console is a list with two elements; the first 
 #'element is a vector of parameter values used by the model, and the second 
 #'element is the landscape in the simulation. The most relevant output will be
@@ -180,10 +181,18 @@ run_farm_sim <- function(mine_output,
                          metabolism = 0,
                          baseline_metabolism = 0,
                          min_age_metabolism  = 1,
-                         max_age_metabolism  = 9){
+                         max_age_metabolism  = 9,
+                         terrain             = NA){
   
-    land <- make_landscape(rows = ydim, cols = xdim, depth = 21, farms = farms);
-  
+    if(is.na(terrain)[1] == FALSE){
+        xdim  <- dim(terrain)[1];
+        ydim  <- dim(terrain)[2];
+        farms <- max(terrain);
+    }
+    
+    land <- make_landscape(terrain = terrain, rows = ydim, cols = xdim, 
+                           depth = 21, farms = farms);
+
     move_distance_T             <- check_is_trait(move_distance);
     food_needed_surv_T          <- check_is_trait(food_needed_surv);
     pesticide_tolerated_surv_T  <- check_is_trait(pesticide_tolerated_surv);

### 2022-06-28 notebook/index.Rmd
Finish up a notebook entry explaining the new custom om landscape feature

@@
 ```{r, echo = FALSE}
-printf(custom_terrain);
+print(custom_terrain);
 ```
@@
                           repro = "asexual", time_steps = 10,
+                          crop_number = 4, pesticide_number = 3,
                           terrain                 = complex_terrain,
+                          food_consume            = c(0.25, 0.25, 0.25, 0),
+                          pesticide_consume       = c(0.1, 0.1, 0),
+                          crop_init               = initial_crops,
+                          crop_rotation_type      = crop_cycle,
+                          pesticide_init          = initial_pesticides,
+                          pesticide_rotation_type = pesticide_cycle,
                           print_inds = FALSE, print_gens = FALSE,
@@
 ```
+
+The output shows that the pests leave the 'food' on the river alone (see layer 5), while the reserve (see layer 4) gets used very much.
+
+```{r, echo = FALSE}
+print(sim[[2]][,,1:5]);
+```
+
+Similarly, the pests will only be affected by pesticides 1 and 2, while 3 is effectively nothing.
+Note that this is of course a trick for modelling landscape areas without food or pesticide, but I do not see a reason to make this more explicit in the function (unless I receive requests due to confusion, but my hope is it is clear from these notes and documentation).
 

### 2022-07-16 src/mine_gmatrix.c
Minor change to the mine_gmatrix function

@@ -1022,7 +1022,7 @@ SEXP mine_gmatrix(SEXP PARAS, SEXP GMATRIX){
       
       /* Add print of highest fitness found */
       if(prnt_out > 0){
-          Rprintf("Gen: %d\t Stress: %f\t Highest: %f\n", 
+          Rprintf("Gen: %d\t Stress: %f\t Lowest: %f\n", 
                   gen, estress, high_fitness[0]);
       }
       gen++;

### 2022-07-16 notebook/ms/ms.Rmd
More writing, up to the landscape. Need to continue explaining pesticide and crop cycling, then pest initialisation and dynamics

@@
 output:
-  pdf_document:
+  bookdown::pdf_document2:
     citation_package: natbib
     fig_caption: yes
+    toc: false
   html_document: default
@@
 
-Abstract
+Abstract {-}
 ===================================================
@@
 
+```{r, echo = FALSE}
+library("resevol");
+```
+
+The resevol package uses individual-based modelling and a quantitative genetics approach to simulate the evolution of a pest population on a changing landscape. 
+A focal goal of the software is to model traits with a pre-specified, but potentially evolving, covariance structure. 
+To achieve this goal, each individual has a genome with $L$ loci that underlie a set of $T$ potentially evolving traits. 
+Pleiotropic loci can vary in their effects on the direction and magnitude of polygenic trait, causing population-wide trait covariance to arise mechanistically from the underlying genetic architecture of individuals.
+To achieve this, two separate steps are necessary.
+First, a genetic algorithm is used to find a network of internal nodes that map standard random normal loci values to covarying traits.
+Values used to map loci to traits are incorporated into individual genomes.
+Second, a population of asexual or sexual individuals is initialised and simulated on a spatially explicit landscape separated into distinct units (e.g., farms).
+Land units can apply one of up to 10 pesticides, and one of up to 10 crops; pesticides and crops rotate independently in a pre-specified way over time.
+The resevol package can thereby model complex and evolving traits in agricultural pests over realistic landscapes that undergo different pesticide use and crop regimes.
+
 Covarying pest quantitative traits
@@
 
-- The issue is that we want mechanistic, not phenomenological, trait covariances
-- Solution is to use a genetic algorithm to find genome structures [@Hamblin2013]
-- Genomes include values that then define a network from alleles to traits
-- Example using the `mine_gmatrix` function, and its various arguments
+The first step of simulation with the resevol package is building individual genomes. 
+This step is separate because building genomes is often time-consuming, and genomes that are built might need to be saved or inspected before actual simulation.
+High computation time is due to the mechanistic nature of how genomes and covarying traits are modelled.
+Instead of imposing a trait covariance structure directly, the objective is to use an evolutionary algorithm to find a network that maps standard random normal values (loci) to covarying values (traits).
+This is useful because it allows genomes to model potentially evolving physiological constraints, and trade-offs among traits, from the bottom up.
+Since multiple networks can potentially map loci to the same trait covariance structure, it is possible to replicate evolution with differently randomised genetic architectures.
+This approach to modelling individual genomes and traits thereby increases the complexity of questions that can be addressed for simulating evolution in agricultural pests.
+
+Conceptually, the relationship between individual genotypes and traits is defined by a network connecting loci to traits through a set of hidden internal nodes (Figure \@ref(fig:network)). 
+Values for loci are randomly drawn from a standard normal distribution $\mathcal{N}\left(0, 1\right)$. 
+Links between loci, internal nodes, and traits, can take any real value and are represented by black arrows in Figure \@ref(fig:network). 
+Traits can take any real value, and are calculated as the summed effects of all preceding nodes (i.e., the blue squares immediately to the left of the traits in Figure \@ref(fig:network)).
+Mathematically, loci are represented by row a vector of length $L$.
+Effects of loci on the first layer of internal nodes (black arrows emanating from loci in Figure \@ref(fig:network)) are represented by an $L \times T$ matrix, and transitions between internal nodes, and between the last set of internal nodes and the final traits, are represented by $T \times T$ matrices.
+For the example in Figure \@ref(fig:network),
+
+$$
+\begin{pmatrix}
+  0.1, & -0.1, & 0.2 \\
+\end{pmatrix} 
+\begin{pmatrix}
+  0.4, &  0.1 \\
+ -1.2, & -0.6 \\
+  0.2, &  2.2 \\
+\end{pmatrix}
+\begin{pmatrix}
+   1.3, & -0.5 \\
+  -0.1  &  0.0 \\
+\end{pmatrix}
+\begin{pmatrix}
+   2.2, &  0.9 \\
+   1.6  &  0.8 \\
+\end{pmatrix} = 
+\begin{pmatrix}
+  0.2998, & 0.1081 \\
+\end{pmatrix}.
+$$
+
+Values mapping loci to traits become part of an individual's genome, so the genome for the individual represented by Figure \@ref(fig:network) is stored in the model as below:
+
+```
+0.1, -0.1, -0.2, 0.4, 0.1, -1.2, -0.6, 0.2, 2.2, 1.3, -0.5, -0.1, 0.0, 2.2, 0.9, 1.6, 0.8
+```
+
+Individuals with different loci can therefore have different covarying traits that are constrained by the network structure encoded in each genome. 
+Individuals can be haploid (as in Figure \@ref(fig:network)) or diploid (in which case, allele values are summed at a locus).
+
+```{r network, echo = FALSE, fig.width = 6, fig.height = 4, fig.cap = "Example network mapping loci (green circles) to traits (red diamonds) through an intermediate set of hidden layers (blue squares) in the mine_gmatrix function.
+Individual genomes in the resevol R package consist of standard random normal values for at loci, real values for black arrows linking nodes, and real values of traits.
+Values shown for loci and arrows are an example for illustration."}
+par(mar = c(0.2, 0.2, 0, 0.2));
+plot(x = 0, y = 0, type = "n", xlim = c(0, 1000), ylim = c(0, 1000), 
+     xaxt = "n", yaxt = "n", bty = "n", xlab = "", ylab = "");
+mtext(side = 1, text = "Internal nodes", cex = 1.25, col = "blue", line = -3);
+mtext(side = 2, text = "Loci", cex = 1.25, col = "darkgreen", line = -2);
+mtext(side = 4, text = "Traits", cex = 1.25, col = "red", line = -2);
+Lx  <- rep(x = 100, times = 3);
+Ly  <- seq(from = 250, to = 750, length = 3);
+points(x = Lx, y = Ly, pch = 20, cex = 8, col = "darkgreen");
+H1x <- rep(x = 400, times = 2);
+H1y <- seq(from = 250, to = 750, length = 2);
+points(x = H1x, y = H1y, pch = 15, cex = 8, col = "blue");
+H2x <- rep(x = 600, times = 2);
+H2y <- seq(from = 250, to = 750, length = 2);
+points(x = H2x, y = H2y, pch = 15, cex = 8, col = "blue");
+Tx  <- rep(x = 900, times = 2);
+Ty  <- seq(from = 250, to = 750, length = 2);
+points(x = Tx, y = Ty, pch = 18, cex = 10, col = "red");
+sep_L_points <- c(-50, 0, 50);
+sep_N_points <- c(-30, 30);
+arrows(x0 = Lx + 38, x1 = H1x[1] - 58, y0 = Ly, y1 = H1y[1] + sep_L_points, 
+       length = 0.08, lwd = 3);
+arrows(x0 = H1x + 52, x1 = H2x[1] - 58, y0 = H1y, y1 = H2y[1] + sep_N_points, 
+       length = 0.08, lwd = 3);
+arrows(x0 = H2x + 52, x1 = Tx[1] - 58, y0 = H2y, y1 = Ty[1] + sep_N_points, 
+       length = 0.08, lwd = 3);
+arrows(x0 = Lx + 38, x1 = H1x[2] - 58, y0 = Ly, y1 = H1y[2] + sep_L_points, 
+       length = 0.08, lwd = 3);
+arrows(x0 = H1x + 52, x1 = H2x[2] - 58, y0 = H1y, y1 = H2y[2] + sep_N_points, 
+       length = 0.08, lwd = 3);
+arrows(x0 = H2x + 52, x1 = Tx[2] - 58, y0 = H2y, y1 = Ty[2] + sep_N_points, 
+       length = 0.08, lwd = 3);
+srnm <- c(0.2, -0.1, 0.1);
+text(x = Lx, y = Ly, labels = srnm, col = "white", cex = 0.9);
+text(x = 200, y = 790, labels = "0.4", srt = 10, cex = 0.8);
+text(x = 200, y = 660, labels = "-1.2", srt = -58, cex = 0.8);
+text(x = 270, y = 695, labels = "0.2", srt = 46, cex = 0.8);
+text(x = 180, y = 480, labels = "0.1", srt = -40, cex = 0.8);
+text(x = 160, y = 340, labels = "-0.6", srt = 54, cex = 0.8);
+text(x = 220, y = 255, labels = "2.2", srt = -12, cex = 0.8);
+text(x = 490, y = 790, labels = "1.3", srt = 10, cex = 0.8);
+text(x = 538, y = 410, labels = "-0.5", srt = -76, cex = 0.8);
+text(x = 460, y = 390, labels = "-0.1", srt = 72, cex = 0.8);
+text(x = 495, y = 260, labels = "0.0", srt = -12, cex = 0.8);
+text(x = 720, y = 790, labels = "2.2", srt = 8, cex = 0.8);
+text(x = 720, y = 630, labels = "0.9", srt = -58, cex = 0.8);
+text(x = 690, y = 400, labels = "1.6", srt = 54, cex = 0.8);
+text(x = 740, y = 260, labels = "0.8", srt = -8, cex = 0.8);
+Loci  <- matrix(data = c(0.1, -0.1, 0.2), nrow = 1);
+Hdn1  <- matrix(data = c(0.4, -1.2, 0.2, 0.1, -0.6, 2.2), nrow = 3, ncol = 2);
+Hdn2  <- matrix(data = c(1.3, -0.1, -0.5, 0.0), nrow = 2, ncol = 2);
+Hdn3  <- matrix(data = c(2.2, 1.6, 0.9, 0.8), nrow = 2, ncol = 2);
+Tvals <- Loci %*% Hdn1 %*% Hdn2 %*% Hdn3;
+text(x = Tx, y = Ty, labels = rev(Tvals), cex = 0.8);
+```
+
+An evolutionary algorithm is used to find appropriate values that produce covarying traits from loci (see Supporting Information for details). 
+Evolutionary algorithms are heuristic tools can simulate adaptive evolution to find solutions for a broad range of problems [@Hamblin2013; @Duthie2018]. 
+In the resevol package, the `mine_gmatrix` function runs an evolutionary algorithm, and requires the argument `gmatrix`, which specifies the desired trait covariance matrix.
+The function runs a genetic algorithm that initialises a population of `npsize` networks (Figure \@ref(fig:network)), and this population evolves until some maximum generation number (`max_gen`) or minimum expected network stress (`term_cri`) is met.
+In a single generation of the algorithm, network values crossover and mutate.
+Next, trait covariances produced for each network are estimated by initalising `indivs` individuals with loci sampled from a standard normal distribution.
+Network stress is calculated as the logged mean squared deviation between estimated covariances and those in `gmatrix`. 
+Tournament selection [@Hamblin2013] is then used to determine the networks for the next generation of the algorithm.
+Networks with the lowest estimated stress have the highest fitness, so these networks are disproportionately represented in the next generation.
+Throughout the evolutionary algorithm, the lowest stress network is saved and output upon network termination.
+The robustness of this network's stress to sets of individuals with different loci values can be tested using the `stress_test` function.
+
+An example run of `mine_gmatrix` with the same number of loci and internal nodes is shown below with traits that do not covary:
+
+```{r, eval = FALSE}
+trait_covs  <- matrix(data = c(1, -0.4, -0.4, 1), nrow = 2, ncol = 2);
+new_network <- mine_gmatrix(loci = 3, layers = 2, gmatrix = trait_covs, 
+                            max_gen = 1000, term_cri = -6.0, prnt_out = FALSE);
+```
+
+```{r, echo = FALSE}
+trait_covs  <- matrix(data = c(1, -0.4, -0.4, 1), nrow = 2, ncol = 2);
+load("new_network.rda");
+```
+
+The code above found a genome that produced the following expected trait covariances:
+
+```{r, echo = FALSE}
+print(new_network[[6]]);
+```
+
+The mean deviation between elements of the above matrix and the identity matrix provided by `trait_covs` is `r mean((new_network[[6]] - new_network[[2]])^2)`. 
+Lower values of `term_cri` and higher values of `max_gen` will result in a lower stress, but this will require more computation time, especially if the number of traits is high.
+Similarly, higher values of `indivs` will result in more accurate estimations of true stress, but this also requires more computation time.
+Additional arguments to `mine_gmatrix` can also be used to improve the performance of the evolutionary algorithm (see Supporting Information).
 
@@
 ===================================================
+
+The full output of `mine_gmatrix` is passed to the `run_farm_sim` function, which initialises and simulates an evolving population of pests on a changing landscape.
+In this section, we explain the landscape, pest ecology, and evolving pest traits.
+
+Landscape
+---------------------------------------------------
+
+Landscapes are spatially explicit and initialised in `run_farm_sim` in one of two ways. 
+First, a landscapes can be built from the arguments `xdim`, `ydim`, and `farms`. 
+These arguments specify the dimensions of the landscape and the number of farms on it.
+Contiguous rectangular farms of roughly equal size are generated on the landscape using a splitline algorithm.
+Second, a custom landscapes can be input using the `terrain` argument, which takes a matrix with elements that includes integers 1 to `farms`.
+Each value defines a unique farm, but values do not need to be contiguous.
+These 'farms' could even model non-farmland (e.g., water, roads), if pesticides and crops on them are invisible to pests (see below).
+This `terrain` customisation therefore allows for a high degree of landscape detail, and offers the potential for modelling real-world landscapes from raster images.
+
+Each farm can hold one pesticide and one crop type at any time step, and up to 10 pesticides and 10 crop types are available.
 
@@
 - Function `run_farm_sim` runs simulations. Note unusual timing structure
+
+Integrating traits into simulations
+===================================================
 

### 2022-07-16 notebook/ms/ms.Rmd
Write more and finish up the landscape section, move on to pest ecology

@@
 
-Conceptually, the relationship between individual genotypes and traits is defined by a network connecting loci to traits through a set of hidden internal nodes (Figure \@ref(fig:network)). 
-Values for loci are randomly drawn from a standard normal distribution $\mathcal{N}\left(0, 1\right)$. 
-Links between loci, internal nodes, and traits, can take any real value and are represented by black arrows in Figure \@ref(fig:network). 
-Traits can take any real value, and are calculated as the summed effects of all preceding nodes (i.e., the blue squares immediately to the left of the traits in Figure \@ref(fig:network)).
-Mathematically, loci are represented by row a vector of length $L$.
-Effects of loci on the first layer of internal nodes (black arrows emanating from loci in Figure \@ref(fig:network)) are represented by an $L \times T$ matrix, and transitions between internal nodes, and between the last set of internal nodes and the final traits, are represented by $T \times T$ matrices.
-For the example in Figure \@ref(fig:network),
-
-$$
-\begin{pmatrix}
-  0.1, & -0.1, & 0.2 \\
-\end{pmatrix} 
-\begin{pmatrix}
-  0.4, &  0.1 \\
- -1.2, & -0.6 \\
-  0.2, &  2.2 \\
-\end{pmatrix}
-\begin{pmatrix}
-   1.3, & -0.5 \\
-  -0.1  &  0.0 \\
-\end{pmatrix}
-\begin{pmatrix}
-   2.2, &  0.9 \\
-   1.6  &  0.8 \\
-\end{pmatrix} = 
-\begin{pmatrix}
-  0.2998, & 0.1081 \\
-\end{pmatrix}.
-$$
-
-Values mapping loci to traits become part of an individual's genome, so the genome for the individual represented by Figure \@ref(fig:network) is stored in the model as below:
-
-```
-0.1, -0.1, -0.2, 0.4, 0.1, -1.2, -0.6, 0.2, 2.2, 1.3, -0.5, -0.1, 0.0, 2.2, 0.9, 1.6, 0.8
-```
-
-Individuals with different loci can therefore have different covarying traits that are constrained by the network structure encoded in each genome. 
-Individuals can be haploid (as in Figure \@ref(fig:network)) or diploid (in which case, allele values are summed at a locus).
-
 ```{r network, echo = FALSE, fig.width = 6, fig.height = 4, fig.cap = "Example network mapping loci (green circles) to traits (red diamonds) through an intermediate set of hidden layers (blue squares) in the mine_gmatrix function.
@@
 
+Conceptually, the relationship between individual genotypes and traits is defined by a network connecting loci to traits through a set of hidden internal nodes (Figure \@ref(fig:network)). 
+Values for loci are randomly drawn from a standard normal distribution $\mathcal{N}\left(0, 1\right)$. 
+Links between loci, internal nodes, and traits, can take any real value and are represented by black arrows in Figure \@ref(fig:network). 
+Traits can take any real value, and are calculated as the summed effects of all preceding nodes (i.e., the blue squares immediately to the left of the traits in Figure \@ref(fig:network)).
+Mathematically, loci are represented by row a vector of length $L$.
+Effects of loci on the first layer of internal nodes (black arrows emanating from loci in Figure \@ref(fig:network)) are represented by an $L \times T$ matrix, and transitions between internal nodes, and between the last set of internal nodes and the final traits, are represented by $T \times T$ matrices.
+For the example in Figure \@ref(fig:network),
+
+$$
+\begin{pmatrix}
+  0.1, & -0.1, & 0.2 \\
+\end{pmatrix} 
+\begin{pmatrix}
+  0.4, &  0.1 \\
+ -1.2, & -0.6 \\
+  0.2, &  2.2 \\
+\end{pmatrix}
+\begin{pmatrix}
+   1.3, & -0.5 \\
+  -0.1  &  0.0 \\
+\end{pmatrix}
+\begin{pmatrix}
+   2.2, &  0.9 \\
+   1.6  &  0.8 \\
+\end{pmatrix} = 
+\begin{pmatrix}
+  0.2998, & 0.1081 \\
+\end{pmatrix}.
+$$
+
+Values mapping loci to traits become part of an individual's genome, so the genome for the individual represented by Figure \@ref(fig:network) is stored in the model as below:
+
+```
+0.1, -0.1, -0.2, 0.4, 0.1, -1.2, -0.6, 0.2, 2.2, 1.3, -0.5, -0.1, 0.0, 2.2, 0.9, 1.6, 0.8
+```
+
+Individuals with different loci can therefore have different covarying traits that are constrained by the network structure encoded in each genome. 
+Individuals can be haploid (as in Figure \@ref(fig:network)) or diploid (in which case, allele values are summed at a locus).
+
 An evolutionary algorithm is used to find appropriate values that produce covarying traits from loci (see Supporting Information for details). 
@@
 
-Each farm can hold one pesticide and one crop type at any time step, and up to 10 pesticides and 10 crop types are available.
+Each farm can hold one pesticide and one crop type at any time step.
+The `crop_init` and `pesticide_init` arguments initialise one of `crop_number` crops and one of `pesticide_number` pesticides for each farm, respectively (maximum of 10 each).
+Initialisation can be random for each farm with equal probability, or can be set using a vector of length `farms` in which vector elements define the initialised crop or pesticide number.
+After initialisation, crops and pesticides rotate once ever `crop_rotation_time` and `pesticide_rotation_time` time steps, respectively.
+Each of these rotation time arguments can take either an integer value from 1-3, or a matrix.
+Integer values specify (1) no rotation, (2) random transition from one type to another, or (3) cycling through each available crop or pesticide in order.
+Square matrices can be used to define the probability that a given crop or pesticide in row $i$ transitions to that in column $j$. 
+Hence, any possible Markov chain can be used to transition between crop or pesticide types on farms.
+
+Pest ecology
+---------------------------------------------------
+
+Individual pests can be modelled to have several reproductive systems and life histories.
 
@@
 
-Integrating traits into simulations
-===================================================
+Pest evolution
+---------------------------------------------------
 

### 2022-07-17 R/sim_farming.R
Fix a spelling error in the documentation

@@ -278,10 +278,10 @@
 #' 0.
 #'@param min_age_metabolism This determines the minimum age at which losses of 
 #' food consumed in previous time steps enacted by metabolism and 
-#' baseline_metabolsim can occur.
+#' baseline_metabolism can occur.
 #'@param max_age_metabolism This determines the maximum age at which losses of 
 #' food consumed in previous time steps enacted by metabolism and 
-#' baseline_metabolsim can occur.
+#' baseline_metabolism can occur.
 #'@param terrain Insert a custom terrain of different farms, which takes the
 #' form of a matrix that includes a sequence of natural numbers in all matrix
 #' elements. For example, if there are 4 farms, then all matrix elements must

### 2022-07-17 notebook/ms/ms.Rmd
Some work on the pest ecology subsection of the manuscript

@@
 ---
-title: 'resevol: an R package for spatially explicit models of pesticide resistance given evolving pest genomes'
+title: 'resevol: an R package for spatially explicit models of pesticide resistance
+  given evolving pest genomes'
 author: |
@@
 date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
-  UK [2]  alexander.duthie@stir.ac.uk,
-  Biological and Environmental Sciences 3A149 University of Stirling Stirling, FK9
-  4LA, UK'
+  UK [2]  alexander.duthie@stir.ac.uk, Biological and Environmental Sciences 3A149
+  University of Stirling Stirling, FK9 4LA, UK'
 output:
+  pdf_document:
+    toc: no
+  html_document: default
   bookdown::pdf_document2:
@@
     fig_caption: yes
-    toc: false
-  html_document: default
+    toc: no
   word_document: default
@@
 
-The full output of `mine_gmatrix` is passed to the `run_farm_sim` function, which initialises and simulates an evolving population of pests on a changing landscape.
+The full output of `mine_gmatrix` is passed to the `run_farm_sim` function, which initialises and simulates an evolving population of pests on a changing landscape for a natural number of time steps (`time_steps`).
 In this section, we explain the landscape, pest ecology, and evolving pest traits.
@@
 Individual pests can be modelled to have several reproductive systems and life histories.
+Pest reproductive system can be specified in the `run_farm_sim` function using the `repro` argument, which accepts arguments "asexual" (haploid), "sexual" (monoecious), and "biparental" (dioecious).
+For "sexual" pests, the `selfing` argument specifies if self-fertilisation is (`TRUE`) or is not (`FALSE`) allowed.
+At the start of a simulation, pests are initialised in a random location.
+Initialised pests are age of zero if `rand_age = FALSE` or a random age from zero to `max_age` if `rand_age = TRUE`.
+Following initialisation, a single time step proceeds with landscape change (see above), pest aging, feeding, pesticide consumption, movement, reproduction, mortality, and immigration.
+Feeding, pesticide consumption, movement, and reproduction all depend on pest age.
+Pests feed and consume pesticide from ages `min_age_feed` to `max_age_feed`, move from ages  `min_age_move` to `max_age_move`, and reproduce from ages `min_age_reproduce` to `max_age_reproduce` (all inclusive).
+Food accumulated is lost during aging if `baseline_metabolism > 0` and pest age is within `min_age_metabolism` and `max_age_metabolism`.
+The option to set minimum and maximum ages events makes it possible to model pests with much different life histories [e.g., @Sudo2018].
+
+```{r, echo=FALSE, fig.height=1.75, fig.width=6, fig.cap = "Overview of simulated events in the resevol R package.
+Note that metabolism, feeding, pesticide consumption, movement, and reproduction are all subject to a minimum and maximum pest age.
+Consequently, simulation order might not reflect the order of events from the perspective of a focal pest (e.g., pests might move from ages 1-2, but only feed from ages 2-4)."}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,100), ylim=c(50,95), xaxt="n", yaxt="n",
+     xlab="",ylab="");
+ibox <- mbox(x0 = 0,  x1 = 10, y0 = 88, y1 = 72);
+polygon(x = ibox$x, y = ibox$y, lwd = 3, border = "black", col = "white");
+cbox <- mbox(x0 = 15, x1 = 25, y0 = 88, y1 = 72);
+polygon(x = cbox$x, y = cbox$y, lwd = 3, border = "black", col = "white");
+ubox <- mbox(x0 = 30, x1 = 40, y0 = 88, y1 = 72);
+polygon(x = ubox$x, y = ubox$y, lwd = 3, border = "black", col = "white");
+nbox <- mbox(x0 = 45, x1 = 55, y0 = 88, y1 = 72);
+polygon(x = nbox$x, y = nbox$y, lwd = 3, border = "black", col = "white");
+fbox <- mbox(x0 = 60, x1 = 70, y0 = 88, y1 = 72);
+polygon(x = fbox$x, y = fbox$y, lwd = 3, border = "black", col = "white");
+tbox <- mbox(x0 = 75, x1 = 85, y0 = 88, y1 = 72);
+polygon(x = tbox$x, y = tbox$y, lwd = 3, border = "black", col = "white");
+tbox <- mbox(x0 = 90, x1 = 100, y0 = 88, y1 = 72);
+polygon(x = tbox$x, y = tbox$y, lwd = 3, border = "black", col = "white");
+text(x=5,  y=84, labels="Initialise", col="black", cex = 0.5);
+text(x=5,  y=80, labels="pests and", col="black", cex = 0.5);
+text(x=5,  y=76, labels="landscape", col="black", cex = 0.5);
+text(x=20, y=84, labels="Rotate", col="black", cex = 0.5);
+text(x=20, y=80, labels="crops and", col="black", cex = 0.5);
+text(x=20, y=76, labels="pesticides", col="black", cex = 0.5);
+
+text(x=35, y=80, labels="Mutation", col="black", cex = 0.5);
+
+text(x=35, y=80, labels="Mutation", col="black", cex = 0.5);
+text(x=50, y=82, labels="Fitness", col="black", cex = 0.5);
+text(x=50, y=78, labels="evaluation", col="black", cex = 0.5);
+text(x=65, y=82, labels="Tournament", col="black", cex = 0.5);
+text(x=65, y=78, labels="selection", col="black", cex = 0.5);
+text(x=80, y=80, labels="Replacement", col="black", cex = 0.5);
+arrows(x0=10, x1=14, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=25, x1=29, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=40, x1=44, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=55, x1=59, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=70, x1=74, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=85, x1=89, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0 = 80, x1 = 80, y0 = 72, y1 = 68, lwd = 2, length = 0);
+arrows(x0 = 80, x1 = 20, y0 = 65, y1 = 65, lwd = 2, length = 0);
+arrows(x0 = 20, x1 = 20, y0 = 65, y1 = 72, lwd = 2, length = 0.05);
+text(x=73, y=63, labels="No", col="black", cex = 0.5);
+rbox <- mbox(x0 = 75, x1 = 85, y0 = 62, y1 = 68);
+polygon(x = rbox$x, y = rbox$y, lwd = 3, border = "black", col = "black");
+text(x=80, y=65, labels="Termination?", col="white", cex = 0.5);
+fbox <- mbox(x0 = 60, x1 = 70, y0 = 62, y1 = 50);
+polygon(x = fbox$x, y = fbox$y, lwd = 3, border = "black", col = "white");
+text(x=65, y=57, labels="Individual", col="black", cex = 0.5);
+text(x=65, y=53, labels="network", col="black", cex = 0.5);
+arrows(x0 = 80, x1 = 80, y0 = 62, y1 = 55, lwd = 2, length = 0);
+arrows(x0 = 80, x1 = 70, y0 = 55, y1 = 55, lwd = 2, length = 0.05);
+text(x=73, y=53, labels="Yes", col="black", cex = 0.5);
+```
+
+Pests feed by
+
+Pesticide is consumed during feeding
+
+Movement is modelled by
+
+Pests reproduce by
+
+Immigration occurs.
 

### 2022-07-19 notebook/index.Rmd
Write a new journal post explaining the new changes in version 0.3.1.0, and why they are useful. Add a bit more to the manuscript draft

@@
 
+   - **JUL** [19](#u-19-JUL-2022)
    - **JUN** [13](#u-13-JUN-2022), [28](#u-28-JUN-2022)
@@
 ================================================================================
+
+> <a name="u-19-JUL-2022">Update: 19 JUL 2022</a>
+
+Version 0.3.0.1 has been submitted to CRAN and is passing [all checks](https://cran.r-project.org/web/checks/check_results_resevol.html).
+I am now a bit under halfway through writing a first draft of the software note.
+This note be submitted as a pre-print, and for publication, and will explain how to use the package.
+It will also replace the current [Get Started](https://bradduthie.github.io/resevol/articles/resevol.html) article and [CRAN vignette](https://cran.r-project.org/web/packages/resevol/vignettes/resevol.html).
+The structure of the publication introduces some key background, the `mine_gmatrix` function and its underlying logic, the ecology and evolution of pests in the model, and a simple example.
+I think that two supporting information documents are also in order.
+The first will explain the evolutionary algorithm in more detail and basically be a version of the [current vignette](https://bradduthie.github.io/resevol/articles/evolutionary_algorithm.html) with maybe some minor revisions.
+The second will be a more advanced example with complex dynamics and traits, presenting simulations like the ones shown in this journal on [26 AUG 2021](#u-26-AUG-2021).
+My hope is that this plus the documentation is sufficient for anyone to start using the package.
+
+Having submitted version 0.3.0.1, I almost immediately discovered the need for an update that is now version 0.3.1.0.
+I will hold off on submitting this one to CRAN just yet, as the update is quite minor.
+The cause was that the `mine_gmatrix` function kept giving a discrepancy between the stress threshold that caused it to terminate and the actual predicted stress of the network it found.
+The reason for this was that the algorithm was terminating due to finding unusually low stress caused sampling error for the standard random normal loci.
+In other words, a luck draw for random loci often caused networks to look better at producing desired trait covariances than they actually were.
+I've addressed this now by first using the mean stress for a whole population of networks to test the termination criteria (`term_cri`) in the `mine_gmatrix` function, rather than the *best* stress out of all networks in the population.
+But the `mine_gmatrix` function still returns the network that produces the lowest found stress.
+
+Second, I have introduced a new diagnostic function to give the distribution of stress predicted from loci.
+
+```{r, eval = FALSE}
+stress_test <- function(mine_output, indivs = 1000, reps = 10){
+
+    target_VCV     <- mine_output[[2]];
+    Loci_to_Traits <- mine_output[[5]];
+    loci           <- mine_output[[1]][1];
+
+    Ln_stress <- rep(x = NA, times = reps);
+    while(reps > 0){
+        sim_loci        <- rnorm(n = indivs * loci);
+        individuals     <- matrix(data = sim_loci, nrow = indivs, ncol = loci);
+        trait_vals      <- individuals %*% Loci_to_Traits;
+        VCV             <- cov(trait_vals);
+        stress          <- mean((target_VCV - VCV) * (target_VCV - VCV));
+        Ln_stress[reps] <- log(stress);
+        reps            <- reps - 1;
+    }
+
+    return(Ln_stress);
+}
+```
+
+This simple function now forms part of the package itself, so users can run the output of `mine_gmatrix` and get a better overall picture on how the chosen network performs.
+The example used in the documentation is below (I've increased the replicates using the argument `reps = 1000`; the default is 10).
+
+```{r}
+gmt       <- matrix(data = 0, nrow = 4, ncol = 4);
+diag(gmt) <- 1;
+mg        <- mine_gmatrix(gmatrix = gmt, loci = 4, layers = 3, indivs = 100, 
+                          npsize = 100, max_gen = 2, prnt_out = FALSE);
+stresses  <- stress_test(mine_output = mg, reps = 1000);
+```
+
+The output `stresses` is just a vector of stresses for some number (`indivs`) of initialised genomes. The histogram below shows the distribution for the above example.
+
+```{r, echo = FALSE}
+hist(stresses, xlab = "Stress values for initialised genomes", 
+     main = "", breaks = 20);
+```
+
+This should at least make it easier for users to decide if they do or do not like the output of `mine_gmatrix` for use in a new set of simulations.
+The above example shows stress that is quite consistent, but if we let the simulation run for longer to find a better solution, the stress variance seems to increase (note that now `max_gen = 400` below).
+
+```{r}
+gmt       <- matrix(data = 0, nrow = 4, ncol = 4);
+diag(gmt) <- 1;
+mg        <- mine_gmatrix(gmatrix = gmt, loci = 4, layers = 3, indivs = 100, 
+                          npsize = 100, max_gen = 400, prnt_out = FALSE);
+stresses  <- stress_test(mine_output = mg, reps = 1000);
+hist(stresses, xlab = "Stress values for initialised genomes", 
+     main = "", breaks = 20);
+```
+
+It might be good to recommend this kind of diagnostic for users, particularly when simulating very complex individuals.
+I think that I will do this for the supporting information showing an advanced example.
 

### 2022-07-19 notebook/ms/ms.Rmd
Write a new journal post explaining the new changes in version 0.3.1.0, and why they are useful. Add a bit more to the manuscript draft
Code Figure 2 for the manuscript draft

--- 2022-07-16 notebook/ms/ms.Rmd
@@
 Individual pests can be modelled to have several reproductive systems and life histories.
+Pest reproductive system can be specified in the `run_farm_sim` function using the `repro` argument, which accepts arguments "asexual" (haploid), "sexual" (monoecious), and "biparental" (dioecious).
+For "sexual" pests, the `selfing` argument specifies if self-fertilisation is (`TRUE`) or is not (`FALSE`) allowed.
+At the start of a simulation, pests are initialised in a random location.
+Initialised pests are age of zero if `rand_age = FALSE` or a random age from zero to `max_age` if `rand_age = TRUE`.
+Following initialisation, a single time step proceeds with landscape change (see above), pest aging, feeding, pesticide consumption, movement, reproduction, mortality, and immigration.
+Feeding, pesticide consumption, movement, and reproduction all depend on pest age.
+Pests feed and consume pesticide from ages `min_age_feed` to `max_age_feed`, move from ages  `min_age_move` to `max_age_move`, and reproduce from ages `min_age_reproduce` to `max_age_reproduce` (all inclusive).
+Food accumulated is lost during aging if `baseline_metabolism > 0` and pest age is within `min_age_metabolism` and `max_age_metabolism`.
+The option to set minimum and maximum ages for events makes it possible to model pests with much different life histories [e.g., @Sudo2018].
+
+```{r, echo=FALSE, fig.height=3, fig.width=8, fig.cap = "Overview of simulated events in the resevol R package.
+Note that metabolism, feeding, pesticide consumption, movement, and reproduction are all subject to a minimum and maximum pest age.
+Consequently, simulation order might not reflect the order of events from the perspective of a focal pest (e.g., pests might move from ages 1-2, but only feed from ages 2-4).
+Crops and pesticides are also not necessarily rotated in each time step (see Landscape).
+Statistics collected within a time step are printed in a CSV file."}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+par(mar=c(0,0,0,0));
+# ===============================================================
+plot(x=0, y=0, type="n", xlim=c(0,100), ylim=c(40,95), xaxt="n", yaxt="n",
+     xlab="",ylab="");
+ibox <- mbox(x0 = 0,  x1 = 10, y0 = 63, y1 = 47);
+polygon(x = ibox$x, y = ibox$y, lwd = 5, border = "black", col = "white");
+cbox <- mbox(x0 = 0, x1 = 10, y0 = 88, y1 = 72);
+polygon(x = cbox$x, y = cbox$y, lwd = 3, border = "black", col = "white");
+ubox <- mbox(x0 = 15, x1 = 25, y0 = 88, y1 = 72);
+polygon(x = ubox$x, y = ubox$y, lwd = 3, border = "black", col = "white");
+nbox <- mbox(x0 = 30, x1 = 40, y0 = 88, y1 = 72);
+polygon(x = nbox$x, y = nbox$y, lwd = 3, border = "black", col = "white");
+fbox <- mbox(x0 = 45, x1 = 55, y0 = 88, y1 = 72);
+polygon(x = fbox$x, y = fbox$y, lwd = 3, border = "black", col = "white");
+tbox <- mbox(x0 = 60, x1 = 70, y0 = 88, y1 = 72);
+polygon(x = tbox$x, y = tbox$y, lwd = 3, border = "black", col = "white");
+tbox <- mbox(x0 = 75, x1 = 85, y0 = 88, y1 = 72);
+polygon(x = tbox$x, y = tbox$y, lwd = 3, border = "black", col = "white");
+tbox <- mbox(x0 = 90, x1 = 100, y0 = 88, y1 = 72);
+polygon(x = tbox$x, y = tbox$y, lwd = 3, border = "black", col = "white");
+ibox <- mbox(x0 = 90,  x1 = 100, y0 = 63, y1 = 47);
+polygon(x = ibox$x, y = ibox$y, lwd = 3, border = "black", col = "white");
+tbox <- mbox(x0 = 75, x1 = 85, y0 = 63, y1 = 47);
+polygon(x = tbox$x, y = tbox$y, lwd = 3, border = "black", col = "white");
+text(x=5,  y=58, labels="Initialise", col="black", cex = 0.75);
+text(x=5,  y=54, labels="pests and", col="black", cex = 0.75);
+text(x=5,  y=50, labels="landscape", col="black", cex = 0.75);
+text(x=5, y=84, labels="Crops and", col="black", cex = 0.75);
+text(x=5, y=80, labels="pesticides", col="black", cex = 0.75);
+text(x=5, y=76, labels="rotated?", col="black", cex = 0.75);
+text(x=20, y=82, labels="Pests", col="black", cex = 0.75);
+text(x=20, y=78, labels="metabolise", col="black", cex = 0.75);
+text(x=35, y=82, labels="Pests", col="black", cex = 0.75);
+text(x=35, y=78, labels="feed", col="black", cex = 0.75);
+text(x=50, y=82, labels="Pesticide", col="black", cex = 0.75);
+text(x=50, y=78, labels="consumed", col="black", cex = 0.75);
+text(x=65, y=82, labels="Pests", col="black", cex = 0.75);
+text(x=65, y=78, labels="move", col="black", cex = 0.75);
+text(x=80, y=82, labels="Pests", col="black", cex = 0.75);
+text(x=80, y=78, labels="reproduce", col="black", cex = 0.75);
+text(x=95, y=82, labels="Pest", col="black", cex = 0.75);
+text(x=95, y=78, labels="mortality", col="black", cex = 0.75);
+text(x=95,  y=58, labels="Pests", col="black", cex = 0.75);
+text(x=95,  y=54, labels="immigrate", col="black", cex = 0.75);
+text(x=80,  y=59, labels="Collect", col="black", cex = 0.75);
+text(x=80,  y=55, labels="statistics", col="black", cex = 0.75);
+text(x=80,  y=51, labels="as CSV", col="black", cex = 0.75);
+arrows(x0=2, x1=2,   y0=63, y1=71.5, lwd=2, length=0.10);
+arrows(x0=10, x1=14.5, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=25, x1=29.5, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=40, x1=44.5, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=55, x1=59.5, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=70, x1=74.5, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=85, x1=89.5, y0=80, y1=80, lwd=2, length=0.10);
+arrows(x0=95, x1=95,   y0=72, y1=63.5, lwd=2, length=0.10);
+arrows(x0=90, x1=85.5, y0=55, y1=55, lwd=2, length=0.10);
+arrows(x0=75, x1=70.5, y0=60, y1=60, lwd=2, length=0.10);
+arrows(x0=60, x1=55.5, y0=60, y1=60, lwd=2, length=0.10);
+rbox <- mbox(x0 = 60, x1 = 70, y0 = 57, y1 = 63);
+polygon(x = rbox$x, y = rbox$y, lwd = 3, border = "black", col = "black");
+text(x=65, y=60, labels="Extinction?", col="white", cex = 0.75);
+rbox <- mbox(x0 = 45, x1 = 55, y0 = 57, y1 = 63);
+polygon(x = rbox$x, y = rbox$y, lwd = 3, border = "black", col = "black");
+text(x=50, y=60, labels="Max. time?", col="white", cex = 0.75);
+text(x = 58.5, y = 58.5, labels = "No", col = "black", cex = 0.75);
+text(x = 43.5, y = 58.5, labels = "No", col = "black", cex = 0.75);
+text(x = 68, y = 54, labels = "Yes", col = "black", cex = 0.75);
+text(x = 53, y = 54, labels = "Yes", col = "black", cex = 0.75);
+arrows(x0=66, x1=66, y0=57, y1=50, lwd=2, length=0.0);
+arrows(x0=51, x1=51, y0=57, y1=50, lwd=2, length=0.0);
+rbox <- mbox(x0 = 30, x1 = 40, y0 = 47, y1 = 53);
+polygon(x = rbox$x, y = rbox$y, lwd = 3, border = "black", col = "black");
+text(x=35, y=50, labels="STOP", col="white", cex = 1);
+rbox <- mbox(x0 = 10, x1 = 16, y0 = 47, y1 = 63);
+polygon(x = rbox$x, y = rbox$y, lwd = 5, border = "black", col = "black");
+text(x=13, y=55, labels="START", col="white", cex = 1, srt = 90);
+arrows(x0=66, x1=40.5, y0=50, y1=50, lwd=2, length=0.10);
+arrows(x0=45, x1=35, y0=60, y1=60, lwd=2, length=0.0);
+arrows(x0=35, x1=35, y0=60, y1=67, lwd=2, length=0.0);
+arrows(x0=35, x1=7, y0=67, y1=67, lwd=2, length=0.0);
+arrows(x0=7, x1=7, y0=67, y1=71.5, lwd=2, length=0.1);
+```
+
+Pests feed by
+
+Pesticide is consumed during feeding
+
+Movement is modelled by
+
+Pests reproduce by
+
+Immigration occurs.
 

### 2022-07-20 R/initialise.R
Fix a minor white space inconsistency in the code

@@ -121,7 +121,7 @@ initialise_inds <- function(mine_output,
     stop("ERROR: Must specify 'repro' as asexual, sexual, or biparental.");
   }
   if(repro == "sexual" | repro == "biparental"){
-    inds      <- build_sexual(mine_output, N, neutral_loci);
+    inds <- build_sexual(mine_output, N, neutral_loci);
   }else{
     inds <- build_asexual(mine_output, N, neutral_loci);
   }

### 2022-07-20 notebook/ms/ms.Rmd
write the ecology section, now moving on to pest evolution

@@
 output:
-  pdf_document:
-    toc: no
-  html_document: default
   bookdown::pdf_document2:
@@
   word_document: default
+  html_document: default
 header-includes:
@@
 - \fancyhead[R]{RESEVOL, AN R PACKAGE}
+- \usepackage{lineno}
+- \linenumbers
+- \modulolinenumbers[2]
 linestretch: 1
@@
 Ly  <- seq(from = 250, to = 750, length = 3);
-points(x = Lx, y = Ly, pch = 20, cex = 8, col = "darkgreen");
 H1x <- rep(x = 400, times = 2);
 H1y <- seq(from = 250, to = 750, length = 2);
-points(x = H1x, y = H1y, pch = 15, cex = 8, col = "blue");
 H2x <- rep(x = 600, times = 2);
 H2y <- seq(from = 250, to = 750, length = 2);
-points(x = H2x, y = H2y, pch = 15, cex = 8, col = "blue");
 Tx  <- rep(x = 900, times = 2);
 Ty  <- seq(from = 250, to = 750, length = 2);
-points(x = Tx, y = Ty, pch = 18, cex = 10, col = "red");
+points(x = Tx, y = Ty, pch = 23, cex = 8, bg = "red", lwd = 2);
 sep_L_points <- c(-50, 0, 50);
 sep_N_points <- c(-30, 30);
+arrows(x0 = H2x + 52, x1 = Tx[1] - 58, y0 = H2y, y1 = Ty[1] + sep_N_points, 
+       length = 0.08, lwd = 3);
+arrows(x0 = H2x + 52, x1 = Tx[2] - 58, y0 = H2y, y1 = Ty[2] + sep_N_points, 
+       length = 0.08, lwd = 3);
 arrows(x0 = Lx + 38, x1 = H1x[1] - 58, y0 = Ly, y1 = H1y[1] + sep_L_points, 
        length = 0.08, lwd = 3);
+points(x = H2x, y = H2y, pch = 22, cex = 9, bg = "blue", lwd = 2);
 arrows(x0 = H1x + 52, x1 = H2x[1] - 58, y0 = H1y, y1 = H2y[1] + sep_N_points, 
-       length = 0.08, lwd = 3);
-arrows(x0 = H2x + 52, x1 = Tx[1] - 58, y0 = H2y, y1 = Ty[1] + sep_N_points, 
-       length = 0.08, lwd = 3);
-arrows(x0 = Lx + 38, x1 = H1x[2] - 58, y0 = Ly, y1 = H1y[2] + sep_L_points, 
        length = 0.08, lwd = 3);
@@
        length = 0.08, lwd = 3);
-arrows(x0 = H2x + 52, x1 = Tx[2] - 58, y0 = H2y, y1 = Ty[2] + sep_N_points, 
+points(x = H1x, y = H1y, pch = 22, cex = 9, bg = "blue", lwd = 2);
+arrows(x0 = Lx + 38, x1 = H1x[2] - 58, y0 = Ly, y1 = H1y[2] + sep_L_points, 
        length = 0.08, lwd = 3);
+points(x = Lx, y = Ly, pch = 21, cex = 6, bg = "darkgreen", lwd = 2);
 srnm <- c(0.2, -0.1, 0.1);
@@
 This `terrain` customisation therefore allows for a high degree of landscape detail, and offers the potential for modelling real-world landscapes from raster images.
+Landscapes are assumed to be a torus, so pests that move off of one edge return on the opposite side of the landscape.
 
@@
 Individual pests can be modelled to have several reproductive systems and life histories.
-Pest reproductive system can be specified in the `run_farm_sim` function using the `repro` argument, which accepts arguments "asexual" (haploid), "sexual" (monoecious), and "biparental" (dioecious).
-For "sexual" pests, the `selfing` argument specifies if self-fertilisation is (`TRUE`) or is not (`FALSE`) allowed.
+Pest reproductive system can be specified in the `run_farm_sim` function using the `repro` argument, which accepts arguments `"asexual"` (haploid), `"sexual"` (monoecious), and `"biparental"` (dioecious).
+For `"sexual"` pests, the `selfing` argument specifies if self-fertilisation is (`TRUE`) or is not (`FALSE`) allowed.
 At the start of a simulation, pests are initialised in a random location.
 Initialised pests are age of zero if `rand_age = FALSE` or a random age from zero to `max_age` if `rand_age = TRUE`.
-Following initialisation, a single time step proceeds with landscape change (see above), pest aging, feeding, pesticide consumption, movement, reproduction, mortality, and immigration.
+Following initialisation, a single time step proceeds with landscape change (see above), pest aging and metabolism, feeding, pesticide consumption, movement, reproduction, mortality, and immigration.
 Feeding, pesticide consumption, movement, and reproduction all depend on pest age.
@@
      xlab="",ylab="");
-ibox <- mbox(x0 = 0,  x1 = 10, y0 = 63, y1 = 47);
+ibox <- mbox(x0 = 0,  x1 = 14, y0 = 63, y1 = 47);
 polygon(x = ibox$x, y = ibox$y, lwd = 5, border = "black", col = "white");
@@
 polygon(x = tbox$x, y = tbox$y, lwd = 3, border = "black", col = "white");
-text(x=5,  y=58, labels="Initialise", col="black", cex = 0.75);
-text(x=5,  y=54, labels="pests and", col="black", cex = 0.75);
-text(x=5,  y=50, labels="landscape", col="black", cex = 0.75);
-text(x=5, y=84, labels="Crops and", col="black", cex = 0.75);
-text(x=5, y=80, labels="pesticides", col="black", cex = 0.75);
-text(x=5, y=76, labels="rotated?", col="black", cex = 0.75);
-text(x=20, y=82, labels="Pests", col="black", cex = 0.75);
-text(x=20, y=78, labels="metabolise", col="black", cex = 0.75);
-text(x=35, y=82, labels="Pests", col="black", cex = 0.75);
-text(x=35, y=78, labels="feed", col="black", cex = 0.75);
-text(x=50, y=82, labels="Pesticide", col="black", cex = 0.75);
-text(x=50, y=78, labels="consumed", col="black", cex = 0.75);
-text(x=65, y=82, labels="Pests", col="black", cex = 0.75);
-text(x=65, y=78, labels="move", col="black", cex = 0.75);
-text(x=80, y=82, labels="Pests", col="black", cex = 0.75);
-text(x=80, y=78, labels="reproduce", col="black", cex = 0.75);
-text(x=95, y=82, labels="Pest", col="black", cex = 0.75);
-text(x=95, y=78, labels="mortality", col="black", cex = 0.75);
-text(x=95,  y=58, labels="Pests", col="black", cex = 0.75);
-text(x=95,  y=54, labels="immigrate", col="black", cex = 0.75);
-text(x=80,  y=59, labels="Collect", col="black", cex = 0.75);
-text(x=80,  y=55, labels="statistics", col="black", cex = 0.75);
-text(x=80,  y=51, labels="as CSV", col="black", cex = 0.75);
+text(x=7,  y=59, labels="Initialise", col="black", cex = 0.8);
+text(x=7,  y=55, labels="pests and", col="black", cex = 0.8);
+text(x=7,  y=51, labels="landscape", col="black", cex = 0.8);
+text(x=5, y=84, labels="Crops and", col="black", cex = 0.8);
+text(x=5, y=80, labels="pesticides", col="black", cex = 0.8);
+text(x=5, y=76, labels="rotated?", col="black", cex = 0.8);
+text(x=20, y=82, labels="Pests", col="black", cex = 0.8);
+text(x=20, y=78, labels="metabolise", col="black", cex = 0.8);
+text(x=35, y=82, labels="Pests", col="black", cex = 0.8);
+text(x=35, y=78, labels="feed", col="black", cex = 0.8);
+text(x=50, y=82, labels="Pesticide", col="black", cex = 0.8);
+text(x=50, y=78, labels="consumed", col="black", cex = 0.8);
+text(x=65, y=82, labels="Pests", col="black", cex = 0.8);
+text(x=65, y=78, labels="move", col="black", cex = 0.8);
+text(x=80, y=82, labels="Pests", col="black", cex = 0.8);
+text(x=80, y=78, labels="reproduce", col="black", cex = 0.8);
+text(x=95, y=82, labels="Pest", col="black", cex = 0.8);
+text(x=95, y=78, labels="mortality", col="black", cex = 0.8);
+text(x=95,  y=57, labels="Pests", col="black", cex = 0.8);
+text(x=95,  y=53, labels="immigrate", col="black", cex = 0.8);
+text(x=80,  y=59, labels="Collect", col="black", cex = 0.8);
+text(x=80,  y=55, labels="statistics", col="black", cex = 0.8);
+text(x=80,  y=51, labels="as CSV", col="black", cex = 0.8);
 arrows(x0=2, x1=2,   y0=63, y1=71.5, lwd=2, length=0.10);
@@
 text(x=35, y=50, labels="STOP", col="white", cex = 1);
-rbox <- mbox(x0 = 10, x1 = 16, y0 = 47, y1 = 63);
+rbox <- mbox(x0 = 14, x1 = 20, y0 = 47, y1 = 63);
 polygon(x = rbox$x, y = rbox$y, lwd = 5, border = "black", col = "black");
-text(x=13, y=55, labels="START", col="white", cex = 1, srt = 90);
+text(x=17, y=55, labels="START", col="white", cex = 1, srt = 90);
 arrows(x0=66, x1=40.5, y0=50, y1=50, lwd=2, length=0.10);
@@
 
-Pests feed by
+In each time step, pests feed in a random order, consuming the crop on their landscape cell.
+Pests consume a maximum amount of the crop as specified by the `food_consume` argument, which takes a vector with as many elements as their are crops (i.e., if `crop_number = 2`, then `food_consume` has two elements, the first and second defining consumption of crops 1 and 2, respectively).
+If crop amount on the landscape cell exceeds pest consumption ability, then pests consume their maximum amount, and this amount is removed from the landscape cell.
+If crop amount is less than pest consumption ability, then pests consume whatever crop is left and crop amount is reduced to zero.
+Pesticide consumption works identically to crop consumption, except that the amount of pesticide on a landscape cell is not decreased.
+Pests simply consume an amount of pesticide as specified by the `pesticide_consume` argument, which also takes a vector with as many elements as their are pesticides.
+Hence, each pest can potentially feed and be affected by the pesticide of their focal landscape cell.
 
-Pesticide is consumed during feeding
+After interacting with their landscape cell, pests can move. 
+Each pest visits a number of cells during movement, as is specified by the parameter `movement_bouts`. 
+Individual movement bouts occur in a random order across pests.
+During a movement bout, a pest can travel to any cell within a value defined by `movement_distance` from their current location, which could include their current location (i.e., moving zero distance).
+Upon arrival to a cell, a pest can feed if they are of an appropriate feeding age and `feed_while_moving = TRUE`.
+The pest also consumes pesticide if they are of the appropriate age and `pesticide_while_moving = TRUE`.
+Having pests feed and consume pesticide in a random order while moving among landscape cells can model a population competing for food and encountering pesticides on a shorter time scale than an individual time step.
 
-Movement is modelled by
-
-Pests reproduce by
-
-Immigration occurs.
-
-- We want to model evolution of pesticide resistance across heterogenous landscape
-- Track entire genomes, traits, and yield over time and across space
-- Use covarying traits from `mine_gmatrix` to make evolving pests
-- Options for 10 crops and 10 pesticides simultaneously, and rotation
-- Function `run_farm_sim` runs simulations. Note unusual timing structure
+After all pests are done moving, pests reproduce.
+Offspring production is possible for asexual, monoecious, or female pests.
+Pest expected offspring number is defined by a fixed parameter if `reproduction_type = "lambda"`, or is calculated from the amount of food consumed if  `reproduction_type = "food_based"`.
+The former requires specifying `lambda_value`, which becomes the rate parameter for sampling offspring number from a Poisson distribution.
+The latter requires specifying a real value for `food_needed_repr`, which is the amount of food needed to produce one offspring.
+For food-based reproduction, the total amount of food consumed is divided by `food_needed_repr`, then floored to determine offspring number.
+Sexual reproduction requires a mate of reproductive age that is either monoecious or male, and within range of the reproducing focal pest (potential including the focal pest, if `selfing = TRUE`).
+A potential mate is within range if it is within an integer number of cells from the focal pest, as defined by `mating_distance` (e.g., if `mating_distance = 0`, then mates must share a landscape cell).
+All available potential mates sire offspring with equal probability, and reproducing pests are assumed to mate multiply (i.e., paternity is a fair raffle for all offspring). <!--- XXX CHECK THIS --->
+If a carrying capacity on birth is set (`K_on_birth > 0`) and total offspring number in the population exceeds this capacity, then offspring are removed at random until they are  within carrying capacity.
+An real value `immigration_rate` specifies the rate parameter for Poisson random sampling of the number of immigrants added to the population at the end of a time step.
+Immigrants are initialised in the same way as pests were at the start of the simulation.
+Hence, any spatial structure or evolution that occurs during the simulation does not affect immigrant locations, genomes, or traits.
 
@@
 ---------------------------------------------------
+
+Pest genomes evolve in a complex and highly mechanistic way.
+Offspring inherit genome values from their parent(s) with the possibility for mutation and recombination; offspring traits are then calculated from their newly created genomes.
+For asexual pests, genomes are haploid and mutation .
 

### 2022-07-20 src/statistics.c
fixed a typo in the notes

@@ -49,7 +49,7 @@ void print_all_pests(double **pests, double *paras, int ts){
 
 
 /* =============================================================================
- * Prints population level statistics in the pet population
+ * Prints population level statistics in the pest population
  *     pests:           The array holding the parent's information
  *     paras:           The paras vector that holds global information
  * ========================================================================== */

### 2022-07-20 notebook/ms/ms.Rmd
finish evolutionary section and a section on output

@@
 Initialised pests are age of zero if `rand_age = FALSE` or a random age from zero to `max_age` if `rand_age = TRUE`.
-Following initialisation, a single time step proceeds with landscape change (see above), pest aging and metabolism, feeding, pesticide consumption, movement, reproduction, mortality, and immigration.
+Following initialisation, a single time step proceeds with landscape change (see above), pest aging and metabolism, feeding, pesticide consumption, movement, reproduction, mortality, and immigration (Figure \@ref(fig:flowchart)).
 Feeding, pesticide consumption, movement, and reproduction all depend on pest age.
@@
 
-```{r, echo=FALSE, fig.height=3, fig.width=8, fig.cap = "Overview of simulated events in the resevol R package.
+```{r flowchart, echo=FALSE, fig.height=3, fig.width=8, fig.cap = "Overview of simulated events in the resevol R package.
 Note that metabolism, feeding, pesticide consumption, movement, and reproduction are all subject to a minimum and maximum pest age.
@@
 A potential mate is within range if it is within an integer number of cells from the focal pest, as defined by `mating_distance` (e.g., if `mating_distance = 0`, then mates must share a landscape cell).
-All available potential mates sire offspring with equal probability, and reproducing pests are assumed to mate multiply (i.e., paternity is a fair raffle for all offspring). <!--- XXX CHECK THIS --->
+All available potential mates sire offspring with equal probability, and reproducing pests are assumed to mate multiply (i.e., paternity is a fair raffle for all offspring).
 If a carrying capacity on birth is set (`K_on_birth > 0`) and total offspring number in the population exceeds this capacity, then offspring are removed at random until they are  within carrying capacity.
@@
 Pest genomes evolve in a complex and highly mechanistic way.
-Offspring inherit genome values from their parent(s) with the possibility for mutation and recombination; offspring traits are then calculated from their newly created genomes.
-For asexual pests, genomes are haploid and mutation .
+Offspring inherit genome values from their parent(s) with the possibility for mutation and recombination; offspring traits are then calculated from their newly initialised genomes.
+For asexually and sexually reproducing pests, genomes are haploid and diploid, respectively. 
+Asexually reproducing pests receive the full genomes of their parent, while sexually reproducing pests receive half of their alleles from each parent.
+Each diploid parent contributes one half of their genome, effectively modelling a genome with a single paired chromosome.
+Crossover occurs at each position of the genome with a probability of `crossover_pr`, so complete recombination is also possible if `crossover_pr = 0.5`.
+For both haploids and diploids, each genome value then mutates independently with a probability of `mutation_pr`, which can be set to any real number between 0 and 1.
+If a genome value mutates, then a new value is randomly sampled from a standard normal distribution.
+If `mutation_type = 0`, then this new value replaces the old value, and if `mutation_type = 1`, then the new value is added to the old value.
+After mutation, genome values are used to calculate trait values.
+
+Evolution of the genetic architecture linking loci to traits can be constrained by disabling mutation genome values linking loci, internal nodes, and traits (i.e., 'network values' represented by arrows in Figure \@ref(fig:network)).
+While mutation at loci (green circles in Figure \@ref(fig:network)) is always possible as long as `mutation_pr > 0`, the number of intermediary layers for which network values can mutate is constrained by `net_mu_layers`.
+If `net_mu_layers = 0`, then no network values can mutate, but higher integer values cause mutation to occur at network value layers from loci to traits (if `net_mu_dir = 1`) or traits to loci (if `net_mu_dir = 0`).
+For example, if `net_mu_layers = 2` and `net_mu_dir = 1`, then the network values linking loci to the first internal node, and the first internal node to the second, can mutate (i.e., first two columns of arrows in \@ref(fig:network), but not those linking the second internal node to traits).
+This allows pest traits to evolve with varying degrees of constraint on the covariance between traits.
+Low `net_mu_layers` values model strong genetic constraints, while high values model high evolvability of trait covariances.
+
+Finally, evolving and covarying traits can be used in place of fixed parameters described in pest ecology.
+This is done by substituting `"Tj"` as argument input in place of a numeric value, where `j` represents the trait number.
+For example, the argument `move_distance = "T1"` will make Trait 1 the movement distance for individuals. 
+The argument `food_consume = c("T2", "T3")` will set Traits 2 and 3 to define the amount of food of types 1 and 2 that can be consumed by a pest, respectively.
+Parameters that can be replaced include `move_distance`, `food_needed_surv`, `pesticide_tolerated_surv`, `food_needed_repr`, `pesticide_tolerated_repr`, `mating_distance`, `lambda_value`, `movement_bouts`, `metabolism`, `food_consume`, and `pesticide_consume`.
+The resevol package can thereby simulate agricultural pests with complex and co-evolving traits, and potentially evolving trait covariances, under a range of possible pest life histories.
+
+Simulation output
+===================================================
+
+Simulation output can be large, so output is printed in two CSV files, both of which are created in the working directory.
+The first file "population_data.csv" prints population level data over time, including population size, mean age, sex ratio, mean food and pesticide consumed of each type, mortality rate, and mean trait values.
+The second file "individuals.csv" prints all information, including full genomes and traits (columns), for every individual (rows) in the population.
+The printing of individual level data is disabled by default.
+It can be turned on for all time steps by setting `print_inds = TRUE`, but this should be done with caution because it can create extremely large files.
+Instead, individual level data can be printed for only the final time step by setting `print_last = TRUE`.
+Output produced by `run_farm_sim` is a list of two elements, which includes a vector of parameter values used in the simulation and the final state of the landscape as an array.
 
@@
 
-- Example of evolving species with different crop and pesticide applications
-
-Conclusions
-===================================================
+Here we demonstrate a simple simulation of resevol with asexually reproducing pests that have three loci and two traits (Figure \@ref(fig:network)).
 

### 2022-07-25 notebook/ms/ms.Rmd
Update manuscript with a section on an example population

@@
     toc: no
+  pdf_document:
+    toc: no
+  html_document: default
   word_document: default
-  html_document: default
 header-includes:
@@
 
-Introduction
+Introduction {#introduction}
 ===================================================
 
-- Pest resistance to pesticide is a serious wicked problem that affects food security.
-- A quantitative genetic approach to pesticide resistance to maintaining resistance.
-- The resevol R package addresses two problems: genetic architecture and landscape selection
-- The R package can facilitate general questions and targetted ones
+- Pest resistance to pesticide is a serious wicked problem that affects food security (I have some references for this, but am probably missing several).
+- A quantitative genetic approach to pesticide resistance to maintaining resistance (has this been done at all?).
 
@@
 
-Covarying pest quantitative traits
+Covarying pest quantitative traits {#mine_gmatrix}
 ===================================================
@@
 
-Simulating landscape-level pesticide resistance
+Simulating landscape-level pesticide resistance {#simulating}
 ===================================================
@@
 
-Landscape
+Landscape {#landscape}
 ---------------------------------------------------
@@
 
-Pest ecology
+Pest ecology {#ecology}
 ---------------------------------------------------
@@
 
-Pest evolution
+Pest evolution {#evolution}
 ---------------------------------------------------
@@
 
-Simulation output
+Simulation output {#output}
 ===================================================
 
-Simulation output can be large, so output is printed in two CSV files, both of which are created in the working directory.
+Simulation output is large, so output is printed in two CSV files, both of which are created in the working directory.
 The first file "population_data.csv" prints population level data over time, including population size, mean age, sex ratio, mean food and pesticide consumed of each type, mortality rate, and mean trait values.
@@
 
-Example of individual-based simulations
+Example of individual-based simulations {#simulation}
 ===================================================
 
-Here we demonstrate a simple simulation of resevol with asexually reproducing pests that have three loci and two traits (Figure \@ref(fig:network)).
+In resevol v0.3, the `run_farm_sim` function includes 68 arguments, which specify a wide range of possible simulation conditions affecting landscape and pest characteristics.
+These arguments are explained the package documentation, and in supporting information that demonstrates an advanced case study with a custom landscape and complex pest genomes and life history.
+Here we focus on a simple simulation with asexually reproducing pests that have three loci and two traits (Figure \@ref(fig:network)).
+We use the the genome generated in `new_network` from [Section 2](#mine_gmatrix), in which traits 1 and 2 have variances of `r new_network[[6]][1,1]` and `r new_network[[6]][2,2]`, respectively, with a covariance of `r new_network[[6]][1,2]`.
 
-Availability
+Traits 1 and 2 will define the rate at which pests consume two separate pesticides, so we model a system in which there is a trade-off for pesticide susceptibility.
+We use a simple $64 \times 64$ cell landscape with nine farms. 
+Each farm grows the same crop and uses one of two randomly intialised pesticides, which are rotated every 16 time steps.
+Hence, we can conceptualise 16 time steps as a single growing season.
+
+In our example, each cell holds four crop units, all of which can potentially be consumed. 
+Pests consume up to one unit of crop on their landscape cell per time step. 
+Pest survival and reproduction is food-based, and pests need to consume one unit of crop by age two to survive and reproduce. 
+Pests initialised at the start of the simulation are randomly assigned an age from 0-4 with equal probabilities.
+Pests have a maximum age of four, and they feed in ages 0-2, move up to two cells in age three and four, and reproduce in age four.
+Pests can consume pesticides in ages 0-2, and if they consume any pesticide, then they will die.
+In each time step, 10 immigrants arrive.
+We simulate 160 time steps (10 growing seasons) using the function below.
+
+```{r, eval = TRUE}
+sim <- run_farm_sim(mine_output = new_network, repro = "asexual", 
+                    pesticide_number = 2, pesticide_init = "random", 
+                    pesticide_consume = c("T1", "T2"), farms = 9,
+                    pesticide_rotation_time = 16, pesticide_rotation_type = 3,
+                    pesticide_tolerated_surv = 0, pesticide_per_cell = 1,
+                    crop_rotation_time = 16, crop_number = 1, crop_per_cell = 4, 
+                    food_consume = 1, reproduction_type = "food_based", 
+                    food_needed_surv = 1, food_needed_repr = 1, max_age = 4, 
+                    min_age_feed = 0, max_age_feed = 2, min_age_move = 3, 
+                    max_age_move = 4, min_age_reproduce = 4, print_gens = FALSE,
+                    max_age_reproduce = 4, age_pesticide_threshold = 2, 
+                    rand_age = TRUE, move_distance = 2, immigration_rate = 10, 
+                    time_steps = 160, print_last = TRUE, xdim = 64, ydim = 64);
+```
+
+Any arguments to `run_farm_sim` not included above are set to default values.
+[Output files](#outputs) can be used to plot ecological and evolutionary dynamics of pests. 
+Figure \@ref(fig:overtime)a shows the pest population change both over and within seasons.
+Due to the specific parameter values chosen, clear patterns reflecting pest cohort emerge.
+At the start of a season when the most crop is available, pests eat and abundance increases.
+As less crop becomes available in a season, pest abundance decreases.
+Mean food consumed varies over the course of a season caused by the varying frequency of pests in different life history stages (Figure \@ref(fig:overtime)c).
+In the first 50 time steps, there is some consumption of pesticides 1 and 2 (Figure \@ref(fig:overtime)b), but both traits rapidly evolve to negative values making both pesticides ineffective (Figure \@ref(fig:overtime)d).
+In this case, the trade-off between pesticide consumption is not strong enough to maintain susceptibility to either pesticide in the population.
+
+```{r overtime, echo=FALSE, fig.width=8, fig.cap = "Agricultural pest ecological and evolutionary dynamics over 160 time steps from an individual-based simulation using the resevol R package.
+Panels show (a) pesticide abundance change, (b) mean of pesticides 1 and 2 consumed per pest, (c) mean food consumed per pest, and (d) mean values of evolving traits 1 and 2 over time.
+Traits 1 and 2 define how much of pesticides 1 and 2, respectively, a pest will consume.
+White and grey vertical stripes indicate seasons of a single crop and pesticide application."}
+mbox <- function(x0, x1, y0, y1){
+    xx <- seq(from=x0, to=x1, length.out = 100);
+    yy <- seq(from=y0, to=y1, length.out = 100);
+    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
+    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
+    return(list(x=xd, y=yd));
+}
+population_data <- read.csv("population_data.csv");
+row_number      <- 161; # dim(population_data)[1];
+time_step       <- population_data[["time_step"]][2:row_number];
+population_size <- population_data[["population_size"]][2:row_number];
+food_consumed   <- population_data[["mean_food_consumed"]][2:row_number];
+p1_consumed     <- population_data[["mean_pesticide1_consumed"]][2:row_number];
+p2_consumed     <- population_data[["mean_pesticide2_consumed"]][2:row_number];
+trait1          <- population_data[["trait1_mean_value"]][2:row_number];
+trait2          <- population_data[["trait2_mean_value"]][2:row_number];
+season          <- seq(from = 0, to = 160, by = 16);
+blocks          <- length(season) - 1;
+par(mfrow = c(2, 2), mar = c(1, 5, 3, 1));
+# Pest abundance plot
+plot(x = time_step, y = population_size, type = "l", lwd = 2, xaxt = "n",
+     xlab = "", ylab = "Pest abundance", cex.lab = 1, ylim = c(0, 4600),
+     cex.axis = 1);
+for(i in 1:blocks){
+    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
+    if(i %% 2 == 0){
+        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
+                col = "grey90");
+    }
+}
+text(x = 5, y = 4300, labels = "a", cex = 2);
+points(x = time_step, y = population_size, type = "l", lwd = 2);
+box();
+# Pesticide consumed
+plot(x = time_step, y = p1_consumed, type = "n", lwd = 2, xlab = "", xaxt = "n",
+     ylab = "Mean pesticide consumed", cex.lab = 1, ylim = c(0, 0.35),
+     cex.axis = 1);
+for(i in 1:blocks){
+    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1, y1 = 3);
+    if(i %% 2 == 0){
+        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
+                col = "grey90");
+    }
+}
+points(x = time_step, y = p1_consumed, type = "l", lwd = 2);
+points(x = time_step, y = p2_consumed, type = "l", lwd = 2, col = "red",
+       lty = "dashed");
+legend(x = 158, y = 0.35, col = c("black", "red"), lty = c("solid", "dashed"),
+       legend = c("Pesticide 1", "Pesticide 2"), cex = 1, lwd = 2, bg = "white",
+       xjust = 1);
+text(x = 5, y = 0.32, labels = "b", cex = 2);
+box();
+par(mar = c(4, 5, 0, 1));
+# Pest food consumption plot
+plot(x = time_step, y = food_consumed, type = "n", lwd = 2, xlab = "Time step",
+     ylab = "Mean food consumed", cex.lab = 1, ylim = c(0, 2),
+     yaxt = "n");
+axis(side = 2, at = c(0, 0.5, 1, 1.5), cex.axis = 1);
+for(i in 1:blocks){
+    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1, y1 = 3);
+    if(i %% 2 == 0){
+      polygon(x = rbox$x, y = rbox$y, lwd = 2, border = NA, 
+              col = "grey90");
+    }
+}
+points(x = time_step, y = food_consumed, type = "l", lwd = 2);
+text(x = 5, y = 0.2, labels = "c", cex = 2);
+box();
+# Pesticide traits
+plot(x = time_step, y = trait1, type = "n", lwd = 2, xlab = "Time step",
+     ylab = "Mean trait value", cex.lab = 1, ylim = c(-0.9, 0),
+     yaxt = "n");
+axis(side = 2, at = c(-0.8, -0.6, -0.4, -0.2), cex.axis = 1);
+for(i in 1:blocks){
+    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1, y1 = 3);
+    if(i %% 2 == 0){
+        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
+                col = "grey90");
+    }
+}
+points(x = time_step, y = trait1, type = "l", lwd = 2);
+points(x = time_step, y = trait2, type = "l", lwd = 2, col = "red",
+       lty = "dashed");
+legend(x = 158, y = 0, col = c("black", "red"), lty = c("solid", "dashed"),
+       legend = c("Trait 1", "Trait 2"), cex = 1, lwd = 2, xjust = 1, 
+       bg = "white");
+text(x = 5, y = -0.82, labels = "d", cex = 2);
+box();
+```
+
+Figure 4 shows the location of pests on the landscape
+
+```{r landscape, echo=FALSE, fig.width=6, fig.height=6, fig.cap = "Last time step"}
+last_time_step <- read.csv("last_time_step.csv", header = FALSE);
+landscape      <- sim[[2]][,,1];
+landscape_vals <- unique(as.vector(landscape));
+reord_land     <- sample(landscape_vals, size = length(landscape_vals));
+for(i in 1:dim(landscape)[1]){
+    for(j in 1:dim(landscape)[2]){
+        old_val         <- landscape[i, j];
+        new_val         <- reord_land[old_val];
+        landscape[i, j] <- new_val;
+    }
+}
+for(i in 1:dim(last_time_step)[1]){
+    xloc <- last_time_step[i, 3] + 1;
+    yloc <- last_time_step[i, 4] + 1;
+    landscape[xloc, yloc] <- 10;
+}
+land_cols      <- c(hcl.colors(9, "YlOrRd", rev = TRUE), "#000000");
+image(landscape, xaxt = "n", yaxt = "n", col = land_cols);
+```
+
+Availability {#availability}
 ===================================================
@@
 
-Conclusions
+Conclusions {#conclusions}
 ===================================================
 
-Acknowledgements
+Acknowledgements {#acknowledgements}
 ===================================================
@@
 
-References
+References {#references}
 ===================================================

### 2022-07-25 notebook/ms/ms.Rmd
Finish the example section, move on to the conclusion then introduction

@@
 
-An evolutionary algorithm is used to find appropriate values that produce covarying traits from loci (see Supporting Information for details). 
+An evolutionary algorithm is used to find appropriate values that produce covarying traits from loci (see Supporting Information 1 for details). 
 Evolutionary algorithms are heuristic tools can simulate adaptive evolution to find solutions for a broad range of problems [@Hamblin2013; @Duthie2018]. 
@@
 Similarly, higher values of `indivs` will result in more accurate estimations of true stress, but this also requires more computation time.
-Additional arguments to `mine_gmatrix` can also be used to improve the performance of the evolutionary algorithm (see Supporting Information).
+Additional arguments to `mine_gmatrix` can also be used to improve the performance of the evolutionary algorithm (see Supporting Information 1).
 
@@
 In resevol v0.3, the `run_farm_sim` function includes 68 arguments, which specify a wide range of possible simulation conditions affecting landscape and pest characteristics.
-These arguments are explained the package documentation, and in supporting information that demonstrates an advanced case study with a custom landscape and complex pest genomes and life history.
+These arguments are explained the package documentation, and Supporting Information 2, which demonstrates an advanced case study with a custom landscape and complex pest genomes and life history.
 Here we focus on a simple simulation with asexually reproducing pests that have three loci and two traits (Figure \@ref(fig:network)).
@@
 Any arguments to `run_farm_sim` not included above are set to default values.
-[Output files](#outputs) can be used to plot ecological and evolutionary dynamics of pests. 
+[Output files](#output) can be used to plot ecological and evolutionary dynamics of pests. 
 Figure \@ref(fig:overtime)a shows the pest population change both over and within seasons.
@@
 In this case, the trade-off between pesticide consumption is not strong enough to maintain susceptibility to either pesticide in the population.
+Figure \@ref(fig:landscape) shows the location of pests on the landscape in the last time of the simulation (left panel), and how different pesticides are being applied on the landscape (right panel).
+Because the number of crops and pesticides, the number and nature of traits, and the size of pest genomes all vary depending on simulation parameters, no plotting functions are introduced in the resevol package.
+Instead, methods and code for producing plots such as those in Figures \@ref(fig:overtime) and \@ref(fig:landscape) are explained in Supporting Information 2.
 
@@
 
-Figure 4 shows the location of pests on the landscape
-
-```{r landscape, echo=FALSE, fig.width=6, fig.height=6, fig.cap = "Last time step"}
+```{r landscape, echo=FALSE, fig.width=6.4, fig.height=3, fig.cap = "Locations of pests (black) across a landscape that includes nine farms (coloured blocks) in the last time step of a simulation using the resevol R package (left panel).
+The right panel shows which farms apply pesticide 1 (dark grey) and 2 (light grey)."}
 last_time_step <- read.csv("last_time_step.csv", header = FALSE);
@@
 land_cols      <- c(hcl.colors(9, "YlOrRd", rev = TRUE), "#000000");
+par(mfrow = c(1, 2), mar = c(0, 0.1, 0, 0.1));
 image(landscape, xaxt = "n", yaxt = "n", col = land_cols);
+box();
+pesticide_position <- sim[[2]][,,12];
+pesticide_cols     <- c("grey20", "grey60");
+image(pesticide_position, xaxt = "n", yaxt = "n", col = pesticide_cols);
+box();
 ```

### 2022-07-26 notebook/ms/ms.Rmd
last draft before sending off to coauthors

@@
 output:
+  word_document: default
+  pdf_document:
+    toc: no
   bookdown::pdf_document2:
@@
     toc: no
-  pdf_document:
-    toc: no
   html_document: default
-  word_document: default
 header-includes:
@@
 
-1. Pesticide resistance is a serious problem.
-2. We introduce the resevol R package, which runs individual-based models of pests with quantitative and covarying traits and three mating systems.
-3. Simulations are on a landscape where crops and pesticides are changed.
-4. We give an example of simulating pests with covarying traits
-5. The resevol R package is open source under GNU Public License; source code and documents are freely available on GitHub.
+1. The evolution of pesticide resistance is a widespread problem with potentially severe consequences for global food security.
+2.
+We introduce the resevol R package, which simulates individual-based models of pests with evolving genomes that produce complex, polygenic, and covarying traits affecting pest life history and pesticide resistance.
+3.
+Simulations are modelled on a spatially-explicit and highly customisable landscape in which crop and pesticide application and rotation can vary, making the package a highly flexible tool for both general and targetted models of pest management and resistance evolution.
+4. We present the key features of the resevol package and demonstrate its use for a simple example simulating pests with two covarying traits.
+5. The resevol R package is open source under GNU Public License. All source code and documentation are available on GitHub.
 
-**Key words:** *pest management*, *food security*, *ecological modelling*
+**Key words:** *pest management*, *food security*, *ecological modelling*, *pesticide*, *resistance evolution*
 
@@
 ===================================================
-
-- Pest resistance to pesticide is a serious wicked problem that affects food security (I have some references for this, but am probably missing several).
-- A quantitative genetic approach to pesticide resistance to maintaining resistance (has this been done at all?).
 
@@
 
-The resevol package uses individual-based modelling and a quantitative genetics approach to simulate the evolution of a pest population on a changing landscape. 
+Insect resistance to pesticides is a wicked and widespread problem [@Gould2018]. 
+Predicting, identifying, and ultimately delaying the evolution of insecticide resistance is therefore critical for ensuring sustainable global food security [@Hawkins2019; @Lykogianni2021]. 
+To manage pesticide resistance, various strategies have been proposed to disrupt or weaken selection.
+These strategies often focus on varying the location and timing of pesticide application, and therefore varying the strength of selection for pesticide resistance, so that pest susceptibility to one or more pesticides is maintained [@Ives2002; @Liang2013; @Carriere2016; @Liang2018; @South2018; @Sudo2018; @Saikai2021].
+Pesticide resistance management has mostly been associated with the effect of single genes. 
+In such cases, resistance alleles have binary effects on phenotype, enabling resistant phenotypes to arise from genetic changes at single or a small number of loci [@Raymond2001; @Ffrench-Constant2004; @Ffrench-Constant2013].
+Nevertheless, resistance attributable to polygenic effects is also well-established [@Via1986; @Firko1990; @Busi2013; @Haridas2018; @Green2020].
+And the relevance of polygenic resistance is likely to further increase given a rising interest in biological controls for sustainable crop protection [e.g., @Green2020; @Elnahal2022]. 
+Developing strategies to maximise the efficacy of these tools is critical, and such strategies should be well-informed by predictions made from detailed, quantitative genetic models.
+Here we introduce the resevol R package as a tool for building such models and simulating pest management.
+
+The resevol package applies individual-based modelling and a quantitative genetics approach to simulate the evolution of a pest population on a changing landscape. 
 A focal goal of the software is to model traits with a pre-specified, but potentially evolving, covariance structure. 
 To achieve this goal, each individual has a genome with $L$ loci that underlie a set of $T$ potentially evolving traits. 
-Pleiotropic loci can vary in their effects on the direction and magnitude of polygenic trait, causing population-wide trait covariance to arise mechanistically from the underlying genetic architecture of individuals.
+Pleiotropic loci can vary in their effects on the direction and magnitude of polygenic traits, causing population-wide trait covariance to arise mechanistically from the underlying genetic architecture of individuals.
 To achieve this, two separate steps are necessary.
@@
 Second, a population of asexual or sexual individuals is initialised and simulated on a spatially explicit landscape separated into distinct units (e.g., farms).
-Land units can apply one of up to 10 pesticides, and one of up to 10 crops; pesticides and crops rotate independently in a pre-specified way over time.
-The resevol package can thereby model complex and evolving traits in agricultural pests over realistic landscapes that undergo different pesticide use and crop regimes.
+Land units can apply one of up to 10 pesticides, and one of up to 10 crops at a given time.
+Pesticides and crops rotate independently within farms over time in a pre-specified way.
+The resevol package can thereby model complex and evolving agricultural pest traits over realistic landscapes that undergo different pesticide use and crop regimes.
 
@@
 
-The first step of simulation with the resevol package is building individual genomes. 
-This step is separate because building genomes is often time-consuming, and genomes that are built might need to be saved or inspected before actual simulation.
+The first step of simulation is building individual genomes. 
+This step is separate because it is computationally intensive, and genomes that are built might need to be inspected and stored.
 High computation time is due to the mechanistic nature of how genomes and covarying traits are modelled.
-Instead of imposing a trait covariance structure directly, the objective is to use an evolutionary algorithm to find a network that maps standard random normal values (loci) to covarying values (traits).
+Instead of imposing a trait covariance structure directly, an evolutionary algorithm is used to find a network that maps standard random normal values (loci) to covarying values (traits).
 This is useful because it allows genomes to model potentially evolving physiological constraints, and trade-offs among traits, from the bottom up.
-Since multiple networks can potentially map loci to the same trait covariance structure, it is possible to replicate evolution with differently randomised genetic architectures.
+Since multiple networks can potentially map loci to the same trait covariance structure, it is possible to replicate evolution with different randomly generated genetic architectures.
 This approach to modelling individual genomes and traits thereby increases the complexity of questions that can be addressed for simulating evolution in agricultural pests.
 
-```{r network, echo = FALSE, fig.width = 6, fig.height = 4, fig.cap = "Example network mapping loci (green circles) to traits (red diamonds) through an intermediate set of hidden layers (blue squares) in the mine_gmatrix function.
-Individual genomes in the resevol R package consist of standard random normal values for at loci, real values for black arrows linking nodes, and real values of traits.
+```{r network, echo = FALSE, fig.width = 6, fig.height = 4, fig.cap = "Example network mapping loci (green circles) to traits (red diamonds) through an intermediate set of hidden layers (blue squares) in the `mine_gmatrix` function.
+Individual genomes in the resevol R package consist of standard random normal values for loci, real values for black arrows linking nodes, and real values for traits.
 Values shown for loci and arrows are an example for illustration."}
@@
 Conceptually, the relationship between individual genotypes and traits is defined by a network connecting loci to traits through a set of hidden internal nodes (Figure \@ref(fig:network)). 
-Values for loci are randomly drawn from a standard normal distribution $\mathcal{N}\left(0, 1\right)$. 
+Values for loci are randomly drawn from a standard normal distribution, $\mathcal{N}\left(0, 1\right)$. 
 Links between loci, internal nodes, and traits, can take any real value and are represented by black arrows in Figure \@ref(fig:network). 
 Traits can take any real value, and are calculated as the summed effects of all preceding nodes (i.e., the blue squares immediately to the left of the traits in Figure \@ref(fig:network)).
-Mathematically, loci are represented by row a vector of length $L$.
+Mathematically, loci are represented by a row vector of length $L$.
 Effects of loci on the first layer of internal nodes (black arrows emanating from loci in Figure \@ref(fig:network)) are represented by an $L \times T$ matrix, and transitions between internal nodes, and between the last set of internal nodes and the final traits, are represented by $T \times T$ matrices.
@@
 
-Values mapping loci to traits become part of an individual's genome, so the genome for the individual represented by Figure \@ref(fig:network) is stored in the model as below:
+Values mapping loci to traits become part of an individual's genome, so the genome for the individual represented by Figure \@ref(fig:network) is stored in the model as shown below:
 
@@
 Individuals with different loci can therefore have different covarying traits that are constrained by the network structure encoded in each genome. 
-Individuals can be haploid (as in Figure \@ref(fig:network)) or diploid (in which case, allele values are summed at a locus).
+Individuals can be haploid (as in Figure \@ref(fig:network)) or diploid (in which case, allele values are summed at homologous loci).
 
 An evolutionary algorithm is used to find appropriate values that produce covarying traits from loci (see Supporting Information 1 for details). 
-Evolutionary algorithms are heuristic tools can simulate adaptive evolution to find solutions for a broad range of problems [@Hamblin2013; @Duthie2018]. 
-In the resevol package, the `mine_gmatrix` function runs an evolutionary algorithm, and requires the argument `gmatrix`, which specifies the desired trait covariance matrix.
-The function runs a genetic algorithm that initialises a population of `npsize` networks (Figure \@ref(fig:network)), and this population evolves until some maximum generation number (`max_gen`) or minimum expected network stress (`term_cri`) is met.
-In a single generation of the algorithm, network values crossover and mutate.
+Evolutionary algorithms are heuristic tools that can simulate adaptive evolution to find solutions for a broad range of problems [@Hamblin2013; @Duthie2018]. 
+In the resevol package, the `mine_gmatrix` function runs an evolutionary algorithm and requires the argument `gmatrix`, which specifies the desired trait covariance matrix.
+The function initialises a population of `npsize` networks (Figure \@ref(fig:network)), and this population evolves until some maximum generation (`max_gen`) or minimum expected network stress (`term_cri`) is met.
+In a single generation of the algorithm, values crossover between networks and mutate.
 Next, trait covariances produced for each network are estimated by initalising `indivs` individuals with loci sampled from a standard normal distribution.
@@
 Networks with the lowest estimated stress have the highest fitness, so these networks are disproportionately represented in the next generation.
-Throughout the evolutionary algorithm, the lowest stress network is saved and output upon network termination.
+Throughout the evolutionary algorithm, the lowest stress network is saved and returned upon network termination.
 The robustness of this network's stress to sets of individuals with different loci values can be tested using the `stress_test` function.
 
-An example run of `mine_gmatrix` with the same number of loci and internal nodes is shown below with traits that do not covary:
+An example run of `mine_gmatrix` with the same number of loci and internal nodes as in Figure \@ref(fig:network) is shown below for traits that have an intended covariance of -0.4:
 
@@
 
-The full output of `mine_gmatrix` is passed to the `run_farm_sim` function, which initialises and simulates an evolving population of pests on a changing landscape for a natural number of time steps (`time_steps`).
+The full output of `mine_gmatrix` is passed to the `run_farm_sim` function, which initialises and simulates an evolving population of pests on a changing landscape for any natural number of time steps (`time_steps`).
 In this section, we explain the landscape, pest ecology, and evolving pest traits.
@@
 
-Landscapes are spatially explicit and initialised in `run_farm_sim` in one of two ways. 
-First, a landscapes can be built from the arguments `xdim`, `ydim`, and `farms`. 
+Landscapes are spatially explicit and initialised in one of two ways. 
+First, a landscape can be built from the arguments `xdim`, `ydim`, and `farms`. 
 These arguments specify the dimensions of the landscape and the number of farms on it.
 Contiguous rectangular farms of roughly equal size are generated on the landscape using a splitline algorithm.
-Second, a custom landscapes can be input using the `terrain` argument, which takes a matrix with elements that includes integers 1 to `farms`.
+Second, a custom landscape can be input using the `terrain` argument, which takes a matrix with elements that include integers 1 to `farms`.
 Each value defines a unique farm, but values do not need to be contiguous.
-These 'farms' could even model non-farmland (e.g., water, roads), if pesticides and crops on them are invisible to pests (see below).
+These 'farms' could even model non-farmland (e.g., water, roads), if pesticides and crops on them are invisible to pests (see Supporting Information 2).
 This `terrain` customisation therefore allows for a high degree of landscape detail, and offers the potential for modelling real-world landscapes from raster images.
@@
 
-Each farm can hold one pesticide and one crop type at any time step.
+Each farm can hold one pesticide and one crop type in any time step.
 The `crop_init` and `pesticide_init` arguments initialise one of `crop_number` crops and one of `pesticide_number` pesticides for each farm, respectively (maximum of 10 each).
-Initialisation can be random for each farm with equal probability, or can be set using a vector of length `farms` in which vector elements define the initialised crop or pesticide number.
+Initialisation can be random for each farm with equal probability, or it can be set using a vector of length `farms` in which vector elements define the initialised crop or pesticide number.
 After initialisation, crops and pesticides rotate once ever `crop_rotation_time` and `pesticide_rotation_time` time steps, respectively.
 Each of these rotation time arguments can take either an integer value from 1-3, or a matrix.
-Integer values specify (1) no rotation, (2) random transition from one type to another, or (3) cycling through each available crop or pesticide in order.
+Integer values specify (1) no rotation, (2) random transition from one type to another, or (3) cycling through each available crop or pesticide in numeric order.
 Square matrices can be used to define the probability that a given crop or pesticide in row $i$ transitions to that in column $j$. 
-Hence, any possible Markov chain can be used to transition between crop or pesticide types on farms.
+Hence, any possible Markov chain can be used to model transition between crop or pesticide types on farms.
 
@@
 Individual pests can be modelled to have several reproductive systems and life histories.
-Pest reproductive system can be specified in the `run_farm_sim` function using the `repro` argument, which accepts arguments `"asexual"` (haploid), `"sexual"` (monoecious), and `"biparental"` (dioecious).
+Pest reproductive system can be specified using the `repro` argument, which accepts `"asexual"` (haploid), `"sexual"` (monoecious), and `"biparental"` (dioecious).
 For `"sexual"` pests, the `selfing` argument specifies if self-fertilisation is (`TRUE`) or is not (`FALSE`) allowed.
@@
 Crops and pesticides are also not necessarily rotated in each time step (see Landscape).
-Statistics collected within a time step are printed in a CSV file."}
+Statistics collected within a time step are printed to a CSV file."}
 mbox <- function(x0, x1, y0, y1){
@@
 In each time step, pests feed in a random order, consuming the crop on their landscape cell.
-Pests consume a maximum amount of the crop as specified by the `food_consume` argument, which takes a vector with as many elements as their are crops (i.e., if `crop_number = 2`, then `food_consume` has two elements, the first and second defining consumption of crops 1 and 2, respectively).
+Pests consume an amount of the crop as specified by the `food_consume` argument, which takes a vector with as many elements as their are crops (i.e., if `crop_number = 2`, then `food_consume` has two elements, the first and second defining consumption of crops 1 and 2, respectively).
 If crop amount on the landscape cell exceeds pest consumption ability, then pests consume their maximum amount, and this amount is removed from the landscape cell.
@@
 
-After all pests are done moving, pests reproduce.
+After all pests finish moving, pests reproduce.
 Offspring production is possible for asexual, monoecious, or female pests.
@@
 All available potential mates sire offspring with equal probability, and reproducing pests are assumed to mate multiply (i.e., paternity is a fair raffle for all offspring).
-If a carrying capacity on birth is set (`K_on_birth > 0`) and total offspring number in the population exceeds this capacity, then offspring are removed at random until they are  within carrying capacity.
-An real value `immigration_rate` specifies the rate parameter for Poisson random sampling of the number of immigrants added to the population at the end of a time step.
-Immigrants are initialised in the same way as pests were at the start of the simulation.
+If a carrying capacity at birth is set (`K_on_birth > 0`) and total offspring number in the population exceeds this capacity, then offspring are removed at random until they are  within carrying capacity.
+A real value `immigration_rate` specifies the rate parameter for Poisson random sampling of the number of immigrants added to the population at the end of a time step.
+Immigrants are initialised in the same way as pests are at the start of the simulation.
 Hence, any spatial structure or evolution that occurs during the simulation does not affect immigrant locations, genomes, or traits.
@@
 Asexually reproducing pests receive the full genomes of their parent, while sexually reproducing pests receive half of their alleles from each parent.
-Each diploid parent contributes one half of their genome, effectively modelling a genome with a single paired chromosome.
+Each diploid parent contributes one half of their genome, effectively modelling a genome with a single chromosome.
 Crossover occurs at each position of the genome with a probability of `crossover_pr`, so complete recombination is also possible if `crossover_pr = 0.5`.
-For both haploids and diploids, each genome value then mutates independently with a probability of `mutation_pr`, which can be set to any real number between 0 and 1.
+For both haploids and diploids, each genome value then mutates independently with a probability of `mutation_pr`, which can be set to any real number from 0-1.
 If a genome value mutates, then a new value is randomly sampled from a standard normal distribution.
@@
 
-Evolution of the genetic architecture linking loci to traits can be constrained by disabling mutation genome values linking loci, internal nodes, and traits (i.e., 'network values' represented by arrows in Figure \@ref(fig:network)).
+Evolution of the genetic architecture linking loci to traits can be constrained by disabling mutation in genome values linking loci, internal nodes, and traits (i.e., 'network values' represented by arrows in Figure \@ref(fig:network)).
 While mutation at loci (green circles in Figure \@ref(fig:network)) is always possible as long as `mutation_pr > 0`, the number of intermediary layers for which network values can mutate is constrained by `net_mu_layers`.
@@
 Finally, evolving and covarying traits can be used in place of fixed parameters described in pest ecology.
-This is done by substituting `"Tj"` as argument input in place of a numeric value, where `j` represents the trait number.
+This is done by substituting `"Tj"` as an argument input in place of a numeric value, where `j` represents the trait number.
 For example, the argument `move_distance = "T1"` will make Trait 1 the movement distance for individuals. 
 The argument `food_consume = c("T2", "T3")` will set Traits 2 and 3 to define the amount of food of types 1 and 2 that can be consumed by a pest, respectively.
-Parameters that can be replaced include `move_distance`, `food_needed_surv`, `pesticide_tolerated_surv`, `food_needed_repr`, `pesticide_tolerated_repr`, `mating_distance`, `lambda_value`, `movement_bouts`, `metabolism`, `food_consume`, and `pesticide_consume`.
+Parameters that can be replaced with evolving traits include `move_distance`, `food_needed_surv`, `pesticide_tolerated_surv`, `food_needed_repr`, `pesticide_tolerated_repr`, `mating_distance`, `lambda_value`, `movement_bouts`, `metabolism`, `food_consume`, and `pesticide_consume`.
 The resevol package can thereby simulate agricultural pests with complex and co-evolving traits, and potentially evolving trait covariances, under a range of possible pest life histories.
@@
 
-Simulation output is large, so output is printed in two CSV files, both of which are created in the working directory.
+Simulation output is typically large, so output is printed in two CSV files, both of which are created in the working directory.
 The first file "population_data.csv" prints population level data over time, including population size, mean age, sex ratio, mean food and pesticide consumed of each type, mortality rate, and mean trait values.
@@
 In resevol v0.3, the `run_farm_sim` function includes 68 arguments, which specify a wide range of possible simulation conditions affecting landscape and pest characteristics.
-These arguments are explained the package documentation, and Supporting Information 2, which demonstrates an advanced case study with a custom landscape and complex pest genomes and life history.
+These arguments are explained the package documentation, and in Supporting Information 2, which demonstrates an advanced case study with a custom landscape and complex pest genomes and life history.
 Here we focus on a simple simulation with asexually reproducing pests that have three loci and two traits (Figure \@ref(fig:network)).
@@
 
-Traits 1 and 2 will define the rate at which pests consume two separate pesticides, so we model a system in which there is a trade-off for pesticide susceptibility.
+Traits 1 and 2 will define the rate at which pests consume two separate pesticides, so we model a system in which there is a potential trade-off for pesticide susceptibility.
 We use a simple $64 \times 64$ cell landscape with nine farms. 
@@
 
-In our example, each cell holds four crop units, all of which can potentially be consumed. 
+In our example, each cell produces four crop units, all of which can potentially be consumed. 
 Pests consume up to one unit of crop on their landscape cell per time step. 
 Pest survival and reproduction is food-based, and pests need to consume one unit of crop by age two to survive and reproduce. 
-Pests initialised at the start of the simulation are randomly assigned an age from 0-4 with equal probabilities.
-Pests have a maximum age of four, and they feed in ages 0-2, move up to two cells in age three and four, and reproduce in age four.
+Pests initialised at the start of the simulation are randomly assigned an age from 0-4 with equal probability.
+Pests have a maximum age of four, and they feed in ages 0-2, move up to two cells in ages 3-4, and reproduce in age four.
 Pests can consume pesticides in ages 0-2, and if they consume any pesticide, then they will die.
-In each time step, 10 immigrants arrive.
-We simulate 160 time steps (10 growing seasons) using the function below.
+In each time step, a mean of 10 immigrants arrive.
+We simulate 160 time steps (10 growing seasons) using the function below:
 
@@
 [Output files](#output) can be used to plot ecological and evolutionary dynamics of pests. 
-Figure \@ref(fig:overtime)a shows the pest population change both over and within seasons.
-Due to the specific parameter values chosen, clear patterns reflecting pest cohort emerge.
+Figure \@ref(fig:overtime)a shows the pest population change both across and within seasons.
+Due to the specific parameter values chosen, clear patterns reflecting pest cohorts emerge.
 At the start of a season when the most crop is available, pests eat and abundance increases.
-As less crop becomes available in a season, pest abundance decreases.
+As less crop remains in a season, pest abundance decreases.
 Mean food consumed varies over the course of a season caused by the varying frequency of pests in different life history stages (Figure \@ref(fig:overtime)c).
 In the first 50 time steps, there is some consumption of pesticides 1 and 2 (Figure \@ref(fig:overtime)b), but both traits rapidly evolve to negative values making both pesticides ineffective (Figure \@ref(fig:overtime)d).
-In this case, the trade-off between pesticide consumption is not strong enough to maintain susceptibility to either pesticide in the population.
-Figure \@ref(fig:landscape) shows the location of pests on the landscape in the last time of the simulation (left panel), and how different pesticides are being applied on the landscape (right panel).
+In this case, the trade-off in pesticide consumption is not strong enough to maintain susceptibility to either pesticide in the population.
+Figure \@ref(fig:landscape) shows the location of pests on the landscape in the last time of the simulation (left panel), and how different pesticides are currently being applied on the landscape (right panel).
 Because the number of crops and pesticides, the number and nature of traits, and the size of pest genomes all vary depending on simulation parameters, no plotting functions are introduced in the resevol package.
@@
 
-Availability {#availability}
+Conclusions {#conclusions}
 ===================================================
 
-The resevol R package can be downloaded from CRAN (https://cran.r-project.org/package=resevol) or GitHub (https://bradduthie.github.io/resevol/). The package is open source under GNU Public License.
-
-Conclusions {#conclusions}
-===================================================
+Insecticide resistance to pesticides is a widespread problem that affects global food security [@Tabashnik2014; @Bradshaw2016; @Lykogianni2021]. 
+Many models have been developed to investigate the evolution of resistance under different ecological and evolutionary conditions [e.g., @Curtis1985; @Mani1985; @Ives2002; @Liang2013; @Liang2018; @Haridas2018;  @South2018; @Sudo2018; @Saikai2021].
+The resevol R package makes it possible to rapidly develop and simulate myriad individual-based models of resistance evolution.
+It also introduces a novel approach to modelling complex pest genetic architecture, using an evolutionary algorithm to generate haploid or diploid loci that map to pest traits with pre-specified covariances (Figure \@ref(fig:network); Supporting Information 1).
+Agricultural landscapes and pest life histories are highly customisable, allowing targetted models that can simulate specific real-world case studies (Supporting Information 2).
+The breadth of possible models that can be simulated with resevol also makes it a useful tool for developing theory on pest management, and more even generally on the evolution and ecology of individuals with complex traits on a changing landscape.
 
@@
 
+Authors contributions {#contributions}
+===================================================
+
+All authors contributed to planning software development and writing. 
+ABD wrote and maintains package code. 
+RM contributed code for data visualisation.
+
+Availability {#availability}
+===================================================
+
+The resevol R package can be downloaded from CRAN (https://cran.r-project.org/package=resevol) or GitHub (https://bradduthie.github.io/resevol/). 
+The package is open source under GNU Public License.
+
 References {#references}

### 2022-07-30 R/initialise.R
Add a new argument to the run_farm_sim function that allows users to customise the mean value of evolving traits. The argument trait_means accepts a vector of the same length as the traits, with the element position being the mean of the corresponding trait. If not specified, the default is that all traits have a mean of zero

@@ -47,6 +47,7 @@
 #'@param baseline_metabolism A fixed baseline rate added to 'metabolism'+
 #'@param min_age_metabolism The minimum age affected by metabolism
 #'@param max_age_metabolism The maximum age affected by metabolism
+#'@param trait_means The means of the custom pest traits
 #'@return A two-dimensional array of individuals for simulation
 #'@examples
 #'gmt       <- matrix(data = 0, nrow = 2, ncol = 2);
@@ -90,7 +91,8 @@ initialise_inds <- function(mine_output,
                             metabolism = 0,
                             baseline_metabolism = 0,
                             min_age_metabolism = 1,
-                            max_age_metabolism = 9){
+                            max_age_metabolism = 9,
+                            trait_means = NULL){
   
   food      <- rep(x = 0, times = 10);
   pesticide <- rep(x = 0, times = 10);
@@ -121,9 +123,9 @@ initialise_inds <- function(mine_output,
     stop("ERROR: Must specify 'repro' as asexual, sexual, or biparental.");
   }
   if(repro == "sexual" | repro == "biparental"){
-    inds <- build_sexual(mine_output, N, neutral_loci);
+    inds <- build_sexual(mine_output, N, neutral_loci, trait_means);
   }else{
-    inds <- build_asexual(mine_output, N, neutral_loci);
+    inds <- build_asexual(mine_output, N, neutral_loci, trait_means);
   }
   
   inds[, 1] <- 1:N; # Sample ID
@@ -149,71 +151,87 @@ initialise_inds <- function(mine_output,
     inds[, 29] <- 2;
     inds[, 30] <- neutral_loci;
   }
-  inds[, 6]  <-  move_distance; # Movement distance
-  inds[, 7]  <- -1; # Mother ID
-  inds[, 8]  <- -1; # Father ID
-  inds[, 9]  <- -1; # Mother row
-  inds[, 10] <- -1; # Father row
-  inds[, 11] <-  0; # Offspring produced
-  inds[, 12] <-  mine_output[[1]][1]; # loci;
-  inds[, 13] <-  dim(mine_output[[2]])[1]; # traits;
[…]

### 2022-07-30 src/parents.c

@@ -9,8 +9,9 @@
 void insert_diploid_traits(double **offspring, double *paras, int offspr){
     
     int i, k, vec_pos, row, col, layer, loci, traits, layers, net_vals;
-    int loci_col, trait_col, layer_col, trait_st, net_st, loci_st;
-    double L1, L2, **loc_layer, ***net, **net_sum, **loci_to_traits, **L, **T;
+    int loci_col, trait_col, layer_col, trait_st, net_st, loci_st, T_mean_pos;
+    double **loc_layer, ***net, **net_sum, **loci_to_traits, **L, **T;
+    double T_mean, L1, L2;
     
     loci_col   = (int) paras[11];  /* Column where the number of loci is held */
     trait_col  = (int) paras[12];  /* Column where the number of traits held  */
@@ -102,7 +103,9 @@ void insert_diploid_traits(double **offspring, double *paras, int offspr){
 
     vec_pos = trait_st; /* Now put the traits back to offspring row */
     for(i = 0; i < traits; i++){
-        offspring[offspr][vec_pos] = T[0][i];
+        T_mean_pos                 = (int) paras[90 + i];
+        T_mean                     = offspring[offspr][T_mean_pos];
+        offspring[offspr][vec_pos] = T[0][i] + T_mean;
         vec_pos++;
     }
 
@@ -148,9 +151,9 @@ void insert_diploid_traits(double **offspring, double *paras, int offspr){
  * ========================================================================== */
 void insert_haploid_traits(double **offspring, double *paras, int offspr){
     
-    int i, k, vec_pos, row, col, layer, loci, traits, layers;
+    int i, k, vec_pos, row, col, layer, loci, traits, layers, T_mean_pos;
     int loci_col, trait_col, layer_col, trait_st, net_st, loci_st;
-    double **loc_layer, ***net, **net_sum, **loci_to_traits, **L, **T;
+    double T_mean, **loc_layer, ***net, **net_sum, **loci_to_traits, **L, **T;
 
     loci_col   = (int) paras[11];  /* Column where the number of loci is held */
     trait_col  = (int) paras[12];  /* Column where the number of traits held  */
@@ -234,7 +237,9 @@ void insert_haploid_traits(double **offspring, double *paras, int offspr){
 
     vec_pos = trait_st; /* Now put the traits back to offspring row */
     for(i = 0; i < traits; i++){
-        offspring[offspr][vec_pos] = T[0][i];
[…]

### 2022-07-30 tests/testthat/test-initialise.R

@@ -73,7 +73,80 @@ test_that("Correctly initialises different individual types", {
 
 
 
-
+test_that("Correctly initialise mean traits of individuals", {
+    skip_on_cran();
+    set.seed(Sys.time());
+    gmt       <- matrix(data = 0, nrow = 2, ncol = 2);
+    diag(gmt) <- 1;
+    mg        <- mine_gmatrix(gmatrix = gmt, loci = 4, layers = 2, indivs = 100, 
+                              npsize = 100, max_gen = 4, prnt_out = FALSE);
+    inds1     <- initialise_inds(mine_output = mg, N = 40, xdim = 40, ydim = 40, 
+                                 repro = "asexual", neutral_loci = 10, 
+                                 max_age = 9, min_age_move = 0, 
+                                 max_age_move = 9, min_age_reproduce = 0, 
+                                 max_age_reproduce = 9, min_age_feed = 0, 
+                                 max_age_feed = 9, food_consume = 0.25, 
+                                 pesticide_consume = 0.1, rand_age = FALSE, 
+                                 move_distance = 1, food_needed_surv = 0.25, 
+                                 pesticide_tolerated_surv = 0.1, 
+                                 food_needed_repr = 0, 
+                                 pesticide_tolerated_repr = 0,
+                                 reproduction_type = "lambda", 
+                                 mating_distance = 1, lambda_value = 1,
+                                 movement_bouts = 1, selfing = TRUE,
+                                 feed_while_moving = FALSE, mortality_type = 0,
+                                 age_food_threshold = NA,
+                                 age_pesticide_threshold = NA, 
+                                 trait_means = c(10, 100));
+    inds2     <- initialise_inds(mine_output = mg, N = 40, xdim = 40, ydim = 40, 
+                                 repro = "sexual", neutral_loci = 10, 
+                                 max_age = 9, min_age_move = 0, 
+                                 max_age_move = 9, min_age_reproduce = 0, 
+                                 max_age_reproduce = 9, min_age_feed = 0, 
+                                 max_age_feed = 9, food_consume = 0.25, 
+                                 pesticide_consume = 0.1, rand_age = FALSE, 
+                                 move_distance = 1, food_needed_surv = 0.25, 
+                                 pesticide_tolerated_surv = 0.1, 
+                                 food_needed_repr = 0, 
+                                 pesticide_tolerated_repr = 0,
[…]

### 2022-07-30 notebook/index.Rmd
Notebook update on the new version 0.3.2.0 with a new argument to set initialised trait means

@@
 library("resevol");
+data(gmatrices);
 ```
@@
 
-   - **JUL** [19](#u-19-JUL-2022)
+   - **JUL** [19](#u-19-JUL-2022), [30](#u-30-JUL-2022)
    - **JUN** [13](#u-13-JUN-2022), [28](#u-28-JUN-2022)
@@
 ================================================================================
+
+> <a name="u-30-JUL-2022">Update: 30 JUL 2022</a>
+
+I don't know why I didn't think of this sooner, but there really should be a feature that allows users to set the initial mean values of traits.
+This is easy because trait means are of course independent of the covariance structure of the traits, so there just needs to be some specified increment to be added to each trait throughout the simulation.
+I have introduced this in a new version 0.3.2.0 with the `trait_means` argument, which simply takes a vector of the means for each customised trait.
+By default, values are set to zero (i.e., what they always were originally), but the argument can also take a vector of the same length as the number of traits, with each element position corresponding to the specific trait.
+That is, `trait_means[1]` is the mean value of `"T1"`, `trait_means[2]` is the mean value of `"T2"`, et cetera.
+An error is given if the user attempts to specify a `trait_means` vector that has a length different from the number of traits.
+Just to show this in action, I will use the resevol mine_gmatrix output `mg_v1`, which has the covariance structure below.
+
+```{r, echo = FALSE}
+print(mg_v1[[6]]);
+```
+
+We have four traits, which I will use to define consumption rate of crops 1 and 2 (`T1` and `T2`, respectively), and the consumption rates of pesticides 1 and 2 (`T3` and `T4`, respectively).
+I will set the trait means to T1 = 1, T2 = 2, T3 = 0.2, and T4 = 0.2 and run a simulation with the code below.
+
+```{r, eval = FALSE}
+sim       <- run_farm_sim(mine_output = mg_v1, N = 1000, xdim = 10, ydim = 10, 
+                          repro = "asexual", time_steps = 20, farms = 12,
+                          crop_number             = 2,
+                          crop_init               = "random",
+                          crop_rotation_type      = 2,
+                          pesticide_number        = 2,
+                          pesticide_init          = "random",
+                          pesticide_rotation_type = 2,
+                          food_consume            = c("T1", "T2"),
+                          pesticide_consume       = c("T3", "T4"),
+                          trait_means             = c(1, 2, 0.2, 0.2),
+                          food_needed_surv        = 1,
+                          food_needed_repr        = 3,
+                          print_last              = TRUE);
+```
+
+At the end of the simulation (after some evolution has occurred), we can look at the evolution of traits and final trait distributions to confirm that they are around what we expect.
+Here is the evolution of the different traits over time.
+
+![](images/2022_JUL_30/traits_over_time.png)
+
+Here are the distributions of each trait, individually.
+
+![](images/2022_JUL_30/trait_distributions.png)
+
+Finally, here is a pairs plot showing the trait correlations.
+
+![](images/2022_JUL_30/pairs.png)
+
+Overall, this is doing what we want it to.
+I have also written a new test into testthat for the package.
+New documentation for the `trait_means` argument has been written, and I will incorporate this option into the working manuscript.
 

### 2022-07-30 notebook/ms/ms.Rmd
Update the website with the manuscript information, which will serve as the main vignette of the package from now on. The next step is to update the second vignette and create a third (which will be supporting information)

@@
 output:
-  word_document: default
+  word_document:
+    reference_docx: docx_template.docx
   pdf_document:
     toc: no
+  html_document: default
   bookdown::pdf_document2:
@@
     toc: no
-  html_document: default
 header-includes:
@@
 The argument `food_consume = c("T2", "T3")` will set Traits 2 and 3 to define the amount of food of types 1 and 2 that can be consumed by a pest, respectively.
-Parameters that can be replaced with evolving traits include `move_distance`, `food_needed_surv`, `pesticide_tolerated_surv`, `food_needed_repr`, `pesticide_tolerated_repr`, `mating_distance`, `lambda_value`, `movement_bouts`, `metabolism`, `food_consume`, and `pesticide_consume`.
+Up to 10 of the following parameters can be replaced with evolving traits: `move_distance`, `food_needed_surv`, `pesticide_tolerated_surv`, `food_needed_repr`, `pesticide_tolerated_repr`, `mating_distance`, `lambda_value`, `movement_bouts`, `metabolism`, `food_consume`, and `pesticide_consume`.
+Mean values of traits can be set using the argument `trait_means`, which accepts a vector of the same length as the number of evolving traits such that indices correspond to trait numbers.
 The resevol package can thereby simulate agricultural pests with complex and co-evolving traits, and potentially evolving trait covariances, under a range of possible pest life histories.
@@
                     rand_age = TRUE, move_distance = 2, immigration_rate = 10, 
-                    time_steps = 160, print_last = TRUE, xdim = 64, ydim = 64);
+                    time_steps = 160, print_last = TRUE, xdim = 64, ydim = 64,
+                    trait_means = c(0.1, 0.1));
 ```
@@
 plot(x = time_step, y = p1_consumed, type = "n", lwd = 2, xlab = "", xaxt = "n",
-     ylab = "Mean pesticide consumed", cex.lab = 1, ylim = c(0, 0.35),
+     ylab = "Mean pesticide consumed", cex.lab = 1, ylim = c(0, 0.4),
      cex.axis = 1);
@@
        lty = "dashed");
-legend(x = 158, y = 0.35, col = c("black", "red"), lty = c("solid", "dashed"),
+legend(x = 158, y = 0.4, col = c("black", "red"), lty = c("solid", "dashed"),
        legend = c("Pesticide 1", "Pesticide 2"), cex = 1, lwd = 2, bg = "white",
        xjust = 1);
-text(x = 5, y = 0.32, labels = "b", cex = 2);
+text(x = 5, y = 0.37, labels = "b", cex = 2);
 box();
@@
 plot(x = time_step, y = trait1, type = "n", lwd = 2, xlab = "Time step",
-     ylab = "Mean trait value", cex.lab = 1, ylim = c(-0.9, 0),
+     ylab = "Mean trait value", cex.lab = 1, ylim = c(-0.9, 0.1),
      yaxt = "n");
-axis(side = 2, at = c(-0.8, -0.6, -0.4, -0.2), cex.axis = 1);
+axis(side = 2, at = c(-0.8, -0.6, -0.4, -0.2, 0.0), cex.axis = 1);
 for(i in 1:blocks){
@@
        lty = "dashed");
-legend(x = 158, y = 0, col = c("black", "red"), lty = c("solid", "dashed"),
+legend(x = 158, y = 0.1, col = c("black", "red"), lty = c("solid", "dashed"),
        legend = c("Trait 1", "Trait 2"), cex = 1, lwd = 2, xjust = 1, 

### 2022-08-01 R/sim_farming.R
Introduce three new edge effects that are possible in the package, leaky, sticky, and reflective edges

@@ -303,6 +303,13 @@
 #' simulation, "T1" and "T2", then trait_means should be a vector of length 2).
 #' Note that mean trait values may change as the population evolves, so the
 #' values in this vector define only the means of the initial population.
+#'@param land_edge This sets what happens at the edge of the landscape. Three 
+#' options are possible, including "torus", "leaky", and "reflect". A torus
+#' results in no edge, such that pests that leave one side of the landscape
+#' end up on the other side. A leaky edge causes pests to leave the landscape
+#' and the simulation entirely, so they are effectively no longer recorded. A 
+#' reflective edge causes pests to bounce at the edge back in the direction from
+#' which they came. The default and recommended edge is a torus.
 #'@return The output in the R console is a list with two elements; the first 
 #'element is a vector of parameter values used by the model, and the second 
 #'element is the landscape in the simulation. The most relevant output will be
@@ -395,7 +402,8 @@ run_farm_sim <- function(mine_output,
                          min_age_metabolism  = 1,
                          max_age_metabolism  = 9,
                          terrain             = NA,
-                         trait_means         = NULL){
+                         trait_means         = NULL,
+                         land_edge           = "torus"){
   
     if(is.na(terrain)[1] == FALSE){
         xdim  <- dim(terrain)[1];
@@ -585,7 +593,8 @@ run_farm_sim <- function(mine_output,
                              get_f_coef               = get_f_coef,
                              get_stats                = get_stats,
                              metabolism               = metabolism,
-                             farms                    = farms);
+                             farms                    = farms,
+                             land_edge                = land_edge);
   
     return(sim_results);
 }
@@ -634,9 +643,10 @@ sim_crops <- function(pests,
                       get_f_coef = FALSE,
                       get_stats  = TRUE,
                       metabolism = 0,
-                      farms = 4
+                      farms = 4,
+                      land_edge = "torus"
                       ){
-  
+    
   N    <- dim(pests)[1];
   W    <- dim(pests)[2];
   X    <- dim(land)[2];
@@ -671,6 +681,19 @@ sim_crops <- function(pests,
   immi <- immigration_rate;
   fcoe <- as.numeric(get_f_coef);
   sttt <- as.numeric(get_stats);
+  ledg <- 0;
+  if(land_edge == "leaky"){
+      ledg <- 1;
+  }
[…]

### 2022-08-01 src/movement.c

@@ -20,7 +20,28 @@ int edge_effect(int pos, int edge_1, int edge_2, int edge_type){
                     pos = pos + edge_2;   
                 }
                 break;
-            default:
+            case 1: /* Leaky edge */
+                if(pos >= edge_2){
+                    pos = -1;
+                }
+                if(pos < edge_1){
+                    pos = -1;
+                }
+            case 2: /* Reflective edge */
+                if(pos >= edge_2){
+                    pos = edge_2 - (pos - edge_2) - 1;
+                }
+                if(pos < edge_1){
+                    pos = edge_1 - (pos - edge_1);
+                }
+            case 3: /* Sticky edge */
+                if(pos >= edge_2){
+                    pos = edge_2 - 1;
+                }
+                if(pos < edge_1){
+                    pos = edge_1;
+                }            
+            default: /* Torus landscape */
                 while(pos >= edge_2){
                     pos = pos - edge_2;   
                 }
@@ -43,14 +64,16 @@ void move(double **pests, double *paras, int ind){
   
   int age_col, move_dist_col, min_age_col, max_age_col, Xpos_col, Ypos_col;
   int move_dist, min_age, max_age, land_type, X_dim, Y_dim, disp_x, disp_y;
-  int Xpos, Ypos, new_xpos, new_ypos, age;
+  int Xpos, Ypos, new_xpos, new_ypos, age, mortality_col, max_rep_col, max_rep;
   
-  Xpos_col        = (int) paras[1];
-  Ypos_col        = (int) paras[2];
-  age_col         = (int) paras[3];
-  move_dist_col   = (int) paras[5];
-  min_age_col     = (int) paras[31];
-  max_age_col     = (int) paras[32];
+  Xpos_col         = (int) paras[1];
+  Ypos_col         = (int) paras[2];
+  age_col          = (int) paras[3];
+  move_dist_col    = (int) paras[5];
+  min_age_col      = (int) paras[31];
+  max_age_col      = (int) paras[32];
+  max_rep_col      = (int) paras[36];
+  mortality_col    = (int) paras[81];
   
   land_type = (int) paras[102];
   X_dim     = (int) paras[103];
@@ -61,6 +84,7 @@ void move(double **pests, double *paras, int ind){
   move_dist = (int) floor(pests[ind][move_dist_col]);
[…]

### 2022-08-01 src/reproduction.c

@@ -1,7 +1,7 @@
 #include "utilities.h"
 
 /* =============================================================================
- * Counts the number of offspring that a single individual should produce
+ * Checks to see if a potential mate is within range of individual in row
  *     pests:           The array holding the parent's information
  *     paras:           The paras vector that holds global information
  *     row:             The row of the individual making the offspring

### 2022-08-01 notebook/ms/ms.Rmd
Begin writing up the second supplemental information

@@
     reference_docx: docx_template.docx
-  pdf_document:
-    toc: no
-  html_document: default
   bookdown::pdf_document2:
@@
     toc: no
+  pdf_document:
+    toc: no
+  html_document: default
 header-includes:
@@
 This `terrain` customisation therefore allows for a high degree of landscape detail, and offers the potential for modelling real-world landscapes from raster images.
-Landscapes are assumed to be a torus, so pests that move off of one edge return on the opposite side of the landscape.
+Edge effects are set using the `land_edge` argument.
+Edge options include `"leaky"` (pests disappear from the landscape), `"reflect"` (pests bounce off of the edge), `"sticky"` (pests stick to the edge), or `"torus"` (pests that move off of one edge return on the opposite side of the landscape).
 
@@
                     time_steps = 160, print_last = TRUE, xdim = 64, ydim = 64,
-                    trait_means = c(0.1, 0.1));
+                    trait_means = c(0.1, 0.1), land_edge = "torus");
 ```

### 2022-08-01 tests/testthat/test-sim_farming.R

@@ -46,8 +46,132 @@ test_that("Simulations are run and output is produced", {
 })
 
 
+test_that("Movement with a reflective boundary", {
+    skip_on_cran();
+    gmt       <- matrix(data = 0, nrow = 2, ncol = 2);
+    diag(gmt) <- 1;
+    mg        <- mine_gmatrix(gmatrix = gmt, loci = 4, layers = 2, indivs = 100, 
+                              npsize = 100, max_gen = 4, prnt_out = FALSE);
+    sim       <- run_farm_sim(mine_output = mg, N = 100, xdim = 40, ydim = 40, 
+                              repro = "sexual", neutral_loci = 1000, 
+                              max_age = 9, min_age_move = 0, max_age_move = 9,
+                              min_age_reproduce = 0, max_age_reproduce = 9, 
+                              min_age_feed = 0, max_age_feed = 9,
+                              food_consume = 0.25, pesticide_consume = 0.1,
+                              rand_age = FALSE, move_distance = 1, 
+                              food_needed_surv = 0.25, 
+                              pesticide_tolerated_surv = 0.1, 
+                              food_needed_repr = 0,
+                              pesticide_tolerated_repr = 0,
+                              reproduction_type = "lambda", mating_distance = 1,
+                              lambda_value = 1,
+                              movement_bouts = 1,
+                              selfing = TRUE, feed_while_moving = FALSE,
+                              mortality_type = 0, age_food_threshold = 0,
+                              age_pesticide_threshold = 0, farms = 4,
+                              time_steps = 4, mutation_pr = 0,
+                              crossover_pr = 0, mutation_type = 0, 
+                              net_mu_layers = 0, net_mu_dir = 0, 
+                              mutation_direction = 0, crop_rotation_type = 2,
+                              crop_rotation_time = 1, 
+                              pesticide_rotation_type = 2, 
+                              pesticide_rotation_time = 1, crop_per_cell = 1,
+                              pesticide_per_cell = 1, crop_sd = 0, 
+                              pesticide_sd = 0, crop_min = 0, crop_max = 1000,
+                              pesticide_min = 0, pesticide_max = 1000,
+                              crop_number = 2, pesticide_number = 1, 
+                              print_inds = FALSE, print_gens = FALSE,
+                              print_last = FALSE, K_on_birth = 1000,
+                              pesticide_start = 0, immigration_rate = 0,
[…]

### 2022-08-02 notebook/ms/ms.Rmd
Fix an error in Figure 1 with the matrix algebra

@@
 output:
-  word_document:
-    reference_docx: docx_template.docx
   bookdown::pdf_document2:
@@
     toc: no
+  word_document:
+    reference_docx: docx_template.docx
   html_document: default
@@
 text(x = 200, y = 790, labels = "0.4", srt = 10, cex = 0.8);
-text(x = 200, y = 660, labels = "-1.2", srt = -58, cex = 0.8);
-text(x = 270, y = 695, labels = "0.2", srt = 46, cex = 0.8);
-text(x = 180, y = 480, labels = "0.1", srt = -40, cex = 0.8);
-text(x = 160, y = 340, labels = "-0.6", srt = 54, cex = 0.8);
+text(x = 200, y = 660, labels = "0.1", srt = -58, cex = 0.8);
+text(x = 270, y = 695, labels = "-1.2", srt = 46, cex = 0.8);
+text(x = 180, y = 480, labels = "-0.6", srt = -40, cex = 0.8);
+text(x = 160, y = 340, labels = "0.2", srt = 54, cex = 0.8);
 text(x = 220, y = 255, labels = "2.2", srt = -12, cex = 0.8);

### 2022-08-06 notebook/ms/evolutionary_algorithm.Rmd
Finish the advanced resevol vignette

---
title: 'resevol: an R package for spatially explicit models of pesticide resistance
  given evolving pest genomes'
subtitle: The evolutionary algorithm (Supporting Information 1)
author: |
  | A. Bradley Duthie&#xb9;&#xb2;, et al.
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2]  alexander.duthie@stir.ac.uk, Biological and Environmental Sciences 3A149
  University of Stirling Stirling, FK9 4LA, UK'
output:
  bookdown::pdf_document2:
    citation_package: natbib
    fig_caption: yes
    toc: no
  word_document:
    reference_docx: docx_template.docx
  pdf_document:
    toc: no
  html_document: default
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[R]{RESEVOL, AN R PACKAGE}
- \usepackage{lineno}
- \linenumbers
- \modulolinenumbers[2]
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: resevol.bib
biblio-style: apalike
---

<a name = "introduction">Introduction</a>
================================================================================

```{r, echo = FALSE}
oldpar <- par();
```

The resevol package models individuals with complex genomes that can include any number of haploid or diploid loci and any number of traits with an arbitrary pre-specified covariance structure. It does this by using a complex network mapping a path from the allele values at each loci to the covarying trait values of individuals.

```{r, echo = FALSE, fig.width = 7, fig.height = 6, fig.cap = "*Network mapping loci to traits through an intermediate set of hidden layers in the mine_gmatrix function*"}
par(mar = c(0.2, 0.2, 0, 0.2));
plot(x = 0, y = 0, type = "n", xlim = c(0, 1000), ylim = c(0, 1000), 
     xaxt = "n", yaxt = "n", bty = "n", xlab = "", ylab = "");
mtext(side = 1, text = "Hidden layers", cex = 2, col = "blue", line = -1);
mtext(side = 2, text = "Loci", cex = 2, col = "darkgreen", line = -2);
mtext(side = 4, text = "Traits", cex = 2, col = "red", line = -1);
Lx  <- rep(x = 100, times = 12);
Ly  <- seq(from = 100, to = 900, length = 12);
points(x = Lx, y = Ly, pch = 20, cex = 6, col = "darkgreen");
H1x <- rep(x = 300, times = 4);
H1y <- seq(from = 100, to = 900, length = 4);
points(x = H1x, y = H1y, pch = 15, cex = 6, col = "blue");
H2x <- rep(x = 500, times = 4);
H2y <- seq(from = 100, to = 900, length = 4);
points(x = H2x, y = H2y, pch = 15, cex = 6, col = "blue");
H3x <- rep(x = 700, times = 4);
H3y <- seq(from = 100, to = 900, length = 4);
points(x = H3x, y = H3y, pch = 15, cex = 6, col = "blue");
Tx  <- rep(x = 900, times = 4);
Ty  <- seq(from = 100, to = 900, length = 4);
points(x = Tx, y = Ty, pch = 18, cex = 8, col = "red");
for(i in 1:4){
    e1 <- seq(from = -36, to = 36, length = 4);
    arrows(x0 = Lx + 26, x1 = H1x[i] - 40, y0 = Ly, y1 = H1y[i] + e1, 
           length = 0.08);
    arrows(x0 = H1x + 36, x1 = H2x[i] - 38, y0 = H1y, y1 = H2y[i] + e1, 
           length = 0.08);
    arrows(x0 = H2x + 36, x1 = H3x[i] - 38, y0 = H2y, y1 = H3y[i] + e1, 
           length = 0.08);
    e2 <- seq(from = -12, to = 12, length = 4)
    arrows(x0 = H3x + 36, x1 = Tx[i] - 38, y0 = H3y, y1 = Ty[i] + e2, 
           length = 0.08);
}
```

The objective of the evolutionary algorithm in resevol is to find a set of network values that produce traits (red diamonds above) with the pre-specified covariance structure given allele values (green circles above) that are sampled from a standard normal distribution (i.e., mean of zero and standard deviation of one). Conceptually, the problem is simple; we need to find values for the black arrows above that produce traits that covary in the way that we want them to covary (within some acceptable margin of error). The `mine_gmatrix` function uses an evolutionary algorithm for identifying sets of values that work. An evolutionary algorithm is an algorithm that works especially well for *I know it when I see it* problems [@Hamblin2013]. @Luke2013 explains the idea behind these algorithms in more detail:

> They're algorithms used to find answers to problems when you have very little to help you: you don’t know beforehand what the optimal solution looks like, you don’t know how to go about finding it in a principled way, you have very little heuristic information to go on, and brute-force search is out of the question because the space is too large. But if you’re given a candidate solution to your problem, you can test it and assess how good it is. That is, you know a good one when you see it.

In the `mine_gmatrix` function of the resevol package, an evolving population of networks like that in Figure 1 above is initialised. Parent networks produce offspring networks with some recombination and mutation, and offspring networks are selected based on how close their trait covariance matrix is to the pre-specified matrix input into the function. The algorithm is inspired by a similar algorithm within the [GMSE R package](https://confoobio.github.io/gmse/index.html) [@Duthie2018].

<a name = "structures">Key data structures used</a>
================================================================================

In the code, the arrows in the above Figure 1 are represented by a set of matrices that map loci values to trait values. There are 12 loci in Figure 1, and 4 nodes in each hidden layer (blue squares). Arrow values between loci and the first hidden layer can then be represented by a matrix with 12 rows and 4 columns (i.e., row 1 holds a value for each of 4 arrows that point to the 4 hidden layer nodes). Note that the values below are initialised randomly, which is how they are initialised in the evolutionary algorithm.

```{r}
arrows_1_dat <- rnorm(n = 4 * 12, mean = 0, sd = 0.1);
arrows_1_mat <- matrix(data = arrows_1_dat, nrow = 12, ncol = 4);
print(arrows_1_mat);
```

We can initialise 12 allele values, one for each locus.

```{r}
loci <- rnorm(n = 12, mean = 0, sd = 1);
```

To then get the value of the first column of four hidden layer nodes (i.e., the first column of blue squares in Figure 1), we can use matrix multiplication.

```{r}
print(loci %*% arrows_1_mat);
```

We can likewise use a 4 by 4 square matrix to represent the values of the arrows from the first column of four hidden layer nodes to the second column of hidden layer nodes.

```{r}
arrows_2_dat <- rnorm(n = 4 * 4, mean = 0, sd = 0.1);
arrows_2_mat <- matrix(data = arrows_2_dat, nrow = 4, ncol = 4);
print(arrows_2_mat);
```

We can then use matrix multiplication to map the 12 allele values to the values of the second column of hidden layer nodes.

```{r}
print(loci %*% arrows_1_mat %*% arrows_2_mat);
```

This pattern can continue, with 4 by 4 square matrices representing the value of arrows between columns of hidden layer nodes, and between the last hidden layer column and traits (note that the number of hidden layer columns can be any natural number, but the number of nodes within a column always equals the number of traits). In the [actual evolutionary algorithm code](https://github.com/bradduthie/resevol/blob/master/src/mine_gmatrix.c), all of these square matrices are themselves held in a large 3D array. But the idea is the same; a vector of allele values is multiplied by multiple matrices until a set of trait values is produced. If multiple vectors of random standard normal allele values are generated, then the traits that they produce from all of this matrix multiplication can be made to covary in some pre-specified way using the evolutionary algorithm.

<a name = "overview">General overview of the evolutionary algorithm</a>
================================================================================

The evolutionary algorithm first initialises a population of networks, with each network having a unique set of values (i.e., black arrows in Figure 1, represented in the code by matrices explained in the [previous selction](#structures)). In each iteration of the evolutionary algorithm, with some probability, networks crossover their values with another randomly selected network. Individual values in each network then mutate with some probability. The fitness of each network is then calculated by comparing its trait covariances with those of a pre-specified covariance matrix. A tournament is then used to select the highest fitness networks, and those selected networks replace the old to comprise the new population. Iterations continue until either a maximum number of iterations is reached or a network is found that produces trait covariances sufficiently close to the pre-specified covariance matrix. Figure 2 below provides a general overview of the evolutionary algorithm.

```{r, echo=FALSE, fig.height=1.75, fig.width=6, fig.cap = "*Conceptual overview of the evolutionary algorithm used in the resevol package.*"}
mbox <- function(x0, x1, y0, y1){
    xx <- seq(from=x0, to=x1, length.out = 100);
    yy <- seq(from=y0, to=y1, length.out = 100);
    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
    return(list(x=xd, y=yd));
}
par(mar=c(0,0,0,0));
# ===============================================================
plot(x=0, y=0, type="n", xlim=c(0,85), ylim=c(50,95), xaxt="n", yaxt="n",
     xlab="",ylab="");
ibox <- mbox(x0 = 0,  x1 = 10, y0 = 90, y1 = 70);
polygon(x = ibox$x, y = ibox$y, lwd = 3, border = "black", col = "white");
cbox <- mbox(x0 = 15, x1 = 25, y0 = 90, y1 = 70);
polygon(x = cbox$x, y = cbox$y, lwd = 3, border = "black", col = "white");
ubox <- mbox(x0 = 30, x1 = 40, y0 = 90, y1 = 70);
polygon(x = ubox$x, y = ubox$y, lwd = 3, border = "black", col = "white");
nbox <- mbox(x0 = 45, x1 = 55, y0 = 90, y1 = 70);
polygon(x = nbox$x, y = nbox$y, lwd = 3, border = "black", col = "white");
fbox <- mbox(x0 = 60, x1 = 70, y0 = 90, y1 = 70);
polygon(x = fbox$x, y = fbox$y, lwd = 3, border = "black", col = "white");
tbox <- mbox(x0 = 75, x1 = 85, y0 = 90, y1 = 70);
polygon(x = tbox$x, y = tbox$y, lwd = 3, border = "black", col = "white");
text(x=5,  y=80, labels="Initialisation", col="black", cex = 0.5);
text(x=20, y=80, labels="Crossover", col="black", cex = 0.5);
text(x=35, y=80, labels="Mutation", col="black", cex = 0.5);
text(x=50, y=82, labels="Fitness", col="black", cex = 0.5);
text(x=50, y=78, labels="evaluation", col="black", cex = 0.5);
text(x=65, y=82, labels="Tournament", col="black", cex = 0.5);
text(x=65, y=78, labels="selection", col="black", cex = 0.5);
text(x=80, y=80, labels="Replacement", col="black", cex = 0.5);
arrows(x0=10, x1=14, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0=25, x1=29, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0=40, x1=44, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0=55, x1=59, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0=70, x1=74, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0 = 80, x1 = 80, y0 = 70, y1 = 68, lwd = 2, length = 0);
arrows(x0 = 80, x1 = 20, y0 = 65, y1 = 65, lwd = 2, length = 0);
arrows(x0 = 20, x1 = 20, y0 = 65, y1 = 70, lwd = 2, length = 0.05);
text(x=73, y=63, labels="No", col="black", cex = 0.5);
rbox <- mbox(x0 = 75, x1 = 85, y0 = 62, y1 = 68);
polygon(x = rbox$x, y = rbox$y, lwd = 3, border = "black", col = "black");
text(x=80, y=65, labels="Termination?", col="white", cex = 0.5);
fbox <- mbox(x0 = 60, x1 = 70, y0 = 60, y1 = 50);
polygon(x = fbox$x, y = fbox$y, lwd = 3, border = "black", col = "white");
text(x=65, y=57, labels="Individual", col="black", cex = 0.5);
text(x=65, y=53, labels="network", col="black", cex = 0.5);
arrows(x0 = 80, x1 = 80, y0 = 62, y1 = 55, lwd = 2, length = 0);
arrows(x0 = 80, x1 = 70, y0 = 55, y1 = 55, lwd = 2, length = 0.05);
text(x=73, y=53, labels="Yes", col="black", cex = 0.5);
```

The steps listed in the box above are explained in more detail below with reference to the arguments applied in the `mine_gmatrix` function that calls it.

<a name = "initialisation">Initialisation</a>
--------------------------------------------------------------------------------

At the start of the evolutionary algorithm, a population of `npsize` networks is initialised. Each individual network in this population is represented by one matrix and one three dimensional array (see an explanation of [the data structures](#structures) above). All the elements of networks are initialised with a random sample from a normal distribution with a mean of 0 and a standard deviation of `sd_ini`.

<a name = "crossover">Crossover</a>
--------------------------------------------------------------------------------

An iteration of the evolutionary algorithm begins with a crossover between the values of networks in the population. For each network in the population, a crossover event will occur in the values linking loci to the first hidden layer (see Figure 1) with a probability of `pr_cross`. And a second independent crossover event will occur in the values linking the first hidden layer to traits, also with a probability of `pr_cross`. The reason that these two crossover events are independent is due to the different dimensions of the underlying arrays (see [key data structures used](#structures) above). A matrix with `loci` rows and a number of columns that matches the number of traits holds the values linking loci to the first hidden layer. A three dimensional array with row and column numbers matching trait number, and a depth matching the number of hidden `layers` (e.g., 3 in Figure 1) holds the remaining values linking the first hidden layer to trait values.

If a crossover event occurs for a focal network, then a contiguous set of values is defined and swapped with another randomly selected network in the population. Dimensions of the contiguous set are selected from a random uniform distribution. For example, given that the network in Figure 1 would be represented by a three dimensional array with 4 rows, 4 columns, and 3 layers, three random integers from 1-4, 1-4, and 1-3 would be sampled twice, respectively, with replacement. If the values selected were 1, 3, and 2 in the first sample, then 3, 3, and 1 in the second sample, then all values from rows 1-3, column 3, and layers 1-2 would be swapped between networks. Conceptually, this is the equivalent of drawing a square around set of arrows in Figure 1 and swapping the arrow values with the values of another network.

<a name = "mutation">Mutation</a>
--------------------------------------------------------------------------------

After crossover occurs, all network values mutate independently at a fixed probability of `mu_pr`. If a mutation event occurs, then a new value is randomly sampled from a normal distribution with a mean of 0 and a standard deviation of `mu_sd`. This value is then added to the existing value in the network.

<a name = "fitness">Fitness evaluation</a>
--------------------------------------------------------------------------------

After mutation, the fitness of each network in the population is evaluated. For each network, a set of `indivs` loci vectors is created, which represents the allele values of `indivs` potential individuals in a simulated population. Elements of each loci vector are randomly sampled from a standard normal distribution. For example, in the network of Figure 1 where `loci = 12`, `loci * indivs` standard normal values would be generated in total. After these `indivs` loci vectors are initialised, values in each vector are mapped to traits using the focal network, thereby producing `indivs` sets of traits from the focal network. These `indivs` sets of traits are then used to calculate the among trait covariances and build a trait covariance matrix for the network. This trait covariance matrix is then compared to the pre-specified `gmatrix` by calculating stress as the mean squared deviation between matrix elements. Lower stress values correspond to higher network fitness.

The stress of each network in the population of `npsize` networks is calculated using the above algorithm. Selection of the next generation of `npsize` networks is then done using a tournament.

<a name = "tournament">Tournament selection</a>
--------------------------------------------------------------------------------

After fitness evaluation, networks in the population compete in a series of tournaments to determine the composition of the next generation of `npsize` networks. Tournament selection is a flexible way to choose the fittest subset of the population [@Hamblin2013]. It starts by randomly selecting `sampleK` networks with replacement to form the competitors in a tournament (note that `sampleK` is constrained to be less than or equal to `npsize`). Of those `sampleK` networks, the `chooseK` networks with the highest fitness (i.e., lowest stress) are set aside to be placed in the new population (note that `chooseK` must be less than or equal to `sampleK`). More tournaments continue until a total of `npsize` new networks are set aside to form the new generation of networks.

<a name = "termination">Termination</a>
--------------------------------------------------------------------------------

Throughout the evolutionary algorithm, the network with the lowest overall stress (from any generation) is retained. The evolutionary algorithm terminates if either the logged stress of the mean network is less than or equal to `term_cri`, or if `max_gen` generations of the evolutionary algorithm have occurred. The mean network stress is used instead of the lowest overall stress because error in randomly generated loci can result in unusually low stress values due to chance, which might not be replicated with a new random sample of loci. When the evolutionary algorithm terminates, only the network with the lowest overall stress is returned.

<a name = "ploidy">Haploid and diploid individuals</a>
================================================================================

The evolutionary algorithm does not distinguish between haploid and diploid genomes Instead, haploid and diploid individuals in resevol simulations are build differently from the mined network described above. For haploid individuals, network values are placed in individual genomes exactly as they are returned by `mine_gmatrix`. Hence, standard normal allele values at loci from haploid genomes will map to predictably covarying traits. For diploid individuals, all network values returned by `mine_gmatrix` are divided by 2, and two copies of each value are then placed in each individual to model diploid genomes. Allele values are then randomly sampled from a normal distribution with a mean of 0 and a standard deviation of `1 / sqrt(2)`, so that summed allele values at homologous loci will have a standard normal distribution. As such, effects of each loci are determined by the sum of homologous alleles. Similarly, homologous network values mapping allele values to traits are also summed, thereby producing the expected trait covariance structure.

<a name = "lit">Literature Cited</a>
================================================================================

```{r, echo = FALSE}
suppressWarnings(par(oldpar));
```

### 2022-08-06 notebook/advanced_vignette_scratch.R

@@ -1,0 +1,263 @@
+mg  <- mine_gmatrix(gmatrix = gmt, loci = 12, indivs = 2000, npsize = 12000, 
+                    max_gen = 5400, sampleK = 1200, chooseK = 6, layers = 1,
+                    mu_pr = 0.2, pr_cross = 0.2, mu_sd = 0.004, 
+                    term_cri = -8);
+
+
+
+library("resevol");
+gmt <- matrix(data = c(1.0, -0.5, 0.2, 0.2, -0.5, 1.0, 0.2, 0.2, 0.2, 
+                       0.2, 0.4, -0.6, 0.2, 0.2, -0.6, 0.4), nrow = 4);
+
+
+set.seed(2022);
+start <- Sys.time();
+mg    <- mine_gmatrix(gmatrix = gmt, loci = 12, indivs = 2000, npsize = 12000, 
+                      max_gen = 5400, sampleK = 1200, chooseK = 6, layers = 2,
+                      mu_pr = 0.1, pr_cross = 0.1, mu_sd = 0.005, 
+                      term_cri = -8, sd_ini = 0.01);
+elaps <- Sys.time() - start;
+print(elaps);
+
+
+
+set.seed(2022);
+start <- Sys.time();
+mg    <- mine_gmatrix(gmatrix = gmt, loci = 12, indivs = 2000, npsize = 12000, 
+                      max_gen = 5400, sampleK = 1200, chooseK = 6, layers = 2,
+                      mu_pr = 0.1, pr_cross = 0.1, mu_sd = 0.01, 
+                      term_cri = -8);
+elaps <- Sys.time() - start;
+print(elaps);
+
+set.seed(2022);
+start <- Sys.time();
+mg    <- mine_gmatrix(gmatrix = gmt, loci = 12, indivs = 2000, npsize = 12000, 
+                      max_gen = 5400, sampleK = 1200, chooseK = 6, layers = 2,
+                      mu_pr = 0.2, pr_cross = 0.2, mu_sd = 0.02, 
+                      term_cri = -8);
+elaps <- Sys.time() - start;
+print(elaps);
+
+set.seed(2022);
+start <- Sys.time();
+mg    <- mine_gmatrix(gmatrix = gmt, loci = 12, indivs = 2000, npsize = 12000, 
+                      max_gen = 5400, sampleK = 1200, chooseK = 6, layers = 2,
[…]

### 2022-08-06 notebook/ms/advanced_techniques.Rmd
Add a new paragraph to the advanced techniques vignette, which focuses on the covariance between pesticide consumption traits

---
title: 'resevol: an R package for spatially explicit models of pesticide resistance
  given evolving pest genomes'
subtitle: Advanced techniques (Supporting Information 2)
author: |
  | A. Bradley Duthie&#xb9;&#xb2;, et al.
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2]  alexander.duthie@stir.ac.uk, Biological and Environmental Sciences 3A149
  University of Stirling Stirling, FK9 4LA, UK'
output:
  bookdown::pdf_document2:
    citation_package: natbib
    fig_caption: yes
    toc: no
  word_document:
    reference_docx: docx_template.docx
  pdf_document:
    toc: no
  html_document: default
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[R]{RESEVOL, AN R PACKAGE}
- \usepackage{lineno}
- \linenumbers
- \modulolinenumbers[2]
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: resevol.bib
biblio-style: apalike
---

<a name = "introduction">Introduction</a>
================================================================================

```{r, echo = FALSE}
oldpar <- par();
```

Here we focus on advanced techniques for simulating pest ecological and evolutionary dynamics using the resevol R package. 
In the main text, we provided a simple example of individual-based simulations to demonstrate how to get started.
This document focuses instead on demonstrating the more advanced options of the package and showcasing what it can do.
The case study will focus on crop and pesticide rotation on a complex, customised landscape that includes farmland, grassland, forest, and water.
The pest species will be sexually reproducing and obligately biparental, and its life history will include an egg and larval stage during which it feeds on crops and consumes pesticide, and a stage during which it moves, mates, and reproduces.
The pest species will include a total of four evolving traits that affect the consumption of two crops and two pesticides.
The objective of simulations will be to test how pest population size changes and traits evolve given different crop and pesticide rotation regimes.
Below, we explain how to model this system in detail, including all of the necessary code for reproducing the example within the text.

<a name = "mining">Initialising pest genomes</a>
================================================================================

First, we need to use the `mine_gmatrix` function to initialise pest genomes. 
In our example, individuals will have four covarying traits resulting from 12 loci.
There will be 6 internal nodes that map loci values to traits (see the Evolutionary Algorithm explanation for details).
The trait covariance structure will be defined as below.

```{r}
library("resevol");
gmt <- matrix(data = c(1.0, -0.5, 0.2, 0.2, -0.5, 1.0, 0.2, 0.2, 0.2, 
                       0.2, 0.4, -0.6, 0.2, 0.2, -0.6, 0.4), nrow = 4);
print(gmt);
```

Rows and columns 1 and 2 will represent traits underlying the consumption rate of crops 1 and 2, respectively. 
Hence, the variation in crop consumption rate is 1 for both crops, while the covariance in consumption rate is -0.5, meaning that there is a trade-off between pest ability to consume crop 1 versus crop 2.
Rows and columns 3 and 4 will represent traits underlying the consumption rate of pesticides 1 and 2, respectively. 
Hence, the variation in pesticide consumption rate is 0.4 for both pesticides, which is lower than it is for crops.
There is a trade-off in the consumption rate of pesticide 1 versus pesticide 2, which is reflected in the covariance of -0.6 in the above matrix.
Finally, there is a positive covariance rate of 0.2 between all crop and all pesticide consumption rates.
What this means is that pests that consume crops quickly also tend to consume pesticides quickly, causing a potential trade-off between the beneficial effects of feeding ability and the negative effects of pesticides.
We can now use the `mine_gmatrix` function to find a network mapping loci to traits that satisfies the above trait covariance structure.
The options used below in `mine_gmatrix` arguments will lead to a computationally intense (and therefore time-consuming) search, particularly due to the large number of networks in the population (`npsize = 12000`), high number of individuals used to test network stress (`indivs = 2000`), high number of hidden nodes (`layers = 6`), and strict stress criteria (`term_cri = -8`).

```{r, eval = FALSE}
set.seed(2022);
mg  <- mine_gmatrix(gmatrix = gmt, loci = 12, indivs = 2000, npsize = 12000, 
                    max_gen = 5400, sampleK = 1200, chooseK = 6, layers = 4,
                    mu_pr = 0.2, pr_cross = 0.2, mu_sd = 0.004, 
                    term_cri = -8);
```

To save time, a genome `mg` produced from the code above has been saved into the resevol R package as `advanced_techniques_eg.rda`, which we can load.
Note that running `mine_gmatrix` can be a time-consuming process, and will often require multiple attempts to get the above parameter settings to produce a desired network.
See @Hamblin2013 for useful advice on parameter value selection for the evolutionary algorithm.
Time-sensitive parameter values specific to the resevol package include `indivs` and `layers`.
Increasing values for both of these parameters can increase run time.
In particular, fewer internal nodes (`layers`) will greatly reduce the time it takes to find a suitable network.
The trade-off against this reduced time is that more layers make it possible to model more highly constrained trait covariances.
Nevertheless, if such constraints are not needed, then only 1-2 `layers` might be necessary.

```{r}
load(system.file("advanced_eg.rda", package = "resevol"));
```

The contents of each list element of `mg` are not important for our purposes, but an explanation is available in the [package documentation](../reference/mine_gmatrix.html). 
What is important is the sixth list element, which holds the estimated covariance structure found by the evolutionary algorithm.

```{r, echo = FALSE}
print(mg[[6]]);
```

We can compare the above covariance structure with the one that we specified in `gmt`. 
The stress of `mg` (i.e., mean squared deviation of `mg` elements from `gmt` elements) is `r mean((mg[[6]] - gmt)^2)`. 
But this is only based on one population of `indivs = 2000` initialised pests. 
We can use the `stress_test` function to see what the distribution of stress is over 1000 such initialised populations of 2000 individuals (Figure 1).

```{r, fig.height = 6, fig.width = 6, fig.align = "left", fig.cap = "Distribution of stress for initialised pest trait covariances in the resevol R package. Stress values are computed using genome values produced by the mine\\_gmatrix function for 1000 replicate populations of initialised pest loci values."}
sim_stress <- stress_test(mine_output = mg, indivs = 2000, reps = 1000);
hist(x = sim_stress, main = "", breaks = 20, xlab = "Initialised stress");
```

If we are satisfied with the distribution of stress, then we can accept `mg` as the genome for our initialised pests. 
The whole contents of `mg` are then passed on to the `run_farm_sim` function, which initialises pests using the genome values in `mg` before running simulations.
Next, because we want to run a simulation on a customised landscape, we will explain how such a landscape can be generated and inserted into the simulation.

<a name = "land">Landscape</a>
================================================================================

Default settings for running simulations with the `run_farm_sim` function create a spatially explicitly landscape with dimensions specified by arguments `xdim` and `ydim`, and a number of farms specified by `farms`.
These arguments can be over-ridden with a matrix that defines landscape cell identities, which is set in the `terrain` argument. 
A very simple custom 12 by 12 landscape might look something like the below.

```{r}
simple_terrain              <- matrix(data = 0, nrow = 12, ncol = 12)
simple_terrain[1:2, 1:3]    <- 1;
simple_terrain[3:7, 1:2]    <- 1;
simple_terrain[8:12, 1:2]   <- 2;
simple_terrain[1:2, 6:9]    <- 3;
simple_terrain[3:6, 7:9]    <- 3;
simple_terrain[1:6, 10:12]  <- 4;
simple_terrain[10:12, 7]    <- 5;
simple_terrain[7:12, 8:12]  <- 5;
simple_terrain[3:12, 3:4]   <- 6;
simple_terrain[7:9, 5]      <- 6;
simple_terrain[1:2, 4:5]    <- 7;
simple_terrain[3:6, 5:6]    <- 7;
simple_terrain[7:9, 6:7]    <- 7;
simple_terrain[10:12, 5:6]  <- 7;
print(simple_terrain)
```

What is important is that the matrix elements only include natural numbers from 1 to the total number of farms (not skipping any numbers). 
In the case of the above, numbers 1-7 all appear on the landscape, so the matrix will work.
We can visualise the matrix more clearly by looking at it as an image (Figure 2).

```{r, fig.width=5, fig.height=5, fig.align = "left", fig.cap = "Image representing a simple landscape for the resevol package created from a 12 by 12 matrix of natural numbers from 1-7."}
par(mar = c(0, 0, 0, 0));
image(t(simple_terrain), xaxt = "n", yaxt = "n", useRaster = TRUE);
```

In the image above, each colour represents a unique farm, and colours correspond to the numbers within the matrix `simple_terrain` (note that the orientation differs between the matrix and the image).
This is a simple example, but matrices can be made large to model highly complex landscapes.
Additionally, not every colour needs to represent a 'farm' per se. 
We can represent other types of terrain such as grassland, woodland, or water with its own natural number.
The trick in this case is to have each of these other types of terrain grow a crop and use a pesticide that has no effect on pests (i.e., cannot be consumed).
If terrain types have no effect on pests, then they can effectively model land that is not used for farming.
We can show this using the matrix `land_eg`, which is also included in `advanced_techniques_eg.rda`.
The dimensions of the landscape modelled in `land_eg` is 128 by 128 cells, and the landscape includes 14 farms (1-14), grassland (15), forest (16), and water (17).
The code below produces an image that shows how these terrain types are distributed over the landscape (Figure 3).

```{r, fig.width=7, fig.height=7, fig.align = "left", fig.cap = "A complex landscape to be used in the resevol R package, including 14 separate farms, grassland, forest, and water. The image is represented in the code by a matrix in which element numbers correspond to different terrain colours (e.g., elements corresponding to water are numbered 17)."}
land_file <- system.file("landscape_eg.csv", package = "resevol");
land_dat  <- read.csv(file = land_file, header = FALSE);
land_eg   <- t(as.matrix(land_dat));
farm_cols <- c("#f4eadc", "#6a4b20", "#cea05f", "#e1c59d", "#a97833", "#cea05f",
               "#f2e6d6", "#6a4b20", "#cc9c59", "#dfc197", "#a27331", "#f0e3d0",
               "#5d421c", "#ca9852");
land_cols <- c(farm_cols, "#00ab41", "#234F1E", "#2832C2");
par(mar = c(0, 0, 0, 0));
image(land_eg, xaxt = "n", yaxt = "n", col = land_cols);
points(x = 0.2, y = 0.05, cex = 9, pch = 20);
text(x = 0.2, y = 0.05, labels = "1", cex = 2, col = "red");
points(x = 0.4, y = 0.1, cex = 9, pch = 20);
text(x = 0.4, y = 0.1, labels = "2", cex = 2, col = "red");
points(x = 0.4, y = 0.1, cex = 9, pch = 20);
text(x = 0.4, y = 0.1, labels = "2", cex = 2, col = "red");
points(x = 0.25, y = 0.27, cex = 9, pch = 20);
text(x = 0.25, y = 0.27, labels = "3", cex = 2, col = "red");
points(x = 0.20, y = 0.43, cex = 9, pch = 20);
text(x = 0.20, y = 0.43, labels = "4", cex = 2, col = "red");
points(x = 0.42, y = 0.48, cex = 9, pch = 20);
text(x = 0.42, y = 0.48, labels = "5", cex = 2, col = "red");
points(x = 0.28, y = 0.58, cex = 9, pch = 20);
text(x = 0.28, y = 0.58, labels = "6", cex = 2, col = "red");
points(x = 0.1, y = 0.8, cex = 9, pch = 20);
text(x = 0.1, y = 0.8, labels = "7", cex = 2, col = "red");
points(x = 0.7, y = 0.05, cex = 9, pch = 20);
text(x = 0.7, y = 0.05, labels = "8", cex = 2, col = "red");
points(x = 0.9, y = 0.2, cex = 9, pch = 20);
text(x = 0.9, y = 0.2, labels = "9", cex = 2, col = "red");
points(x = 0.85, y = 0.4, cex = 9, pch = 20);
text(x = 0.85, y = 0.4, labels = "10", cex = 2, col = "red");
points(x = 0.92, y = 0.58, cex = 9, pch = 20);
text(x = 0.92, y = 0.58, labels = "11", cex = 2, col = "red");
points(x = 0.88, y = 0.76, cex = 9, pch = 20);
text(x = 0.88, y = 0.76, labels = "12", cex = 2, col = "red");
points(x = 0.86, y = 0.93, cex = 9, pch = 20);
text(x = 0.86, y = 0.93, labels = "13", cex = 2, col = "red");
points(x = 0.52, y = 0.91, cex = 9, pch = 20);
text(x = 0.52, y = 0.91, labels = "14", cex = 2, col = "red");
points(x = 0.05, y = 0.59, cex = 7, pch = 20);
text(x = 0.05, y = 0.59, labels = "15", cex = 1.7, col = "white");
points(x = 0.46, y = 0.28, cex = 7, pch = 20);
text(x = 0.46, y = 0.28, labels = "16", cex = 1.7, col = "white");
points(x = 0.62, y = 0.36, cex = 7, pch = 20);
text(x = 0.62, y = 0.36, labels = "17", cex = 1.7, col = "white");
```

This `land_eg` matrix can be used in `run_farm_sim` by setting `terrain = land_eg`. 
The dimensions of the landscape will then automatically be set to `xdim = 128` and `ydim = 128`, and the number of farms will be set to `farms = 17`.
To ensure that we do not have pests consuming crops or pesticides from non-farm areas, we just need to set pest consumption options accordingly.
Next, we describe how this can be done by defining how crops and pesticides will be initialised and rotated on the landscape.

<a name = "simulation">Running simulations</a>
================================================================================

We will run the same simulated population using the `mg` pest genome and `land_eg` landscape [described above](#land) under two different pesticide regimes. 
In both regimes, the same pesticide is applied to farms on each side of the river (e.g., farms 1-7 apply pesticide 1 and farms 8-14 apply pesticide 2).
In simulation 1, the pesticide is never rotated.
In simulation 2, farms rotate between pesticides 1 and 2 every 9 time steps.
First, we will demonstrate how to customise crop and pesticide rotation for simulation 1, then run the simulation, and finally show the output. 
Second, we will repeat the process for simulation 2 and demonstrate how comparisons can be made between the two simulations.
We will then compare output from the two simulations.

<a name = "sim1">Simulation 1</a>
--------------------------------------------------------------------------------

First, we focus on how to set [custom pesticide and crop rotation](#rotation1) for simulations.
Next, we run [simulation 1](#run_farm_sim1) with the `run_farm_sim` function.
Finally, we demonstrate how to produce [output](#output_sim1) for simulation 1.

### <a name = "rotation1">Custom pesticide and crop rotation</a>

The `run_farm_sim` arguments `crop_init` and `crop_rotation_type` specify starting crops for each farm and how crops rotate, respectively.
Equivalent arguments `pesticide_init` and `pesticide_rotation_type` similarly specify starting pesticides and pesticide rotations, respectively.
We can set each argument to a number as a short-hand way of specifying these parameters.
For example, the default initialisation `crop_init = "random"` simply causes each farm to start the simulation with a random crop.
Setting values of `crop_rotation_type` to 1, 2, or 3 causes no rotation, random rotation, or cycling through crop numbers in sequence, respectively.
But we can generate more specific initialisation and rotation types using a vector to initialise crops and pesticides, and using a matrix to define rotation.
For initialisation, a vector elements correspond to farm number and element values correspond to the crop initialised.
For our example with 17 unique zones of terrain on the landscape, we can define initialised crop choice as follows.

```{r}
initial_crop <- c(1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 2, 3, 3, 3);
```

That is the first 7 farms on one side of the river all start growing crop 1. 
Farms 8-14 instead all start growing crop 2.
Crop 3 is used as a dummy for grassland (15), forest (16), and water (17).
We explain how to ensure that 15-17 is never used by pests below.
Meanwhile, we can specify a matrix for the crop rotation regime.
This regime will cause farmers to rotate between crops 1 and 2, but zones 15-17 to always use 'crop' 3.

```{r}
rotate_crop       <- matrix(data = 0, nrow = 3, ncol = 3);
rotate_crop[1, 2] <- 1;
rotate_crop[2, 1] <- 1;
rotate_crop[3, 3] <- 1;
print(rotate_crop);
```

In the above matrix `rotate_crop_sim1`, matrix elements define the probability of transitioning from one crop type (rows) to another crop type (columns).
Hence, in `rotate_crop_sim1`, farms that have been applying crop 1 will always switch to crop 2, and vice versa.
Landscape zones applying crop 3 will never transition to any other crop, nor will any other crop transition to type 3.
We can set initial pesticides and pesticide rotation in the same way.
First we set initial pesticide use.

```{r}
initial_pesticide <- c(1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 2, 3, 3, 3);
```

Next, we define pesticide rotation.

```{r}
rotate_pesticide           <- matrix(data = 0, nrow = 3, ncol = 3);
rotate_pesticide[1, 2]     <- 1;
rotate_pesticide[2, 1]     <- 1;
rotate_pesticide[3, 3]     <- 1;
print(rotate_pesticide);
```

We therefore have set farms to always use one of two crops and one of two pesticides.
Everything that is not a farm will use crop 3 and pesticide 3.
Next, we bring everything together and show how pests, the landscape, and the rotation of crops and pesticides can be used in `run_farm_sim` for simulations.

### <a name = "run_farm_sim1">Running simulation 1</a>

```{r, eval = FALSE}
set.seed(2022);
sim1       <- run_farm_sim(mine_output             = mg, 
                          terrain                  = land_eg,
                          crop_init                = initial_crop,
                          crop_rotation_type       = rotate_crop,
                          pesticide_init           = initial_pesticide,
                          pesticide_rotation_type  = rotate_pesticide,
                          food_consume             = c("T1", "T2", 0),
                          pesticide_consume        = c("T3", "T4", 0),
                          crop_number              = 3,
                          pesticide_number         = 3,
                          trait_means              = c(2, 2, 0.0, 0.0),
                          max_age                  = 6,
                          min_age_feed             = 0,
                          max_age_feed             = 2,
                          min_age_move             = 3,
                          max_age_move             = 6,
                          min_age_metabolism       = 3,
                          max_age_metabolism       = 6,
                          metabolism               = 0.5,
                          food_needed_surv         = 1,
                          reproduction_type        = "food_based",
                          food_needed_repr         = 2,
                          N                        = 1000, 
                          repro                    = "biparental",
                          mating_distance          = 4,
                          rand_age                 = TRUE,
                          pesticide_tolerated_surv = 2,
                          movement_bouts           = 4,
                          move_distance            = 2,
                          crop_per_cell            = 10,
                          crop_sd                  = 0,
                          pesticide_per_cell       = 1,
                          pesticide_sd             = 0,
                          crop_rotation_time       = 18,
                          pesticide_rotation_time  = 9,
                          time_steps               = 240,
                          pesticide_start          = 81,
                          immigration_rate         = 100,
                          land_edge                = "reflect",
                          mutation_pr              = 0.01,
                          crossover_pr             = 0.01,
                          print_gens              = TRUE,
                          print_last              = TRUE);
```

Immediately below the first six arguments, we have the arguments `food_consume`, `pesticide_consume`, `crop_number`, and `pesticide_number`. 
The way that these arguments are set ensures that pests only interact with farm cells.
The rate of food consumption for the first two crops is defined by the values of traits 1 and 2 and set as `"T1"` and `"T2"`, respectively.
The rate of consumption for crop 3 is set to 0, meaning that no food can be eaten on these cells.
Similarly, the rate at which pesticides 1 and 2 are consumed is defined by the values of traits 3 and 4, which are set as `"T3"` and `"T4"`, respectively.
The rate of consumption for pesticide 3 is set to 0 so that pesticide is not consumed on these cells.

We next set the mean values of traits at the start of the simulation with the argument `trait_means`. 
Mean consumption rate of both crops is set to 1, while mean pesticide consumption rate is set to 0.1.
Pests live up to 6 time steps (`max_age = 6`), and are defined to start feeding upon birth (`min_age_feed = 0`) and stop at age 2 (`max_age_feed = 2`). 
At age 3, pests start to move (`min_age_move = 3`) and metabolise food (`min_age_metabolism = 3`) at a rate of `metabolism = 0.5`.
The `metabolism` argument defines the amount of consumed food lost during the time step, which can affect pests when food consumed affects survival (`food_needed_surv`) and reproduction (`reproduction_type`).
By setting `reproduction_type = "food_based"` and `food_needed_surv = 0`, we model a system in which pests burn the food consumed from ages 0-2 at a rate of 0.5 units per time step after age 3, then die when their reserves drop below 0.
By setting `food_needed_repr = 1`, we ensure that only pests that have 1 or more units of food stored can reproduce.
Note that if we desired, we could also set `food_needed_surv`, `food_needed_repr`, and `metabolism` to be evolving traits.

We initialise the population with `N = 1000` obligately biparental (`repro = "biparental"`) pests, and can find a mate within `mating_distance = 4` cells of their own cell.
All pests are initialised at a random age (`rand_age = TRUE`).
Pests will die if they consume more than 0.1 total units of pesticide (`pesticide_tolerated_surv = 0.1`).
In each time step, pests that are able to move will do so an average of 4 times on the landscape (`movement_bouts = 4`), moving up to 2 cells in any direction during each bout (`movement_distance = 2`).
Recall that since pests consume food and pesticide only between ages 0-2, then start moving at age 3, they no food or pesticide consumption occurs during movement (if applicable, it could be turned on with `feed_while_moving = TRUE` and `pesticide_while_moving = TRUE`).
Each farm cell produces 10 units of crops (`crop_per_cell = 10` and `crop_sd = 0`) and 1 unit of pesticide (`pesticide_per_cell = 1` and `pesticide_sd = 0`).
Crops are rotated every 18 time steps (`crop_rotation_time = 18`), and pesticides are set to rotate every 9 time steps (`pesticide_rotation_time = 9`), but this pesticide rotation has no effect because the same pesticide is applied on rotation (as defined by `rotate_pesticide_sim1`).
This can be conceptualised as modelling a 18 time step growing season in which pesticides are re-applied at the start and halfway point of a season.
Note that it is critical to consider that crops are therefore refreshed every 18 time steps, not every time step.
Hence, any food consumed by a pest will be lost from a cell and not replenished until crop rotation occurs.
Because more than one pest can occupy a single cell, crop loss can occur quickly depending on how parameter values are set.

Lastly, we set the number of time steps to `time_steps = 240`, and we set `pesticide_start = 81`, which means that pesticides are not applied at all until after a burn-in of 120 time steps.
In each time step, an average of 100 immigrants enter the population (`immigration_rate = 100`).
We set the landscape edge type to `land_edge = "reflect"` to model a reflective edge in which pests that attempt to leave one side of the landscape bounce back toward the centre.
Pest genome mutation rate and crossover rate at a locus are set to `mutation_pr = 0.01` and `crossover_pr = 0.01`, respectively.
We print out the dynamics of the evolving population over time, and all of the individual data from the last time step, to a CSV file (individual data are not included in the package due to its size).
All values not mentioned are set to default parameter values, which can be found in the [package documentation](../reference/run_farm_sim.html).

### <a name = "output_sim1">Simulation 1 output</a>

When running the function `run_farm_sim` above, the population size of the pest will be printed in the R console in each time step.
This is primarily because simulations can take a long time, and printing makes it possible to estimate how much time is remaining.
Printing to the console can be turned off by setting `print_gens = FALSE`.
Once the simulation has finished, two CSV files will be created in the working directory.
The file "population_data.csv" saves population data over time, and "individuals.csv" saves every characteristic of all individuals in the last time step (including full individual genomes).
Here we show how to work with the most relevant information from these two files to make inferences about pest population and evolutionary dynamics.

We will start with the population level output that is printed to "population_data.csv", which has been renamed "population_data_sim1.csv".

```{r}
population_data_file_sim1 <- system.file("population_data_sim1.csv", 
                                    package = "resevol");
population_data_sim1      <- read.csv(file = population_data_file_sim1);
```

This file includes population-level parameters of population size and mean pest age, sex, food consumed, pesticide consumed, mortality rate, and trait values reported for each time step.
If `get_f_coef = TRUE` in `run_farm_sim`, then mean inbreeding coefficients are also reported.

```{r}
print(head(population_data_sim1));
```

Because the structure of the data frame above depends on the number of crops, pesticides, and evolving traits simulated, there are no pre-set functions in the resevol package for plotting.
Here we include code for plotting population size, consumption of the two food types, two pesticides types, and evolving trait means.
Note that because crop and pesticide 3 are dummy variables (not consumed at all), we do not need to plot these.
Figure 4 shows how population density of pests changes over time.

```{r, fig.height = 7, fig.width = 7, fig.align = "left", fig.cap = "Pest abundance over time in a population simulated using the resevol R package in which pesticides are not rotated over time. The grey shaded regions show individual crop seasons; pesticides are applied at the start and midpoint of each seasons, beginning at the time step indicated by the red vertical line."}
mbox <- function(x0, x1, y0, y1){
    xx <- seq(from=x0, to=x1, length.out = 100);
    yy <- seq(from=y0, to=y1, length.out = 100);
    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
    return(list(x=xd, y=yd));
}
season          <- seq(from = 0, to = 240, by = 18);
blocks          <- length(season) - 1;
plot(x = population_data_sim1[["time_step"]], type = "n",
     y = population_data_sim1[["population_size"]], cex.lab = 1.25, 
     cex.axis = 1.25, ylab = "Pest population abundance", xlab = "Time step", 
     ylim = c(0, max(population_data_sim1[["population_size"]])));
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 21000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
     y = population_data_sim1[["population_size"]], type = "l", lwd = 2);
abline(v = 81, lwd = 3, col = "red");
box();
```

Figure 5 shows how the mean consumption of different foods and pesticides changes over time.

```{r, fig.height = 7, fig.width = 7, fig.align = "left", fig.cap = "Pest consumption over time in a population simulated using the resevol R package in which pesticides are not rotated over time. Panels show (a) mean amount of crop 1 and (b) crop 2 consumed per pest, and (c) mean amount of pesticide 1 and (d) pesticide 2 consumed per pest. Grey shaded regions show individual crop seasons; pesticides are applied at the start and midpoint of each seasons, beginning at the time step 100."}
par(mfrow = c(2, 2), mar = c(1, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(0, 0.8),
     y = population_data_sim1[["mean_food1_consumed"]], cex.lab = 1.25, 
     ylab = "Mean crop 1 consumed", xlab = "Time step", xaxt = "n");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
       y = population_data_sim1[["mean_food1_consumed"]], type = "l", lwd = 2);
text(x = 5, y = 0.79, labels = "a", cex = 2);
box();
par(mar = c(1, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(0, 0.8),
     y = population_data_sim1[["mean_food2_consumed"]], cex.lab = 1.25 ,
     ylab = "Mean crop 2 consumed", xlab = "Time step", xaxt = "n");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
       y = population_data_sim1[["mean_food2_consumed"]], type = "l", lwd = 2);
text(x = 5, y = 0.79, labels = "b", cex = 2);
box();
par(mar = c(4, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(0, 0.8),
     y = population_data_sim1[["mean_pesticide1_consumed"]], cex.lab = 1.25, 
     ylab = "Mean pesticide 1 consumed", xlab = "Time step");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
       y = population_data_sim1[["mean_pesticide1_consumed"]], type = "l", lwd = 2);
box();
text(x = 5, y = 0.78, labels = "c", cex = 2);
par(mar = c(4, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(0, 0.8),
     y = population_data_sim1[["mean_pesticide2_consumed"]], cex.lab = 1.25, 
     ylab = "Mean pesticide 2 consumed", xlab = "Time step");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
       y = population_data_sim1[["mean_pesticide2_consumed"]], type = "l", 
       lwd = 2);
box();
text(x = 5, y = 0.78, labels = "d", cex = 2);
```

Similarly, we can visualise how all four traits evolve over time (Figure 6).

```{r, fig.height = 7, fig.width = 7, fig.align = "left", fig.cap = "Mean values of pest traits T1 (crop 1 consumption rate), T2 (crop 2 consumption rate), T3 (pesticide 1 consumption rate), and T4 (pesticide 2 consumption rate) in a population simulated using the resevol R package in which pesticides are not rotated over time. Grey shaded regions show individual crop seasons; pesticides are applied at the start and midpoint of each seasons, beginning at the time step 100."}
par(mfrow = c(2, 2), mar = c(1, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(2, 3.6),
     y = population_data_sim1[["trait1_mean_value"]], cex.lab = 1.25, 
     ylab = "Mean crop 1 consumption trait", xlab = "Time step", xaxt = "n");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
     y = population_data_sim1[["trait1_mean_value"]], type = "l", lwd = 2);
text(x = 5, y = 3.56, labels = "a", cex = 2);
box();
par(mar = c(1, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(2, 3.6),
     y = population_data_sim1[["trait2_mean_value"]], cex.lab = 1.25 ,
     ylab = "Mean crop 2 consumption trait", xlab = "Time step", xaxt = "n");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
       y = population_data_sim1[["trait2_mean_value"]], type = "l", lwd = 2);
text(x = 5, y = 3.56, labels = "b", cex = 2);
box();
par(mar = c(4, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(-0.8, 1.4),
     y = population_data_sim1[["trait3_mean_value"]], cex.lab = 1.25, 
     ylab = "Mean pesticide 1 consumption trait", xlab = "Time step");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
       y = population_data_sim1[["trait3_mean_value"]], type = "l", lwd = 2);
box();
text(x = 5, y = 1.34, labels = "c", cex = 2);
par(mar = c(4, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(-0.8, 1.4),
     y = population_data_sim1[["trait4_mean_value"]], cex.lab = 1.25, 
     ylab = "Mean pesticide 2 consumption trait", xlab = "Time step");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
       y = population_data_sim1[["trait4_mean_value"]], type = "l", lwd = 2);
box();
text(x = 5, y = 1.34, labels = "d", cex = 2);
```

Next, we look at the file that includes the complete information for all individuals in the last time step of the simulation. 
Using this data set, we can examine the distribution of pests on the landscape, and the covariance of evolving pest traits. 
Note that information has been removed from the file below to avoid a very large file size. But in this output "last_time_step.csv", traits always begin in column 102, meaning that individual trait values for traits 1, 2, 3, and 4 will be in columns 102, 103, 104, and 105, respectively. 
In this example, the CSV output file "last_time_step.csv" has been renamed "last_time_step_sim1.csv".
To minimise the size of the resevol R package, all data columns except pest locations and traits have been removed from example outputs, so pest locations are in columns 1-2 (not 3-4) and traits are in columns 3-6 (not 102-105).
The last time step output is read into R below.

```{r}
last_time_step_file_sim1 <- system.file("last_time_step_sim1.csv", 
                                        package = "resevol");
pop_last_time_step_sim1  <- read.csv(file = last_time_step_file_sim1);
```

We can plot the distribution over the landscape using the code below.
The code identifies the x-locations and y-locations of each pest `i` in `pop_last_time_step`, and these landscape cells are replaced with the number 18.
A new land colour for black `#000000` is then included, and the map is reproduced as an image in R.
Figure 7 shows the same map as above, but with the numbering removed to make the pests (black points) more visible.

```{r, fig.width=7, fig.height=7, fig.align = "left", fig.cap = "A complex landscape for simulating the ecology of pests and the evolution of pesticide resistance on farmland in the resevol R package. Terrain includes farms (brown colours), grassland (light green), forest (dark green), and water (blue). Black points show the locations of individual pests after 240 time steps for a simulation in which pesticides are not rotated over time."}
par(mar = c(0, 0, 0, 0));
landscape    <- land_eg;
for(i in 1:dim(pop_last_time_step_sim1)[1]){
    xloc <- pop_last_time_step_sim1[i, 1] + 1;
    yloc <- pop_last_time_step_sim1[i, 2] + 1;
    landscape[xloc, yloc] <- 18;
}
land_cols <- c(land_cols, "#000000");
image(landscape, xaxt = "n", yaxt = "n", col = land_cols);
head(pop_last_time_step_sim1);
print(land_cols);
```

The spatial pattern of pests over the landscape can be useful for visualising pest population dynamics.
In the case of the above, pest density is higher on some (e.g., 1, 8, 9, 10, and 15) farms than others (e.g., 3, 12, and 13), and this pest density could be interpreted with respect to individual farm crop and pesticide rotation regimes.
Hence, for the four evolving traits, a covariance matrix of pest traits can be calculated at the end of the simulation.

```{r, eval = TRUE}
end_trait_covs <- cov(pop_last_time_step_sim1[,3:6]);
print(end_trait_covs);
```

The above covariance matrix shows that after 240 time steps, the covariance between consumption of pesticides 1 and 2 has decreased from `r mg[[6]][3, 4]` to `r end_trait_covs[3, 4]`, suggesting that the trade-off for resisting pesticides had become stronger.
In contrast, the covariance between consumption of crops 1 and 2 has increased from `r mg[[6]][1, 2]` to `r end_trait_covs[1, 2]`, suggesting that the trade-off in ability to consume crops of each type has weakened.

<a name = "sim2">Simulation 2</a>
--------------------------------------------------------------------------------

We now repeat the simulations such that pesticide application is rotated every 9 time steps.
To make this rotation, we only need to redefine `rotate_pesticide`.

```{r}
rotate_pesticide           <- matrix(data = 0, nrow = 3, ncol = 3);
rotate_pesticide[1, 2]     <- 1;
rotate_pesticide[2, 1]     <- 1;
rotate_pesticide[3, 3]     <- 1;
```

Now, every 9 time steps, crops using pesticide 1 will switch to pesticide 2, and crops using pesticide 2 will switch to pesticide 1.
We can re-run the same simulation with the same genome (`mg`), landscape (`land_eg`), and simulation seed (`2022`) used in simulation 1.
Hence, the only thing changing between the two simulations is the pesticide rotation.

```{r, eval = FALSE}
set.seed(2022);
sim2       <- run_farm_sim(mine_output             = mg, 
                          terrain                  = land_eg,
                          crop_init                = initial_crop,
                          crop_rotation_type       = rotate_crop,
                          pesticide_init           = initial_pesticide,
                          pesticide_rotation_type  = rotate_pesticide,
                          food_consume             = c("T1", "T2", 0),
                          pesticide_consume        = c("T3", "T4", 0),
                          crop_number              = 3,
                          pesticide_number         = 3,
                          trait_means              = c(2, 2, 0.0, 0.0),
                          max_age                  = 6,
                          min_age_feed             = 0,
                          max_age_feed             = 2,
                          min_age_move             = 3,
                          max_age_move             = 6,
                          min_age_metabolism       = 3,
                          max_age_metabolism       = 6,
                          metabolism               = 0.5,
                          food_needed_surv         = 1,
                          reproduction_type        = "food_based",
                          food_needed_repr         = 2,
                          N                        = 1000, 
                          repro                    = "biparental",
                          mating_distance          = 4,
                          rand_age                 = TRUE,
                          pesticide_tolerated_surv = 2,
                          movement_bouts           = 4,
                          move_distance            = 2,
                          crop_per_cell            = 10,
                          crop_sd                  = 0,
                          pesticide_per_cell       = 1,
                          pesticide_sd             = 0,
                          crop_rotation_time       = 18,
                          pesticide_rotation_time  = 9,
                          time_steps               = 240,
                          pesticide_start          = 81,
                          immigration_rate         = 100,
                          land_edge                = "reflect",
                          mutation_pr              = 0.01,
                          crossover_pr             = 0.01,
                          print_gens              = TRUE,
                          print_last              = TRUE);
```

We can look at the same outputs for simulation 2 that we did for simulation 1.
We first read in the output for simulation 2.

```{r}
population_data_file_sim2 <- system.file("population_data_sim2.csv", 
                                    package = "resevol");
population_data_sim2      <- read.csv(file = population_data_file_sim2);
```

To avoid repetition, the code for plots will not be included for simulation 2.
Figure 8 shows population abundance over time for simulation 2.

```{r, echo = FALSE, fig.height = 7, fig.width = 7, fig.align = "left", fig.cap = "Pest abundance over time in a population simulated using the resevol R package in which pesticides rotates every 9 time steps. The grey shaded regions show individual crop seasons; pesticides are applied at the start and midpoint of each seasons, beginning at the time step indicated by the red vertical line."}
mbox <- function(x0, x1, y0, y1){
    xx <- seq(from=x0, to=x1, length.out = 100);
    yy <- seq(from=y0, to=y1, length.out = 100);
    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
    return(list(x=xd, y=yd));
}
season          <- seq(from = 0, to = 240, by = 18);
blocks          <- length(season) - 1;
plot(x = population_data_sim2[["time_step"]], type = "n",
     y = population_data_sim2[["population_size"]], cex.lab = 1.25, 
     cex.axis = 1.25, ylab = "Pest population abundance", xlab = "Time step", 
     ylim = c(0, max(population_data_sim2[["population_size"]])));
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 21000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim2[["time_step"]], 
     y = population_data_sim2[["population_size"]], type = "l", lwd = 2);
abline(v = 81, lwd = 3, col = "red");
box();
```

After pesticides are applied at time step 81, pest population size decreases and does not recover.
Figure 9 shows crop and pesticide consumption over time.

```{r, echo = FALSE, fig.height = 7, fig.width = 7, fig.align = "left", fig.cap = "Pest consumption over time in a population simulated using the resevol R package in which pesticides rotate every 9 time steps. Panels show (a) mean amount of crop 1 and (b) crop 2 consumed per pest, and (c) mean amount of pesticide 1 and (d) pesticide 2 consumed per pest. Grey shaded regions show individual crop seasons; pesticides are applied at the start and midpoint of each seasons, beginning at the time step 100."}
par(mfrow = c(2, 2), mar = c(1, 4, 1, 1));
plot(x = population_data_sim2[["time_step"]], type = "n", ylim = c(0, 0.8),
     y = population_data_sim2[["mean_food1_consumed"]], cex.lab = 1.25, 
     ylab = "Mean crop 1 consumed", xlab = "Time step", xaxt = "n");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim2[["time_step"]], 
       y = population_data_sim2[["mean_food1_consumed"]], type = "l", lwd = 2);
text(x = 5, y = 0.79, labels = "a", cex = 2);
box();
par(mar = c(1, 4, 1, 1));
plot(x = population_data_sim2[["time_step"]], type = "n", ylim = c(0, 0.8),
     y = population_data_sim2[["mean_food2_consumed"]], cex.lab = 1.25 ,
     ylab = "Mean crop 2 consumed", xlab = "Time step", xaxt = "n");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim2[["time_step"]], 
       y = population_data_sim2[["mean_food2_consumed"]], type = "l", lwd = 2);
text(x = 5, y = 0.79, labels = "b", cex = 2);
box();
par(mar = c(4, 4, 1, 1));
plot(x = population_data_sim2[["time_step"]], type = "n", ylim = c(0, 0.8),
     y = population_data_sim2[["mean_pesticide1_consumed"]], cex.lab = 1.25, 
     ylab = "Mean pesticide 1 consumed", xlab = "Time step");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim2[["time_step"]], 
       y = population_data_sim2[["mean_pesticide1_consumed"]], type = "l", lwd = 2);
box();
text(x = 5, y = 0.78, labels = "c", cex = 2);
par(mar = c(4, 4, 1, 1));
plot(x = population_data_sim2[["time_step"]], type = "n", ylim = c(0, 0.8),
     y = population_data_sim2[["mean_pesticide2_consumed"]], cex.lab = 1.25, 
     ylab = "Mean pesticide 2 consumed", xlab = "Time step");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim2[["time_step"]], 
       y = population_data_sim2[["mean_pesticide2_consumed"]], type = "l", 
       lwd = 2);
box();
text(x = 5, y = 0.78, labels = "d", cex = 2);
```

Figure 10 shows how pest traits evolve over time when pesticides are rotated.

```{r, echo = FALSE, fig.height = 7, fig.width = 7, fig.align = "left", fig.cap = "Mean values of pest traits T1 (crop 1 consumption rate), T2 (crop 2 consumption rate), T3 (pesticide 1 consumption rate), and T4 (pesticide 2 consumption rate) in a population simulated using the resevol R package in which pesticides rotate every 9 time steps. Grey shaded regions show individual crop seasons; pesticides are applied at the start and midpoint of each seasons, beginning at the time step 100."}
par(mfrow = c(2, 2), mar = c(1, 4, 1, 1));
plot(x = population_data_sim2[["time_step"]], type = "n", ylim = c(2, 3.6),
     y = population_data_sim2[["trait1_mean_value"]], cex.lab = 1.25, 
     ylab = "Mean crop 1 consumption trait", xlab = "Time step", xaxt = "n");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim2[["time_step"]], 
     y = population_data_sim2[["trait1_mean_value"]], type = "l", lwd = 2);
text(x = 5, y = 3.56, labels = "a", cex = 2);
box();
par(mar = c(1, 4, 1, 1));
plot(x = population_data_sim2[["time_step"]], type = "n", ylim = c(2, 3.6),
     y = population_data_sim2[["trait2_mean_value"]], cex.lab = 1.25 ,
     ylab = "Mean crop 2 consumption trait", xlab = "Time step", xaxt = "n");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim2[["time_step"]], 
       y = population_data_sim2[["trait2_mean_value"]], type = "l", lwd = 2);
text(x = 5, y = 3.56, labels = "b", cex = 2);
box();
par(mar = c(4, 4, 1, 1));
plot(x = population_data_sim2[["time_step"]], type = "n", ylim = c(-0.8, 1.4),
     y = population_data_sim2[["trait3_mean_value"]], cex.lab = 1.25, 
     ylab = "Mean pesticide 1 consumption trait", xlab = "Time step");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim2[["time_step"]], 
       y = population_data_sim2[["trait3_mean_value"]], type = "l", lwd = 2);
box();
text(x = 5, y = 1.34, labels = "c", cex = 2);
par(mar = c(4, 4, 1, 1));
plot(x = population_data_sim2[["time_step"]], type = "n", ylim = c(-0.8, 1.4),
     y = population_data_sim2[["trait4_mean_value"]], cex.lab = 1.25, 
     ylab = "Mean pesticide 2 consumption trait", xlab = "Time step");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim2[["time_step"]], 
       y = population_data_sim2[["trait4_mean_value"]], type = "l", lwd = 2);
box();
text(x = 5, y = 1.34, labels = "d", cex = 2);
```

We can observe the spatial distribution of pests on the landscape in the final time step (Figure 11).

```{r, echo = FALSE, fig.width=7, fig.height=7, fig.align = "left", fig.cap = "A complex landscape for simulating the ecology of pests and the evolution of pesticide resistance on farmland in the resevol R package. Terrain includes farms (brown colours), grassland (light green), forest (dark green), and water (blue). Black points show the locations of individual pests after 240 time steps for a simulation in which pesticides are not rotated over time."}
last_time_step_file_sim2 <- system.file("last_time_step_sim2.csv", 
                                        package = "resevol");
pop_last_time_step_sim2  <- read.csv(file = last_time_step_file_sim2);
par(mar = c(0, 0, 0, 0));
landscape    <- land_eg;
for(i in 1:dim(pop_last_time_step_sim2)[1]){
    xloc <- pop_last_time_step_sim2[i, 1] + 1;
    yloc <- pop_last_time_step_sim2[i, 2] + 1;
    landscape[xloc, yloc] <- 18;
}
image(landscape, xaxt = "n", yaxt = "n", col = land_cols);
```

Finally, we can examine the trait covariances in the final time step below.

```{r, eval = TRUE, echo = FALSE}
end_trait_covs2 <- cov(pop_last_time_step_sim2[,3:6]);
print(end_trait_covs2);
```

We can use the outputs above to contrast the population dynamics of pests under a regime that lacks (simulation 1) or does not lack (simulation 2) the rotation of pesticide types.

<a name = "conclusion">Conclusion</a>
================================================================================

From the output above, we can conclude that when farms rotate the pesticide that they apply between pesticide 1 and pesticide 2, it has a substantial effect on pest density and evolution.
In the absence of pesticide rotation, the pest population quickly recovers after pesticides are first applied (Figure 4).
Consequently, pesticides have very little effect on the amount of crops consumed (Figure 5a,b), and while there is an initial spike in the consumption of both pesticides (Figure 5c,d), this consumption quickly declines.
Despite the positive covariance between crop consumption and pesticide consumption, and the negative covariances between consumption of crop and pesticide types, these trade-offs appear to have only a modest effect on reducing actual crop consumption and maintaining or increasing pesticide consumption (Figure 6).
Consequently, some farms experience high densities of pests that are resistant to the applied pesticides (Figure 7).

In contrast, when pesticides are rotated every 9 time steps, the pest population density drops after the onset of pesticide application and does not recover (Figure 8).
The amount of each crop consumed per pests also drops (Figure 9a, b), while the amount of pesticide 1 and 2 consumed is maintained at a much higher rate than occurred in the absence of pesticide rotation (Figure 9c, d).
Because selection against pesticide consumption was not consistent for a specific farm, and there is a trade-off between pest consumption of pesticide 1 versus 2, local adaptation to a single pesticide did not occur (Figure 10).
Overall, farms therefore did not experience high densities of pests, and pest resistance to pesticide is well-managed (Figure 11).

Trait covariance differences between simulations 1 and 2 also highlight the effect that pesticide rotation had on pesticide resistance evolution.
When no pesticide rotation occurred, the trade-off between consumption rate of pesticides 1 and 2 was reflected in a trait covariance of `r end_trait_covs[3, 4]`.
But when pesticides were rotated, the covariance was instead `r end_trait_covs2[3, 4]`.
This means that in simulation 1, pests that are highly resistant to pesticide 1 were not highly resistant to pesticide 2, and vice versa.
In other words, the lack of pesticide rotation resulted in local adaptation, with some pests specialising on resistance to one of the two pesticides.
In simulation 2, pests maintained at least some resistance to both pesticides.
Local adaptation was not possible because different pesticides were applied in sequence on any given farm.

This advanced example was not rigorous, but it illustrates how a rigorous simulation of pesticide resistance evolution could be designed with the resevol R package.
To make robust predictions, multiple replicate simulations would need to be run for the same set of simulation conditions (in this case, for simulation 1 and 2) to account for stochastic effects in the model.
For example, we might repeat the above [initalisation](#mining) of pest genomes to obtain 20 genomes with similar covariance structure, then run [simulation 1](#sim1) and [simulation 2](#sim2) 20 times to get more robust predictions about pest population dynamics.
We might also need to consider a range of parameter values if some values are unknown or uncertain (e.g., pest movement or trait means).
To develop theory, we can also contrast simulations with more fundamental differences.
For example, we might test how mating system affects pesticide resistance evolution by running the same set of simulations for `repro = "asexual"`, `repro = "sexual"`, and `repro = "biparental"`.
The resevol R package is therefore a highly flexible and powerful tool for running complex models tailored to specific systems, and for developing theory on pest ecology, evolution, and management.

<a name = "lit">Literature Cited</a>
================================================================================

```{r, echo = FALSE}
suppressWarnings(par(oldpar));
```

### 2022-08-06 notebook/ms/advanced_techniques.Rmd
Update the advanced vignette to fix some spelling errors and parameter errors

@@
 The pest species will include a total of four evolving traits that affect the consumption of two crops and two pesticides.
-The objective of simulations will be to test how pest population size changes and traits evolve given different crop and pesticide rotation regimes.
+The objective of simulations will be to test how pest population size changes and traits evolve given different pesticide rotation regimes.
 Below, we explain how to model this system in detail, including all of the necessary code for reproducing the example within the text.
@@
 In our example, individuals will have four covarying traits resulting from 12 loci.
-There will be 6 internal nodes that map loci values to traits (see the Evolutionary Algorithm explanation for details).
+There will be 4 internal nodes that map loci values to traits (see the Evolutionary Algorithm explanation for details).
 The trait covariance structure will be defined as below.
@@
 Rows and columns 3 and 4 will represent traits underlying the consumption rate of pesticides 1 and 2, respectively. 
-Hence, the variation in pesticide consumption rate is 0.4 for both pesticides, which is lower than it is for crops.
+Hence, the variation in pesticide consumption rate is 0.4 for both pesticides, which is lower than what it is for crops.
 There is a trade-off in the consumption rate of pesticide 1 versus pesticide 2, which is reflected in the covariance of -0.6 in the above matrix.
-Finally, there is a positive covariance rate of 0.2 between all crop and all pesticide consumption rates.
+Finally, there is a positive covariance of 0.2 between all crop and all pesticide consumption rates.
 What this means is that pests that consume crops quickly also tend to consume pesticides quickly, causing a potential trade-off between the beneficial effects of feeding ability and the negative effects of pesticides.
 We can now use the `mine_gmatrix` function to find a network mapping loci to traits that satisfies the above trait covariance structure.
-The options used below in `mine_gmatrix` arguments will lead to a computationally intense (and therefore time-consuming) search, particularly due to the large number of networks in the population (`npsize = 12000`), high number of individuals used to test network stress (`indivs = 2000`), high number of hidden nodes (`layers = 6`), and strict stress criteria (`term_cri = -8`).
+The options used below in `mine_gmatrix` arguments will lead to a computationally intense (and therefore time-consuming) search, particularly due to the large number of networks in the population (`npsize = 12000`), high number of individuals used to test network stress (`indivs = 2000`), high number of hidden nodes (`layers = 4`), strict stress criteria (`term_cri = -8`), and high maximum generation number (`max_gen = 5400`).
 
@@
 Increasing values for both of these parameters can increase run time.
-In particular, fewer internal nodes (`layers`) will greatly reduce the time it takes to find a suitable network.
-The trade-off against this reduced time is that more layers make it possible to model more highly constrained trait covariances.
-Nevertheless, if such constraints are not needed, then only 1-2 `layers` might be necessary.
 
@@
 
-The contents of each list element of `mg` are not important for our purposes, but an explanation is available in the [package documentation](../reference/mine_gmatrix.html). 
+The contents of each list element of `mg` are not important for our purposes, but an explanation is available in the package documentation. 
 What is important is the sixth list element, which holds the estimated covariance structure found by the evolutionary algorithm.
@@
 If we are satisfied with the distribution of stress, then we can accept `mg` as the genome for our initialised pests. 
-The whole contents of `mg` are then passed on to the `run_farm_sim` function, which initialises pests using the genome values in `mg` before running simulations.
-Next, because we want to run a simulation on a customised landscape, we will explain how such a landscape can be generated and inserted into the simulation.
+The whole contents of `mg` will then be passed on to the `run_farm_sim` function, which initialises pests using the genome values in `mg` before running simulations.
+Before doing any of this, because we want to run a simulation on a customised landscape, we will explain how such a landscape can be generated and inserted into the simulation.
 
@@
 
-In the image above, each colour represents a unique farm, and colours correspond to the numbers within the matrix `simple_terrain` (note that the orientation differs between the matrix and the image).
+In Figure 2, each colour represents a unique farm, and colours correspond to the numbers within the matrix `simple_terrain` (note that the orientation differs between the matrix and the image).
 This is a simple example, but matrices can be made large to model highly complex landscapes.
@@
 If terrain types have no effect on pests, then they can effectively model land that is not used for farming.
-We can show this using the matrix `land_eg`, which is also included in `advanced_techniques_eg.rda`.
-The dimensions of the landscape modelled in `land_eg` is 128 by 128 cells, and the landscape includes 14 farms (1-14), grassland (15), forest (16), and water (17).
+We can show this using the matrix `land_eg`, which is also included in the resevol package.
+The dimensions of the landscape modelled in `land_eg` are 128 by 128 cells, and the landscape includes 14 farms (1-14), grassland (15), forest (16), and water (17).
 The code below produces an image that shows how these terrain types are distributed over the landscape (Figure 3).
@@
 
-We will run the same simulated population using the `mg` pest genome and `land_eg` landscape [described above](#land) under two different pesticide regimes. 
-In both regimes, the same pesticide is applied to farms on each side of the river (e.g., farms 1-7 apply pesticide 1 and farms 8-14 apply pesticide 2).
+We will simulate the same population twice using the `mg` pest genome and `land_eg` landscape [described above](#land) under two different pesticide regimes, simulation 1 and simulation 2. 
+In both regimes, the same pesticide is applied to farms that occupy the same side of the river (e.g., farms 1-7 apply pesticide 1 and farms 8-14 apply pesticide 2).
 In simulation 1, the pesticide is never rotated.
@@
 Second, we will repeat the process for simulation 2 and demonstrate how comparisons can be made between the two simulations.
-We will then compare output from the two simulations.
 
@@
 Next, we run [simulation 1](#run_farm_sim1) with the `run_farm_sim` function.
-Finally, we demonstrate how to produce [output](#output_sim1) for simulation 1.
+Finally, we demonstrate how to interpret [output](#output_sim1) for simulation 1.
 
@@
 But we can generate more specific initialisation and rotation types using a vector to initialise crops and pesticides, and using a matrix to define rotation.
-For initialisation, a vector elements correspond to farm number and element values correspond to the crop initialised.
+For initialisation, vector elements correspond to farm number and element values correspond to the crop initialised.
 For our example with 17 unique zones of terrain on the landscape, we can define initialised crop choice as follows.
@@
 
-That is the first 7 farms on one side of the river all start growing crop 1. 
-Farms 8-14 instead all start growing crop 2.
+That is, the first 7 farms on one side of the river all start off growing crop 1. 
+Farms 8-14 all start off growing crop 2.
 Crop 3 is used as a dummy for grassland (15), forest (16), and water (17).
@@
 
-In the above matrix `rotate_crop_sim1`, matrix elements define the probability of transitioning from one crop type (rows) to another crop type (columns).
-Hence, in `rotate_crop_sim1`, farms that have been applying crop 1 will always switch to crop 2, and vice versa.
+In the above `rotate_crop`, matrix elements define the probability of transitioning from one crop type (rows) to another crop type (columns).
+Hence, in `rotate_crop`, farms that have been applying crop 1 will always switch to crop 2, and vice versa.
 Landscape zones applying crop 3 will never transition to any other crop, nor will any other crop transition to type 3.
@@
 
-Next, we define pesticide rotation.
+Next, we define pesticide rotation. Recall that in simulation 1, pesticides will not rotate and farms will maintain the same pesticide type over all time steps.
 
@@
 rotate_pesticide           <- matrix(data = 0, nrow = 3, ncol = 3);
-rotate_pesticide[1, 2]     <- 1;
-rotate_pesticide[2, 1]     <- 1;
+rotate_pesticide[1, 1]     <- 1;
+rotate_pesticide[2, 2]     <- 1;
 rotate_pesticide[3, 3]     <- 1;
@@
 ### <a name = "run_farm_sim1">Running simulation 1</a>
+
+The code below runs simulation 1 using the `run_farm_sim` function.
 
@@
 We next set the mean values of traits at the start of the simulation with the argument `trait_means`. 
-Mean consumption rate of both crops is set to 1, while mean pesticide consumption rate is set to 0.1.
+Mean consumption rate of both crops is set to 2, while mean pesticide consumption rate is set to 0 (note, negative consumption rate values are possible, but are realised as no consumption).
 Pests live up to 6 time steps (`max_age = 6`), and are defined to start feeding upon birth (`min_age_feed = 0`) and stop at age 2 (`max_age_feed = 2`). 
@@
 The `metabolism` argument defines the amount of consumed food lost during the time step, which can affect pests when food consumed affects survival (`food_needed_surv`) and reproduction (`reproduction_type`).
-By setting `reproduction_type = "food_based"` and `food_needed_surv = 0`, we model a system in which pests burn the food consumed from ages 0-2 at a rate of 0.5 units per time step after age 3, then die when their reserves drop below 0.
-By setting `food_needed_repr = 1`, we ensure that only pests that have 1 or more units of food stored can reproduce.
+By setting `reproduction_type = "food_based"` and `food_needed_surv = 1`, we model a system in which pests burn the food consumed from ages 0-2 at a rate of 0.5 units per time step after age 3, then die when their reserves drop below 1.
+By setting `food_needed_repr = 2`, we ensure that only pests that have 2 or more units of food stored can reproduce.
 Note that if we desired, we could also set `food_needed_surv`, `food_needed_repr`, and `metabolism` to be evolving traits.
@@
 All pests are initialised at a random age (`rand_age = TRUE`).
-Pests will die if they consume more than 0.1 total units of pesticide (`pesticide_tolerated_surv = 0.1`).
+Pests will die if they consume more than 2 total units of pesticide (`pesticide_tolerated_surv = 2`).
 In each time step, pests that are able to move will do so an average of 4 times on the landscape (`movement_bouts = 4`), moving up to 2 cells in any direction during each bout (`movement_distance = 2`).
-Recall that since pests consume food and pesticide only between ages 0-2, then start moving at age 3, they no food or pesticide consumption occurs during movement (if applicable, it could be turned on with `feed_while_moving = TRUE` and `pesticide_while_moving = TRUE`).
+Recall that since pests consume food and pesticide only between ages 0-2, then start moving at age 3, no food or pesticide consumption occurs during movement (if applicable, it could be turned on with `feed_while_moving = TRUE` and `pesticide_while_moving = TRUE`).
 Each farm cell produces 10 units of crops (`crop_per_cell = 10` and `crop_sd = 0`) and 1 unit of pesticide (`pesticide_per_cell = 1` and `pesticide_sd = 0`).
-Crops are rotated every 18 time steps (`crop_rotation_time = 18`), and pesticides are set to rotate every 9 time steps (`pesticide_rotation_time = 9`), but this pesticide rotation has no effect because the same pesticide is applied on rotation (as defined by `rotate_pesticide_sim1`).
-This can be conceptualised as modelling a 18 time step growing season in which pesticides are re-applied at the start and halfway point of a season.
+Crops are rotated every 18 time steps (`crop_rotation_time = 18`), and pesticides are set to rotate every 9 time steps (`pesticide_rotation_time = 9`), but this pesticide rotation has no effect in simulation 1 because the same pesticide is applied upon rotation (as defined by `rotate_pesticide`).
+This can be conceptualised as modelling an 18 time step growing season in which pesticides are re-applied at the start and halfway point of a season.
 Note that it is critical to consider that crops are therefore refreshed every 18 time steps, not every time step.
@@
 
-Lastly, we set the number of time steps to `time_steps = 240`, and we set `pesticide_start = 81`, which means that pesticides are not applied at all until after a burn-in of 120 time steps.
+Lastly, we set the number of time steps to `time_steps = 240`, and we set `pesticide_start = 81`, which means that pesticides are not applied at all until after a burn-in of 81 time steps.
 In each time step, an average of 100 immigrants enter the population (`immigration_rate = 100`).
@@
 Pest genome mutation rate and crossover rate at a locus are set to `mutation_pr = 0.01` and `crossover_pr = 0.01`, respectively.
-We print out the dynamics of the evolving population over time, and all of the individual data from the last time step, to a CSV file (individual data are not included in the package due to its size).
-All values not mentioned are set to default parameter values, which can be found in the [package documentation](../reference/run_farm_sim.html).
+We print out the dynamics of the evolving population over time, and all of the individual data from the last time step, to two separate CSV files (individual data are not included in the package due to its size).
+All arguments of `run_farm_sim` not mentioned are set to default parameter values, which can be found in the package documentation.
 
@@
 
-When running the function `run_farm_sim` above, the population size of the pest will be printed in the R console in each time step.
+When running the function `run_farm_sim`, the population size of the pest will be printed in the R console in each time step.
 This is primarily because simulations can take a long time, and printing makes it possible to estimate how much time is remaining.
@@
 This file includes population-level parameters of population size and mean pest age, sex, food consumed, pesticide consumed, mortality rate, and trait values reported for each time step.
-If `get_f_coef = TRUE` in `run_farm_sim`, then mean inbreeding coefficients are also reported.
+If `get_f_coef = TRUE` in `run_farm_sim`, then mean inbreeding coefficients of pests are also reported.
 
@@
 Next, we look at the file that includes the complete information for all individuals in the last time step of the simulation. 
-Using this data set, we can examine the distribution of pests on the landscape, and the covariance of evolving pest traits. 
-Note that information has been removed from the file below to avoid a very large file size.
-But in this output "last_time_step.csv", traits always begin in column 102, meaning that individual trait values for traits 1, 2, 3, and 4 will be in columns 102, 103, 104, and 105, respectively.
+Using this data set, we can examine the distribution of pests on the landscape, and the realised covariance of evolving pest traits. 
+In this output "last_time_step.csv", traits always begin in column 102, meaning that individual trait values for traits 1, 2, 3, and 4 will be in columns 102, 103, 104, and 105, respectively. 
 In this example, the CSV output file "last_time_step.csv" has been renamed "last_time_step_sim1.csv".
@@
 We can plot the distribution over the landscape using the code below.
-The code identifies the x-locations and y-locations of each pest `i` in `pop_last_time_step`, and these landscape cells are replaced with the number 18.
+The code identifies the x-locations and y-locations of each pest `i` in `pop_last_time_step_sim1`, and these landscape cells are replaced with the number 18.
 A new land colour for black `#000000` is then included, and the map is reproduced as an image in R.
@@
 The spatial pattern of pests over the landscape can be useful for visualising pest population dynamics.
-In the case of the above, pest density is higher on some (e.g., 1, 8, 9, 10, and 15) farms than others (e.g., 3, 12, and 13), and this pest density could be interpreted with respect to individual farm crop and pesticide rotation regimes.
+In the case of the above, pest density is higher on some farms (e.g., 1, 8, 9, 10, and 15) than others (e.g., 3, 12, and 13), and this pest density could be interpreted with respect to individual farm crop and pesticide rotation regimes.
 Hence, for the four evolving traits, a covariance matrix of pest traits can be calculated at the end of the simulation.
@@
 
-To avoid repetition, the code for plots will not be included for simulation 2.
+To avoid needless repetition, the code for plots will not be included for simulation 2.
 Figure 8 shows population abundance over time for simulation 2.
@@
 In the absence of pesticide rotation, the pest population quickly recovers after pesticides are first applied (Figure 4).
-Consequently, pesticides have very little effect on the amount of crops consumed (Figure 5a,b), and while there is an initial spike in the consumption of both pesticides (Figure 5c,d), this consumption quickly declines.
+Consequently, pesticides have very little long-term effect on the amount of crops consumed (Figure 5a,b), and while there is an initial spike in the consumption of both pesticides (Figure 5c,d), this consumption quickly declines.
 Despite the positive covariance between crop consumption and pesticide consumption, and the negative covariances between consumption of crop and pesticide types, these trade-offs appear to have only a modest effect on reducing actual crop consumption and maintaining or increasing pesticide consumption (Figure 6).
@@
 Trait covariance differences between simulations 1 and 2 also highlight the effect that pesticide rotation had on pesticide resistance evolution.
-When no pesticide rotation occurred, the trade-off between consumption rate of pesticides 1 and 2 was reflected in a trait covariance of `r end_trait_covs[3, 4]`.
-But when pesticides were rotated, the covariance was instead `r end_trait_covs2[3, 4]`.
+When no pesticide rotation occurred, the trade-off between consumption rate of pesticides 1 and 2 was reflected in a realised trait covariance of `r end_trait_covs[3, 4]`.
+But when pesticides were rotated, the realised covariance was instead `r end_trait_covs2[3, 4]`.
 This means that in simulation 1, pests that are highly resistant to pesticide 1 were not highly resistant to pesticide 2, and vice versa.
@@
 To make robust predictions, multiple replicate simulations would need to be run for the same set of simulation conditions (in this case, for simulation 1 and 2) to account for stochastic effects in the model.
-For example, we might repeat the above [initalisation](#mining) of pest genomes to obtain 20 genomes with similar covariance structure, then run [simulation 1](#sim1) and [simulation 2](#sim2) 20 times to get more robust predictions about pest population dynamics.
+For example, we might repeat the above [initalisation](#mining) of pest genomes to obtain 20 genomes with similar covariance structures, then run [simulation 1](#sim1) and [simulation 2](#sim2) 20 times to get more robust predictions about pest population dynamics.
 We might also need to consider a range of parameter values if some values are unknown or uncertain (e.g., pest movement or trait means).
-To develop theory, we can also contrast simulations with more fundamental differences.
+To develop theory, we could also contrast simulations with more fundamental differences.
 For example, we might test how mating system affects pesticide resistance evolution by running the same set of simulations for `repro = "asexual"`, `repro = "sexual"`, and `repro = "biparental"`.

### 2022-11-19 notebook/ms/advanced_techniques.Rmd
Here I'm committing some change that appears to have been made to the advanced techniques, I believe to the author details

@@
 author: |
-  | A. Bradley Duthie&#xb9;&#xb2;, et al.
+  | A. Bradley Duthie&#xb9; &#179;, Rosie Mangan&#xb9;, Chintamani Rose McKeon&#xb9;,
+  | Matthew C. Tinsley&#xb9;, and Luc F. Bussi&egrave;re&#xb2;
 date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,

### 2022-11-19 notebook/ms/evolutionary_algorithm.Rmd
Here I'm committing some change that appears to have been made to the evolutionary algorithms SI, I believe to the author details

@@
 output:
+  word_document:
+    reference_docx: docx_template.docx
   bookdown::pdf_document2:
@@
     toc: no
-  word_document:
-    reference_docx: docx_template.docx
   pdf_document:

### 2022-11-19 notebook/ms/ms.Rmd
I'm initialising feedback from @luc-bussiere and @MattTinsleyStirling that I have used to create a revision to the manuscript

@@
 author: |
-  | A. Bradley Duthie&#xb9;&#xb2;, et al.
+  | A. Bradley Duthie$^{1,4}$, Rosie Mangan$^{1}$, C. Rose McKeon$^{1}$,
+  | Matthew C. Tinsley$^{1}$, and Luc F. Bussi&egrave;re$^{2,3}$
 date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
-  UK [2]  alexander.duthie@stir.ac.uk, Biological and Environmental Sciences 3A149
+  UK [2]Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden [3] Gothenburgh Global Biodiversity Centre, Gothenburg, Sweden [4] alexander.duthie@stir.ac.uk, Biological and Environmental Sciences 3A149
   University of Stirling Stirling, FK9 4LA, UK'
@@
 3.
-Simulations are modelled on a spatially-explicit and highly customisable landscape in which crop and pesticide application and rotation can vary, making the package a highly flexible tool for both general and targetted models of pest management and resistance evolution.
+Simulations are modelled on a spatially-explicit and highly customisable landscape in which crop and pesticide application and rotation can vary, making the package a highly flexible tool for both general and tactical models of pest management and resistance evolution.
 4. We present the key features of the resevol package and demonstrate its use for a simple example simulating pests with two covarying traits.
@@
 The resevol package applies individual-based modelling and a quantitative genetics approach to simulate the evolution of a pest population on a changing landscape. 
-A focal goal of the software is to model traits with a pre-specified, but potentially evolving, covariance structure. 
+Multiple traits determine the overall fitness of any pest genotype. 
+For example, while alleles conferring resistance to a particular pesticide enhance fitness in the presence of that pesticide, they could also impose reproductive fitness costs.
+Such trade-offs can be quantified by the genetic covariance between traits.
+A focal goal of the software is to model traits with a pre-specified, but potentially evolving, genetic covariance structure. 
 To achieve this goal, each individual has a genome with $L$ loci that underlie a set of $T$ potentially evolving traits. 
@@
 Second, a population of asexual or sexual individuals is initialised and simulated on a spatially explicit landscape separated into distinct units (e.g., farms).
-Land units can apply one of up to 10 pesticides, and one of up to 10 crops at a given time.
+Land units can apply one of up to 10 pesticides, and one of up to 10 landscape types at a given time (for simplicity, here we interpret landscape types to be crop types).
 Pesticides and crops rotate independently within farms over time in a pre-specified way.
@@
 In the resevol package, the `mine_gmatrix` function runs an evolutionary algorithm and requires the argument `gmatrix`, which specifies the desired trait covariance matrix.
-The function initialises a population of `npsize` networks (Figure \@ref(fig:network)), and this population evolves until some maximum generation (`max_gen`) or minimum expected network stress (`term_cri`) is met.
-In a single generation of the algorithm, values crossover between networks and mutate.
+The function initialises a population of `npsize` separate, and potentially unique, networks (i.e., `npsize` copies of a network like the one shown in Figure \@ref(fig:network)), and this population evolves until some maximum iteration (`max_gen`) or minimum expected network stress (`term_cri`) is met.
+In a single iteration of the algorithm, values mutate and crossover occurs between networks (i.e., some fo the `npsize` networks swap values with some probability).
 Next, trait covariances produced for each network are estimated by initalising `indivs` individuals with loci sampled from a standard normal distribution.
 Network stress is calculated as the logged mean squared deviation between estimated covariances and those in `gmatrix`. 
-Tournament selection [@Hamblin2013] is then used to determine the networks for the next generation of the algorithm.
-Networks with the lowest estimated stress have the highest fitness, so these networks are disproportionately represented in the next generation.
+Tournament selection [@Hamblin2013] is then used to determine the networks for the next iteration of the algorithm.
+Networks with the lowest estimated stress have the highest fitness, so these networks are disproportionately represented in the next iteration.
 Throughout the evolutionary algorithm, the lowest stress network is saved and returned upon network termination.
@@
 Initialisation can be random for each farm with equal probability, or it can be set using a vector of length `farms` in which vector elements define the initialised crop or pesticide number.
-After initialisation, crops and pesticides rotate once ever `crop_rotation_time` and `pesticide_rotation_time` time steps, respectively.
+After initialisation, crops and pesticides rotate once every `crop_rotation_time` and `pesticide_rotation_time` time steps, respectively.
 Each of these rotation time arguments can take either an integer value from 1-3, or a matrix.
@@
 In each time step, pests feed in a random order, consuming the crop on their landscape cell.
-Pests consume an amount of the crop as specified by the `food_consume` argument, which takes a vector with as many elements as their are crops (i.e., if `crop_number = 2`, then `food_consume` has two elements, the first and second defining consumption of crops 1 and 2, respectively).
+Each pest consumes an amount of crop as specified by the `food_consume` argument, which takes a vector with as many elements as their are crops (i.e., if `crop_number = 2`, then `food_consume` has two elements, the first and second defining consumption of crops 1 and 2, respectively).
 If crop amount on the landscape cell exceeds pest consumption ability, then pests consume their maximum amount, and this amount is removed from the landscape cell.
@@
 Pesticide consumption works identically to crop consumption, except that the amount of pesticide on a landscape cell is not decreased.
+Hence, each pest can potentially feed and be affected by the pesticide of their focal landscape cell. 
 Pests simply consume an amount of pesticide as specified by the `pesticide_consume` argument, which also takes a vector with as many elements as their are pesticides.
-Hence, each pest can potentially feed and be affected by the pesticide of their focal landscape cell.
+Pesticide consumption affects pest survival and reproduction using `pesticide_tolerated_surv` and `pesticide_tolerated_repr` arguments.
 
@@
 Each diploid parent contributes one half of their genome, effectively modelling a genome with a single chromosome.
-Crossover occurs at each position of the genome with a probability of `crossover_pr`, so complete recombination is also possible if `crossover_pr = 0.5`.
+Crossover occurs at each position of the genome with a probability of `crossover_pr`.
+When a crossover event occurs, alleles are swapped between the two chromosomes, so complete recombination is also possible if `crossover_pr = 0.5`.
 For both haploids and diploids, each genome value then mutates independently with a probability of `mutation_pr`, which can be set to any real number from 0-1.
@@
 In resevol v0.3, the `run_farm_sim` function includes 68 arguments, which specify a wide range of possible simulation conditions affecting landscape and pest characteristics.
-These arguments are explained the package documentation, and in Supporting Information 2, which demonstrates an advanced case study with a custom landscape and complex pest genomes and life history.
+These arguments are explained in the package documentation, and in Supporting Information 2, which demonstrates an advanced case study with a custom landscape and complex pest genomes and life history.
 Here we focus on a simple simulation with asexually reproducing pests that have three loci and two traits (Figure \@ref(fig:network)).
@@
 
-Traits 1 and 2 will define the rate at which pests consume two separate pesticides, so we model a system in which there is a potential trade-off for pesticide susceptibility.
+Traits 1 and 2 will define the realised rate of uptake of the two separate pesticides, so we model a system in which there is a potential trade-off for pesticide susceptibility.
 We use a simple $64 \times 64$ cell landscape with nine farms. 
@@
 Pests have a maximum age of four, and they feed in ages 0-2, move up to two cells in ages 3-4, and reproduce in age four.
-Pests can consume pesticides in ages 0-2, and if they consume any pesticide, then they will die.
+Pests can uptake (i.e., 'consume') pesticides in ages 0-2, and if they consume any pesticide, then they will die.
 In each time step, a mean of 10 immigrants arrive.
@@
 ```{r overtime, echo=FALSE, fig.width=8, fig.cap = "Agricultural pest ecological and evolutionary dynamics over 160 time steps from an individual-based simulation using the resevol R package.
-Panels show (a) pesticide abundance change, (b) mean of pesticides 1 and 2 consumed per pest, (c) mean food consumed per pest, and (d) mean values of evolving traits 1 and 2 over time.
-Traits 1 and 2 define how much of pesticides 1 and 2, respectively, a pest will consume.
+Panels show (a) pesticide abundance change, (b) mean realised amount of pesticides 1 and 2 uptaken per pest, (c) mean food consumed per pest, and (d) mean value of evolving traits 1 and 2 underlying pest uptake over time.
+Note that only pests with positive values for Traits 1 or 2 can uptake Pesticide 1 or 2, respectively.
+Pests with negative trait values will be unaffected by corresponding pesticides (i.e., pesticide consumption is a threshold trait), hence the difference between realised pesticide consumption (b) and the traits underlying it (d).
 White and grey vertical stripes indicate seasons of a single crop and pesticide application."}
@@
 Agricultural landscapes and pest life histories are highly customisable, allowing targetted models that can simulate specific real-world case studies (Supporting Information 2).
-The breadth of possible models that can be simulated with resevol also makes it a useful tool for developing theory on pest management, and more even generally on the evolution and ecology of individuals with complex traits on a changing landscape.
+The breadth of possible models that can be simulated with resevol also makes it a useful tool for developing theory on pest management, and even more generally on the evolution and ecology of individuals with complex traits on a heterogeneous landscape.
 
@@
 ABD wrote and maintains package code. 
-RM contributed code for data visualisation.
+CRM contributed code for data visualisation.
 

### 2022-11-24 R/sim_farming.R
Add a new feature in the package to allow landscape cell values to increase (or decrease) thereby simulating crop growth over the course of a season (between crop rotation times), which should resolve Issue #47 and address the comment of @MattTinsleyStirling in the draft manuscript. All the tests indicate that this is functional (I've written two automated tests), and all checks are passed

@@ -310,6 +310,19 @@
 #' and the simulation entirely, so they are effectively no longer recorded. A 
 #' reflective edge causes pests to bounce at the edge back in the direction from
 #' which they came. The default and recommended edge is a torus.
+#'@param crop_growth This sets the amount by which landscape cell values
+#' increase for each crop from one time step to the next. This might be used to
+#' model the increase in biomass caused by plant growth on the landscape cell.
+#' The way in which growth is enacted is set by crop_growth_type. Values for
+#' crop_growth can either be a single number (in which case, the growth amount
+#' applies to all crop types) or a vector of the same length as crop_types (in
+#' which case, vector indices give the amount of growth for each crop number).
+#'@param crop_growth_type This clarifies whether crop growth should be 
+#' "none" (default), "geometric", or "linear". If geometric, then the value of a
+#' landscape cell increases to cell_value * (1 + crop_growth). If linear, then
+#' a fixed value crop_growth is added to the existing cell value. Note that for
+#' linear growth, this fixed value is only added if the existing cell value is
+#' greater than zero (else it is assumed that there is no crop to grow).
 #'@return The output in the R console is a list with two elements; the first 
 #'element is a vector of parameter values used by the model, and the second 
 #'element is the landscape in the simulation. The most relevant output will be
@@ -403,7 +416,9 @@ run_farm_sim <- function(mine_output,
                          max_age_metabolism  = 9,
                          terrain             = NA,
                          trait_means         = NULL,
-                         land_edge           = "torus"){
+                         land_edge           = "torus",
+                         crop_growth         = 0,
+                         crop_growth_type    = "none"){
   
     if(is.na(terrain)[1] == FALSE){
         xdim  <- dim(terrain)[1];
@@ -488,6 +503,9 @@ run_farm_sim <- function(mine_output,
     }
     if(pesticide_number > 10){
         stop("Cannot have more than 10 pesticides");
+    }
+    if(length(crop_growth) != 1 & length(crop_growth) != crop_number){
+        stop("crop_growth must be of length 1 or 'crop_number'.")
     }
     
     food_consume <- as.list(food_consume);
@@ -594,7 +612,9 @@ run_farm_sim <- function(mine_output,
[…]

### 2022-11-24 src/land_change.c

@@ -293,6 +293,102 @@ void change_crop_choice(double **C_init, double **C_change, double *paras){
     }
 }
 
+
+/* =============================================================================
+ * Grow the crop amount on each cell
+ *     land:  The landscape array to be adjusted
+ *     grow:  The vector defining how much growth per crop per time step
+ *     paras: The paras vector that holds global information
+ * ========================================================================== */
+void grow_crops(double ***land, double *grow, double *paras){
+    
+    int i, j, xdim, ydim, crop_change_type, crop_number;
+    int crop1_col, crop2_col, crop3_col, crop4_col, crop5_col, crop6_col;
+    int crop7_col, crop8_col, crop9_col, crop10_col;
+    double *grow_temp;
+    
+    grow_temp   = (double *) malloc(10 * sizeof(double));
+    
+    xdim             = (int) paras[103];
+    ydim             = (int) paras[104];
+    crop1_col        = (int) paras[118];
+    crop2_col        = (int) paras[119];
+    crop3_col        = (int) paras[120];
+    crop4_col        = (int) paras[121];
+    crop5_col        = (int) paras[122];
+    crop6_col        = (int) paras[123];
+    crop7_col        = (int) paras[124];
+    crop8_col        = (int) paras[125];
+    crop9_col        = (int) paras[126];
+    crop10_col       = (int) paras[127];
+    crop_number      = (int) paras[156];
+    crop_change_type = (int) paras[173];
+    
+    for(i = 0; i < 10; i++){
+        if(i < crop_number){
+            grow_temp[i] = grow[i];
+        }else{
+            grow_temp[i] = 0.0;
+        }
+    }
+    
+    if(crop_change_type == 1){
+        for(i = 0; i < xdim; i++){
+            for(j = 0; j < ydim; j++){
+                land[i][j][crop1_col]  *= (1 + grow_temp[0]);
+                land[i][j][crop2_col]  *= (1 + grow_temp[1]);
+                land[i][j][crop3_col]  *= (1 + grow_temp[2]);
+                land[i][j][crop4_col]  *= (1 + grow_temp[3]);
+                land[i][j][crop5_col]  *= (1 + grow_temp[4]);
+                land[i][j][crop6_col]  *= (1 + grow_temp[5]);
+                land[i][j][crop7_col]  *= (1 + grow_temp[6]);
+                land[i][j][crop8_col]  *= (1 + grow_temp[7]);
[…]

### 2022-11-24 src/land_change.h

@@ -18,5 +18,8 @@ void init_crop(double ***land, double *paras, double **C_init);
 
 void change_crop_choice(double **C_init, double **C_change, double *paras);
 
+void grow_crops(double ***land, double *grow, double *paras);
+
 void land_change(double ***land, double *paras, int ts, double **C_init,
-                 double **C_change, double **P_init, double **P_change);
+                 double **C_change, double **P_init, double **P_change,
+                 double *grow);

### 2022-11-24 src/resevol_init.c

@@ -10,12 +10,12 @@
 /* .Call calls */
 extern SEXP build_ownership(SEXP);
 extern SEXP mine_gmatrix(SEXP, SEXP);
-extern SEXP sim_farming(SEXP, SEXP, SEXP, SEXP, SEXP, SEXP, SEXP);
+extern SEXP sim_farming(SEXP, SEXP, SEXP, SEXP, SEXP, SEXP, SEXP, SEXP);
 
 static const R_CallMethodDef CallEntries[] = {
     {"build_ownership",         (DL_FUNC) &build_ownership,     1},
     {"mine_gmatrix",            (DL_FUNC) &mine_gmatrix,        2},
-    {"sim_farming",             (DL_FUNC) &sim_farming,         7},
+    {"sim_farming",             (DL_FUNC) &sim_farming,         8},
     {NULL, NULL, 0}
 };
 

### 2022-11-24 src/sim_farming.c

@@ -22,9 +22,10 @@
  *      PROT:    Pesticide rotation matrix
  *      CINIT:   Initial crop choice for each farmer
  *      PINIT:   Initial pesticide choice for each farmer
+ *      CGROW:   Rate of growth of each crop type
  * ===========================================================================*/
 SEXP sim_farming(SEXP IND, SEXP LAND, SEXP PARAS, SEXP CROT, SEXP PROT,
-                 SEXP CINIT, SEXP PINIT){
+                 SEXP CINIT, SEXP PINIT, SEXP CGROW){
  
     /* SOME STANDARD DECLARATIONS OF KEY VARIABLES AND POINTERS               */
     /* ====================================================================== */
@@ -44,6 +45,7 @@ SEXP sim_farming(SEXP IND, SEXP LAND, SEXP PARAS, SEXP CROT, SEXP PROT,
     int    immigrants;       /* Immigrants in a time step */
     int    protected_n;      /* Number of protected R objects */
     int    len_PARAS;        /* Length of the parameters vector */
+    int    len_CGROW;        /* Length of the crop growth vector */
     int    print_gen;        /* Should the generations be printed */
     int    get_stats;        /* Should print a CSV with statistics */
     int    *dim_IND;         /* Dimensions of the individual array */
@@ -61,7 +63,9 @@ SEXP sim_farming(SEXP IND, SEXP LAND, SEXP PARAS, SEXP CROT, SEXP PROT,
     double *PROT_ptr;
     double *CINIT_ptr;
     double *PINIT_ptr;
+    double *CGROW_ptr;
     double *paras;
+    double *grow;          /* Vector on crop growth per time step */
     double **pests;        /* The pests array */
     double **offspring;    /* The offspring of pests within a time step */
     double **new_pests;    /* The pest array at the end of a time step */
@@ -112,7 +116,12 @@ SEXP sim_farming(SEXP IND, SEXP LAND, SEXP PARAS, SEXP CROT, SEXP PROT,
     protected_n++;
     PINIT_ptr = REAL(PINIT);
     
+    PROTECT( CGROW = AS_NUMERIC(CGROW) );
+    protected_n++;
+    CGROW_ptr = REAL(CGROW);
+    
     len_PARAS   = GET_LENGTH(PARAS);
[…]

### 2022-11-24 tests/testthat/test-sim_farming.R

@@ -41,7 +41,7 @@ test_that("Simulations are run and output is produced", {
                               pesticide_start = 0, immigration_rate = 0,
                               get_f_coef = FALSE, get_stats = FALSE,
                               metabolism = 0);
-    expect_equal(length(sim[[1]]), 173);
+    expect_equal(length(sim[[1]]), 174);
     expect_equal(dim(sim[[2]]), c(40, 40, 21));
 })
 
@@ -85,7 +85,7 @@ test_that("Movement with a reflective boundary", {
                               pesticide_start = 0, immigration_rate = 0,
                               get_f_coef = FALSE, get_stats = FALSE,
                               metabolism = 0, land_edge = "reflect");
-    expect_equal(length(sim[[1]]), 173);
+    expect_equal(length(sim[[1]]), 174);
     expect_equal(dim(sim[[2]]), c(40, 40, 21));
 })
 
@@ -129,7 +129,7 @@ test_that("Movement with a leaky boundary", {
                               pesticide_start = 0, immigration_rate = 0,
                               get_f_coef = FALSE, get_stats = FALSE,
                               metabolism = 0, land_edge = "leaky");
-    expect_equal(length(sim[[1]]), 173);
+    expect_equal(length(sim[[1]]), 174);
     expect_equal(dim(sim[[2]]), c(40, 40, 21));
 })
 
@@ -172,6 +172,107 @@ test_that("Movement with a sticky boundary", {
                               pesticide_start = 0, immigration_rate = 0,
                               get_f_coef = FALSE, get_stats = FALSE,
                               metabolism = 0, land_edge = "sticky");
-    expect_equal(length(sim[[1]]), 173);
+    expect_equal(length(sim[[1]]), 174);
     expect_equal(dim(sim[[2]]), c(40, 40, 21));
 })
+
+test_that("Geometric crop growth", {
+    skip_on_cran();
+    gmt       <- matrix(data = 0, nrow = 2, ncol = 2);
+    diag(gmt) <- 1;
+    mg        <- mine_gmatrix(gmatrix = gmt, loci = 4, layers = 2, indivs = 100, 
+                              npsize = 100, max_gen = 4, prnt_out = FALSE);
+    sim       <- run_farm_sim(mine_output = mg, N = 100, xdim = 10, ydim = 10, 
+                              repro = "sexual", neutral_loci = 1000, 
+                              max_age = 30, min_age_move = 0, max_age_move = 30,
[…]

### 2022-11-24 notebook/index.Rmd
Record the latest changes regarding Issue #47 in the journal and update the news and Description file

@@
 
+   - **NOV** [24](#u-24-NOV-2022)
    - **JUL** [19](#u-19-JUL-2022), [30](#u-30-JUL-2022)
@@
 ================================================================================
+
+> <a name="u-24-NOV-2022">Update: 24 NOV 2022</a>
+
+During the revision process of the software note, Matt suggested another feature that I really should have thought of making earlier.
+In version 0.3.3.0, after crop rotation, landscape cell values are initialised with whatever value is used for `crop_per_cell`.
+This means that there is no way to simulate crop growth on a landscape in between crop rotations.
+In version 0.3.4.0, I have fixed this with two new arguments, `crop_growth` and `crop_growth_type`.
+
+The argument `crop_growth` indicates how much a cell value representing a given crop should change from one time step to the next (before the potential effects of feeding by pests).
+By default, this value is `crop_growth = 0`, meaning that all biomass on a cell will be initialised at the start of a new crop rotation.
+Nonzero values will affect the change of a cell's crop value in a way dependent up on `crop_growth_type`.
+Note that `crop_growth` can either take a single value, or a vector of the same length as `crop_number`.
+If, for example, there are two crops (`crop_number = 2`), then we could set `crop_growth = c(0.1, 0.4)`, meaning that crop 1 would increase by a value of 0.1 each time step, and crop 2 would increase by a value of 0.4 each time step.
+If we set a `crop_growth` to a single value (e.g., `crop_growth = 0.1`), then all crops grow at the same rate of `crop_growth`.
+
+The argument `crop_growth_type` specifies how growth is enacted (i.e., how the value on the landscape changes).
+The default option is `crop_growth_type = "none"`, in which case, no growth occurs.
+If `crop_growth_type = "geometric"`, then the value of `crop_growth` specifies the proportion by which the landscape value increases in a time step,
+
+$$value_{t+1} = value_{t} \left(1 + crop\_growth \right).$$
+
+So if we initialise a crop amount at 2 on the landscape cell, and `crop_growth = 0.1`, then in the next time step (unless some crop is eaten by pests), we can predict that the crop amount will be `r 2 * (1 + 0.1)` (i.e., an increase of 10 per cent).
+
+If `crop_growth_type = "linear"`, then the increment `crop_growth` is simply added to the existing cell value in each time step.
+
+$$value_{t+1} = value_{t} + crop\_growth.$$
+
+My hope is that these options will be useful for a lot of purposes, but there is plenty of room for expanding the package with different rules for crop growth.
 

### 2022-11-24 notebook/ms/ms.Rmd
Add in a quick new paragraph in the manuscript to address comments from @MattTinsleyStirling and resolve Issue #47

@@
 First, a landscape can be built from the arguments `xdim`, `ydim`, and `farms`. 
-These arguments specify the dimensions of the landscape and the number of farms on it.
+These arguments specify the cell dimensions and number of farms on the landscape.
 Contiguous rectangular farms of roughly equal size are generated on the landscape using a splitline algorithm.
@@
 Hence, any possible Markov chain can be used to model transition between crop or pesticide types on farms.
+Upon rotation, crop and pesticide values are reset on each farm.
+
+Values for `crop_per_cell` and `pesticide_per_cell` determine the quantity of crops and pesticides initialised per cell upon crop or pesticide rotation, respectively.
+Between crop rotations, crop values can increase each time step by a proportion or increment `crop_growth`, which depends on `crop_growth_type`.
+These arguments can be used to model crop growth over a season.
 

### 2023-07-17 notebook/presentation/BES2022.Rmd
Presentation for British Ecological Society 2022

---
title: "resevol: an R package for spatially explicit models of pesticide resistance
  given evolving pest genomes"
author: |
  | A. Bradley Duthie$^{1}$, Rosie Mangan$^{1}$, C. Rose McKeon$^{1}$,
  | Matthew C. Tinsley$^{1}$, and Luc F. Bussi&egrave;re$^{2,3}$
date: '[1] University of Stirling [2] University of Gothenburg [3] Gothenburgh Global Biodiversity Centre alexander.duthie@stir.ac.uk, @bradduthie@ecoevo.social http://bradduthie.github.io/talks/BES2022.pdf'
output:
  beamer_presentation:
    theme: "default"
    colortheme: "default"
    fonttheme: "default"
  ioslides_presentation: default
  slidy_presentation: default
header-includes:
- \usepackage{hyperref}
- \usepackage{amsmath}
colorlinks: true
linkcolor: blue
urlcolor: blue
---

## Pesticide resistance evolution

\Large

\begin{columns}
\begin{column}{0.3\textwidth}

```{r, out.width = "100%", echo = FALSE}
library("knitr");
include_graphics("img/part_1.png");
```

\end{column}

\begin{column}{0.7\textwidth}

\textbf{Impact on agriculture}

\begin{itemize}
\item Resistance wicked \& widespread$^{1}$
\item Obstacle to food security$^{2}$
\end{itemize}

\vspace{2mm}\pause

\textbf{Manage resistance evolution}

\begin{itemize}
\item Vary biopesticide application$^{3,4}$
\item Polygenic resistance traits$^{5}$
\item Negative cross-resistance$^{6}$
\end{itemize}

\end{column}
\end{columns}

\footnotetext[1]{Gould et al. 2018. \textit{Ecol. Appl.} 13:1791-1805.}
\footnotetext[2]{Bradshaw et al. 2016. \textit{Nat. Commun.} 7:12986.}
\footnotetext[3]{Carroll et al. 2014. \textit{Science} 346:1245993.}
\footnotetext[4]{Saikai et al. 2021. \textit{Pest Manag. Sci.} 77:273-284.}
\footnotetext[5]{Boots, M. 2011. \textit{Am. Nat.} 178:214-220.}
\footnotetext[6]{Via, S. 1986. \textit{Pesticide resistance}. 222-235.}

## Agricultural heterogeneity to overcome resistance evolution

```{r, echo = FALSE, fig.width = 6.2, fig.height = 3.1}
library("resevol");
land           <- make_landscape(rows = 100, cols = 100, farms = 9);    
landscape      <- land[,,1];
landscape_vals <- unique(as.vector(landscape));
reord_land     <- sample(landscape_vals, size = length(landscape_vals));
for(i in 1:dim(landscape)[1]){
    for(j in 1:dim(landscape)[2]){
        old_val         <- landscape[i, j];
        new_val         <- reord_land[old_val];
        landscape[i, j] <- new_val;
    }
}
land_cols      <- hcl.colors(9, "YlOrRd", rev = TRUE);
par(mfrow = c(1, 2), mar = c(0, 0.1, 0, 0.1));
image(landscape, xaxt = "n", yaxt = "n", col = land_cols);
box();
pesticid_A         <- rbinom(n = 9, size = 1, prob = 0.5);
pesticide_cols     <- c("grey20", "grey60");
land_pesticide     <- landscape;
for(i in 1:dim(landscape)[1]){
    for(j in 1:dim(landscape)[2]){
        farm_number <- landscape[i, j];
        if(pesticid_A[farm_number] > 0){
            land_pesticide[i, j] <- 0;
        }else{
            land_pesticide[i, j] <- 1;
        }
    }
}
image(land_pesticide, xaxt = "n", yaxt = "n", col = pesticide_cols);
box();
```

\LARGE

\begin{columns}
\begin{column}{0.5\textwidth} 
\begin{center}
Unique farms
\end{center}
\end{column}
\begin{column}{0.5\textwidth} 
\begin{center}
Biopesticide applied
\end{center}
\end{column}
\end{columns}

##

```{r, out.width = "100%", echo = FALSE}
include_graphics("img/Endorse_project.png");
```

##

```{r, out.width = "100%", echo = FALSE}
include_graphics("img/Endorse_people2.png");
```

## Modelling as a proof of concept before scaling up

```{r, out.width = "100%", echo = FALSE}
include_graphics("img/real_farmland.jpg");
```

\footnotetext[1]{Image: \href{https://commons.wikimedia.org/wiki/File:ISS022-E-83025_-_View_of_Earth.jpg}{Public Domain}}

## Agent-based modelling

\Large

\textbf{Simulate a complex system \textit{in silico} with code}
\vspace{8mm}
\normalsize

\begin{columns}
\begin{column}{0.5\textwidth}

```{r, out.width = "100%", echo = FALSE}
par(mar = c(0, 0, 0, 0));
land_cols <- c(land_cols, "black")
xx <- sample(x = 1:100, size = 1000, replace = TRUE);
yy <- sample(x = 1:100, size = 1000, replace = TRUE);
for(i in 1:1000){
    x_pt <- xx[i];
    y_pt <- yy[i];
    landscape[x_pt, y_pt] <- 10;
}
image(landscape, xaxt = "n", yaxt = "n", col = land_cols);
```

\end{column}

\begin{column}{0.5\textwidth} 
\begin{itemize}
\item Agents as discrete entities
\item Individual variation
\item Spatially explicit landscapes
\item Stochastic processes
\item High model complexity
\end{itemize}
\end{column}

\end{columns}

## resevol: An R package for modelling agricultural pests

```{r, out.width = "100%", echo = FALSE}
include_graphics("img/website.png");
```

\footnotetext[1]{Duthie AB, McKeon CR (2022). resevol: Simulate Agricultural Production and Evolution of Pesticide Resistance. R package version 0.3.3.0, https://bradduthie.github.io/resevol/. }

## resevol: An R package for modelling agricultural pests

\begin{columns}
\begin{column}{0.7\textwidth} 
\textbf{Complex individuals}
\begin{itemize}
\item Complete genomes
\item Highly polygenic traits
\item Set trait covariances
\item Flexible life-history
\item Flexible mating system
\end{itemize}
\end{column}
\begin{column}{0.3\textwidth}

```{r, out.width = "70%", echo = FALSE}
include_graphics("img/individual.png");
```

\end{column}

\end{columns}

\footnotetext[1]{\textbf{Image}: \href{https://commons.wikimedia.org/wiki/File:DNA_-_The_Noun_Project.svg
}{Public domain}}

## resevol: An R package for modelling agricultural pests

\begin{columns}
\begin{column}{0.5\textwidth} 
\textbf{Complex individuals}
\begin{itemize}
\item Complete genomes
\item Highly polygenic traits
\item Set trait covariances
\item Flexible life-history
\item Flexible mating system
\end{itemize}
\end{column}
\begin{column}{0.5\textwidth} 
\textbf{Dynamic landscapes}
\begin{itemize}
\item Custom land dimensions
\item Custom pesticide rotation
\item Custom crop rotation
\item Individual farms tracked
\item Raster landscape maps
\end{itemize}
\end{column}

\end{columns}

\footnotetext[1]{Duthie AB, McKeon CR (2022). resevol: Simulate Agricultural Production and Evolution of Pesticide Resistance. R package version 0.3.3.0, https://bradduthie.github.io/resevol/.}
\footnotetext[2]{Duthie, AB, R Mangan, CR McKeon, MC Tinsley, LF Bussi\`{e}re. 2022. resevol: an R package for spatially explicit models of pesticide resistance given evolving pest genomes. bioRxiv 2022.08.22.504740; doi: https://doi.org/10.1101/2022.08.22.504740}

## A genetic architecture for pre-specified trait covariances

\vspace{6mm}

```{r, echo = FALSE, fig.width = 6, fig.height = 4}
par(mar = c(0.2, 0.2, 0, 0.2));
plot(x = 0, y = 0, type = "n", xlim = c(0, 1000), ylim = c(0, 1000), 
     xaxt = "n", yaxt = "n", bty = "n", xlab = "", ylab = "");
mtext(side = 1, text = "Internal nodes", cex = 1.25, col = "blue", line = -3);
mtext(side = 2, text = "Loci", cex = 1.25, col = "darkgreen", line = -2);
mtext(side = 4, text = "Traits", cex = 1.25, col = "red", line = -2);
Lx  <- rep(x = 100, times = 3);
Ly  <- seq(from = 250, to = 750, length = 3);
H1x <- rep(x = 400, times = 2);
H1y <- seq(from = 250, to = 750, length = 2);
H2x <- rep(x = 600, times = 2);
H2y <- seq(from = 250, to = 750, length = 2);
Tx  <- rep(x = 900, times = 2);
Ty  <- seq(from = 250, to = 750, length = 2);
points(x = Tx, y = Ty, pch = 23, cex = 8, bg = "red", lwd = 2);
sep_L_points <- c(-50, 0, 50);
sep_N_points <- c(-30, 30);
arrows(x0 = H2x + 52, x1 = Tx[1] - 58, y0 = H2y, y1 = Ty[1] + sep_N_points, 
       length = 0.08, lwd = 3);
arrows(x0 = H2x + 52, x1 = Tx[2] - 58, y0 = H2y, y1 = Ty[2] + sep_N_points, 
       length = 0.08, lwd = 3);
arrows(x0 = Lx + 38, x1 = H1x[1] - 58, y0 = Ly, y1 = H1y[1] + sep_L_points, 
       length = 0.08, lwd = 3);
points(x = H2x, y = H2y, pch = 22, cex = 9, bg = "blue", lwd = 2);
arrows(x0 = H1x + 52, x1 = H2x[1] - 58, y0 = H1y, y1 = H2y[1] + sep_N_points, 
       length = 0.08, lwd = 3);
arrows(x0 = H1x + 52, x1 = H2x[2] - 58, y0 = H1y, y1 = H2y[2] + sep_N_points, 
       length = 0.08, lwd = 3);
points(x = H1x, y = H1y, pch = 22, cex = 9, bg = "blue", lwd = 2);
arrows(x0 = Lx + 38, x1 = H1x[2] - 58, y0 = Ly, y1 = H1y[2] + sep_L_points, 
       length = 0.08, lwd = 3);
points(x = Lx, y = Ly, pch = 21, cex = 6, bg = "darkgreen", lwd = 2);
srnm <- c(0.2, -0.1, 0.1);
text(x = Lx, y = Ly, labels = srnm, col = "white", cex = 0.9);
text(x = 200, y = 790, labels = "0.4", srt = 10, cex = 0.8);
text(x = 200, y = 660, labels = "0.1", srt = -58, cex = 0.8);
text(x = 270, y = 695, labels = "-1.2", srt = 46, cex = 0.8);
text(x = 180, y = 480, labels = "-0.6", srt = -40, cex = 0.8);
text(x = 160, y = 340, labels = "0.2", srt = 54, cex = 0.8);
text(x = 220, y = 255, labels = "2.2", srt = -12, cex = 0.8);
text(x = 490, y = 790, labels = "1.3", srt = 10, cex = 0.8);
text(x = 538, y = 410, labels = "-0.5", srt = -76, cex = 0.8);
text(x = 460, y = 390, labels = "-0.1", srt = 72, cex = 0.8);
text(x = 495, y = 260, labels = "0.0", srt = -12, cex = 0.8);
text(x = 720, y = 790, labels = "2.2", srt = 8, cex = 0.8);
text(x = 720, y = 630, labels = "0.9", srt = -58, cex = 0.8);
text(x = 690, y = 400, labels = "1.6", srt = 54, cex = 0.8);
text(x = 740, y = 260, labels = "0.8", srt = -8, cex = 0.8);
Loci  <- matrix(data = c(0.1, -0.1, 0.2), nrow = 1);
Hdn1  <- matrix(data = c(0.4, -1.2, 0.2, 0.1, -0.6, 2.2), nrow = 3, ncol = 2);
Hdn2  <- matrix(data = c(1.3, -0.1, -0.5, 0.0), nrow = 2, ncol = 2);
Hdn3  <- matrix(data = c(2.2, 1.6, 0.9, 0.8), nrow = 2, ncol = 2);
Tvals <- Loci %*% Hdn1 %*% Hdn2 %*% Hdn3;
text(x = Tx, y = Ty, labels = rev(Tvals), cex = 0.8);
```

\footnotetext[1]{resevol R package vignette: \href{https://bradduthie.github.io/resevol/articles/evolutionary_algorithm.html}{https://bradduthie.github.io/resevol/articles/evolutionary\_algorithm.html}}

## Evolutionary algorithm to find genetic architecture

\Large

Initialise a population of multiple potential networks from loci to traits.

\vspace{6mm}

```{r, echo=FALSE, fig.height=1.75, fig.width=6}
mbox <- function(x0, x1, y0, y1){
    xx <- seq(from=x0, to=x1, length.out = 100);
    yy <- seq(from=y0, to=y1, length.out = 100);
    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
    return(list(x=xd, y=yd));
}
par(mar=c(0,0,0,0));
# ===============================================================
plot(x=0, y=0, type="n", xlim=c(0,85), ylim=c(50,95), xaxt="n", yaxt="n",
     xlab="",ylab="");
ibox <- mbox(x0 = 0,  x1 = 10, y0 = 90, y1 = 70);
polygon(x = ibox$x, y = ibox$y, lwd = 3, border = "black", col = "white");
cbox <- mbox(x0 = 15, x1 = 25, y0 = 90, y1 = 70);
polygon(x = cbox$x, y = cbox$y, lwd = 3, border = "black", col = "white");
ubox <- mbox(x0 = 30, x1 = 40, y0 = 90, y1 = 70);
polygon(x = ubox$x, y = ubox$y, lwd = 3, border = "black", col = "white");
nbox <- mbox(x0 = 45, x1 = 55, y0 = 90, y1 = 70);
polygon(x = nbox$x, y = nbox$y, lwd = 3, border = "black", col = "white");
fbox <- mbox(x0 = 60, x1 = 70, y0 = 90, y1 = 70);
polygon(x = fbox$x, y = fbox$y, lwd = 3, border = "black", col = "white");
tbox <- mbox(x0 = 75, x1 = 85, y0 = 90, y1 = 70);
polygon(x = tbox$x, y = tbox$y, lwd = 3, border = "black", col = "white");
text(x=5,  y=80, labels="Initialisation", col="black", cex = 0.65);
text(x=20, y=80, labels="Crossover", col="black", cex = 0.65);
text(x=35, y=80, labels="Mutation", col="black", cex = 0.65);
text(x=50, y=82, labels="Fitness", col="black", cex = 0.65);
text(x=50, y=78, labels="evaluation", col="black", cex = 0.65);
text(x=65, y=82, labels="Tournament", col="black", cex = 0.65);
text(x=65, y=78, labels="selection", col="black", cex = 0.65);
text(x=80, y=80, labels="Replace", col="black", cex = 0.65);
arrows(x0=10, x1=14, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0=25, x1=29, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0=40, x1=44, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0=55, x1=59, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0=70, x1=74, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0 = 80, x1 = 80, y0 = 70, y1 = 68, lwd = 2, length = 0);
arrows(x0 = 80, x1 = 20, y0 = 65, y1 = 65, lwd = 2, length = 0);
arrows(x0 = 20, x1 = 20, y0 = 65, y1 = 70, lwd = 2, length = 0.05);
text(x=73, y=63, labels="No", col="black", cex = 0.5);
rbox <- mbox(x0 = 75, x1 = 85, y0 = 62, y1 = 68);
polygon(x = rbox$x, y = rbox$y, lwd = 3, border = "black", col = "black");
text(x=80, y=65, labels="Termination?", col="white", cex = 0.65);
fbox <- mbox(x0 = 60, x1 = 70, y0 = 60, y1 = 50);
polygon(x = fbox$x, y = fbox$y, lwd = 3, border = "black", col = "white");
text(x=65, y=57, labels="Individual", col="black", cex = 0.65);
text(x=65, y=53, labels="network", col="black", cex = 0.65);
arrows(x0 = 80, x1 = 80, y0 = 62, y1 = 55, lwd = 2, length = 0);
arrows(x0 = 80, x1 = 70, y0 = 55, y1 = 55, lwd = 2, length = 0.05);
text(x=73, y=53, labels="Yes", col="black", cex = 0.5);
```

\vspace{6mm}

Evolve values between nodes, mapping loci to traits.

\footnotetext[1]{resevol R package vignette: \href{https://bradduthie.github.io/resevol/articles/evolutionary_algorithm.html}{https://bradduthie.github.io/resevol/articles/evolutionary\_algorithm.html}}

## Evolutionary algorithm to find genetic architecture

```{r, eval = FALSE}
gmt <- matrix(data = c( 1.0, -0.5, -0.5, 2.0), 
              nrow = 2);
```

\vspace{2mm}
\hrule
\vspace{2mm}

```{r, eval = FALSE}
mg  <- mine_gmatrix(gmatrix = gmt, loci = 3, 
                    indivs = 2000, npsize = 12000, 
                    max_gen = 5400, sampleK = 1200, 
                    chooseK = 6, layers = 4, 
                    mu_pr = 0.2, pr_cross = 0.2, 
                    mu_sd = 0.004, term_cri = -8);
```

## Constructing a landscape: Default options

\Large

\begin{columns}
\begin{column}{0.5\textwidth}

```{r, out.width = "100%", echo = FALSE}
image(landscape, xaxt = "n", yaxt = "n", col = land_cols);
```

\end{column}

\begin{column}{0.5\textwidth} 
\begin{itemize}
\item Specify `farms`, `xdim`, `ydim`.
\item Shortest splitline algorithm
\item Land edge: torus, leaky, reflect
\end{itemize}
\end{column}
\end{columns}

\vspace{8mm}

Each farm can use up to one crop and one pesticide, and the timing and rotation rules for crops and pesticide allow any Markov process.

## Constructing a landscape: Advanced options

```{r, echo = FALSE}
land_file <- system.file("landscape_eg.csv", package = "resevol");
land_dat  <- read.csv(file = land_file, header = FALSE);
land_eg   <- t(as.matrix(land_dat));
farm_cols <- c("#f4eadc", "#6a4b20", "#cea05f", "#e1c59d", "#a97833", "#cea05f",
               "#f2e6d6", "#6a4b20", "#cc9c59", "#dfc197", "#a27331", "#f0e3d0",
               "#5d421c", "#ca9852");
land_cols <- c(farm_cols, "#00ab41", "#234F1E", "#2832C2");
par(mar = c(0, 0, 0, 0));
image(land_eg, xaxt = "n", yaxt = "n", col = land_cols);
points(x = 0.2, y = 0.05, cex = 9, pch = 20);
text(x = 0.2, y = 0.05, labels = "1", cex = 2, col = "red");
points(x = 0.4, y = 0.1, cex = 9, pch = 20);
text(x = 0.4, y = 0.1, labels = "2", cex = 2, col = "red");
points(x = 0.4, y = 0.1, cex = 9, pch = 20);
text(x = 0.4, y = 0.1, labels = "2", cex = 2, col = "red");
points(x = 0.25, y = 0.27, cex = 9, pch = 20);
text(x = 0.25, y = 0.27, labels = "3", cex = 2, col = "red");
points(x = 0.20, y = 0.43, cex = 9, pch = 20);
text(x = 0.20, y = 0.43, labels = "4", cex = 2, col = "red");
points(x = 0.42, y = 0.48, cex = 9, pch = 20);
text(x = 0.42, y = 0.48, labels = "5", cex = 2, col = "red");
points(x = 0.28, y = 0.58, cex = 9, pch = 20);
text(x = 0.28, y = 0.58, labels = "6", cex = 2, col = "red");
points(x = 0.1, y = 0.8, cex = 9, pch = 20);
text(x = 0.1, y = 0.8, labels = "7", cex = 2, col = "red");
points(x = 0.7, y = 0.05, cex = 9, pch = 20);
text(x = 0.7, y = 0.05, labels = "8", cex = 2, col = "red");
points(x = 0.9, y = 0.2, cex = 9, pch = 20);
text(x = 0.9, y = 0.2, labels = "9", cex = 2, col = "red");
points(x = 0.85, y = 0.4, cex = 9, pch = 20);
text(x = 0.85, y = 0.4, labels = "10", cex = 2, col = "red");
points(x = 0.92, y = 0.58, cex = 9, pch = 20);
text(x = 0.92, y = 0.58, labels = "11", cex = 2, col = "red");
points(x = 0.88, y = 0.76, cex = 9, pch = 20);
text(x = 0.88, y = 0.76, labels = "12", cex = 2, col = "red");
points(x = 0.86, y = 0.93, cex = 9, pch = 20);
text(x = 0.86, y = 0.93, labels = "13", cex = 2, col = "red");
points(x = 0.52, y = 0.91, cex = 9, pch = 20);
text(x = 0.52, y = 0.91, labels = "14", cex = 2, col = "red");
points(x = 0.05, y = 0.59, cex = 7, pch = 20);
text(x = 0.05, y = 0.59, labels = "15", cex = 1.7, col = "white");
points(x = 0.46, y = 0.28, cex = 7, pch = 20);
text(x = 0.46, y = 0.28, labels = "16", cex = 1.7, col = "white");
points(x = 0.62, y = 0.36, cex = 7, pch = 20);
text(x = 0.62, y = 0.36, labels = "17", cex = 1.7, col = "white");
```

## Pest life-history events defined within an age range

\Large

Within a time step, order of events are specified.

\vspace{5mm}

```{r, echo=FALSE, fig.height=3, fig.width=8}
mbox <- function(x0, x1, y0, y1){
    xx <- seq(from=x0, to=x1, length.out = 100);
    yy <- seq(from=y0, to=y1, length.out = 100);
    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
    return(list(x=xd, y=yd));
}
par(mar=c(0,0,0,0));
# ===============================================================
plot(x=0, y=0, type="n", xlim=c(0,100), ylim=c(40,95), xaxt="n", yaxt="n",
     xlab="",ylab="");
ibox <- mbox(x0 = 0,  x1 = 14, y0 = 63, y1 = 47);
polygon(x = ibox$x, y = ibox$y, lwd = 5, border = "black", col = "white");
cbox <- mbox(x0 = 0, x1 = 10, y0 = 88, y1 = 72);
polygon(x = cbox$x, y = cbox$y, lwd = 3, border = "black", col = "white");
ubox <- mbox(x0 = 15, x1 = 25, y0 = 88, y1 = 72);
polygon(x = ubox$x, y = ubox$y, lwd = 3, border = "black", col = "white");
nbox <- mbox(x0 = 30, x1 = 40, y0 = 88, y1 = 72);
polygon(x = nbox$x, y = nbox$y, lwd = 3, border = "black", col = "white");
fbox <- mbox(x0 = 45, x1 = 55, y0 = 88, y1 = 72);
polygon(x = fbox$x, y = fbox$y, lwd = 3, border = "black", col = "white");
tbox <- mbox(x0 = 60, x1 = 70, y0 = 88, y1 = 72);
polygon(x = tbox$x, y = tbox$y, lwd = 3, border = "black", col = "white");
tbox <- mbox(x0 = 75, x1 = 85, y0 = 88, y1 = 72);
polygon(x = tbox$x, y = tbox$y, lwd = 3, border = "black", col = "white");
tbox <- mbox(x0 = 90, x1 = 100, y0 = 88, y1 = 72);
polygon(x = tbox$x, y = tbox$y, lwd = 3, border = "black", col = "white");
ibox <- mbox(x0 = 90,  x1 = 100, y0 = 63, y1 = 47);
polygon(x = ibox$x, y = ibox$y, lwd = 3, border = "black", col = "white");
tbox <- mbox(x0 = 75, x1 = 85, y0 = 63, y1 = 47);
polygon(x = tbox$x, y = tbox$y, lwd = 3, border = "black", col = "white");
text(x=7,  y=59, labels="Initialise", col="black", cex = 0.8);
text(x=7,  y=55, labels="pests and", col="black", cex = 0.8);
text(x=7,  y=51, labels="landscape", col="black", cex = 0.8);
text(x=5, y=84, labels="Crops and", col="black", cex = 0.8);
text(x=5, y=80, labels="pesticides", col="black", cex = 0.8);
text(x=5, y=76, labels="rotated?", col="black", cex = 0.8);
text(x=20, y=82, labels="Pests", col="black", cex = 0.8);
text(x=20, y=78, labels="metabolise", col="black", cex = 0.8);
text(x=35, y=82, labels="Pests", col="black", cex = 0.8);
text(x=35, y=78, labels="feed", col="black", cex = 0.8);
text(x=50, y=82, labels="Pesticide", col="black", cex = 0.8);
text(x=50, y=78, labels="uptake", col="black", cex = 0.8);
text(x=65, y=82, labels="Pests", col="black", cex = 0.8);
text(x=65, y=78, labels="move", col="black", cex = 0.8);
text(x=80, y=82, labels="Pests", col="black", cex = 0.8);
text(x=80, y=78, labels="reproduce", col="black", cex = 0.8);
text(x=95, y=82, labels="Pest", col="black", cex = 0.8);
text(x=95, y=78, labels="mortality", col="black", cex = 0.8);
text(x=95,  y=57, labels="Pests", col="black", cex = 0.8);
text(x=95,  y=53, labels="immigrate", col="black", cex = 0.8);
text(x=80,  y=59, labels="Collect", col="black", cex = 0.8);
text(x=80,  y=55, labels="statistics", col="black", cex = 0.8);
text(x=80,  y=51, labels="as CSV", col="black", cex = 0.8);
arrows(x0=2, x1=2,   y0=63, y1=71.5, lwd=2, length=0.10);
arrows(x0=10, x1=14.5, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0=25, x1=29.5, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0=40, x1=44.5, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0=55, x1=59.5, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0=70, x1=74.5, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0=85, x1=89.5, y0=80, y1=80, lwd=2, length=0.10);
arrows(x0=95, x1=95,   y0=72, y1=63.5, lwd=2, length=0.10);
arrows(x0=90, x1=85.5, y0=55, y1=55, lwd=2, length=0.10);
arrows(x0=75, x1=70.5, y0=60, y1=60, lwd=2, length=0.10);
arrows(x0=60, x1=55.5, y0=60, y1=60, lwd=2, length=0.10);
rbox <- mbox(x0 = 60, x1 = 70, y0 = 57, y1 = 63);
polygon(x = rbox$x, y = rbox$y, lwd = 3, border = "black", col = "black");
text(x=65, y=60, labels="Extinction?", col="white", cex = 0.75);
rbox <- mbox(x0 = 45, x1 = 55, y0 = 57, y1 = 63);
polygon(x = rbox$x, y = rbox$y, lwd = 3, border = "black", col = "black");
text(x=50, y=60, labels="Max. time?", col="white", cex = 0.75);
text(x = 58.5, y = 58.5, labels = "No", col = "black", cex = 0.75);
text(x = 43.5, y = 58.5, labels = "No", col = "black", cex = 0.75);
text(x = 68, y = 54, labels = "Yes", col = "black", cex = 0.75);
text(x = 53, y = 54, labels = "Yes", col = "black", cex = 0.75);
arrows(x0=66, x1=66, y0=57, y1=50, lwd=2, length=0.0);
arrows(x0=51, x1=51, y0=57, y1=50, lwd=2, length=0.0);
rbox <- mbox(x0 = 30, x1 = 40, y0 = 47, y1 = 53);
polygon(x = rbox$x, y = rbox$y, lwd = 3, border = "black", col = "black");
text(x=35, y=50, labels="STOP", col="white", cex = 1);
rbox <- mbox(x0 = 14, x1 = 20, y0 = 47, y1 = 63);
polygon(x = rbox$x, y = rbox$y, lwd = 5, border = "black", col = "black");
text(x=17, y=55, labels="START", col="white", cex = 1, srt = 90);
arrows(x0=66, x1=40.5, y0=50, y1=50, lwd=2, length=0.10);
arrows(x0=45, x1=35, y0=60, y1=60, lwd=2, length=0.0);
arrows(x0=35, x1=35, y0=60, y1=67, lwd=2, length=0.0);
arrows(x0=35, x1=7, y0=67, y1=67, lwd=2, length=0.0);
arrows(x0=7, x1=7, y0=67, y1=71.5, lwd=2, length=0.1);
```

\vspace{3mm}

Specific events can be restricted to age ranges:

Feed: 0-2, Move: 3-8, Mate: 7-9

## Running a simulation in resevol

\small

```{r, eval = FALSE}
run_farm_sim(mine_output = mg, repro = "asexual", 
             pesticide_number = 2, time_steps = 160, 
             farms = 9, pesticide_init = "random", 
             pesticide_consume = c("T1", "T2"), 
             pesticide_rotation_time = 16, xdim = 64,
             pesticide_rotation_type = 3, ydim = 64,
             pesticide_tolerated_surv = 0, max_age = 4, 
             pesticide_per_cell = 1, crop_number = 1, 
             crop_rotation_time = 16, crop_per_cell = 4, 
             food_consume = 1, food_needed_surv = 1,
             reproduction_type = "food_based", 
             food_needed_repr = 1, land_edge = "torus",
             min_age_feed = 0, max_age_feed = 2, 
             min_age_move = 3, max_age_move = 4, 
             min_age_reproduce = 4, print_gens = FALSE,
             max_age_reproduce = 4, rand_age = TRUE, 
             age_pesticide_threshold = 2, 
             immigration_rate = 10, move_distance = 2,
             print_last = TRUE, trait_means = c(0.1, 0.1));
```

## Simulating ecology, evolution, crop yields

```{r, echo = FALSE}
load(system.file("advanced_eg.rda", package = "resevol"));
initial_crop <- c(1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 2, 3, 3, 3);
rotate_crop       <- matrix(data = 0, nrow = 3, ncol = 3);
rotate_crop[1, 2] <- 1;
rotate_crop[2, 1] <- 1;
rotate_crop[3, 3] <- 1;
initial_pesticide <- c(1, 1, 1, 1, 1, 1, 1, 2, 2, 2, 2, 2, 2, 2, 3, 3, 3);
rotate_pesticide           <- matrix(data = 0, nrow = 3, ncol = 3);
rotate_pesticide[1, 1]     <- 1;
rotate_pesticide[2, 2]     <- 1;
rotate_pesticide[3, 3]     <- 1;
population_data_file_sim1 <- system.file("population_data_sim1.csv", 
                                         package = "resevol");
population_data_sim1      <- read.csv(file = population_data_file_sim1);
```

```{r, echo = FALSE}
mbox <- function(x0, x1, y0, y1){
    xx <- seq(from=x0, to=x1, length.out = 100);
    yy <- seq(from=y0, to=y1, length.out = 100);
    xd <- c(rep(x0, 100), xx, rep(x1,100), rev(xx));
    yd <- c(yy, rep(y1,100), rev(yy), rep(y0, 100));
    return(list(x=xd, y=yd));
}
season          <- seq(from = 0, to = 240, by = 18);
blocks          <- length(season) - 1;
plot(x = population_data_sim1[["time_step"]], type = "n",
     y = population_data_sim1[["population_size"]], cex.lab = 1.25, 
     cex.axis = 1.25, ylab = "Pest population abundance", xlab = "Time step", 
     ylim = c(0, max(population_data_sim1[["population_size"]])));
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 21000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
     y = population_data_sim1[["population_size"]], type = "l", lwd = 2);
abline(v = 81, lwd = 3, col = "red");
box();
```

## Simulating ecology, evolution, crop yields

```{r, echo = FALSE}
par(mfrow = c(2, 2), mar = c(1, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(0, 0.8),
     y = population_data_sim1[["mean_food1_consumed"]], cex.lab = 1.25, 
     ylab = "Mean crop 1 consumed", xlab = "Time step", xaxt = "n");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
       y = population_data_sim1[["mean_food1_consumed"]], type = "l", lwd = 2);
text(x = 5, y = 0.79, labels = "a", cex = 2);
box();
par(mar = c(1, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(0, 0.8),
     y = population_data_sim1[["mean_food2_consumed"]], cex.lab = 1.25 ,
     ylab = "Mean crop 2 consumed", xlab = "Time step", xaxt = "n");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
       y = population_data_sim1[["mean_food2_consumed"]], type = "l", lwd = 2);
text(x = 5, y = 0.79, labels = "b", cex = 2);
box();
par(mar = c(4, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(0, 0.8),
     y = population_data_sim1[["mean_pesticide1_consumed"]], cex.lab = 1.25, 
     ylab = "Mean pesticide 1 consumed", xlab = "Time step");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
       y = population_data_sim1[["mean_pesticide1_consumed"]], type = "l", lwd = 2);
box();
text(x = 5, y = 0.78, labels = "c", cex = 2);
par(mar = c(4, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(0, 0.8),
     y = population_data_sim1[["mean_pesticide2_consumed"]], cex.lab = 1.25, 
     ylab = "Mean pesticide 2 consumed", xlab = "Time step");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
       y = population_data_sim1[["mean_pesticide2_consumed"]], type = "l", 
       lwd = 2);
box();
text(x = 5, y = 0.78, labels = "d", cex = 2);
```

## Simulating ecology, evolution, crop yields

```{r, echo = FALSE}
par(mfrow = c(2, 2), mar = c(1, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(2, 3.6),
     y = population_data_sim1[["trait1_mean_value"]], cex.lab = 1.25, 
     ylab = "Mean crop 1 consumption trait", xlab = "Time step", xaxt = "n");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
     y = population_data_sim1[["trait1_mean_value"]], type = "l", lwd = 2);
text(x = 5, y = 3.56, labels = "a", cex = 2);
box();
par(mar = c(1, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(2, 3.6),
     y = population_data_sim1[["trait2_mean_value"]], cex.lab = 1.25 ,
     ylab = "Mean crop 2 consumption trait", xlab = "Time step", xaxt = "n");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
       y = population_data_sim1[["trait2_mean_value"]], type = "l", lwd = 2);
text(x = 5, y = 3.56, labels = "b", cex = 2);
box();
par(mar = c(4, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(-0.8, 1.4),
     y = population_data_sim1[["trait3_mean_value"]], cex.lab = 1.25, 
     ylab = "Mean pesticide 1 consumption trait", xlab = "Time step");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
       y = population_data_sim1[["trait3_mean_value"]], type = "l", lwd = 2);
box();
text(x = 5, y = 1.34, labels = "c", cex = 2);
par(mar = c(4, 4, 1, 1));
plot(x = population_data_sim1[["time_step"]], type = "n", ylim = c(-0.8, 1.4),
     y = population_data_sim1[["trait4_mean_value"]], cex.lab = 1.25, 
     ylab = "Mean pesticide 2 consumption trait", xlab = "Time step");
for(i in 1:blocks){
    rbox <- mbox(x0 = season[i], x1 = season[i + 1], y0 = -1000, y1 = 6000);
    if(i %% 2 == 0){
        polygon(x = rbox$x, y = rbox$y, lwd = 3, border = NA, 
                col = "grey90");
    }
}
points(x = population_data_sim1[["time_step"]], 
       y = population_data_sim1[["trait4_mean_value"]], type = "l", lwd = 2);
box();
text(x = 5, y = 1.34, labels = "d", cex = 2);
```

## Simulating ecology, evolution, crop yields

```{r, echo = FALSE}
last_time_step_file_sim1 <- system.file("last_time_step_sim1.csv", 
                                        package = "resevol");
pop_last_time_step_sim1  <- read.csv(file = last_time_step_file_sim1);
par(mar = c(0, 0, 0, 0));
landscape    <- land_eg;
for(i in 1:dim(pop_last_time_step_sim1)[1]){
    xloc <- pop_last_time_step_sim1[i, 1] + 1;
    yloc <- pop_last_time_step_sim1[i, 2] + 1;
    landscape[xloc, yloc] <- 18;
}
land_cols <- c(land_cols, "#000000");
image(landscape, xaxt = "n", yaxt = "n", col = land_cols);
```

## The resevol R package: summary

\Large

\begin{itemize}
\item Simulations of pest ecology, resistance evolution, and agricultural production
\item Covarying polygenetic traits
\item Flexible pest life-histories, mating systems, traits
\item Highly customisable, detailed landscapes
\item Free to use$^{1}$, open source$^{2}$ (GPL >= 2), documented$^{1,3}$, actively maintained$^{4}$
\end{itemize}

\footnotetext[1]{\textbf{Website}: \href{https://bradduthie.github.io/resevol/}{https://bradduthie.github.io/resevol/}}
\footnotetext[2]{\textbf{Code}: \href{https://github.com/bradduthie/resevol}{https://github.com/bradduthie/resevol}}
\footnotetext[3]{\textbf{Lab Notebook}: \href{https://bradduthie.github.io/resevol/notebook/}{https://bradduthie.github.io/resevol/notebook/}}
\footnotetext[4]{\textbf{Email}: alexander.duthie@stir.ac.uk \textbf{Mastodon}: @bradduthie@ecoevo.social}

### 2023-07-17 notebook/presentation/resevol_presentation.Rmd
Presentation for British Ecological Society 2022

--- 2023-07-17 notebook/presentation/BES2022.Rmd
@@
   | Matthew C. Tinsley$^{1}$, and Luc F. Bussi&egrave;re$^{2,3}$
-date: '[1] University of Stirling [2] University of Gothenburg [3] Gothenburgh Global Biodiversity Centre alexander.duthie@stir.ac.uk, @bradduthie@ecoevo.social http://bradduthie.github.io/talks/BES2022.pdf'
+date: '[1] University of Stirling [2] University of Gothenburg [3] Gothenburgh Global Biodiversity Centre alexander.duthie@stir.ac.uk, @bradduthie@ecoevo.social http://bradduthie.github.io/talks/resevol.pdf'
 output:
@@
 
-\vspace{2mm}\pause
+\vspace{8mm}\pause
 
@@
 \begin{itemize}
-\item Vary biopesticide application$^{3,4}$
-\item Polygenic resistance traits$^{5}$
-\item Negative cross-resistance$^{6}$
+\item Weaken selection for resistance$^{3}$
+\item Varying pesticide application$^{4}$
 \end{itemize}
@@
 \footnotetext[4]{Saikai et al. 2021. \textit{Pest Manag. Sci.} 77:273-284.}
-\footnotetext[5]{Boots, M. 2011. \textit{Am. Nat.} 178:214-220.}
-\footnotetext[6]{Via, S. 1986. \textit{Pesticide resistance}. 222-235.}
 
-## Agricultural heterogeneity to overcome resistance evolution
+## Negatively correlated cross-resistance between pesticides
+
+```{r, echo = FALSE, fig.width = 4.65, fig.height = 4.65}
+var1 <- runif(n = 120, min = 0, max = 1);
+var2 <- rnorm(n = 120, sd = 0.2) - var1;
+par(mgp=c(0.2,1,0))  
+plot(x = var1, y = var2, xaxt = "n", yaxt = "n", pch = 20, cex = 1.25,
+     xlab = "Resistance to pesticide A", ylab = "Resistance to pesticide B",
+     cex.lab = 1.25);
+par(mgp=c(3,1,0))
+```
+
+\footnotetext[1]{Chapman \& Penman. 1979. \textit{Nature} 281:298-299.}
+
+## Enhancing Diversity to Overcome Resistance Evolution
 
@@
 \begin{center}
-Biopesticide applied
+Pesticide applied
 \end{center}
@@
 include_graphics("img/Endorse_project.png");
+```
+
+##
+
+```{r, out.width = "100%", echo = FALSE}
+include_graphics("img/Endorse_people.png");
 ```
@@
 \begin{columns}
+\begin{column}{0.5\textwidth} 
+\textbf{Complex individuals}
+\begin{itemize}
+\item Complete genomes
+\item Highly polygenic traits
+\item Set trait covariances
+\item Flexible life-history
+\item Flexible mating system
+\end{itemize}
+\end{column}
+\pause
+\begin{column}{0.5\textwidth} 
+\textbf{Dynamic landscapes}
+\begin{itemize}
+\item Custom land dimensions
+\item Custom pesticide rotation
+\item Custom crop rotation
+\item Individual farms tracked
+\item Raster landscape maps
+\end{itemize}
+\end{column}
+
+\end{columns}
+
+\footnotetext[1]{Duthie AB, McKeon CR (2022). resevol: Simulate Agricultural Production and Evolution of Pesticide Resistance. R package version 0.3.3.0, https://bradduthie.github.io/resevol/.}
+\footnotetext[2]{Duthie, AB, R Mangan, CR McKeon, MC Tinsley, LF Bussi\`{e}re.
+2022.
+resevol: an R package for spatially explicit models of pesticide resistance given evolving pest genomes.
+bioRxiv 2022.08.22.504740; doi: https://doi.org/10.1101/2022.08.22.504740}
+
+## resevol: An R package for modelling agricultural pests
+
+\begin{columns}
 \begin{column}{0.7\textwidth} 
@@
 }{Public domain}}
-
-## resevol: An R package for modelling agricultural pests
-
-\begin{columns}
-\begin{column}{0.5\textwidth} 
-\textbf{Complex individuals}
-\begin{itemize}
-\item Complete genomes
-\item Highly polygenic traits
-\item Set trait covariances
-\item Flexible life-history
-\item Flexible mating system
-\end{itemize}
-\end{column}
-\begin{column}{0.5\textwidth} 
-\textbf{Dynamic landscapes}
-\begin{itemize}
-\item Custom land dimensions
-\item Custom pesticide rotation
-\item Custom crop rotation
-\item Individual farms tracked
-\item Raster landscape maps
-\end{itemize}
-\end{column}
-
-\end{columns}
-
-\footnotetext[1]{Duthie AB, McKeon CR (2022). resevol: Simulate Agricultural Production and Evolution of Pesticide Resistance. R package version 0.3.3.0, https://bradduthie.github.io/resevol/.}
-\footnotetext[2]{Duthie, AB, R Mangan, CR McKeon, MC Tinsley, LF Bussi\`{e}re.
-2022.
-resevol: an R package for spatially explicit models of pesticide resistance given evolving pest genomes.
-bioRxiv 2022.08.22.504740; doi: https://doi.org/10.1101/2022.08.22.504740}
 
@@
 
-## Constructing a landscape: Default options
+## Evolutionary algorithm to find genetic architecture
 
-\Large
+\LARGE
 
@@
 \begin{column}{0.5\textwidth} 
+\textbf{Pre-specified covariances}
 
-```{r, out.width = "100%", echo = FALSE}
-image(landscape, xaxt = "n", yaxt = "n", col = land_cols);
-```
+\begin{align}
+\begin{bmatrix}
+ 1.0 & -0.5 \\
+-0.5 &  2.0
+\end{bmatrix}
+\end{align}
+
+\end{column}
+\begin{column}{0.5\textwidth} 
+\textbf{Covariances produced}
+
+\begin{align}
+\begin{bmatrix}
+ 1.007 & -0.436 \\
+-0.436 &  1.906
+\end{bmatrix}
+\end{align}
 
@@
 
-\begin{column}{0.5\textwidth} 
-\begin{itemize}
-\item Specify `farms`, `xdim`, `ydim`.
-\item Shortest splitline algorithm
-\item Land edge: torus, leaky, reflect
-\end{itemize}
-\end{column}
 \end{columns}
-
-\vspace{8mm}
-
-Each farm can use up to one crop and one pesticide, and the timing and rotation rules for crops and pesticide allow any Markov process.
-
-## Constructing a landscape: Advanced options
-
-```{r, echo = FALSE}
-land_file <- system.file("landscape_eg.csv", package = "resevol");
-land_dat  <- read.csv(file = land_file, header = FALSE);
-land_eg   <- t(as.matrix(land_dat));
-farm_cols <- c("#f4eadc", "#6a4b20", "#cea05f", "#e1c59d", "#a97833", "#cea05f",
-               "#f2e6d6", "#6a4b20", "#cc9c59", "#dfc197", "#a27331", "#f0e3d0",
-               "#5d421c", "#ca9852");
-land_cols <- c(farm_cols, "#00ab41", "#234F1E", "#2832C2");
-par(mar = c(0, 0, 0, 0));
-image(land_eg, xaxt = "n", yaxt = "n", col = land_cols);
-points(x = 0.2, y = 0.05, cex = 9, pch = 20);
-text(x = 0.2, y = 0.05, labels = "1", cex = 2, col = "red");
-points(x = 0.4, y = 0.1, cex = 9, pch = 20);
-text(x = 0.4, y = 0.1, labels = "2", cex = 2, col = "red");
-points(x = 0.4, y = 0.1, cex = 9, pch = 20);
-text(x = 0.4, y = 0.1, labels = "2", cex = 2, col = "red");
-points(x = 0.25, y = 0.27, cex = 9, pch = 20);
-text(x = 0.25, y = 0.27, labels = "3", cex = 2, col = "red");
-points(x = 0.20, y = 0.43, cex = 9, pch = 20);
-text(x = 0.20, y = 0.43, labels = "4", cex = 2, col = "red");
-points(x = 0.42, y = 0.48, cex = 9, pch = 20);
-text(x = 0.42, y = 0.48, labels = "5", cex = 2, col = "red");
-points(x = 0.28, y = 0.58, cex = 9, pch = 20);
-text(x = 0.28, y = 0.58, labels = "6", cex = 2, col = "red");
-points(x = 0.1, y = 0.8, cex = 9, pch = 20);
-text(x = 0.1, y = 0.8, labels = "7", cex = 2, col = "red");
-points(x = 0.7, y = 0.05, cex = 9, pch = 20);
-text(x = 0.7, y = 0.05, labels = "8", cex = 2, col = "red");
-points(x = 0.9, y = 0.2, cex = 9, pch = 20);
-text(x = 0.9, y = 0.2, labels = "9", cex = 2, col = "red");
-points(x = 0.85, y = 0.4, cex = 9, pch = 20);
-text(x = 0.85, y = 0.4, labels = "10", cex = 2, col = "red");
-points(x = 0.92, y = 0.58, cex = 9, pch = 20);
-text(x = 0.92, y = 0.58, labels = "11", cex = 2, col = "red");
-points(x = 0.88, y = 0.76, cex = 9, pch = 20);
-text(x = 0.88, y = 0.76, labels = "12", cex = 2, col = "red");
-points(x = 0.86, y = 0.93, cex = 9, pch = 20);
-text(x = 0.86, y = 0.93, labels = "13", cex = 2, col = "red");
-points(x = 0.52, y = 0.91, cex = 9, pch = 20);
-text(x = 0.52, y = 0.91, labels = "14", cex = 2, col = "red");
-points(x = 0.05, y = 0.59, cex = 7, pch = 20);
-text(x = 0.05, y = 0.59, labels = "15", cex = 1.7, col = "white");
-points(x = 0.46, y = 0.28, cex = 7, pch = 20);
-text(x = 0.46, y = 0.28, labels = "16", cex = 1.7, col = "white");
-points(x = 0.62, y = 0.36, cex = 7, pch = 20);
-text(x = 0.62, y = 0.36, labels = "17", cex = 1.7, col = "white");
-```
 
@@
 
-\vspace{3mm}
+\vspace{3mm}\pause
 
@@
 Feed: 0-2, Move: 3-8, Mate: 7-9
+
+## Constructing a landscape: Default options
+
+\Large
+
+\begin{columns}
+\begin{column}{0.5\textwidth}
+
+```{r, out.width = "100%", echo = FALSE}
+image(landscape, xaxt = "n", yaxt = "n", col = land_cols);
+```
+
+\end{column}
+
+\begin{column}{0.5\textwidth} 
+\begin{itemize}
+\item Specify `farms`, `xdim`, `ydim`.
+\item Shortest splitline algorithm
+\item Land edge: torus, leaky, reflect
+\end{itemize}
+\end{column}
+\end{columns}
+
+\vspace{8mm}
+
+Each farm can use up to one crop and one pesticide, and the timing and rotation rules for crops and pesticide allow any Markov process.
+
+## Constructing a landscape: Advanced options
+
+```{r, echo = FALSE}
+land_file <- system.file("landscape_eg.csv", package = "resevol");
+land_dat  <- read.csv(file = land_file, header = FALSE);
+land_eg   <- t(as.matrix(land_dat));
+farm_cols <- c("#f4eadc", "#6a4b20", "#cea05f", "#e1c59d", "#a97833", "#cea05f",
+               "#f2e6d6", "#6a4b20", "#cc9c59", "#dfc197", "#a27331", "#f0e3d0",
+               "#5d421c", "#ca9852");
+land_cols <- c(farm_cols, "#00ab41", "#234F1E", "#2832C2");
+par(mar = c(0, 0, 0, 0));
+image(land_eg, xaxt = "n", yaxt = "n", col = land_cols);
+points(x = 0.2, y = 0.05, cex = 9, pch = 20);
+text(x = 0.2, y = 0.05, labels = "1", cex = 2, col = "red");
+points(x = 0.4, y = 0.1, cex = 9, pch = 20);
+text(x = 0.4, y = 0.1, labels = "2", cex = 2, col = "red");
+points(x = 0.4, y = 0.1, cex = 9, pch = 20);
+text(x = 0.4, y = 0.1, labels = "2", cex = 2, col = "red");
+points(x = 0.25, y = 0.27, cex = 9, pch = 20);
+text(x = 0.25, y = 0.27, labels = "3", cex = 2, col = "red");
+points(x = 0.20, y = 0.43, cex = 9, pch = 20);
+text(x = 0.20, y = 0.43, labels = "4", cex = 2, col = "red");
+points(x = 0.42, y = 0.48, cex = 9, pch = 20);
+text(x = 0.42, y = 0.48, labels = "5", cex = 2, col = "red");
+points(x = 0.28, y = 0.58, cex = 9, pch = 20);
+text(x = 0.28, y = 0.58, labels = "6", cex = 2, col = "red");
+points(x = 0.1, y = 0.8, cex = 9, pch = 20);
+text(x = 0.1, y = 0.8, labels = "7", cex = 2, col = "red");
+points(x = 0.7, y = 0.05, cex = 9, pch = 20);
+text(x = 0.7, y = 0.05, labels = "8", cex = 2, col = "red");
+points(x = 0.9, y = 0.2, cex = 9, pch = 20);
+text(x = 0.9, y = 0.2, labels = "9", cex = 2, col = "red");
+points(x = 0.85, y = 0.4, cex = 9, pch = 20);
+text(x = 0.85, y = 0.4, labels = "10", cex = 2, col = "red");
+points(x = 0.92, y = 0.58, cex = 9, pch = 20);
+text(x = 0.92, y = 0.58, labels = "11", cex = 2, col = "red");
+points(x = 0.88, y = 0.76, cex = 9, pch = 20);
+text(x = 0.88, y = 0.76, labels = "12", cex = 2, col = "red");
+points(x = 0.86, y = 0.93, cex = 9, pch = 20);
+text(x = 0.86, y = 0.93, labels = "13", cex = 2, col = "red");
+points(x = 0.52, y = 0.91, cex = 9, pch = 20);
+text(x = 0.52, y = 0.91, labels = "14", cex = 2, col = "red");
+points(x = 0.05, y = 0.59, cex = 7, pch = 20);
+text(x = 0.05, y = 0.59, labels = "15", cex = 1.7, col = "white");
+points(x = 0.46, y = 0.28, cex = 7, pch = 20);
+text(x = 0.46, y = 0.28, labels = "16", cex = 1.7, col = "white");
+points(x = 0.62, y = 0.36, cex = 7, pch = 20);
+text(x = 0.62, y = 0.36, labels = "17", cex = 1.7, col = "white");
+```
 
@@
 \begin{itemize}
-\item Simulations of pest ecology, resistance evolution, and agricultural production
-\item Covarying polygenetic traits
-\item Flexible pest life-histories, mating systems, traits
+\item Rapidly develop and simulation models of pest ecology and resistance evolution
+\item Introduces a novel approach to modelling complex pest genetic architecture
+\item Flexible pest life-histories, mating systems, trait covariances
 \item Highly customisable, detailed landscapes
-\item Free to use$^{1}$, open source$^{2}$ (GPL >= 2), documented$^{1,3}$, actively maintained$^{4}$
+\item Free to use$^{1}$, open source$^{2}$ (GPL >= 2), documented$^{1,3}$
 \end{itemize}
 
-\footnotetext[1]{\textbf{Website}: \href{https://bradduthie.github.io/resevol/}{https://bradduthie.github.io/resevol/}}
-\footnotetext[2]{\textbf{Code}: \href{https://github.com/bradduthie/resevol}{https://github.com/bradduthie/resevol}}
-\footnotetext[3]{\textbf{Lab Notebook}: \href{https://bradduthie.github.io/resevol/notebook/}{https://bradduthie.github.io/resevol/notebook/}}
-\footnotetext[4]{\textbf{Email}: alexander.duthie@stir.ac.uk \textbf{Mastodon}: @bradduthie@ecoevo.social}
+\footnotetext[1]{Website: \href{https://bradduthie.github.io/resevol/}{https://bradduthie.github.io/resevol/}}
+\footnotetext[2]{Code: \href{https://github.com/bradduthie/resevol}{https://github.com/bradduthie/resevol}}
+\footnotetext[3]{Lab Notebook: \href{https://bradduthie.github.io/resevol/notebook/}{https://bradduthie.github.io/resevol/notebook/}}

### 2023-07-17 notebook/ms/ms.Rmd
Make some minor edits to the latest version of the manuscript before updating some references

@@
     toc: no
+  word_document:
+    reference_docx: docx_template.docx
   pdf_document:
     toc: no
-  word_document:
-    reference_docx: docx_template.docx
   html_document: default
@@
 - \modulolinenumbers[2]
-linestretch: 1
+linestretch: 2
 link-citations: yes
@@
 
-1. The evolution of pesticide resistance is a widespread problem with potentially severe consequences for global food security.
-2.
+The evolution of pesticide resistance is a widespread problem with potentially severe consequences for global food security.
 We introduce the resevol R package, which simulates individual-based models of pests with evolving genomes that produce complex, polygenic, and covarying traits affecting pest life history and pesticide resistance.
-3.
 Simulations are modelled on a spatially-explicit and highly customisable landscape in which crop and pesticide application and rotation can vary, making the package a highly flexible tool for both general and tactical models of pest management and resistance evolution.
-4. We present the key features of the resevol package and demonstrate its use for a simple example simulating pests with two covarying traits.
-5. The resevol R package is open source under GNU Public License. All source code and documentation are available on GitHub.
+We present the key features of the resevol package and demonstrate its use for a simple example simulating pests with two covarying traits. 
+The resevol R package is open source under GNU Public License. All source code and documentation are available on GitHub.
 
@@
 Such trade-offs can be quantified by the genetic covariance between traits.
-A focal goal of the software is to model traits with a pre-specified, but potentially evolving, genetic covariance structure. 
+A focal goal of the resevol package is to model traits with a pre-specified, but potentially evolving, genetic covariance structure. 
 To achieve this goal, each individual has a genome with $L$ loci that underlie a set of $T$ potentially evolving traits. 
@@
 To achieve this, two separate steps are necessary.
-First, a genetic algorithm is used to find a network of internal nodes that map standard random normal loci values to covarying traits.
+First, an evolutionary algorithm is used to find a network of internal nodes that map standard random normal loci values to covarying traits.
 Values used to map loci to traits are incorporated into individual genomes.
@@
 The function initialises a population of `npsize` separate, and potentially unique, networks (i.e., `npsize` copies of a network like the one shown in Figure \@ref(fig:network)), and this population evolves until some maximum iteration (`max_gen`) or minimum expected network stress (`term_cri`) is met.
-In a single iteration of the algorithm, values mutate and crossover occurs between networks (i.e., some fo the `npsize` networks swap values with some probability).
+In a single iteration of the algorithm, values mutate and crossover occurs between networks (i.e., some of the `npsize` networks swap values with some probability).
 Next, trait covariances produced for each network are estimated by initalising `indivs` individuals with loci sampled from a standard normal distribution.
@@
 After initialisation, crops and pesticides rotate once every `crop_rotation_time` and `pesticide_rotation_time` time steps, respectively.
-Each of these rotation time arguments can take either an integer value from 1-3, or a matrix.
+The arguments `crop_rotation_type` and `pesticide_rotation_type` specify how crops and pesticides are rotated, respectively.
+Each of these rotation type arguments can take either an integer value from 1-3, or a matrix.
 Integer values specify (1) no rotation, (2) random transition from one type to another, or (3) cycling through each available crop or pesticide in numeric order.
@@
 At the start of a simulation, pests are initialised in a random location.
-Initialised pests are age of zero if `rand_age = FALSE` or a random age from zero to `max_age` if `rand_age = TRUE`.
+Initialised pests are of age zero if `rand_age = FALSE` or a random age from zero to `max_age` if `rand_age = TRUE`.
 Following initialisation, a single time step proceeds with landscape change (see above), pest aging and metabolism, feeding, pesticide consumption, movement, reproduction, mortality, and immigration (Figure \@ref(fig:flowchart)).

### 2023-07-17 notebook/ms/ms.Rmd
Update the manuscript with references and a few additional poitns of clarification

@@
 In such cases, resistance alleles have binary effects on phenotype, enabling resistant phenotypes to arise from genetic changes at single or a small number of loci [@Raymond2001; @Ffrench-Constant2004; @Ffrench-Constant2013].
-Nevertheless, resistance attributable to polygenic effects is also well-established [@Via1986; @Firko1990; @Busi2013; @Haridas2018; @Green2020].
-And the relevance of polygenic resistance is likely to further increase given a rising interest in biological controls for sustainable crop protection [e.g., @Green2020; @Elnahal2022]. 
+Nevertheless, resistance attributable to polygenic effects is also well-established [@Via1986; @Firko1990; @Busi2013; @Haridas2018; @Green2020; @Mangan2023].
+And the relevance of polygenic resistance is likely to further increase given a rising interest in biological controls for sustainable crop protection [e.g., @Green2020; @Elnahal2022; @Mangan2023]. 
 Developing strategies to maximise the efficacy of these tools is critical, and such strategies should be well-informed by predictions made from detailed, quantitative genetic models.
-Here we introduce the resevol R package as a tool for building such models and simulating pest management.
+Here we introduce the resevol R package as a tool for building individual-based models and simulating pest management [@Daly2022].
 
@@
 Each value defines a unique farm, but values do not need to be contiguous.
-These 'farms' could even model non-farmland (e.g., water, roads), if pesticides and crops on them are invisible to pests (see Supporting Information 2).
-This `terrain` customisation therefore allows for a high degree of landscape detail, and offers the potential for modelling real-world landscapes from raster images.
+These 'farms' could even model non-farmland (e.g., water, roads), if pesticides and crops on them are invisible to pests (see Supporting Information 2), or they could more broadly be interpreted as heterogenous landscape properties across an arbitrary scale [e.g., @Newman2023].
+This `terrain` customisation therefore allows for a high degree of landscape detail, and offers the potential for modelling real-world landscapes from raster images [e.g., @Millington2021].
 Edge effects are set using the `land_edge` argument.
@@
 Square matrices can be used to define the probability that a given crop or pesticide in row $i$ transitions to that in column $j$. 
-Hence, any possible Markov chain can be used to model transition between crop or pesticide types on farms.
+Hence, any possible Markov chain can be used to model transition between crop or pesticide types on farms, potentially integrating real crop use or pesticide use patterns [@Zhang2020].
 Upon rotation, crop and pesticide values are reset on each farm.

### 2023-07-17 notebook/ms/ms.Rmd
Update PDF for biorxiv

@@
 - \fancyhead[R]{RESEVOL, AN R PACKAGE}
-- \usepackage{lineno}
-- \linenumbers
-- \modulolinenumbers[2]
-linestretch: 2
+linestretch: 1
 link-citations: yes

### 2023-07-17 R/gmatrices.R
Update the website with the new text

@@ -4,6 +4,7 @@
 #' with a covariance structure of the identity matrix (replicate 1)
 #'
 #' @format A list of 8 elements used in individual-based model initialisation
+#' @keywords internal
 "mg_n1"
 
 #' Sample mine_gmatrix outputs
@@ -12,6 +13,7 @@
 #' with a covariance structure of the identity matrix (replicate 2)
 #'
 #' @format A list of 8 elements used in individual-based model initialisation
+#' @keywords internal
 "mg_n2"
 
 #' Sample mine_gmatrix outputs
@@ -20,6 +22,7 @@
 #' with a covariance structure of the identity matrix (replicate 3)
 #'
 #' @format A list of 8 elements used in individual-based model initialisation
+#' @keywords internal
 "mg_n3"
 
 #' Sample mine_gmatrix outputs
@@ -28,6 +31,7 @@
 #' with a covariance structure  of the identity matrix (replicate 4)
 #'
 #' @format A list of 8 elements used in individual-based model initialisation
+#' @keywords internal
 "mg_n4"
 
 #' Sample mine_gmatrix outputs
@@ -36,6 +40,7 @@
 #' with a covariance structure of the identity matrix (replicate 5)
 #'
 #' @format A list of 8 elements used in individual-based model initialisation
+#' @keywords internal
 "mg_n5"
 
 
@@ -46,6 +51,7 @@
 #' (replicate 1)
 #'
 #' @format A list of 8 elements used in individual-based model initialisation
+#' @keywords internal
 "mg_v1"
 
 #' Sample mine_gmatrix outputs
@@ -55,6 +61,7 @@
 #' (replicate 2)
 #'
 #' @format A list of 8 elements used in individual-based model initialisation
+#' @keywords internal
 "mg_v2"
 
 
@@ -65,6 +72,7 @@
 #' (replicate 3)
 #'
 #' @format A list of 8 elements used in individual-based model initialisation
+#' @keywords internal
 "mg_v3"
 
 
@@ -76,6 +84,7 @@
 #' (replicate 4)
 #'
 #' @format A list of 8 elements used in individual-based model initialisation
+#' @keywords internal
 "mg_v4"
 
 
@@ -86,6 +95,7 @@
 #' (replicate 5)
 #'
 #' @format A list of 8 elements used in individual-based model initialisation
+#' @keywords internal
 "mg_v5"
 
 

### 2023-07-17 R/initialise.R

@@ -56,6 +56,7 @@
 #'                          npsize = 100, max_gen = 4, prnt_out = FALSE);
 #'inds      <- initialise_inds(mine_output = mg, N = 40, repro = "asexual");
 #'@importFrom stats rnorm
+#'@keywords internal
 #'@export
 initialise_inds <- function(mine_output, 
                             N = 1000, 

### 2023-07-17 R/landscape.R

@@ -25,6 +25,7 @@
 #'land <- make_landscape(rows = 10, cols = 10, depth = 2, farms = 4)
 #'@useDynLib resevol
 #'@importFrom stats rnorm rpois runif
+#'@keywords internal
 #'@export
 make_landscape <- function(terrain = NA, rows = NA, cols = NA, depth = 21, 
                            farms = 4, public_land = 0, farm_var = 0){

### 2023-07-17 docs/notebook/images/2021_AUG_19/sim.R

@@ -1,124 +1,0 @@
-library("helicoverpa");
-set.seed(Sys.time());
-mg  <- readRDS("mg.rds");
-sim <- run_farm_sim(mine_output              = mg,
-                    N                        = 1000,
-                    neutral_loci             = 1000,
-                    xdim                     = 100,
-                    ydim                     = 100,
-                    repro                    = "sexual",
-                    max_age                  = 6,
-                    selfing                  = TRUE,
-                    food_consume             = c("T1", "T2"),
-                    pesticide_consume        = c("T3", "T4"),
-                    food_needed_surv         = 1,
-                    food_needed_repr         = 1,
-                    reproduction_type        = "food_based",
-                    pesticide_tolerated_surv = 0,
-                    pesticide_rotation_type  = 1,
-                    crop_rotation_type       = 1,
-                    min_age_reproduce        = 4,
-                    max_age_feed             = 3,
-                    lambda_value             = 1.5,
-                    farms                    = 24,
-                    time_steps               = 480,
-                    mutation_pr              = 0.001,
-                    crossover_pr             = 0.1,
-                    net_mu_layers            = 0,
-                    crop_rotation_time       = 12,
-                    pesticide_rotation_time  = 12,
-                    crop_per_cell            = 8,
-                    pesticide_per_cell       = 0.4,
-                    crop_number              = 2,
-                    pesticide_number         = 2,
-                    print_inds               = FALSE, # Careful with this one
-                    K_on_birth               = 1000000,
-                    min_age_move             = 4,
-                    age_food_threshold       = 2,
-                    min_age_feed             = 1,
-                    pesticide_start          = 100,
-                    print_last               = TRUE,
[…]

### 2023-07-17 docs/notebook/images/2021_AUG_26/sim.R

@@ -1,124 +1,0 @@
-library("helicoverpa");
-set.seed(Sys.time());
-mg  <- readRDS("mg_n1.rds");
-sim <- run_farm_sim(mine_output              = mg,
-                    N                        = 1000,
-                    neutral_loci             = 1000,
-                    xdim                     = 100,
-                    ydim                     = 100,
-                    repro                    = "asexual",
-                    max_age                  = 6,
-                    selfing                  = TRUE,
-                    food_consume             = c("T1", "T2"),
-                    pesticide_consume        = c("T3", "T4"),
-                    food_needed_surv         = 1,
-                    food_needed_repr         = 1,
-                    reproduction_type        = "food_based",
-                    pesticide_tolerated_surv = 0,
-                    pesticide_rotation_type  = 2,
-                    crop_rotation_type       = 2,
-                    min_age_reproduce        = 4,
-                    max_age_feed             = 3,
-                    lambda_value             = 1.5,
-                    farms                    = 24,
-                    time_steps               = 480,
-                    mutation_pr              = 0.001,
-                    crossover_pr             = 0.1,
-                    net_mu_layers            = 0,
-                    crop_rotation_time       = 12,
-                    pesticide_rotation_time  = 12,
-                    crop_per_cell            = 8,
-                    pesticide_per_cell       = 0.4,
-                    crop_number              = 2,
-                    pesticide_number         = 2,
-                    print_inds               = FALSE, # Careful with this one
-                    K_on_birth               = 1000000,
-                    min_age_move             = 4,
-                    age_food_threshold       = 2,
-                    min_age_feed             = 1,
-                    pesticide_start          = 100,
-                    print_last               = TRUE,
[…]

### 2023-07-17 docs/notebook/images/2021_AUG_26/sim.R
Put the notebook back into the docs folder

@@ -1,0 +1,124 @@
+library("helicoverpa");
+set.seed(Sys.time());
+mg  <- readRDS("mg_n1.rds");
+sim <- run_farm_sim(mine_output              = mg,
+                    N                        = 1000,
+                    neutral_loci             = 1000,
+                    xdim                     = 100,
+                    ydim                     = 100,
+                    repro                    = "asexual",
+                    max_age                  = 6,
+                    selfing                  = TRUE,
+                    food_consume             = c("T1", "T2"),
+                    pesticide_consume        = c("T3", "T4"),
+                    food_needed_surv         = 1,
+                    food_needed_repr         = 1,
+                    reproduction_type        = "food_based",
+                    pesticide_tolerated_surv = 0,
+                    pesticide_rotation_type  = 2,
+                    crop_rotation_type       = 2,
+                    min_age_reproduce        = 4,
+                    max_age_feed             = 3,
+                    lambda_value             = 1.5,
+                    farms                    = 24,
+                    time_steps               = 480,
+                    mutation_pr              = 0.001,
+                    crossover_pr             = 0.1,
+                    net_mu_layers            = 0,
+                    crop_rotation_time       = 12,
+                    pesticide_rotation_time  = 12,
+                    crop_per_cell            = 8,
+                    pesticide_per_cell       = 0.4,
+                    crop_number              = 2,
+                    pesticide_number         = 2,
+                    print_inds               = FALSE, # Careful with this one
+                    K_on_birth               = 1000000,
+                    min_age_move             = 4,
+                    age_food_threshold       = 2,
+                    min_age_feed             = 1,
+                    pesticide_start          = 100,
+                    print_last               = TRUE,
[…]

### 2023-11-30 notebook/ms/advanced_techniques.Rmd
Some edits to the manuscript prior to acceptance from PLoS Computational Biology

@@
   bookdown::pdf_document2:
-    citation_package: natbib
     fig_caption: yes
@@
 bibliography: resevol.bib
-biblio-style: apalike
+csl: plos-computational-biology.csl
+biblio-style: acm
 ---
+
+\renewcommand\thefigure{\Alph{figure}}
 
@@
 Note that running `mine_gmatrix` can be a time-consuming process, and will often require multiple attempts to get the above parameter settings to produce a desired network.
-See @Hamblin2013 for useful advice on parameter value selection for the evolutionary algorithm.
+See Hamblin @Hamblin2013 for useful advice on parameter value selection for the evolutionary algorithm.
 Time-sensitive parameter values specific to the resevol package include `indivs` and `layers`.
@@
 But this is only based on one population of `indivs = 2000` initialised pests. 
-We can use the `stress_test` function to see what the distribution of stress is over 1000 such initialised populations of 2000 individuals (Figure 1).
+We can use the `stress_test` function to see what the distribution of stress is over 1000 such initialised populations of 2000 individuals (Figure A).
 
@@
 In the case of the above, numbers 1-7 all appear on the landscape, so the matrix will work.
-We can visualise the matrix more clearly by looking at it as an image (Figure 2).
+We can visualise the matrix more clearly by looking at it as an image (Figure B).
 
@@
 
-In Figure 2, each colour represents a unique farm, and colours correspond to the numbers within the matrix `simple_terrain` (note that the orientation differs between the matrix and the image).
+In Figure B, each colour represents a unique farm, and colours correspond to the numbers within the matrix `simple_terrain` (note that the orientation differs between the matrix and the image).
 This is a simple example, but matrices can be made large to model highly complex landscapes.
@@
 The dimensions of the landscape modelled in `land_eg` are 128 by 128 cells, and the landscape includes 14 farms (1-14), grassland (15), forest (16), and water (17).
-The code below produces an image that shows how these terrain types are distributed over the landscape (Figure 3).
+The code below produces an image that shows how these terrain types are distributed over the landscape (Figure C).
 
@@
 Note that because crop and pesticide 3 are dummy variables (not consumed at all), we do not need to plot these.
-Figure 4 shows how population density of pests changes over time.
+Figure D shows how population density of pests changes over time.
 
@@
 
-Figure 5 shows how the mean consumption of different foods and pesticides changes over time.
+Figure E shows how the mean consumption of different foods and pesticides changes over time.
 
@@
 
-Similarly, we can visualise how all four traits evolve over time (Figure 6).
+Similarly, we can visualise how all four traits evolve over time (Figure F).
 
@@
 A new land colour for black `#000000` is then included, and the map is reproduced as an image in R.
-Figure 7 shows the same map as above, but with the numbering removed to make the pests (black points) more visible.
+Figure G shows the same map as above, but with the numbering removed to make the pests (black points) more visible.
 
@@
 To avoid needless repetition, the code for plots will not be included for simulation 2.
-Figure 8 shows population abundance over time for simulation 2.
+Figure H shows population abundance over time for simulation 2.
 
@@
 After pesticides are applied at time step 81, pest population size decreases and does not recover.
-Figure 9 shows crop and pesticide consumption over time.
+Figure I shows crop and pesticide consumption over time.
 
@@
 
-Figure 10 shows how pest traits evolve over time when pesticides are rotated.
+Figure J shows how pest traits evolve over time when pesticides are rotated.
 
@@
 
-We can observe the spatial distribution of pests on the landscape in the final time step (Figure 11).
+We can observe the spatial distribution of pests on the landscape in the final time step (Figure K).
 
@@
 From the output above, we can conclude that when farms rotate the pesticide that they apply between pesticide 1 and pesticide 2, it has a substantial effect on pest density and evolution.
-In the absence of pesticide rotation, the pest population quickly recovers after pesticides are first applied (Figure 4).
-Consequently, pesticides have very little long-term effect on the amount of crops consumed (Figure 5a,b), and while there is an initial spike in the consumption of both pesticides (Figure 5c,d), this consumption quickly declines.
-Despite the positive covariance between crop consumption and pesticide consumption, and the negative covariances between consumption of crop and pesticide types, these trade-offs appear to have only a modest effect on reducing actual crop consumption and maintaining or increasing pesticide consumption (Figure 6).
-Consequently, some farms experience high densities of pests that are resistant to the applied pesticides (Figure 7).
+In the absence of pesticide rotation, the pest population quickly recovers after pesticides are first applied (Figure D).
+Consequently, pesticides have very little long-term effect on the amount of crops consumed (Figure Ea,b), and while there is an initial spike in the consumption of both pesticides (Figure Ec,d), this consumption quickly declines.
+Despite the positive covariance between crop consumption and pesticide consumption, and the negative covariances between consumption of crop and pesticide types, these trade-offs appear to have only a modest effect on reducing actual crop consumption and maintaining or increasing pesticide consumption (Figure F).
+Consequently, some farms experience high densities of pests that are resistant to the applied pesticides (Figure G).
 
-In contrast, when pesticides are rotated every 9 time steps, the pest population density drops after the onset of pesticide application and does not recover (Figure 8).
-The amount of each crop consumed per pests also drops (Figure 9a, b), while the amount of pesticide 1 and 2 consumed is maintained at a much higher rate than occurred in the absence of pesticide rotation (Figure 9c, d).
-Because selection against pesticide consumption was not consistent for a specific farm, and there is a trade-off between pest consumption of pesticide 1 versus 2, local adaptation to a single pesticide did not occur (Figure 10).
-Overall, farms therefore did not experience high densities of pests, and pest resistance to pesticide is well-managed (Figure 11).
+In contrast, when pesticides are rotated every 9 time steps, the pest population density drops after the onset of pesticide application and does not recover (Figure H).
+The amount of each crop consumed per pests also drops (Figure Ia, b), while the amount of pesticide 1 and 2 consumed is maintained at a much higher rate than occurred in the absence of pesticide rotation (Figure Ic, d).
+Because selection against pesticide consumption was not consistent for a specific farm, and there is a trade-off between pest consumption of pesticide 1 versus 2, local adaptation to a single pesticide did not occur (Figure J).
+Overall, farms therefore did not experience high densities of pests, and pest resistance to pesticide is well-managed (Figure K).
 

### 2023-11-30 notebook/ms/evolutionary_algorithm.Rmd
Some edits to the manuscript prior to acceptance from PLoS Computational Biology

@@
 output:
+  bookdown::pdf_document2:
+    fig_caption: yes
+    toc: no
   word_document:
     reference_docx: docx_template.docx
-  bookdown::pdf_document2:
-    citation_package: natbib
-    fig_caption: yes
-    toc: no
   pdf_document:
@@
 An evolutionary algorithm is an algorithm that works especially well for *I know it when I see it* problems [@Hamblin2013].
-@Luke2013 explains the idea behind these algorithms in more detail:
+Luke @Luke2013 explains the idea behind these algorithms in more detail:
 
@@
 
-In the `mine_gmatrix` function of the resevol package, an evolving population of networks like that in Figure 1 above is initialised.
+In the `mine_gmatrix` function of the resevol package, an evolving population of networks like that in Figure A above is initialised.
 Parent networks produce offspring networks with some recombination and mutation, and offspring networks are selected based on how close their trait covariance matrix is to the pre-specified matrix input into the function.
@@
 
-In the code, the arrows in the above Figure 1 are represented by a set of matrices that map loci values to trait values.
-There are 12 loci in Figure 1, and 4 nodes in each hidden layer (blue squares).
+In the code, the arrows in the above Figure A are represented by a set of matrices that map loci values to trait values.
+There are 12 loci in Figure A, and 4 nodes in each hidden layer (blue squares).
 Arrow values between loci and the first hidden layer can then be represented by a matrix with 12 rows and 4 columns (i.e., row 1 holds a value for each of 4 arrows that point to the 4 hidden layer nodes).
@@
 
-To then get the value of the first column of four hidden layer nodes (i.e., the first column of blue squares in Figure 1), we can use matrix multiplication.
+To then get the value of the first column of four hidden layer nodes (i.e., the first column of blue squares in Figure A), we can use matrix multiplication.
 
@@
 
-The evolutionary algorithm first initialises a population of networks, with each network having a unique set of values (i.e., black arrows in Figure 1, represented in the code by matrices explained in the [previous selction](#structures)).
+The evolutionary algorithm first initialises a population of networks, with each network having a unique set of values (i.e., black arrows in Figure A, represented in the code by matrices explained in the [previous selction](#structures)).
 In each iteration of the evolutionary algorithm, with some probability, networks crossover their values with another randomly selected network.
@@
 Iterations continue until either a maximum number of iterations is reached or a network is found that produces trait covariances sufficiently close to the pre-specified covariance matrix.
-Figure 2 below provides a general overview of the evolutionary algorithm.
+Figure B below provides a general overview of the evolutionary algorithm.
 
@@
 An iteration of the evolutionary algorithm begins with a crossover between the values of networks in the population.
-For each network in the population, a crossover event will occur in the values linking loci to the first hidden layer (see Figure 1) with a probability of `pr_cross`.
+For each network in the population, a crossover event will occur in the values linking loci to the first hidden layer (see Figure A) with a probability of `pr_cross`.
 And a second independent crossover event will occur in the values linking the first hidden layer to traits, also with a probability of `pr_cross`.
@@
 A matrix with `loci` rows and a number of columns that matches the number of traits holds the values linking loci to the first hidden layer.
-A three dimensional array with row and column numbers matching trait number, and a depth matching the number of hidden `layers` (e.g., 3 in Figure 1) holds the remaining values linking the first hidden layer to trait values.
+A three dimensional array with row and column numbers matching trait number, and a depth matching the number of hidden `layers` (e.g., 3 in Figure A) holds the remaining values linking the first hidden layer to trait values.
 
@@
 Dimensions of the contiguous set are selected from a random uniform distribution.
-For example, given that the network in Figure 1 would be represented by a three dimensional array with 4 rows, 4 columns, and 3 layers, three random integers from 1-4, 1-4, and 1-3 would be sampled twice, respectively, with replacement.
+For example, given that the network in Figure A would be represented by a three dimensional array with 4 rows, 4 columns, and 3 layers, three random integers from 1-4, 1-4, and 1-3 would be sampled twice, respectively, with replacement.
 If the values selected were 1, 3, and 2 in the first sample, then 3, 3, and 1 in the second sample, then all values from rows 1-3, column 3, and layers 1-2 would be swapped between networks.
-Conceptually, this is the equivalent of drawing a square around set of arrows in Figure 1 and swapping the arrow values with the values of another network.
+Conceptually, this is the equivalent of drawing a square around set of arrows in Figure A and swapping the arrow values with the values of another network.
 
@@
 Elements of each loci vector are randomly sampled from a standard normal distribution.
-For example, in the network of Figure 1 where `loci = 12`, `loci * indivs` standard normal values would be generated in total.
+For example, in the network of Figure A where `loci = 12`, `loci * indivs` standard normal values would be generated in total.
 After these `indivs` loci vectors are initialised, values in each vector are mapped to traits using the focal network, thereby producing `indivs` sets of traits from the focal network.

### 2023-11-30 notebook/ms/ms.Rmd
Some edits to the manuscript prior to acceptance from PLoS Computational Biology

@@
 author: |
-  | A. Bradley Duthie$^{1,4}$, Rosie Mangan$^{1}$, C. Rose McKeon$^{1}$,
+  | A. Bradley Duthie$^{1,*}$, Rosie Mangan$^{1}$, C. Rose McKeon$^{1}$,
   | Matthew C. Tinsley$^{1}$, and Luc F. Bussi&egrave;re$^{2,3}$
 date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
-  UK [2]Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden [3] Gothenburgh Global Biodiversity Centre, Gothenburg, Sweden [4] alexander.duthie@stir.ac.uk, Biological and Environmental Sciences 3A149
-  University of Stirling Stirling, FK9 4LA, UK'
+  UK [2]Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden [3] Gothenburgh Global Biodiversity Centre, Gothenburg, Sweden [*] alexander.duthie@stir.ac.uk'
 output:
@@
     toc: no
+    keep_tex: true
   word_document:
@@
 - \fancyhead[R]{RESEVOL, AN R PACKAGE}
-linestretch: 1
+- \usepackage{lineno}
+- \linenumbers
+- \modulolinenumbers[2]
+linestretch: 2
 link-citations: yes
 linkcolor: blue
+csl: plos-computational-biology.csl
 bibliography: resevol.bib
-biblio-style: apalike
+biblio-style: acm
 ---
@@
 library("resevol");
+options(tinytex.clean = FALSE);
 ```
@@
 
+Design and Implementation
+===================================================
+
 Covarying pest quantitative traits {#mine_gmatrix}
-===================================================
+---------------------------------------------------
 
@@
 
-Conceptually, the relationship between individual genotypes and traits is defined by a network connecting loci to traits through a set of hidden internal nodes (Figure \@ref(fig:network)). 
+Conceptually, the relationship between individual genotypes and traits is defined by a network connecting loci to traits through a set of hidden internal nodes (Fig \@ref(fig:network)). 
 Values for loci are randomly drawn from a standard normal distribution, $\mathcal{N}\left(0, 1\right)$. 
-Links between loci, internal nodes, and traits, can take any real value and are represented by black arrows in Figure \@ref(fig:network). 
-Traits can take any real value, and are calculated as the summed effects of all preceding nodes (i.e., the blue squares immediately to the left of the traits in Figure \@ref(fig:network)).
+Links between loci, internal nodes, and traits, can take any real value and are represented by black arrows in Fig \@ref(fig:network). 
+Traits can take any real value, and are calculated as the summed effects of all preceding nodes (i.e., the blue squares immediately to the left of the traits in Fig \@ref(fig:network)).
 Mathematically, loci are represented by a row vector of length $L$.
-Effects of loci on the first layer of internal nodes (black arrows emanating from loci in Figure \@ref(fig:network)) are represented by an $L \times T$ matrix, and transitions between internal nodes, and between the last set of internal nodes and the final traits, are represented by $T \times T$ matrices.
-For the example in Figure \@ref(fig:network),
+Effects of loci on the first layer of internal nodes (black arrows emanating from loci in Fig \@ref(fig:network)) are represented by an $L \times T$ matrix, and transitions between internal nodes, and between the last set of internal nodes and the final traits, are represented by $T \times T$ matrices.
+For the example in Fig \@ref(fig:network),
 
@@
 
-Values mapping loci to traits become part of an individual's genome, so the genome for the individual represented by Figure \@ref(fig:network) is stored in the model as shown below:
+Values mapping loci to traits become part of an individual's genome, so the genome for the individual represented by Fig \@ref(fig:network) is stored in the model as shown below:
 
@@
 Individuals with different loci can therefore have different covarying traits that are constrained by the network structure encoded in each genome. 
-Individuals can be haploid (as in Figure \@ref(fig:network)) or diploid (in which case, allele values are summed at homologous loci).
+Individuals can be haploid (as in Fig \@ref(fig:network)) or diploid (in which case, allele values are summed at homologous loci).
 
@@
 In the resevol package, the `mine_gmatrix` function runs an evolutionary algorithm and requires the argument `gmatrix`, which specifies the desired trait covariance matrix.
-The function initialises a population of `npsize` separate, and potentially unique, networks (i.e., `npsize` copies of a network like the one shown in Figure \@ref(fig:network)), and this population evolves until some maximum iteration (`max_gen`) or minimum expected network stress (`term_cri`) is met.
+The function initialises a population of `npsize` separate, and potentially unique, networks (i.e., `npsize` copies of a network like the one shown in Fig \@ref(fig:network)), and this population evolves until some maximum iteration (`max_gen`) or minimum expected network stress (`term_cri`) is met.
 In a single iteration of the algorithm, values mutate and crossover occurs between networks (i.e., some of the `npsize` networks swap values with some probability).
@@
 
-An example run of `mine_gmatrix` with the same number of loci and internal nodes as in Figure \@ref(fig:network) is shown below for traits that have an intended covariance of -0.4:
+An example run of `mine_gmatrix` with the same number of loci and internal nodes as in Fig \@ref(fig:network) is shown below for traits that have an intended covariance of -0.4:
 
@@
 Simulating landscape-level pesticide resistance {#simulating}
-===================================================
+---------------------------------------------------
 
@@
 
-Landscape {#landscape}
----------------------------------------------------
+### Landscape {#landscape}
 
@@
 
-Pest ecology {#ecology}
----------------------------------------------------
+### Pest ecology {#ecology}
 
@@
 Initialised pests are of age zero if `rand_age = FALSE` or a random age from zero to `max_age` if `rand_age = TRUE`.
-Following initialisation, a single time step proceeds with landscape change (see above), pest aging and metabolism, feeding, pesticide consumption, movement, reproduction, mortality, and immigration (Figure \@ref(fig:flowchart)).
+Following initialisation, a single time step proceeds with landscape change (see above), pest aging and metabolism, feeding, pesticide consumption, movement, reproduction, mortality, and immigration (Fig \@ref(fig:flowchart)).
 Feeding, pesticide consumption, movement, and reproduction all depend on pest age.
@@
 
-Pest evolution {#evolution}
----------------------------------------------------
+### Pest evolution {#evolution}
 
@@
 
-Evolution of the genetic architecture linking loci to traits can be constrained by disabling mutation in genome values linking loci, internal nodes, and traits (i.e., 'network values' represented by arrows in Figure \@ref(fig:network)).
-While mutation at loci (green circles in Figure \@ref(fig:network)) is always possible as long as `mutation_pr > 0`, the number of intermediary layers for which network values can mutate is constrained by `net_mu_layers`.
+Evolution of the genetic architecture linking loci to traits can be constrained by disabling mutation in genome values linking loci, internal nodes, and traits (i.e., 'network values' represented by arrows in Fig \@ref(fig:network)).
+While mutation at loci (green circles in Fig \@ref(fig:network)) is always possible as long as `mutation_pr > 0`, the number of intermediary layers for which network values can mutate is constrained by `net_mu_layers`.
 If `net_mu_layers = 0`, then no network values can mutate, but higher integer values cause mutation to occur at network value layers from loci to traits (if `net_mu_dir = 1`) or traits to loci (if `net_mu_dir = 0`).
@@
 Simulation output {#output}
-===================================================
+---------------------------------------------------
 
@@
 
-Example of individual-based simulations {#simulation}
+Results {#simulation}
 ===================================================
@@
 These arguments are explained in the package documentation, and in Supporting Information 2, which demonstrates an advanced case study with a custom landscape and complex pest genomes and life history.
-Here we focus on a simple simulation with asexually reproducing pests that have three loci and two traits (Figure \@ref(fig:network)).
+Here we focus on a simple simulation with asexually reproducing pests that have three loci and two traits (Fig \@ref(fig:network)).
 We use the the genome generated in `new_network` from [Section 2](#mine_gmatrix), in which traits 1 and 2 have variances of `r new_network[[6]][1,1]` and `r new_network[[6]][2,2]`, respectively, with a covariance of `r new_network[[6]][1,2]`.
@@
 [Output files](#output) can be used to plot ecological and evolutionary dynamics of pests. 
-Figure \@ref(fig:overtime)a shows the pest population change both across and within seasons.
+Fig \@ref(fig:overtime)a shows the pest population change both across and within seasons.
 Due to the specific parameter values chosen, clear patterns reflecting pest cohorts emerge.
@@
 As less crop remains in a season, pest abundance decreases.
-Mean food consumed varies over the course of a season caused by the varying frequency of pests in different life history stages (Figure \@ref(fig:overtime)c).
-In the first 50 time steps, there is some consumption of pesticides 1 and 2 (Figure \@ref(fig:overtime)b), but both traits rapidly evolve to negative values making both pesticides ineffective (Figure \@ref(fig:overtime)d).
+Mean food consumed varies over the course of a season caused by the varying frequency of pests in different life history stages (Fig \@ref(fig:overtime)c).
+In the first 50 time steps, there is some consumption of pesticides 1 and 2 (Fig \@ref(fig:overtime)b), but both traits rapidly evolve to negative values making both pesticides ineffective (Fig \@ref(fig:overtime)d).
 In this case, the trade-off in pesticide consumption is not strong enough to maintain susceptibility to either pesticide in the population.
-Figure \@ref(fig:landscape) shows the location of pests on the landscape in the last time of the simulation (left panel), and how different pesticides are currently being applied on the landscape (right panel).
+Fig \@ref(fig:landscape) shows the location of pests on the landscape in the last time of the simulation (left panel), and how different pesticides are currently being applied on the landscape (right panel).
 Because the number of crops and pesticides, the number and nature of traits, and the size of pest genomes all vary depending on simulation parameters, no plotting functions are introduced in the resevol package.
-Instead, methods and code for producing plots such as those in Figures \@ref(fig:overtime) and \@ref(fig:landscape) are explained in Supporting Information 2.
+Instead, methods and code for producing plots such as those in Fig \@ref(fig:overtime) and \@ref(fig:landscape) are explained in Supporting Information 2.
 
@@
 
-Conclusions {#conclusions}
+Availability and Future Directions {#conclusions}
 ===================================================
@@
 The resevol R package makes it possible to rapidly develop and simulate myriad individual-based models of resistance evolution.
-It also introduces a novel approach to modelling complex pest genetic architecture, using an evolutionary algorithm to generate haploid or diploid loci that map to pest traits with pre-specified covariances (Figure \@ref(fig:network); Supporting Information 1).
+It also introduces a novel approach to modelling complex pest genetic architecture, using an evolutionary algorithm to generate haploid or diploid loci that map to pest traits with pre-specified covariances (Fig \@ref(fig:network); Supporting Information 1).
 Agricultural landscapes and pest life histories are highly customisable, allowing targetted models that can simulate specific real-world case studies (Supporting Information 2).
@@
 
-Acknowledgements {#acknowledgements}
-===================================================
-
-This software was developed as part of the project for Enhancing Diversity to Overcome Resistance Evolution (ENDORSE). 
-The ENDORSE project is a joint Newton funded international partnership between the Biotechnology and Biological Sciences Research Council (BBSRC) in the UK and the S&#xE3;o Paulo Research Foundation (FAPESP) in Brazil under BBSRC award reference BB/S018956/1 and FAPESP award reference 2018/21089-3.
-ENDORSE is a partnership among Universidade Estadual Paulista (UNESP), the University of Stirling (UoS), and the Centre for Agricultural and Biosciences International (CABI).
+The resevol R package can be downloaded from CRAN (https://cran.r-project.org/package=resevol) or GitHub (https://bradduthie.github.io/resevol/). 
+The package is open source under GNU Public License.
 
@@
 
-Availability {#availability}
+Financial Disclosure statement {#financial}
 ===================================================
 
-The resevol R package can be downloaded from CRAN (https://cran.r-project.org/package=resevol) or GitHub (https://bradduthie.github.io/resevol/). 
-The package is open source under GNU Public License.
+This software was developed as part of the project for Enhancing Diversity to Overcome Resistance Evolution (ENDORSE) led by Luc Bussi&egrave;re, Ricardo Polanczyk, and Matthew Tinsley (Co-Investigators: Nils Bunnefeld, Yelitza Colmenarez, Nat&aacute;lia Corniani, Renata de Lima, Brad Duthie, Steve Edgington, Leonardo Fraceto, Belinda Luke, and Rosie Mangan).
+The ENDORSE project is a joint Newton funded international partnership between the Biotechnology and Biological Sciences Research Council (BBSRC) in the UK and the S&atilde;o Paulo Research Foundation (FAPESP) in Brazil under BBSRC award reference BB/S018956/1 and FAPESP award reference 2018/21089-3.
+ENDORSE is a partnership among Universidade Estadual Paulista (UNESP), the University of Stirling (UoS), and the Centre for Agricultural and Biosciences International (CABI). 
+The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
+
+Competing interests {#competing}
+===================================================
+
+The authors have declared that no competing interests exist.
 
@@
 ===================================================
+
+Supporting information {#SI}
+===================================================
+
+S1 Text.
+The evolutionary algorithm of the resevol R package, including key data structures used, a general overview of the evolutionary algorithm, and details concerning haploid and diploid individuals.
+(PDF)
+
+S2 Text. Advanced techniques for using the resevol R package, including initialising pest genomes and running simulations. (PDF)

### 2023-11-30 notebook/ms/ms.tex
Some edits to the manuscript prior to acceptance from PLoS Computational Biology

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={resevol: an R package for spatially explicit models of pesticide resistance given evolving pest genomes},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[R]{RESEVOL, AN R PACKAGE}
\usepackage{lineno}
\linenumbers
\modulolinenumbers[2]
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{acm}

\title{resevol: an R package for spatially explicit models of pesticide resistance given evolving pest genomes}
\author{A. Bradley Duthie\(^{1,*}\), Rosie Mangan\(^{1}\), C. Rose McKeon\(^{1}\),\\
Matthew C. Tinsley\(^{1}\), and Luc F. Bussière\(^{2,3}\)}
\date{{[}1{]} Biological and Environmental Sciences, University of Stirling, Stirling, UK {[}2{]}Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden {[}3{]} Gothenburgh Global Biodiversity Centre, Gothenburg, Sweden {[}*{]} \href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}}

\begin{document}
\maketitle

\setstretch{2}
\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\hypertarget{abstract}{%
\section*{Abstract}\label{abstract}}
\addcontentsline{toc}{section}{Abstract}

The evolution of pesticide resistance is a widespread problem with potentially severe consequences for global food security.
We introduce the resevol R package, which simulates individual-based models of pests with evolving genomes that produce complex, polygenic, and covarying traits affecting pest life history and pesticide resistance.
Simulations are modelled on a spatially-explicit and highly customisable landscape in which crop and pesticide application and rotation can vary, making the package a highly flexible tool for both general and tactical models of pest management and resistance evolution.
We present the key features of the resevol package and demonstrate its use for a simple example simulating pests with two covarying traits.
The resevol R package is open source under GNU Public License. All source code and documentation are available on GitHub.

\textbf{Key words:} \emph{pest management}, \emph{food security}, \emph{ecological modelling}, \emph{pesticide}, \emph{resistance evolution}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Insect resistance to pesticides is a wicked and widespread problem \citep{Gould2018}.
Predicting, identifying, and ultimately delaying the evolution of insecticide resistance is therefore critical for ensuring sustainable global food security \citep{Hawkins2019, Lykogianni2021}.
To manage pesticide resistance, various strategies have been proposed to disrupt or weaken selection.
These strategies often focus on varying the location and timing of pesticide application, and therefore varying the strength of selection for pesticide resistance, so that pest susceptibility to one or more pesticides is maintained \citep{Ives2002, Liang2013, Carriere2016, Liang2018, South2018, Sudo2018, Saikai2021}.
Pesticide resistance management has mostly been associated with the effect of single genes.
In such cases, resistance alleles have binary effects on phenotype, enabling resistant phenotypes to arise from genetic changes at single or a small number of loci \citep{Raymond2001, Ffrench-Constant2004, Ffrench-Constant2013}.
Nevertheless, resistance attributable to polygenic effects is also well-established \citep{Via1986, Firko1990, Busi2013, Haridas2018, Green2020, Mangan2023}.
And the relevance of polygenic resistance is likely to further increase given a rising interest in biological controls for sustainable crop protection \citep[e.g.,][]{Green2020, Elnahal2022, Mangan2023}.
Developing strategies to maximise the efficacy of these tools is critical, and such strategies should be well-informed by predictions made from detailed, quantitative genetic models.
Here we introduce the resevol R package as a tool for building individual-based models and simulating pest management \citep{Daly2022}.

The resevol package applies individual-based modelling and a quantitative genetics approach to simulate the evolution of a pest population on a changing landscape.
Multiple traits determine the overall fitness of any pest genotype.
For example, while alleles conferring resistance to a particular pesticide enhance fitness in the presence of that pesticide, they could also impose reproductive fitness costs.
Such trade-offs can be quantified by the genetic covariance between traits.
A focal goal of the resevol package is to model traits with a pre-specified, but potentially evolving, genetic covariance structure.
To achieve this goal, each individual has a genome with \(L\) loci that underlie a set of \(T\) potentially evolving traits.
Pleiotropic loci can vary in their effects on the direction and magnitude of polygenic traits, causing population-wide trait covariance to arise mechanistically from the underlying genetic architecture of individuals.
To achieve this, two separate steps are necessary.
First, an evolutionary algorithm is used to find a network of internal nodes that map standard random normal loci values to covarying traits.
Values used to map loci to traits are incorporated into individual genomes.
Second, a population of asexual or sexual individuals is initialised and simulated on a spatially explicit landscape separated into distinct units (e.g., farms).
Land units can apply one of up to 10 pesticides, and one of up to 10 landscape types at a given time (for simplicity, here we interpret landscape types to be crop types).
Pesticides and crops rotate independently within farms over time in a pre-specified way.
The resevol package can thereby model complex and evolving agricultural pest traits over realistic landscapes that undergo different pesticide use and crop regimes.

\hypertarget{design-and-implementation}{%
\section{Design and Implementation}\label{design-and-implementation}}

\hypertarget{mine_gmatrix}{%
\subsection{Covarying pest quantitative traits}\label{mine_gmatrix}}

The first step of simulation is building individual genomes.
This step is separate because it is computationally intensive, and genomes that are built might need to be inspected and stored.
High computation time is due to the mechanistic nature of how genomes and covarying traits are modelled.
Instead of imposing a trait covariance structure directly, an evolutionary algorithm is used to find a network that maps standard random normal values (loci) to covarying values (traits).
This is useful because it allows genomes to model potentially evolving physiological constraints, and trade-offs among traits, from the bottom up.
Since multiple networks can potentially map loci to the same trait covariance structure, it is possible to replicate evolution with different randomly generated genetic architectures.
This approach to modelling individual genomes and traits thereby increases the complexity of questions that can be addressed for simulating evolution in agricultural pests.

\begin{figure}
\centering
\includegraphics{ms_files/figure-latex/network-1.pdf}
\caption{\label{fig:network}Example network mapping loci (green circles) to traits (red diamonds) through an intermediate set of hidden layers (blue squares) in the \texttt{mine\_gmatrix} function. Individual genomes in the resevol R package consist of standard random normal values for loci, real values for black arrows linking nodes, and real values for traits. Values shown for loci and arrows are an example for illustration.}
\end{figure}

Conceptually, the relationship between individual genotypes and traits is defined by a network connecting loci to traits through a set of hidden internal nodes (Fig \ref{fig:network}).
Values for loci are randomly drawn from a standard normal distribution, \(\mathcal{N}\left(0, 1\right)\).
Links between loci, internal nodes, and traits, can take any real value and are represented by black arrows in Fig \ref{fig:network}.
Traits can take any real value, and are calculated as the summed effects of all preceding nodes (i.e., the blue squares immediately to the left of the traits in Fig \ref{fig:network}).
Mathematically, loci are represented by a row vector of length \(L\).
Effects of loci on the first layer of internal nodes (black arrows emanating from loci in Fig \ref{fig:network}) are represented by an \(L \times T\) matrix, and transitions between internal nodes, and between the last set of internal nodes and the final traits, are represented by \(T \times T\) matrices.
For the example in Fig \ref{fig:network},

\[
\begin{pmatrix}
  0.1, & -0.1, & 0.2 \\
\end{pmatrix} 
\begin{pmatrix}
  0.4, &  0.1 \\
 -1.2, & -0.6 \\
  0.2, &  2.2 \\
\end{pmatrix}
\begin{pmatrix}
   1.3, & -0.5 \\
  -0.1  &  0.0 \\
\end{pmatrix}
\begin{pmatrix}
   2.2, &  0.9 \\
   1.6  &  0.8 \\
\end{pmatrix} = 
\begin{pmatrix}
  0.2998, & 0.1081 \\
\end{pmatrix}.
\]

Values mapping loci to traits become part of an individual's genome, so the genome for the individual represented by Fig \ref{fig:network} is stored in the model as shown below:

\begin{verbatim}
0.1, -0.1, -0.2, 0.4, 0.1, -1.2, -0.6, 0.2, 2.2, 1.3, -0.5, -0.1, 0.0, 2.2, 0.9, 1.6, 0.8
\end{verbatim}

Individuals with different loci can therefore have different covarying traits that are constrained by the network structure encoded in each genome.
Individuals can be haploid (as in Fig \ref{fig:network}) or diploid (in which case, allele values are summed at homologous loci).

An evolutionary algorithm is used to find appropriate values that produce covarying traits from loci (see Supporting Information 1 for details).
Evolutionary algorithms are heuristic tools that can simulate adaptive evolution to find solutions for a broad range of problems \citep{Hamblin2013, Duthie2018}.
In the resevol package, the \texttt{mine\_gmatrix} function runs an evolutionary algorithm and requires the argument \texttt{gmatrix}, which specifies the desired trait covariance matrix.
The function initialises a population of \texttt{npsize} separate, and potentially unique, networks (i.e., \texttt{npsize} copies of a network like the one shown in Fig \ref{fig:network}), and this population evolves until some maximum iteration (\texttt{max\_gen}) or minimum expected network stress (\texttt{term\_cri}) is met.
In a single iteration of the algorithm, values mutate and crossover occurs between networks (i.e., some of the \texttt{npsize} networks swap values with some probability).
Next, trait covariances produced for each network are estimated by initalising \texttt{indivs} individuals with loci sampled from a standard normal distribution.
Network stress is calculated as the logged mean squared deviation between estimated covariances and those in \texttt{gmatrix}.
Tournament selection \citep{Hamblin2013} is then used to determine the networks for the next iteration of the algorithm.
Networks with the lowest estimated stress have the highest fitness, so these networks are disproportionately represented in the next iteration.
Throughout the evolutionary algorithm, the lowest stress network is saved and returned upon network termination.
The robustness of this network's stress to sets of individuals with different loci values can be tested using the \texttt{stress\_test} function.

An example run of \texttt{mine\_gmatrix} with the same number of loci and internal nodes as in Fig \ref{fig:network} is shown below for traits that have an intended covariance of -0.4:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{trait\_covs  }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\AttributeTok{data =} \FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\SpecialCharTok{{-}}\FloatTok{0.4}\NormalTok{, }\SpecialCharTok{{-}}\FloatTok{0.4}\NormalTok{, }\DecValTok{1}\NormalTok{), }\AttributeTok{nrow =} \DecValTok{2}\NormalTok{, }\AttributeTok{ncol =} \DecValTok{2}\NormalTok{);}
\NormalTok{new\_network }\OtherTok{\textless{}{-}} \FunctionTok{mine\_gmatrix}\NormalTok{(}\AttributeTok{loci =} \DecValTok{3}\NormalTok{, }\AttributeTok{layers =} \DecValTok{2}\NormalTok{, }\AttributeTok{gmatrix =}\NormalTok{ trait\_covs, }
                            \AttributeTok{max\_gen =} \DecValTok{1000}\NormalTok{, }\AttributeTok{term\_cri =} \SpecialCharTok{{-}}\FloatTok{6.0}\NormalTok{, }\AttributeTok{prnt\_out =} \ConstantTok{FALSE}\NormalTok{);}
\end{Highlighting}
\end{Shaded}

The code above found a genome that produced the following expected trait covariances:

\begin{verbatim}
##            [,1]       [,2]
## [1,]  1.0008181 -0.3925799
## [2,] -0.3925799  1.0271340
\end{verbatim}

The mean deviation between elements of the above matrix and the identity matrix provided by \texttt{trait\_covs} is \ensuremath{2.1176024\times 10^{-4}}.
Lower values of \texttt{term\_cri} and higher values of \texttt{max\_gen} will result in a lower stress, but this will require more computation time, especially if the number of traits is high.
Similarly, higher values of \texttt{indivs} will result in more accurate estimations of true stress, but this also requires more computation time.
Additional arguments to \texttt{mine\_gmatrix} can also be used to improve the performance of the evolutionary algorithm (see Supporting Information 1).

\hypertarget{simulating}{%
\subsection{Simulating landscape-level pesticide resistance}\label{simulating}}

The full output of \texttt{mine\_gmatrix} is passed to the \texttt{run\_farm\_sim} function, which initialises and simulates an evolving population of pests on a changing landscape for any natural number of time steps (\texttt{time\_steps}).
In this section, we explain the landscape, pest ecology, and evolving pest traits.

\hypertarget{landscape}{%
\subsubsection{Landscape}\label{landscape}}

Landscapes are spatially explicit and initialised in one of two ways.
First, a landscape can be built from the arguments \texttt{xdim}, \texttt{ydim}, and \texttt{farms}.
These arguments specify the cell dimensions and number of farms on the landscape.
Contiguous rectangular farms of roughly equal size are generated on the landscape using a splitline algorithm.
Second, a custom landscape can be input using the \texttt{terrain} argument, which takes a matrix with elements that include integers 1 to \texttt{farms}.
Each value defines a unique farm, but values do not need to be contiguous.
These `farms' could even model non-farmland (e.g., water, roads), if pesticides and crops on them are invisible to pests (see Supporting Information 2), or they could more broadly be interpreted as heterogenous landscape properties across an arbitrary scale \citep[e.g.,][]{Newman2023}.
This \texttt{terrain} customisation therefore allows for a high degree of landscape detail, and offers the potential for modelling real-world landscapes from raster images \citep[e.g.,][]{Millington2021}.
Edge effects are set using the \texttt{land\_edge} argument.
Edge options include \texttt{"leaky"} (pests disappear from the landscape), \texttt{"reflect"} (pests bounce off of the edge), \texttt{"sticky"} (pests stick to the edge), or \texttt{"torus"} (pests that move off of one edge return on the opposite side of the landscape).

Each farm can hold one pesticide and one crop type in any time step.
The \texttt{crop\_init} and \texttt{pesticide\_init} arguments initialise one of \texttt{crop\_number} crops and one of \texttt{pesticide\_number} pesticides for each farm, respectively (maximum of 10 each).
Initialisation can be random for each farm with equal probability, or it can be set using a vector of length \texttt{farms} in which vector elements define the initialised crop or pesticide number.
After initialisation, crops and pesticides rotate once every \texttt{crop\_rotation\_time} and \texttt{pesticide\_rotation\_time} time steps, respectively.
The arguments \texttt{crop\_rotation\_type} and \texttt{pesticide\_rotation\_type} specify how crops and pesticides are rotated, respectively.
Each of these rotation type arguments can take either an integer value from 1-3, or a matrix.
Integer values specify (1) no rotation, (2) random transition from one type to another, or (3) cycling through each available crop or pesticide in numeric order.
Square matrices can be used to define the probability that a given crop or pesticide in row \(i\) transitions to that in column \(j\).
Hence, any possible Markov chain can be used to model transition between crop or pesticide types on farms, potentially integrating real crop use or pesticide use patterns \citep{Zhang2020}.
Upon rotation, crop and pesticide values are reset on each farm.

Values for \texttt{crop\_per\_cell} and \texttt{pesticide\_per\_cell} determine the quantity of crops and pesticides initialised per cell upon crop or pesticide rotation, respectively.
Between crop rotations, crop values can increase each time step by a proportion or increment \texttt{crop\_growth}, which depends on \texttt{crop\_growth\_type}.
These arguments can be used to model crop growth over a season.

\hypertarget{ecology}{%
\subsubsection{Pest ecology}\label{ecology}}

Individual pests can be modelled to have several reproductive systems and life histories.
Pest reproductive system can be specified using the \texttt{repro} argument, which accepts \texttt{"asexual"} (haploid), \texttt{"sexual"} (monoecious), and \texttt{"biparental"} (dioecious).
For \texttt{"sexual"} pests, the \texttt{selfing} argument specifies if self-fertilisation is (\texttt{TRUE}) or is not (\texttt{FALSE}) allowed.
At the start of a simulation, pests are initialised in a random location.
Initialised pests are of age zero if \texttt{rand\_age\ =\ FALSE} or a random age from zero to \texttt{max\_age} if \texttt{rand\_age\ =\ TRUE}.
Following initialisation, a single time step proceeds with landscape change (see above), pest aging and metabolism, feeding, pesticide consumption, movement, reproduction, mortality, and immigration (Fig \ref{fig:flowchart}).
Feeding, pesticide consumption, movement, and reproduction all depend on pest age.
Pests feed and consume pesticide from ages \texttt{min\_age\_feed} to \texttt{max\_age\_feed}, move from ages \texttt{min\_age\_move} to \texttt{max\_age\_move}, and reproduce from ages \texttt{min\_age\_reproduce} to \texttt{max\_age\_reproduce} (all inclusive).
Food accumulated is lost during aging if \texttt{baseline\_metabolism\ \textgreater{}\ 0} and pest age is within \texttt{min\_age\_metabolism} and \texttt{max\_age\_metabolism}.
The option to set minimum and maximum ages for events makes it possible to model pests with much different life histories \citep[e.g.,][]{Sudo2018}.

\begin{figure}
\centering
\includegraphics{ms_files/figure-latex/flowchart-1.pdf}
\caption{\label{fig:flowchart}Overview of simulated events in the resevol R package. Note that metabolism, feeding, pesticide consumption, movement, and reproduction are all subject to a minimum and maximum pest age. Consequently, simulation order might not reflect the order of events from the perspective of a focal pest (e.g., pests might move from ages 1-2, but only feed from ages 2-4). Crops and pesticides are also not necessarily rotated in each time step (see Landscape). Statistics collected within a time step are printed to a CSV file.}
\end{figure}

In each time step, pests feed in a random order, consuming the crop on their landscape cell. Each pest consumes an amount of crop as specified by the \texttt{food\_consume} argument, which takes a vector with as many elements as their are crops (i.e., if \texttt{crop\_number\ =\ 2}, then \texttt{food\_consume} has two elements, the first and second defining consumption of crops 1 and 2, respectively).
If crop amount on the landscape cell exceeds pest consumption ability, then pests consume their maximum amount, and this amount is removed from the landscape cell.
If crop amount is less than pest consumption ability, then pests consume whatever crop is left and crop amount is reduced to zero.
Pesticide consumption works identically to crop consumption, except that the amount of pesticide on a landscape cell is not decreased.
Hence, each pest can potentially feed and be affected by the pesticide of their focal landscape cell.
Pests simply consume an amount of pesticide as specified by the \texttt{pesticide\_consume} argument, which also takes a vector with as many elements as their are pesticides.
Pesticide consumption affects pest survival and reproduction using \texttt{pesticide\_tolerated\_surv} and \texttt{pesticide\_tolerated\_repr} arguments.

After interacting with their landscape cell, pests can move.
Each pest visits a number of cells during movement, as is specified by the parameter \texttt{movement\_bouts}.
Individual movement bouts occur in a random order across pests.
During a movement bout, a pest can travel to any cell within a value defined by \texttt{movement\_distance} from their current location, which could include their current location (i.e., moving zero distance).
Upon arrival to a cell, a pest can feed if they are of an appropriate feeding age and \texttt{feed\_while\_moving\ =\ TRUE}.
The pest also consumes pesticide if they are of the appropriate age and \texttt{pesticide\_while\_moving\ =\ TRUE}.
Having pests feed and consume pesticide in a random order while moving among landscape cells can model a population competing for food and encountering pesticides on a shorter time scale than an individual time step.

After all pests finish moving, pests reproduce.
Offspring production is possible for asexual, monoecious, or female pests.
Pest expected offspring number is defined by a fixed parameter if \texttt{reproduction\_type\ =\ "lambda"}, or is calculated from the amount of food consumed if \texttt{reproduction\_type\ =\ "food\_based"}.
The former requires specifying \texttt{lambda\_value}, which becomes the rate parameter for sampling offspring number from a Poisson distribution.
The latter requires specifying a real value for \texttt{food\_needed\_repr}, which is the amount of food needed to produce one offspring.
For food-based reproduction, the total amount of food consumed is divided by \texttt{food\_needed\_repr}, then floored to determine offspring number.
Sexual reproduction requires a mate of reproductive age that is either monoecious or male, and within range of the reproducing focal pest (potential including the focal pest, if \texttt{selfing\ =\ TRUE}).
A potential mate is within range if it is within an integer number of cells from the focal pest, as defined by \texttt{mating\_distance} (e.g., if \texttt{mating\_distance\ =\ 0}, then mates must share a landscape cell).
All available potential mates sire offspring with equal probability, and reproducing pests are assumed to mate multiply (i.e., paternity is a fair raffle for all offspring).
If a carrying capacity at birth is set (\texttt{K\_on\_birth\ \textgreater{}\ 0}) and total offspring number in the population exceeds this capacity, then offspring are removed at random until they are within carrying capacity.
A real value \texttt{immigration\_rate} specifies the rate parameter for Poisson random sampling of the number of immigrants added to the population at the end of a time step.
Immigrants are initialised in the same way as pests are at the start of the simulation.
Hence, any spatial structure or evolution that occurs during the simulation does not affect immigrant locations, genomes, or traits.

\hypertarget{evolution}{%
\subsubsection{Pest evolution}\label{evolution}}

Pest genomes evolve in a complex and highly mechanistic way.
Offspring inherit genome values from their parent(s) with the possibility for mutation and recombination; offspring traits are then calculated from their newly initialised genomes.
For asexually and sexually reproducing pests, genomes are haploid and diploid, respectively.
Asexually reproducing pests receive the full genomes of their parent, while sexually reproducing pests receive half of their alleles from each parent.
Each diploid parent contributes one half of their genome, effectively modelling a genome with a single chromosome.
Crossover occurs at each position of the genome with a probability of \texttt{crossover\_pr}.
When a crossover event occurs, alleles are swapped between the two chromosomes, so complete recombination is also possible if \texttt{crossover\_pr\ =\ 0.5}.
For both haploids and diploids, each genome value then mutates independently with a probability of \texttt{mutation\_pr}, which can be set to any real number from 0-1.
If a genome value mutates, then a new value is randomly sampled from a standard normal distribution.
If \texttt{mutation\_type\ =\ 0}, then this new value replaces the old value, and if \texttt{mutation\_type\ =\ 1}, then the new value is added to the old value.
After mutation, genome values are used to calculate trait values.

Evolution of the genetic architecture linking loci to traits can be constrained by disabling mutation in genome values linking loci, internal nodes, and traits (i.e., `network values' represented by arrows in Fig \ref{fig:network}).
While mutation at loci (green circles in Fig \ref{fig:network}) is always possible as long as \texttt{mutation\_pr\ \textgreater{}\ 0}, the number of intermediary layers for which network values can mutate is constrained by \texttt{net\_mu\_layers}.
If \texttt{net\_mu\_layers\ =\ 0}, then no network values can mutate, but higher integer values cause mutation to occur at network value layers from loci to traits (if \texttt{net\_mu\_dir\ =\ 1}) or traits to loci (if \texttt{net\_mu\_dir\ =\ 0}).
For example, if \texttt{net\_mu\_layers\ =\ 2} and \texttt{net\_mu\_dir\ =\ 1}, then the network values linking loci to the first internal node, and the first internal node to the second, can mutate (i.e., first two columns of arrows in \ref{fig:network}, but not those linking the second internal node to traits).
This allows pest traits to evolve with varying degrees of constraint on the covariance between traits.
Low \texttt{net\_mu\_layers} values model strong genetic constraints, while high values model high evolvability of trait covariances.

Finally, evolving and covarying traits can be used in place of fixed parameters described in pest ecology.
This is done by substituting \texttt{"Tj"} as an argument input in place of a numeric value, where \texttt{j} represents the trait number.
For example, the argument \texttt{move\_distance\ =\ "T1"} will make Trait 1 the movement distance for individuals.
The argument \texttt{food\_consume\ =\ c("T2",\ "T3")} will set Traits 2 and 3 to define the amount of food of types 1 and 2 that can be consumed by a pest, respectively.
Up to 10 of the following parameters can be replaced with evolving traits: \texttt{move\_distance}, \texttt{food\_needed\_surv}, \texttt{pesticide\_tolerated\_surv}, \texttt{food\_needed\_repr}, \texttt{pesticide\_tolerated\_repr}, \texttt{mating\_distance}, \texttt{lambda\_value}, \texttt{movement\_bouts}, \texttt{metabolism}, \texttt{food\_consume}, and \texttt{pesticide\_consume}.
Mean values of traits can be set using the argument \texttt{trait\_means}, which accepts a vector of the same length as the number of evolving traits such that indices correspond to trait numbers.
The resevol package can thereby simulate agricultural pests with complex and co-evolving traits, and potentially evolving trait covariances, under a range of possible pest life histories.

\hypertarget{output}{%
\subsection{Simulation output}\label{output}}

Simulation output is typically large, so output is printed in two CSV files, both of which are created in the working directory.
The first file ``population\_data.csv'' prints population level data over time, including population size, mean age, sex ratio, mean food and pesticide consumed of each type, mortality rate, and mean trait values.
The second file ``individuals.csv'' prints all information, including full genomes and traits (columns), for every individual (rows) in the population.
The printing of individual level data is disabled by default.
It can be turned on for all time steps by setting \texttt{print\_inds\ =\ TRUE}, but this should be done with caution because it can create extremely large files.
Instead, individual level data can be printed for only the final time step by setting \texttt{print\_last\ =\ TRUE}.
Output produced by \texttt{run\_farm\_sim} is a list of two elements, which includes a vector of parameter values used in the simulation and the final state of the landscape as an array.

\hypertarget{simulation}{%
\section{Results}\label{simulation}}

In resevol v0.3, the \texttt{run\_farm\_sim} function includes 68 arguments, which specify a wide range of possible simulation conditions affecting landscape and pest characteristics.
These arguments are explained in the package documentation, and in Supporting Information 2, which demonstrates an advanced case study with a custom landscape and complex pest genomes and life history.
Here we focus on a simple simulation with asexually reproducing pests that have three loci and two traits (Fig \ref{fig:network}).
We use the the genome generated in \texttt{new\_network} from \protect\hyperlink{mine_gmatrix}{Section 2}, in which traits 1 and 2 have variances of 1.0008181 and 1.027134, respectively, with a covariance of -0.3925799.

Traits 1 and 2 will define the realised rate of uptake of the two separate pesticides, so we model a system in which there is a potential trade-off for pesticide susceptibility.
We use a simple \(64 \times 64\) cell landscape with nine farms.
Each farm grows the same crop and uses one of two randomly intialised pesticides, which are rotated every 16 time steps.
Hence, we can conceptualise 16 time steps as a single growing season.

In our example, each cell produces four crop units, all of which can potentially be consumed.
Pests consume up to one unit of crop on their landscape cell per time step.
Pest survival and reproduction is food-based, and pests need to consume one unit of crop by age two to survive and reproduce.
Pests initialised at the start of the simulation are randomly assigned an age from 0-4 with equal probability.
Pests have a maximum age of four, and they feed in ages 0-2, move up to two cells in ages 3-4, and reproduce in age four.
Pests can uptake (i.e., `consume') pesticides in ages 0-2, and if they consume any pesticide, then they will die.
In each time step, a mean of 10 immigrants arrive.
We simulate 160 time steps (10 growing seasons) using the function below:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sim }\OtherTok{\textless{}{-}} \FunctionTok{run\_farm\_sim}\NormalTok{(}\AttributeTok{mine\_output =}\NormalTok{ new\_network, }\AttributeTok{repro =} \StringTok{"asexual"}\NormalTok{, }
                    \AttributeTok{pesticide\_number =} \DecValTok{2}\NormalTok{, }\AttributeTok{pesticide\_init =} \StringTok{"random"}\NormalTok{, }
                    \AttributeTok{pesticide\_consume =} \FunctionTok{c}\NormalTok{(}\StringTok{"T1"}\NormalTok{, }\StringTok{"T2"}\NormalTok{), }\AttributeTok{farms =} \DecValTok{9}\NormalTok{,}
                    \AttributeTok{pesticide\_rotation\_time =} \DecValTok{16}\NormalTok{, }\AttributeTok{pesticide\_rotation\_type =} \DecValTok{3}\NormalTok{,}
                    \AttributeTok{pesticide\_tolerated\_surv =} \DecValTok{0}\NormalTok{, }\AttributeTok{pesticide\_per\_cell =} \DecValTok{1}\NormalTok{,}
                    \AttributeTok{crop\_rotation\_time =} \DecValTok{16}\NormalTok{, }\AttributeTok{crop\_number =} \DecValTok{1}\NormalTok{, }\AttributeTok{crop\_per\_cell =} \DecValTok{4}\NormalTok{, }
                    \AttributeTok{food\_consume =} \DecValTok{1}\NormalTok{, }\AttributeTok{reproduction\_type =} \StringTok{"food\_based"}\NormalTok{, }
                    \AttributeTok{food\_needed\_surv =} \DecValTok{1}\NormalTok{, }\AttributeTok{food\_needed\_repr =} \DecValTok{1}\NormalTok{, }\AttributeTok{max\_age =} \DecValTok{4}\NormalTok{, }
                    \AttributeTok{min\_age\_feed =} \DecValTok{0}\NormalTok{, }\AttributeTok{max\_age\_feed =} \DecValTok{2}\NormalTok{, }\AttributeTok{min\_age\_move =} \DecValTok{3}\NormalTok{, }
                    \AttributeTok{max\_age\_move =} \DecValTok{4}\NormalTok{, }\AttributeTok{min\_age\_reproduce =} \DecValTok{4}\NormalTok{, }\AttributeTok{print\_gens =} \ConstantTok{FALSE}\NormalTok{,}
                    \AttributeTok{max\_age\_reproduce =} \DecValTok{4}\NormalTok{, }\AttributeTok{age\_pesticide\_threshold =} \DecValTok{2}\NormalTok{, }
                    \AttributeTok{rand\_age =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{move\_distance =} \DecValTok{2}\NormalTok{, }\AttributeTok{immigration\_rate =} \DecValTok{10}\NormalTok{, }
                    \AttributeTok{time\_steps =} \DecValTok{160}\NormalTok{, }\AttributeTok{print\_last =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{xdim =} \DecValTok{64}\NormalTok{, }\AttributeTok{ydim =} \DecValTok{64}\NormalTok{,}
                    \AttributeTok{trait\_means =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.1}\NormalTok{, }\FloatTok{0.1}\NormalTok{), }\AttributeTok{land\_edge =} \StringTok{"torus"}\NormalTok{);}
\end{Highlighting}
\end{Shaded}

Any arguments to \texttt{run\_farm\_sim} not included above are set to default values.
\protect\hyperlink{output}{Output files} can be used to plot ecological and evolutionary dynamics of pests.
Fig \ref{fig:overtime}a shows the pest population change both across and within seasons.
Due to the specific parameter values chosen, clear patterns reflecting pest cohorts emerge.
At the start of a season when the most crop is available, pests eat and abundance increases.
As less crop remains in a season, pest abundance decreases.
Mean food consumed varies over the course of a season caused by the varying frequency of pests in different life history stages (Fig \ref{fig:overtime}c).
In the first 50 time steps, there is some consumption of pesticides 1 and 2 (Fig \ref{fig:overtime}b), but both traits rapidly evolve to negative values making both pesticides ineffective (Fig \ref{fig:overtime}d).
In this case, the trade-off in pesticide consumption is not strong enough to maintain susceptibility to either pesticide in the population.
Fig \ref{fig:landscape} shows the location of pests on the landscape in the last time of the simulation (left panel), and how different pesticides are currently being applied on the landscape (right panel).
Because the number of crops and pesticides, the number and nature of traits, and the size of pest genomes all vary depending on simulation parameters, no plotting functions are introduced in the resevol package.
Instead, methods and code for producing plots such as those in Fig \ref{fig:overtime} and \ref{fig:landscape} are explained in Supporting Information 2.

\begin{figure}
\centering
\includegraphics{ms_files/figure-latex/overtime-1.pdf}
\caption{\label{fig:overtime}Agricultural pest ecological and evolutionary dynamics over 160 time steps from an individual-based simulation using the resevol R package. Panels show (a) pesticide abundance change, (b) mean realised amount of pesticides 1 and 2 uptaken per pest, (c) mean food consumed per pest, and (d) mean value of evolving traits 1 and 2 underlying pest uptake over time. Note that only pests with positive values for Traits 1 or 2 can uptake Pesticide 1 or 2, respectively. Pests with negative trait values will be unaffected by corresponding pesticides (i.e., pesticide consumption is a threshold trait), hence the difference between realised pesticide consumption (b) and the traits underlying it (d). White and grey vertical stripes indicate seasons of a single crop and pesticide application.}
\end{figure}

\begin{figure}
\centering
\includegraphics{ms_files/figure-latex/landscape-1.pdf}
\caption{\label{fig:landscape}Locations of pests (black) across a landscape that includes nine farms (coloured blocks) in the last time step of a simulation using the resevol R package (left panel). The right panel shows which farms apply pesticide 1 (dark grey) and 2 (light grey).}
\end{figure}

\hypertarget{conclusions}{%
\section{Availability and Future Directions}\label{conclusions}}

Insecticide resistance to pesticides is a widespread problem that affects global food security \citep{Tabashnik2014, Bradshaw2016, Lykogianni2021}.
Many models have been developed to investigate the evolution of resistance under different ecological and evolutionary conditions \citep[e.g.,][]{Curtis1985, Mani1985, Ives2002, Liang2013, Liang2018, Haridas2018, South2018, Sudo2018, Saikai2021}.
The resevol R package makes it possible to rapidly develop and simulate myriad individual-based models of resistance evolution.
It also introduces a novel approach to modelling complex pest genetic architecture, using an evolutionary algorithm to generate haploid or diploid loci that map to pest traits with pre-specified covariances (Fig \ref{fig:network}; Supporting Information 1).
Agricultural landscapes and pest life histories are highly customisable, allowing targetted models that can simulate specific real-world case studies (Supporting Information 2).
The breadth of possible models that can be simulated with resevol also makes it a useful tool for developing theory on pest management, and even more generally on the evolution and ecology of individuals with complex traits on a heterogeneous landscape.

The resevol R package can be downloaded from CRAN (\url{https://cran.r-project.org/package=resevol}) or GitHub (\url{https://bradduthie.github.io/resevol/}).
The package is open source under GNU Public License.

\hypertarget{contributions}{%
\section{Authors contributions}\label{contributions}}

All authors contributed to planning software development and writing.
ABD wrote and maintains package code.
CRM contributed code for data visualisation.

\hypertarget{financial}{%
\section{Financial Disclosure statement}\label{financial}}

This software was developed as part of the project for Enhancing Diversity to Overcome Resistance Evolution (ENDORSE) led by Luc Bussière, Ricardo Polanczyk, and Matthew Tinsley (Co-Investigators: Nils Bunnefeld, Yelitza Colmenarez, Natália Corniani, Renata de Lima, Brad Duthie, Steve Edgington, Leonardo Fraceto, Belinda Luke, and Rosie Mangan).
The ENDORSE project is a joint Newton funded international partnership between the Biotechnology and Biological Sciences Research Council (BBSRC) in the UK and the São Paulo Research Foundation (FAPESP) in Brazil under BBSRC award reference BB/S018956/1 and FAPESP award reference 2018/21089-3.
ENDORSE is a partnership among Universidade Estadual Paulista (UNESP), the University of Stirling (UoS), and the Centre for Agricultural and Biosciences International (CABI).
The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.

\hypertarget{competing}{%
\section{Competing interests}\label{competing}}

The authors have declared that no competing interests exist.

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{SI}{%
\section{Supporting information}\label{SI}}

S1 Text. The evolutionary algorithm of the resevol R package, including key data structures used, a general overview of the evolutionary algorithm, and details concerning haploid and diploid individuals. (PDF)

S2 Text. Advanced techniques for using the resevol R package, including initialising pest genomes and running simulations. (PDF)

  \bibliography{resevol.bib}

\end{document}

### 2023-11-30 notebook/ms/plos_latex_ms.tex
Some edits to the manuscript prior to acceptance from PLoS Computational Biology

--- 2023-11-30 notebook/ms/ms.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+% Template for PLoS
+% Version 3.6 Aug 2022
 %
-\documentclass[
-]{article}
+% % % % % % % % % % % % % % % % % % % % % %
+%
+% -- IMPORTANT NOTE
+%
+% This template contains comments intended 
+% to minimize problems and delays during our production 
+% process. Please follow the template instructions
+% whenever possible.
+%
+% % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Once your paper is accepted for publication, 
+% PLEASE REMOVE ALL TRACKED CHANGES in this file 
+% and leave only the final text of your manuscript. 
+% PLOS recommends the use of latexdiff to track changes during review, as this will help to maintain a clean tex file.
+% Visit https://www.ctan.org/pkg/latexdiff?lang=en for info or contact us at latex@plos.org.
+%
+%
+% There are no restrictions on package use within the LaTeX files except that no packages listed in the template may be deleted.
+%
+% Please do not include colors or graphics in the text.
+%
+% The manuscript LaTeX source should be contained within a single file (do not use \input, \externaldocument, or similar commands).
+%
+% % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- FIGURES AND TABLES
+%
+% Please include tables/figure captions directly after the paragraph where they are first cited in the text.
+%
+% DO NOT INCLUDE GRAPHICS IN YOUR MANUSCRIPT
+% - Figures should be uploaded separately from your manuscript file. 
+% - Figures generated using LaTeX should be extracted and removed from the PDF before submission. 
+% - Figures containing multiple panels/subfigures must be combined into one image file before submission.
+% For figure citations, please use "Fig" instead of "Figure".
+% See http://journals.plos.org/plosone/s/figures for PLOS figure guidelines.
+%
+% Tables should be cell-based and may not contain:
+% - spacing/line breaks within cells to alter layout or alignment
+% - do not nest tabular environments (no tabular environments within tabular environments)
+% - no graphics or colored text (cell background color/shading OK)
+% See http://journals.plos.org/plosone/s/tables for table guidelines.
+%
+% For tables that exceed the width of the text column, use the adjustwidth environment as illustrated in the example table in text below.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+%
+% -- EQUATIONS, MATH SYMBOLS, SUBSCRIPTS, AND SUPERSCRIPTS
+%
+% IMPORTANT
+% Below are a few tips to help format your equations and other special characters according to our specifications.
+For more tips to help reduce the possibility of formatting errors during conversion, please see our LaTeX guidelines at http://journals.plos.org/plosone/s/latex
+%
+% For inline equations, please be sure to include all portions of an equation in the math environment.
+For example, x$^2$ is incorrect; this should be formatted as $x^2$ (or $\mathrm{x}^2$ if the romanized font is desired).
+%
+% Do not include text that is not math in the math environment. For example, CO2 should be written as CO\textsubscript{2} instead of CO$_2$.
+%
+% Please add line breaks to long display equations when possible in order to fit size of the column. 
+%
+% For inline equations, please do not include punctuation (commas, etc) within the math environment unless this is part of the equation.
+%
+% When adding superscript or subscripts outside of brackets/braces, please group using {}.  For example, change "[U(D,E,\gamma)]^2" to "{[U(D,E,\gamma)]}^2". 
+%
+% Do not use \cal for caligraphic font.  Instead, use \mathcal{}
+%
+% % % % % % % % % % % % % % % % % % % % % % % % 
+%
+% Please contact latex@plos.org with any questions.
+%
+% % % % % % % % % % % % % % % % % % % % % % % %
+
+\documentclass[10pt,letterpaper]{article}
+\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}
+
+% amsmath and amssymb packages, useful for mathematical formulas and symbols
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{setspace}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
+%\usepackage{natbib}
+
+% Use adjustwidth environment to exceed column width (see example table in text)
+\usepackage{changepage}
+
+% textcomp package and marvosym package for additional characters
+\usepackage{textcomp,marvosym}
+
+% cite package, to clean up citations in the main text. Do not remove.
+\usepackage{cite}
+
+% Use nameref to cite supporting information files (see Supporting Information section for more info)
+\usepackage{nameref,hyperref}
+
+% line numbers
+\usepackage[right]{lineno}
+
+% ligatures disabled
+\usepackage[nopatch=eqnum]{microtype}
+\DisableLigatures[f]{encoding = *, family = * }
+
+% color can be used to apply background shading to table cells only
+\usepackage[table]{xcolor}
+
+% array package and thick rules for tables
+\usepackage{array}
+
+% create "+" rule type for thick vertical lines
+\newcolumntype{+}{!{\vrule width 2pt}}
+
+% create \thickcline for thick horizontal lines of variable length
+\newlength\savedwidth
+\newcommand\thickcline[1]{%
+  \noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+  \cline{#1}%
+  \noalign{\vskip\arrayrulewidth}%
+  \noalign{\global\arrayrulewidth\savedwidth}%
+}
+
+% \thickhline command for thick horizontal lines that span the table
+\newcommand\thickhline{\noalign{\global\savedwidth\arrayrulewidth\global\arrayrulewidth 2pt}%
+\hline
+\noalign{\global\arrayrulewidth\savedwidth}}
+
+% Remove comment for double spacing
+%\usepackage{setspace} 
+%\doublespacing
+
+% Text layout
+\raggedright
+\setlength{\parindent}{0.5cm}
+\textwidth 5.25in 
+\textheight 8.75in
+
+% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
+% Captions will be left justified
+\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
+\renewcommand{\figurename}{Fig}
+
+% Use the PLoS provided BiBTeX style
+\bibliographystyle{plos2015}
+
+% Remove brackets from numbering in List of References
 \makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
+\renewcommand{\@biblabel}[1]{\quad#1.}
 \makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={resevol: an R package for spatially explicit models of pesticide resistance given evolving pest genomes},
-  colorlinks=true,
-  linkcolor={blue},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={Blue},
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage[margin=1in]{geometry}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{5}
-\usepackage{fancyhdr}
+
+% Header and Footer with logo
+\usepackage{lastpage,fancyhdr,graphicx}
+\usepackage{epstopdf}
+%\pagestyle{myheadings}
 \pagestyle{fancy}
-\fancyhead[R]{RESEVOL, AN R PACKAGE}
-\usepackage{lineno}
-\linenumbers
-\modulolinenumbers[2]
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\usepackage[]{natbib}
-\bibliographystyle{acm}
+\fancyhf{}
+%\setlength{\headheight}{27.023pt}
+%\lhead{\includegraphics[width=2.0in]{PLOS-submission.eps}}
+\rfoot{\thepage/\pageref{LastPage}}
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
+\fancyheadoffset[L]{2.25in}
+\fancyfootoffset[L]{2.25in}
+\lfoot{\today}
 
-\title{resevol: an R package for spatially explicit models of pesticide resistance given evolving pest genomes}
-\author{A. Bradley Duthie\(^{1,*}\), Rosie Mangan\(^{1}\), C. Rose McKeon\(^{1}\),\\
-Matthew C. Tinsley\(^{1}\), and Luc F. Bussière\(^{2,3}\)}
-\date{{[}1{]} Biological and Environmental Sciences, University of Stirling, Stirling, UK {[}2{]}Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden {[}3{]} Gothenburgh Global Biodiversity Centre, Gothenburg, Sweden {[}*{]} \href{mailto:alexander.duthie@stir.ac.uk}{\nolinkurl{alexander.duthie@stir.ac.uk}}}
+%% Include all macros below
+
+\newcommand{\lorem}{{\bf LOREM}}
+\newcommand{\ipsum}{{\bf IPSUM}}
+
+%% END MACROS SECTION
 
 \begin{document}
-\maketitle
+\vspace*{0.2in}
 
-\setstretch{2}
-\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}
+% Title must be 250 characters or less.
+\begin{flushleft}
+{\Large
+\textbf\newline{resevol: An R package for spatially explicit models of pesticide resistance given evolving pest genomes} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
+}
+\newline
+% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
+\\
+A. Bradley Duthie\textsuperscript{1*},
+Rosie Mangan\textsuperscript{1},
+C. Rose McKeon\textsuperscript{1},
+Matthew C. Tinsley\textsuperscript{1},
+Luc F. Bussi\`{e}re\textsuperscript{2,3}
+\\
+\bigskip
+\textbf{1} Biological and Environmental Sciences, University of Stirling, Stirling, UK
+\\
+\textbf{2} Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden
+\\
+\textbf{3} Gothenburgh Global Biodiversity Centre, Gothenburg, Sweden
+\\
+\bigskip
 
-\hypertarget{abstract}{%
-\section*{Abstract}\label{abstract}}
-\addcontentsline{toc}{section}{Abstract}
+% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
+% 
+% Remove or comment out the author notes below if they aren't used.
+%
+% Primary Equal Contribution Note
+%\Yinyang These authors contributed equally to this work.
 
+% Additional Equal Contribution Note
+% Also use this double-dagger symbol for special authorship notes, such as senior authorship.
+%\ddag These authors also contributed equally to this work.
+
+% Current address notes
+%\textcurrency Current Address: Dept/Program/Center, Institution Name, City, State, Country % change symbol to "\textcurrency a" if more than one current address note
+% \textcurrency b Insert second current address 
+% \textcurrency c Insert third current address
+
+% Deceased author note
+%\dag Deceased
+
+% Group/Consortium Author Note
+%\textpilcrow Membership list can be found in the Acknowledgments section.
+
+% Use the asterisk to denote corresponding authorship and provide email address in note below.
+* alexander.duthie@stir.ac.uk
+
+\end{flushleft}
+
+% Please keep the abstract below 300 words
+\section*{Abstract}
 The evolution of pesticide resistance is a widespread problem with potentially severe consequences for global food security.
@@
 
-\textbf{Key words:} \emph{pest management}, \emph{food security}, \emph{ecological modelling}, \emph{pesticide}, \emph{resistance evolution}
+% Please keep the Author Summary between 150 and 200 words
+% Use first person. PLOS ONE authors please skip this step. 
+% Author Summary not valid for PLOS ONE submissions.   
+%\section*{Author summary}
+%Lorem ipsum dolor sit amet, consectetur adipiscing elit.
+Curabitur eget porta erat.
+Morbi consectetur est vel gravida pretium.
+Suspendisse ut dui eu ante cursus gravida non sed sem.
+Nullam sapien tellus, commodo id velit id, eleifend volutpat quam.
+Phasellus mauris velit, dapibus finibus elementum vel, pulvinar non tellus.
+Nunc pellentesque pretium diam, quis maximus dolor faucibus id.
+Nunc convallis sodales ante, ut ullamcorper est egestas vitae.
+Nam sit amet enim ultrices, ultrices elit pulvinar, volutpat risus.
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+\linenumbers
 
-Insect resistance to pesticides is a wicked and widespread problem \citep{Gould2018}.
-Predicting, identifying, and ultimately delaying the evolution of insecticide resistance is therefore critical for ensuring sustainable global food security \citep{Hawkins2019, Lykogianni2021}.
+% Use "Eq" instead of "Equation" for equation citations.
+\section*{Introduction}
+Insect resistance to pesticides is a wicked and widespread problem {[}\protect\hyperlink{ref-Gould2018}{1}{]}.
+Predicting, identifying, and ultimately delaying the evolution of insecticide resistance is therefore critical for ensuring sustainable global food security {[}\protect\hyperlink{ref-Hawkins2019}{2},\protect\hyperlink{ref-Lykogianni2021}{3}{]}.
 To manage pesticide resistance, various strategies have been proposed to disrupt or weaken selection.
-These strategies often focus on varying the location and timing of pesticide application, and therefore varying the strength of selection for pesticide resistance, so that pest susceptibility to one or more pesticides is maintained \citep{Ives2002, Liang2013, Carriere2016, Liang2018, South2018, Sudo2018, Saikai2021}.
+These strategies often focus on varying the location and timing of pesticide application, and therefore varying the strength of selection for pesticide resistance, so that pest susceptibility to one or more pesticides is maintained {[}\protect\hyperlink{ref-Ives2002}{4}--\protect\hyperlink{ref-Saikai2021}{10}{]}.
 Pesticide resistance management has mostly been associated with the effect of single genes.
-In such cases, resistance alleles have binary effects on phenotype, enabling resistant phenotypes to arise from genetic changes at single or a small number of loci \citep{Raymond2001, Ffrench-Constant2004, Ffrench-Constant2013}.
-Nevertheless, resistance attributable to polygenic effects is also well-established \citep{Via1986, Firko1990, Busi2013, Haridas2018, Green2020, Mangan2023}.
-And the relevance of polygenic resistance is likely to further increase given a rising interest in biological controls for sustainable crop protection \citep[e.g.,][]{Green2020, Elnahal2022, Mangan2023}.
+In such cases, resistance alleles have binary effects on phenotype, enabling resistant phenotypes to arise from genetic changes at single or a small number of loci {[}\protect\hyperlink{ref-Raymond2001}{11}--\protect\hyperlink{ref-Ffrench-Constant2013}{13}{]}.
+Nevertheless, resistance attributable to polygenic effects is also well-established {[}\protect\hyperlink{ref-Via1986}{14}--\protect\hyperlink{ref-Mangan2023}{19}{]}.
+And the relevance of polygenic resistance is likely to further increase given a rising interest in biological controls for sustainable crop protection {[}e.g., \protect\hyperlink{ref-Green2020}{18},\protect\hyperlink{ref-Mangan2023}{19},\protect\hyperlink{ref-Elnahal2022}{20}{]}.
 Developing strategies to maximise the efficacy of these tools is critical, and such strategies should be well-informed by predictions made from detailed, quantitative genetic models.
-Here we introduce the resevol R package as a tool for building individual-based models and simulating pest management \citep{Daly2022}.
+Here we introduce the resevol R package as a tool for building individual-based models and simulating pest management {[}\protect\hyperlink{ref-Daly2022}{21}{]}.
 
@@
 This approach to modelling individual genomes and traits thereby increases the complexity of questions that can be addressed for simulating evolution in agricultural pests.
-
-\begin{figure}
-\centering
-\includegraphics{ms_files/figure-latex/network-1.pdf}
-\caption{\label{fig:network}Example network mapping loci (green circles) to traits (red diamonds) through an intermediate set of hidden layers (blue squares) in the \texttt{mine\_gmatrix} function.
-Individual genomes in the resevol R package consist of standard random normal values for loci, real values for black arrows linking nodes, and real values for traits.
-Values shown for loci and arrows are an example for illustration.}
-\end{figure}
 
@@
 \begin{verbatim}
-0.1, -0.1, -0.2, 0.4, 0.1, -1.2, -0.6, 0.2, 2.2, 1.3, -0.5, -0.1, 0.0, 2.2, 0.9, 1.6, 0.8
+0.1, -0.1, -0.2, 0.4, 0.1, -1.2, -0.6, 0.2, 2.2, 1.3, -0.5, -0.1, 0.0, 
+2.2, 0.9, 1.6, 0.8
 \end{verbatim}
@@
 
-An evolutionary algorithm is used to find appropriate values that produce covarying traits from loci (see Supporting Information 1 for details).
-Evolutionary algorithms are heuristic tools that can simulate adaptive evolution to find solutions for a broad range of problems \citep{Hamblin2013, Duthie2018}.
+\begin{figure}[!h]
+\caption{{\bf Example resevol network.}
+Example network mapping loci (green circles) to traits (red diamonds) through an intermediate set of hidden layers (blue squares) in the \texttt{mine\_gmatrix} function.
+Individual genomes in the resevol R package consist of standard random normal values for loci, real values for black arrows linking nodes, and real values for traits.
+Values shown for loci and arrows are an example for illustration.}
+\label{fig:network}
+\end{figure}
+
+An evolutionary algorithm is used to find appropriate values that produce covarying traits from loci (see \nameref{S1_Text} for details).
+Evolutionary algorithms are heuristic tools that can simulate adaptive evolution to find solutions for a broad range of problems {[}\protect\hyperlink{ref-Hamblin2013}{22},\protect\hyperlink{ref-Duthie2018}{23}{]}.
 In the resevol package, the \texttt{mine\_gmatrix} function runs an evolutionary algorithm and requires the argument \texttt{gmatrix}, which specifies the desired trait covariance matrix.
@@
 Network stress is calculated as the logged mean squared deviation between estimated covariances and those in \texttt{gmatrix}.
-Tournament selection \citep{Hamblin2013} is then used to determine the networks for the next iteration of the algorithm.
+Tournament selection {[}\protect\hyperlink{ref-Hamblin2013}{22}{]} is then used to determine the networks for the next iteration of the algorithm.
 Networks with the lowest estimated stress have the highest fitness, so these networks are disproportionately represented in the next iteration.
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{trait\_covs  }\OtherTok{\textless{}{-}} \FunctionTok{matrix}\NormalTok{(}\AttributeTok{data =} \FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{, }\SpecialCharTok{{-}}\FloatTok{0.4}\NormalTok{, }\SpecialCharTok{{-}}\FloatTok{0.4}\NormalTok{, }\DecValTok{1}\NormalTok{), }\AttributeTok{nrow =} \DecValTok{2}\NormalTok{, }\AttributeTok{ncol =} \DecValTok{2}\NormalTok{);}
-\NormalTok{new\_network }\OtherTok{\textless{}{-}} \FunctionTok{mine\_gmatrix}\NormalTok{(}\AttributeTok{loci =} \DecValTok{3}\NormalTok{, }\AttributeTok{layers =} \DecValTok{2}\NormalTok{, }\AttributeTok{gmatrix =}\NormalTok{ trait\_covs, }
-                            \AttributeTok{max\_gen =} \DecValTok{1000}\NormalTok{, }\AttributeTok{term\_cri =} \SpecialCharTok{{-}}\FloatTok{6.0}\NormalTok{, }\AttributeTok{prnt\_out =} \ConstantTok{FALSE}\NormalTok{);}
-\end{Highlighting}
-\end{Shaded}
+\begin{verbatim}
+trait_covs  <- matrix(data = c(1, -0.4, -0.4, 1), nrow = 2, ncol = 2);
+new_network <- mine_gmatrix(loci = 3, layers = 2, gmatrix = trait_covs, 
+                            max_gen = 1000, term_cri = -6.0, 
+                            prnt_out = FALSE);
+\end{verbatim}
 
@@
 Similarly, higher values of \texttt{indivs} will result in more accurate estimations of true stress, but this also requires more computation time.
-Additional arguments to \texttt{mine\_gmatrix} can also be used to improve the performance of the evolutionary algorithm (see Supporting Information 1).
+Additional arguments to \texttt{mine\_gmatrix} can also be used to improve the performance of the evolutionary algorithm (see \nameref{S1_Text}).
 
@@
 Each value defines a unique farm, but values do not need to be contiguous.
-These `farms' could even model non-farmland (e.g., water, roads), if pesticides and crops on them are invisible to pests (see Supporting Information 2), or they could more broadly be interpreted as heterogenous landscape properties across an arbitrary scale \citep[e.g.,][]{Newman2023}.
-This \texttt{terrain} customisation therefore allows for a high degree of landscape detail, and offers the potential for modelling real-world landscapes from raster images \citep[e.g.,][]{Millington2021}.
+These `farms' could even model non-farmland (e.g., water, roads), if pesticides and crops on them are invisible to pests (see \nameref{S2_Text}), or they could more broadly be interpreted as heterogenous landscape properties across an arbitrary scale {[}e.g., \protect\hyperlink{ref-Newman2023}{24}{]}.
+This \texttt{terrain} customisation therefore allows for a high degree of landscape detail, and offers the potential for modelling real-world landscapes from raster images {[}e.g., \protect\hyperlink{ref-Millington2021}{25}{]}.
 Edge effects are set using the \texttt{land\_edge} argument.
@@
 Square matrices can be used to define the probability that a given crop or pesticide in row \(i\) transitions to that in column \(j\).
-Hence, any possible Markov chain can be used to model transition between crop or pesticide types on farms, potentially integrating real crop use or pesticide use patterns \citep{Zhang2020}.
+Hence, any possible Markov chain can be used to model transition between crop or pesticide types on farms, potentially integrating real crop use or pesticide use patterns {[}\protect\hyperlink{ref-Zhang2020}{26}{]}.
 Upon rotation, crop and pesticide values are reset on each farm.
@@
 Food accumulated is lost during aging if \texttt{baseline\_metabolism\ \textgreater{}\ 0} and pest age is within \texttt{min\_age\_metabolism} and \texttt{max\_age\_metabolism}.
-The option to set minimum and maximum ages for events makes it possible to model pests with much different life histories \citep[e.g.,][]{Sudo2018}.
+The option to set minimum and maximum ages for events makes it possible to model pests with much different life histories {[}e.g., \protect\hyperlink{ref-Sudo2018}{9}{]}.
 
-\begin{figure}
-\centering
-\includegraphics{ms_files/figure-latex/flowchart-1.pdf}
-\caption{\label{fig:flowchart}Overview of simulated events in the resevol R package.
+\begin{figure}[!h]
+\caption{{\bf Overview of resevol events.}
+Overview of simulated events in the resevol R package.
 Note that metabolism, feeding, pesticide consumption, movement, and reproduction are all subject to a minimum and maximum pest age.
@@
 Statistics collected within a time step are printed to a CSV file.}
+\label{fig:flowchart}
 \end{figure}
@@
 In resevol v0.3, the \texttt{run\_farm\_sim} function includes 68 arguments, which specify a wide range of possible simulation conditions affecting landscape and pest characteristics.
-These arguments are explained in the package documentation, and in Supporting Information 2, which demonstrates an advanced case study with a custom landscape and complex pest genomes and life history.
+These arguments are explained in the package documentation, and in \nameref{S2_Text}, which demonstrates an advanced case study with a custom landscape and complex pest genomes and life history.
 Here we focus on a simple simulation with asexually reproducing pests that have three loci and two traits (Fig \ref{fig:network}).
-We use the the genome generated in \texttt{new\_network} from \protect\hyperlink{mine_gmatrix}{Section 2}, in which traits 1 and 2 have variances of 1.0008181 and 1.027134, respectively, with a covariance of -0.3925799.
+We use the the genome generated in \texttt{new\_network} from \protect\hyperlink{mine_gmatrix}{Section 1.1}, in which traits 1 and 2 have variances of 1.0008181 and 1.027134, respectively, with a covariance of -0.3925799.
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{sim }\OtherTok{\textless{}{-}} \FunctionTok{run\_farm\_sim}\NormalTok{(}\AttributeTok{mine\_output =}\NormalTok{ new\_network, }\AttributeTok{repro =} \StringTok{"asexual"}\NormalTok{, }
-                    \AttributeTok{pesticide\_number =} \DecValTok{2}\NormalTok{, }\AttributeTok{pesticide\_init =} \StringTok{"random"}\NormalTok{, }
-                    \AttributeTok{pesticide\_consume =} \FunctionTok{c}\NormalTok{(}\StringTok{"T1"}\NormalTok{, }\StringTok{"T2"}\NormalTok{), }\AttributeTok{farms =} \DecValTok{9}\NormalTok{,}
-                    \AttributeTok{pesticide\_rotation\_time =} \DecValTok{16}\NormalTok{, }\AttributeTok{pesticide\_rotation\_type =} \DecValTok{3}\NormalTok{,}
-                    \AttributeTok{pesticide\_tolerated\_surv =} \DecValTok{0}\NormalTok{, }\AttributeTok{pesticide\_per\_cell =} \DecValTok{1}\NormalTok{,}
-                    \AttributeTok{crop\_rotation\_time =} \DecValTok{16}\NormalTok{, }\AttributeTok{crop\_number =} \DecValTok{1}\NormalTok{, }\AttributeTok{crop\_per\_cell =} \DecValTok{4}\NormalTok{, }
-                    \AttributeTok{food\_consume =} \DecValTok{1}\NormalTok{, }\AttributeTok{reproduction\_type =} \StringTok{"food\_based"}\NormalTok{, }
-                    \AttributeTok{food\_needed\_surv =} \DecValTok{1}\NormalTok{, }\AttributeTok{food\_needed\_repr =} \DecValTok{1}\NormalTok{, }\AttributeTok{max\_age =} \DecValTok{4}\NormalTok{, }
-                    \AttributeTok{min\_age\_feed =} \DecValTok{0}\NormalTok{, }\AttributeTok{max\_age\_feed =} \DecValTok{2}\NormalTok{, }\AttributeTok{min\_age\_move =} \DecValTok{3}\NormalTok{, }
-                    \AttributeTok{max\_age\_move =} \DecValTok{4}\NormalTok{, }\AttributeTok{min\_age\_reproduce =} \DecValTok{4}\NormalTok{, }\AttributeTok{print\_gens =} \ConstantTok{FALSE}\NormalTok{,}
-                    \AttributeTok{max\_age\_reproduce =} \DecValTok{4}\NormalTok{, }\AttributeTok{age\_pesticide\_threshold =} \DecValTok{2}\NormalTok{, }
-                    \AttributeTok{rand\_age =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{move\_distance =} \DecValTok{2}\NormalTok{, }\AttributeTok{immigration\_rate =} \DecValTok{10}\NormalTok{, }
-                    \AttributeTok{time\_steps =} \DecValTok{160}\NormalTok{, }\AttributeTok{print\_last =} \ConstantTok{TRUE}\NormalTok{, }\AttributeTok{xdim =} \DecValTok{64}\NormalTok{, }\AttributeTok{ydim =} \DecValTok{64}\NormalTok{,}
-                    \AttributeTok{trait\_means =} \FunctionTok{c}\NormalTok{(}\FloatTok{0.1}\NormalTok{, }\FloatTok{0.1}\NormalTok{), }\AttributeTok{land\_edge =} \StringTok{"torus"}\NormalTok{);}
-\end{Highlighting}
-\end{Shaded}
+\begin{verbatim}
+sim <- run_farm_sim(mine_output = new_network, repro = "asexual", 
+                    pesticide_number = 2, pesticide_init = "random", 
+                    pesticide_consume = c("T1", "T2"), farms = 9,
+                    pesticide_rotation_time = 16, 
+                    pesticide_rotation_type = 3,
+                    pesticide_tolerated_surv = 0, 
+                    pesticide_per_cell = 1,
+                    crop_rotation_time = 16, crop_number = 1, 
+                    crop_per_cell = 4, food_consume = 1, 
+                    reproduction_type = "food_based", 
+                    food_needed_surv = 1, food_needed_repr = 1, 
+                    max_age = 4, min_age_feed = 0, max_age_feed = 2, 
+                    min_age_move = 3, max_age_move = 4, 
+                    min_age_reproduce = 4, print_gens = FALSE,
+                    max_age_reproduce = 4, age_pesticide_threshold = 2, 
+                    rand_age = TRUE, move_distance = 2, 
+                    immigration_rate = 10, time_steps = 160, 
+                    print_last = TRUE, xdim = 64, ydim = 64,
+                    trait_means = c(0.1, 0.1), land_edge = "torus");
+\end{verbatim}
 
@@
 Because the number of crops and pesticides, the number and nature of traits, and the size of pest genomes all vary depending on simulation parameters, no plotting functions are introduced in the resevol package.
-Instead, methods and code for producing plots such as those in Fig \ref{fig:overtime} and \ref{fig:landscape} are explained in Supporting Information 2.
+Instead, methods and code for producing plots such as those in Fig \ref{fig:overtime} and \ref{fig:landscape} are explained in \nameref{S2_Text}.
 
-\begin{figure}
-\centering
-\includegraphics{ms_files/figure-latex/overtime-1.pdf}
-\caption{\label{fig:overtime}Agricultural pest ecological and evolutionary dynamics over 160 time steps from an individual-based simulation using the resevol R package.
+\begin{figure}[!h]
+\caption{{\bf Example resevol population dynamics.}
+Agricultural pest ecological and evolutionary dynamics over 160 time steps from an individual-based simulation using the resevol R package.
 Panels show (a) pesticide abundance change, (b) mean realised amount of pesticides 1 and 2 uptaken per pest, (c) mean food consumed per pest, and (d) mean value of evolving traits 1 and 2 underlying pest uptake over time.
@@
 White and grey vertical stripes indicate seasons of a single crop and pesticide application.}
+\label{fig:overtime}
 \end{figure}
 
-\begin{figure}
-\centering
-\includegraphics{ms_files/figure-latex/landscape-1.pdf}
-\caption{\label{fig:landscape}Locations of pests (black) across a landscape that includes nine farms (coloured blocks) in the last time step of a simulation using the resevol R package (left panel).
+\begin{figure}[!h]
+\caption{{\bf Pest spatial distributions in resevol simulation.}
+Locations of pests (black) across a landscape that includes nine farms (coloured blocks) in the last time step of a simulation using the resevol R package (left panel).
 The right panel shows which farms apply pesticide 1 (dark grey) and 2 (light grey).}
+\label{fig:landscape}
 \end{figure}
@@
 
-Insecticide resistance to pesticides is a widespread problem that affects global food security \citep{Tabashnik2014, Bradshaw2016, Lykogianni2021}.
-Many models have been developed to investigate the evolution of resistance under different ecological and evolutionary conditions \citep[e.g.,][]{Curtis1985, Mani1985, Ives2002, Liang2013, Liang2018, Haridas2018, South2018, Sudo2018, Saikai2021}.
+Insecticide resistance to pesticides is a widespread problem that affects global food security {[}\protect\hyperlink{ref-Lykogianni2021}{3},\protect\hyperlink{ref-Tabashnik2014}{27},\protect\hyperlink{ref-Bradshaw2016}{28}{]}.
+Many models have been developed to investigate the evolution of resistance under different ecological and evolutionary conditions {[}\protect\hyperlink{ref-Ives2002}{4},\protect\hyperlink{ref-Liang2013}{5},\protect\hyperlink{ref-Liang2018}{7}--\protect\hyperlink{ref-Saikai2021}{10},\protect\hyperlink{ref-Haridas2018}{17},e.g., \protect\hyperlink{ref-Curtis1985}{29},\protect\hyperlink{ref-Mani1985}{30}{]}.
 The resevol R package makes it possible to rapidly develop and simulate myriad individual-based models of resistance evolution.
-It also introduces a novel approach to modelling complex pest genetic architecture, using an evolutionary algorithm to generate haploid or diploid loci that map to pest traits with pre-specified covariances (Fig \ref{fig:network}; Supporting Information 1).
-Agricultural landscapes and pest life histories are highly customisable, allowing targetted models that can simulate specific real-world case studies (Supporting Information 2).
+It also introduces a novel approach to modelling complex pest genetic architecture, using an evolutionary algorithm to generate haploid or diploid loci that map to pest traits with pre-specified covariances (Fig \ref{fig:network}; \nameref{S1_Text}).
+Agricultural landscapes and pest life histories are highly customisable, allowing targetted models that can simulate specific real-world case studies (\nameref{S2_Text}).
 The breadth of possible models that can be simulated with resevol also makes it a useful tool for developing theory on pest management, and even more generally on the evolution and ecology of individuals with complex traits on a heterogeneous landscape.
 
-The resevol R package can be downloaded from CRAN (\url{https://cran.r-project.org/package=resevol}) or GitHub (\url{https://bradduthie.github.io/resevol/}).
-The package is open source under GNU Public License.
+\nolinenumbers
 
-\hypertarget{contributions}{%
-\section{Authors contributions}\label{contributions}}
+% Either type in your references using
+% \begin{thebibliography}{}
+% \bibitem{}
+% Text
+% \end{thebibliography}
+%
+% or
+%
+% Compile your BiBTeX database using our plos2015.bst
+% style file and paste the contents of your .bbl file
+% here. See http://journals.plos.org/plosone/s/latex for 
+% step-by-step instructions.
+% 
+\begin{thebibliography}{30}
 
-All authors contributed to planning software development and writing.
-ABD wrote and maintains package code.
-CRM contributed code for data visualisation.
+\bibitem{ref-Gould2018}
+\hypertarget{ref-Gould2018}{Gould} F, Brown ZS, Kuzma J.
+{Wicked evolution: Can we address the sociobiological dilemma of pesticide resistance?} Science.
+2018;360: 728--732.
+doi:\href{https://doi.org/10.1126/science.aar3780}{10.1126/science.aar3780}
 
-\hypertarget{financial}{%
-\section{Financial Disclosure statement}\label{financial}}
+\bibitem{ref-Hawkins2019}
+\hypertarget{ref-Hawkins2019}{Hawkins} NJ, Bass C, Dixon A, Neve P.
+{The evolutionary origins of pesticide resistance}.
+Biological Reviews.
+2019;94: 135--155.
+doi:\href{https://doi.org/10.1111/brv.12440}{10.1111/brv.12440}
 
-This software was developed as part of the project for Enhancing Diversity to Overcome Resistance Evolution (ENDORSE) led by Luc Bussière, Ricardo Polanczyk, and Matthew Tinsley (Co-Investigators: Nils Bunnefeld, Yelitza Colmenarez, Natália Corniani, Renata de Lima, Brad Duthie, Steve Edgington, Leonardo Fraceto, Belinda Luke, and Rosie Mangan).
-The ENDORSE project is a joint Newton funded international partnership between the Biotechnology and Biological Sciences Research Council (BBSRC) in the UK and the São Paulo Research Foundation (FAPESP) in Brazil under BBSRC award reference BB/S018956/1 and FAPESP award reference 2018/21089-3.
-ENDORSE is a partnership among Universidade Estadual Paulista (UNESP), the University of Stirling (UoS), and the Centre for Agricultural and Biosciences International (CABI).
-The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.
+\bibitem{ref-Lykogianni2021}
+\hypertarget{ref-Lykogianni2021}{Lykogianni} M, Bempelou E, Karamaouna F, Aliferis KA.
+{Do pesticides promote or hinder sustainability in agriculture?
+The challenge of sustainable use of pesticides in modern agriculture}.
+Science of the Total Environment.
+2021;795: 148625.
+doi:\href{https://doi.org/10.1016/j.scitotenv.2021.148625}{10.1016/j.scitotenv.2021.148625}
 
-\hypertarget{competing}{%
-\section{Competing interests}\label{competing}}
+\bibitem{ref-Ives2002}
+\hypertarget{ref-Ives2002}{Ives} AR, Andow DA.
+{Evolution of resistance to Bt crops: Directional selection in structured environments}.
+Ecology Letters.
+2002;5: 792--801.
+doi:\href{https://doi.org/10.1046/j.1461-0248.2002.00392.x}{10.1046/j.1461-0248.2002.00392.x}
 
-The authors have declared that no competing interests exist.
+\bibitem{ref-Liang2013}
+\hypertarget{ref-Liang2013}{Liang} J, Tang S, Nieto JJ, Cheke RA.
+{Analytical methods for detecting pesticide switches with evolution of pesticide resistance}.
+Mathematical Biosciences.
+2013;245: 249--257.
+doi:\href{https://doi.org/10.1016/j.mbs.2013.07.008}{10.1016/j.mbs.2013.07.008}
 
-\hypertarget{references}{%
-\section{References}\label{references}}
+\bibitem{ref-Carriere2016}
+Carri\`{e}re Y, Fabrick JA, Tabashnik BE.
+{Can Pyramids and Seed Mixtures Delay Resistance to Bt Crops?} Trends in Biotechnology.
+2016;34: 291--302.
+doi:\href{https://doi.org/10.1016/j.tibtech.2015.12.011}{10.1016/j.tibtech.2015.12.011}
 
-\hypertarget{SI}{%
-\section{Supporting information}\label{SI}}
+\bibitem{ref-Liang2018}
+\hypertarget{ref-Liang2018}{Liang} J, Tang S, Cheke RA.
+{A discrete host-parasitoid model with development of pesticide resistance and ipm strategies}.
+Journal of Biological Dynamics.
+2018;12: 1059--1078.
+doi:\href{https://doi.org/10.1080/17513758.2018.1556351}{10.1080/17513758.2018.1556351}
 
-S1 Text.
-The evolutionary algorithm of the resevol R package, including key data structures used, a general overview of the evolutionary algorithm, and details concerning haploid and diploid individuals.
+\bibitem{ref-South2018}
+8.
+South A, Hastings IM.
+{Insecticide resistance evolution with mixtures and sequences: A model-based explanation}.
+Malaria Journal.
+2018;17: 1--20.
+doi:\href{https://doi.org/10.1186/s12936-018-2203-y}{10.1186/s12936-018-2203-y}
+
+\bibitem{ref-Sudo2018}
+\hypertarget{ref-Sudo2018}{Sudo} M, Takahashi D, Andow DA, Suzuki Y, Yamanaka T.
+{Optimal management strategy of insecticide resistance under various insect life histories: Heterogeneous timing of selection and interpatch dispersal}.
+Evolutionary Applications.
+2018;11: 271--283.
+doi:\href{https://doi.org/10.1111/eva.12550}{10.1111/eva.12550}
+
+\bibitem{ref-Saikai2021}
+\hypertarget{ref-Saikai2021}{Saikai} Y, Hurley TM, Mitchell PD.
+{An agent-based model of insect resistance management and mitigation for Bt maize: a social science perspective}.
+Pest Management Science.
+2021;77: 273--284.
+doi:\href{https://doi.org/10.1002/ps.6016}{10.1002/ps.6016}
+
+\bibitem{ref-Raymond2001}
+\hypertarget{ref-Raymond2001}{Raymond} M, Berticat C, Weill M, Pasteur N, Chevillon C.
+{Insecticide resistance in the mosquito \textit{Culex pipiens}: What have we learned about adaptation?} Genetica.
+2001;103: 239--248.
+doi:\href{https://doi.org/10.1023/A}{10.1023/A}
+
+\bibitem{ref-Ffrench-Constant2004}
+Ffrench-Constant RH, Daborn PJ, Le Goff G.
+{The genetics and genomics of insecticide resistance}.
+Trends in Genetics.
+2004;20: 163--170.
+doi:\href{https://doi.org/10.1016/j.tig.2004.01.003}{10.1016/j.tig.2004.01.003}
+
+\bibitem{ref-Ffrench-Constant2013}
+\hypertarget{ref-Ffrench-Constant2013}{Ffrench-Constant} RH.
+{The molecular genetics of insecticide resistance}.
+Genetics.
+2013;194: 807--815.
+doi:\href{https://doi.org/10.1534/genetics.112.141895}{10.1534/genetics.112.141895}
+
+\bibitem{ref-Via1986}
+\hypertarget{ref-Via1986}{Via} S.
+{Quantitative genetic models and the evolution of pesticide resistance}.
+Pesticide resistance: Strategies and tactics for management.
+Washington, DC: National Academy Press; 1986.
+pp.
+222--235.
+
+\bibitem{ref-Firko1990}
+Firko MJ, Hayes JL. {Quantitative genetic tools for insecticide resistance risk assessment: estimating the heritability of resistance}. Journal of Economic Entomology. 1990;83: 657--654.
+
+\bibitem{ref-Busi2013}
+Busi R, Neve P, Powles S.
+{Evolved polygenic herbicide resistance in \textit{Lolium rigidum} by low-dose herbicide selection within standing genetic variation}.
+Evolutionary Applications.
+2013;6: 231--242.
+doi:\href{https://doi.org/10.1111/j.1752-4571.2012.00282.x}{10.1111/j.1752-4571.2012.00282.x}
+
+\bibitem{ref-Haridas2018}
+\hypertarget{ref-Haridas2018}{Haridas} CV, Tenhumberg B.
+{Modeling effects of ecological factors on evolution of polygenic pesticide resistance}.
+Journal of Theoretical Biology.
+2018;456: 224--232.
+doi:\href{https://doi.org/10.1016/j.jtbi.2018.07.034}{10.1016/j.jtbi.2018.07.034}
+
+\bibitem{ref-Green2020}
+\hypertarget{ref-Green2020}{Green} KK, Stenberg JA, Lankinen Å.
+{Making sense of Integrated Pest Management (IPM) in the light of evolution}.
+Evolutionary Applications.
+2020;13: 1791--1805.
+doi:\href{https://doi.org/10.1111/eva.13067}{10.1111/eva.13067}
+
+\bibitem{ref-Mangan2023}
+\hypertarget{ref-Mangan2023}{Mangan} R, Bussière LF, Polanczyk RA, Tinsley MC.
+{Increasing ecological heterogeneity can constrain biopesticide resistance evolution}.
+Trends in Ecology and Evolution.
+2023;38: 605--614.
+doi:\href{https://doi.org/10.1016/j.tree.2023.01.012}{10.1016/j.tree.2023.01.012}
+
+\bibitem{ref-Elnahal2022}
+\hypertarget{ref-Elnahal2022}{Elnahal} AS, El-Saadony MT, Saad AM, Desoky ESM, El-Tahan AM, Rady MM, et al.
+{The use of microbial inoculants for biological control, plant growth promotion, and sustainable agriculture: A review}.
+Springer Netherlands; 2022.
+pp.
+759--792.
+doi:\href{https://doi.org/10.1007/s10658-021-02393-7}{10.1007/s10658-021-02393-7}
+
+\bibitem{ref-Daly2022}
+\hypertarget{ref-Daly2022}{Daly} AJ, De Visscher L, Baetens JM, De Baets B.
+{Quo vadis, agent-based modelling tools?} Environmental Modelling and Software.
+2022;157: 105514.
+doi:\href{https://doi.org/10.1016/j.envsoft.2022.105514}{10.1016/j.envsoft.2022.105514}
+
+\bibitem{ref-Hamblin2013}
+\hypertarget{ref-Hamblin2013}{Hamblin} S.
+{On the practical usage of genetic algorithms in ecology and evolution}.
+Methods in Ecology and Evolution.
+2013;4: 184--194.
+doi:\href{https://doi.org/10.1111/2041-210X.12000}{10.1111/2041-210X.12000}
+
+\bibitem{ref-Duthie2018}
+\hypertarget{ref-Duthie2018}{Duthie} AB, Cusack JJ, Jones IL, Nilsen EB, Pozo RA, Rakotonarivo OS, et al.
+{GMSE: an R package for generalised management strategy evaluation}.
+Methods in Ecology and Evolution.
+2018;9: 2396--2401.
+doi:\href{https://doi.org/10.1101/221432}{10.1101/221432}
+
+\bibitem{ref-Newman2023}
+\hypertarget{ref-Newman2023}{Newman} DR, Saurette DD, Cockburn JMH, Dragut L, Lindsay JB.
+{Assessing spatially heterogeneous scale representation with applied digital soil mapping}.
+Environmental Modelling and Software.
+2023;160: 105612.
+doi:\href{https://doi.org/10.1016/j.envsoft.2022.105612}{10.1016/j.envsoft.2022.105612}
+
+\bibitem{ref-Millington2021}
+\hypertarget{ref-Millington2021}{Millington} JDA, Katerinchuk V, Silva RFB da, Victoria D de C, Batistella M.
+{Modelling drivers of Brazilian agricultural change in a telecoupled world}.
+Environmental Modelling and Software.
+2021;139: 105024.
+doi:\href{https://doi.org/10.1016/j.envsoft.2021.105024}{10.1016/j.envsoft.2021.105024}
+
+\bibitem{ref-Zhang2020}
+\hypertarget{ref-Zhang2020}{Zhang} C, Di L, Yang Z, Lin L, Hao P.
+{AgKit4EE: A toolkit for agricultural land use modeling of the conterminous United States based on Google Earth Engine}.
+Environmental Modelling and Software.
+2020;129: 104694.
+doi:\href{https://doi.org/10.1016/j.envsoft.2020.104694}{10.1016/j.envsoft.2020.104694}
+
+\bibitem{ref-Tabashnik2014}
+\hypertarget{ref-Tabashnik2014}{Tabashnik} BE, Mota-Sanchez D, Whalon ME, Hollingworth RM, Carri\`{e}re Y.
+{Defining terms for proactive management of resistance to Bt crops and pesticides}.
+Journal of Economic Entomology.
+2014;107: 496--507.
+doi:\href{https://doi.org/10.1603/EC13458}{10.1603/EC13458}
+
+\bibitem{ref-Bradshaw2016}
+\hypertarget{ref-Bradshaw2016}{Bradshaw} CJA, Leroy B, Bellard C, Roiz D, Albert C, Fournier A, et al.
+{Massive yet grossly underestimated global costs of invasive insects}.
+Nature Communications.
+2016;7: 12986.
+doi:\href{https://doi.org/10.1038/ncomms12986}{10.1038/ncomms12986}
+
+\bibitem{ref-Curtis1985}
+\hypertarget{ref-Curtis1985}{Curtis} CF.
+{Theoretical models of the use of insecticide mixtures for the management of resistance}.
+Bulletin of Entomological Research.
+1985;75: 259--266.
+doi:\href{https://doi.org/10.1017/S0007485300014346}{10.1017/S0007485300014346}
+
+\bibitem{ref-Mani1985}
+\hypertarget{ref-Mani1985}{Mani} GS.
+{Evolution of resistance in the presence of two insecticides.} Genetics.
+1985;109: 761--783.
+doi:\href{https://doi.org/10.1093/genetics/109.4.761}{10.1093/genetics/109.4.761}
+
+\end{thebibliography}
+
+\section*{Supporting information}
+
+% Include only the SI item label in the paragraph heading. Use the \nameref{label} command to cite SI items in the text.
+\paragraph*{S1 Text.}
+\label{S1_Text}
+{\bf The evolutionary algorithm} The evolutionary algorithm of the resevol R package, including key data structures used, a general overview of the evolutionary algorithm, and details concerning haploid and diploid individuals.
 (PDF)
 
-S2 Text. Advanced techniques for using the resevol R package, including initialising pest genomes and running simulations. (PDF)
-
-  \bibliography{resevol.bib}
+\paragraph*{S2 Text.}
+\label{S2_Text}
+{\bf Advanced techniques}  Advanced techniques for using the resevol R package, including initialising pest genomes and running simulations. (PDF)
 
